---
title: "Multivariate Confirmation Analysis - Non-college grads"
editor: visual
---

```{r include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
library(htmlwidgets)
library(RColorBrewer)
library(readxl)
library(janitor)
library(lubridate)
library(systemfonts)
reset_font_cache()
library(ggtext)
library(gmodels)
library(fastDummies)
library(car)
library(glmnet)
library(glmnetUtils)
library(pscl)
library(sjPlot)
library(ggforce)
library(broom)
library(forcats)
library(stringr)
```

```{r themes and shapefiles, include=FALSE}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", linewidth  = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.key.size = unit(1, "lines"),
        legend.margin = margin(0,0,0,0),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.text = element_text(margin = margin(l = 2)),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", linewidth = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))


theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
    select(5,6) %>%
    unique() %>%
    mutate(edr = str_replace(edr, "  ", " "),
           planning.region = str_replace(planning.region, " Minnesota", ""),
           planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
           edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

counties.regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
  rename(mif = `MIF Region`) %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Name = str_replace(Name, "Mcleod", "McLeod"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc) ,
         edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"))

counties.regions.1 <- counties.regions %>%
  mutate(statefp = "27",
         project.pr = ifelse(edr %in% c("EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 4 - West Central"), "Northwest",
                             ifelse(edr == "EDR 3 - Arrowhead", "Northeast",
                                    ifelse(edr %in% c("EDR 5 - North Central", "EDR 7E- East Central", "EDR 7W- Central"), "Central",
                                           ifelse(edr %in% c("EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 8 - Southwest"), "Southwest",
                                                  ifelse(edr %in% c("EDR 9 - South Central", "EDR 10 - Southeast"), "Southern", as.character(planning.region)))))),
         project.pr = fct_relevel(project.pr, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southern"))


color.ruca <- c("Entirely rural" = "#009933", "Town/rural mix" = "#99CC33", "Urban/town/rural mix" = "#CC9966", "Entirely urban" = "#754C29", "Minnesota" = "black")

color.pr <- c("Northwest" = 	"#4575b4", "Northeast" = "grey", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.edr.simple <- c("EDR 1" = "#b3cde3", "EDR 2" = "#8c96c6", "EDR 3" = "#fe9929", "EDR 4" = "#8856a7", "EDR 5" = "#810f7c", "EDR 6E" = "#e5f5f9", "EDR 6W" = "#bdc9e1", "EDR 7E" = "#99d8c9", "EDR 7W" = "#2ca25f", "EDR 8" = "#74a9cf", "EDR 9" = "#0570b0", "EDR 10" = "#d7301f", "EDR 11" = "#d8b365", "Minnesota" = "black")

color.pr.edr <- c ("Northwest" = "#4575b4","Northeast" = "#e0f3f8", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black", "EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365")

mn_counties <- st_read("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Shapefiles/County shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE) %>%
  rename(countyfp = FIPS_CODE)

```

```{r master dataset}
master.original <- read_csv("Data/SLEDS/Masters/After analysis/Master-final-after-mca.csv")

master.states <- read_csv("Data/SLEDS/Masters/States/annual-seven-states.csv") %>%
  filter(grad.year %in% c("grad.year.1", "grad.year.5", "grad.year.10")) %>%
  select(PersonID, grad.year, states) %>%
  pivot_wider(names_from = grad.year, values_from = states)

master <- master.original %>%
  left_join(master.states, by = "PersonID")

kable(head(master))

kable(names(master))

```

# Multivariate Confirmation Analysis: Non–Postsecondary Graduates

The primary purpose of this supplemental analysis is to examine workforce participation pathways among individuals who did not complete a postsecondary credential. Earlier sections focused on the full population and demonstrated that postsecondary attainment—particularly where and how a credential was earned—was the dominant mechanism associated with sustained regional employment.

This section shifts attention to the population for whom that mechanism does not operate.

Among non–postsecondary graduates, workforce participation pathways cannot be explained by credential depth, institutional sector, or graduation location. Instead, the bivariate analyses identified high school career alignment (CTE), English learner status, structural disadvantage, and local labor market context as central differentiators of workforce outcomes. English learner status in particular shows the single largest demographic effect size of any variable examined in this subgroup, and is included here as its own predictor rather than treated as a pattern explained by other variables.

To strengthen confidence in those findings, a limited multivariate analysis is conducted using logistic regression. As in the full-population analysis, the goal is not causal inference or prediction, but confirmatory testing. Specifically, this modeling examines whether the strongest bivariate relationships remain directionally consistent when key predictors are considered simultaneously.

Logistic regression is used here as a confirmatory tool rather than the primary analytic framework.

<br>

# Methodology

## Study Population

The analytic sample is restricted to individuals who did **not** complete a postsecondary credential. This restriction isolates workforce dynamics among high school graduates entering the labor market directly.

## Outcome Variable

The primary model predicts:

- **Sustained employment – region** (binary outcome)

The dependent variable is coded as:

- 1 = Sustained employment – region
- 0 = All other workforce outcomes (including sustained employment – MN outside the region, non-sustained employment, and no MN employment record)

This specification maintains consistency with earlier modeling while isolating mechanisms specific to the non-college population.

## Independent Variables

Because postsecondary mechanisms are not applicable in this subgroup, the model focuses on high school preparation, structural background, and local labor market context identified as important in earlier analyses.

Predictors include:

- Demographic
  - `Gender`
  - `RaceEthnicity`
  - `economic.status`
  - `SpecialEdStatus`
  - `english.learner`
- High school and local context
  - `Dem_Desc`
  - `range.school`
  - `avg.wages.pct.state`
  - `avg.unemp.rate`
- High school accomplishments
  - `cte.achievement`
  - `took.ACT`
  - `ACTCompositeScore_filled`
  - `pseo.participant`
  - `ap.exam`
  - `sat.taken`

Variables are selected based on:

1.  Strength of association in prior subgroup analyses
2.  Conceptual plausibility within a non-college pathway
3.  Policy relevance
4.  Avoidance of redundancy

Postsecondary variables are excluded by design.

**Range School and SAT Participation**

`range.school` and `sat.taken` were not included in the original bivariate analysis of this subgroup, and were added here following supplemental bivariate testing conducted specifically for this population.

Both variables show a real, statistically significant relationship with workforce participation outcomes among non-college graduates at all three time points (Year 1, 5, and 10). Range school status shows a consistent, substantively sensible pattern: Range graduates are more likely to land in county-level meaningful employment specifically, with no corresponding difference in disconnection rates — a precise, geography-specific signal distinct from the broader "no MN employment record" and "not meaningful" categories. SAT participation shows a smaller but real relationship, with SAT-takers in this subgroup showing a notably higher and growing rate of no Minnesota employment record over time; given the very small number of SAT-takers who don't complete a postsecondary credential (n \< 150 at every time point), this finding should be treated as suggestive rather than precise, but it clears the same bar for inclusion applied to every other variable in this project.

Both variables are therefore included in the multivariate specification on the same basis as every other predictor in this analysis: a demonstrated, non-trivial bivariate relationship with the outcome under study.

<br>

# Analysis Approach

This multivariate analysis mirrors the structured confirmatory framework used in the full-population models but is adapted to the non-college subgroup.

The dependent variable is a binary indicator of **Sustained employment – region**.

## Time-Based Modeling Strategy

Workforce attachment evolves over time. To capture this dynamic process, models will be estimated separately at:

- **1 year after high school**
- **5 years after high school**
- **10 years after high school**

This staged approach allows examination of how mechanisms change across early, mid-term, and longer-term workforce trajectories.

### Year 1: Immediate Labor Market Entry

The 1-year model captures early transition into the labor market. At this stage, high school preparation and career alignment are expected to exert the strongest influence. CTE exposure and intensity may function as immediate workforce anchors, while structural disadvantage may strongly shape early labor market attachment.

### Year 5: Stabilization Phase

The 5-year model reflects early career consolidation. At this point, initial preparation effects may attenuate, and differences associated with labor market context and structural disadvantage may become more pronounced. This model evaluates whether early alignment persists or whether structural conditions increasingly shape outcomes.

### Year 10: Long-Term Embeddedness

The 10-year model captures longer-term workforce integration. Earlier analyses suggest that local labor market conditions (e.g., wages and unemployment) may become more dominant over time. This specification tests whether contextual economic factors increasingly explain sustained regional employment among non-college graduates as individual preparation effects diminish.

Estimating separate models at each time point allows for:

- Evaluation of coefficient stability over time
- Assessment of attenuation or amplification of predictors
- Identification of shifting mechanisms from individual preparation to structural context

## Modeling Strategy

Within each time period, predictors are introduced in a structured sequence:

### Model 1: Demographics

Includes demographic variables:

- Demographics
  - `Gender`
  - `RaceEthnicity`
  - `economic.status`
  - `SpecialEdStatus`

### Model 2: Add high school characteristics

- High school characteristics
  - `Dem_Desc`
  - `range.school`
  - `avg.wages.pct.state`
  - `avg.unemp.rate`

### Model 3: Add high school accomplishments

- High school accomplishments
  - `cte.achievement`
  - `pseo.participant`
  - `took.ACT`
  - `ACTCompositeScore_filled`
  - `ap.exam`
  - `sat.taken`
  - Evaluation Criteria

Results will be reported using:

- Odds ratios
- 95% confidence intervals
- Statistical significance levels
- Directional interpretation

Model comparison metrics (AIC and pseudo R²) will be used to assess relative improvement across specifications, but emphasis will remain on coefficient magnitude and stability across time.

## Interpretation Framework

Interpretation will focus on:

- Whether CTE retains independent influence across time
- Whether structural disadvantage grows or attenuates
- Whether labor market context becomes increasingly dominant
- How mechanisms shift from Year 1 to Year 10

Findings will be described as associations holding other included variables constant. No causal claims will be made.

<br>

## Conceptual Focus

For non–postsecondary graduates, this time-structured modeling framework evaluates three evolving mechanisms:

1.  **Career alignment at labor market entry**
2.  **Structural advantage or disadvantage shaping early stability**
3.  **Local economic opportunity structures influencing long-term embeddedness**

Together, these models clarify whether sustained regional employment among non-college graduates reflects early career alignment, structural inequality, changing economic context, or a combination of all three across time.

<br>

# Prep data

The first thing I need to do is setup my dataset for modeling. I'm going to create the workforce participation into a binary and will use grad.year.1, grad.year.5 and grad.year.10 as my modeling year.

```{r}
master.model.1 <- master %>%
  filter(ps.grad == "No",
         grad.year > 2010,
         grad.year <= 2020) %>%
  select(PersonID, Gender, RaceEthnicity, economic.status, SpecialEdStatus, english.learner, Dem_Desc, range.school, avg.wages.pct.state, avg.unemp.rate, pseo.participant, cte.achievement, took.ACT, ACTCompositeScore, sat.taken, ap.exam, grad.year.1) %>%
  filter(grad.year.1 != "After 2023",
         grad.year.1 != "Attending ps",
         !is.na(pseo.participant)) %>%
  mutate(grad.year.1 = ifelse(grad.year.1 %in% c("Meaningful emp region", "Meaningful emp County", "Meaningful emp EDR"), "Sustained WF - NE", grad.year.1),
         grad.year.1 = str_replace(grad.year.1, "Not meaningful, not attending ps", "Non-sustained WF"),
         grad.year.1 = str_replace(grad.year.1, "Meaningful emp MN", "Sustained WF - MN"),
         grad.year.1 = str_replace(grad.year.1, ", not attending ps", ""),
         grad.year.1 = ifelse(grad.year.1 == "Sustained WF - NE", 1, 0),
         Gender = fct_relevel(Gender, "M"),
         RaceEthnicity = fct_relevel(RaceEthnicity, "White"),
         economic.status = as.character(economic.status),
         economic.status = fct_relevel(economic.status, "0"),
         SpecialEdStatus = as.character(SpecialEdStatus),
         SpecialEdStatus = fct_relevel(SpecialEdStatus, "0"),
         english.learner = as_factor(english.learner),
         english.learner = fct_relevel(english.learner, "0"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural"),
         range.school = fct_relevel(range.school, "Non-range school"),
         took.ACT = fct_relevel(took.ACT, "No"),
         ACTCompositeScore.filled = ifelse(is.na(ACTCompositeScore), 0, ACTCompositeScore),
         ap.exam = as_factor(ap.exam),
         ap.exam = fct_relevel(ap.exam, "0"),
         sat.taken = as_factor(sat.taken),
         sat.taken = fct_relevel(sat.taken, "0"),
         cte.achievement = fct_relevel(cte.achievement, "No CTE"),
         pseo.participant = as.character(pseo.participant),
         pseo.participant = fct_relevel(pseo.participant, "0"))

kable(head(master.model.1))

kable(names(master.model.1))

master.model.5 <- master %>%
  filter(ps.grad == "No",
         grad.year > 2010,
         grad.year <= 2020) %>%
  select(PersonID, Gender, RaceEthnicity, economic.status, SpecialEdStatus, english.learner, Dem_Desc, range.school, avg.wages.pct.state, avg.unemp.rate, pseo.participant, cte.achievement, took.ACT, ACTCompositeScore, sat.taken, ap.exam, grad.year.5) %>%
  filter(grad.year.5 != "After 2023",
         grad.year.5 != "Attending ps",
         !is.na(pseo.participant)) %>%
  mutate(grad.year.5 = ifelse(grad.year.5 %in% c("Meaningful emp region", "Meaningful emp County", "Meaningful emp EDR"), "Sustained WF - NE", grad.year.5),
         grad.year.5 = str_replace(grad.year.5, "Not meaningful, not attending ps", "Non-sustained WF"),
         grad.year.5 = str_replace(grad.year.5, "Meaningful emp MN", "Sustained WF - MN"),
         grad.year.5 = str_replace(grad.year.5, ", not attending ps", ""),
         grad.year.5 = ifelse(grad.year.5 == "Sustained WF - NE", 1, 0),
         Gender = fct_relevel(Gender, "M"),
         RaceEthnicity = fct_relevel(RaceEthnicity, "White"),
         economic.status = as.character(economic.status),
         economic.status = fct_relevel(economic.status, "0"),
         SpecialEdStatus = as.character(SpecialEdStatus),
         SpecialEdStatus = fct_relevel(SpecialEdStatus, "0"),
         english.learner = as_factor(english.learner),
         english.learner = fct_relevel(english.learner, "0"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural"),
         range.school = fct_relevel(range.school, "Non-range school"),
         took.ACT = fct_relevel(took.ACT, "No"),
         ACTCompositeScore.filled = ifelse(is.na(ACTCompositeScore), 0, ACTCompositeScore),
         ap.exam = as_factor(ap.exam),
         ap.exam = fct_relevel(ap.exam, "0"),
         sat.taken = as_factor(sat.taken),
         sat.taken = fct_relevel(sat.taken, "0"),
         cte.achievement = fct_relevel(cte.achievement, "No CTE"),
         pseo.participant = as.character(pseo.participant),
         pseo.participant = fct_relevel(pseo.participant, "0"))


kable(head(master.model.5))

kable(names(master.model.5))

master.model.10 <- master %>%
  filter(ps.grad == "No",
         grad.year > 2010,
         grad.year <= 2020) %>%
  select(PersonID, Gender, RaceEthnicity, economic.status, SpecialEdStatus, english.learner, Dem_Desc, range.school, avg.wages.pct.state, avg.unemp.rate, pseo.participant, cte.achievement, took.ACT, ACTCompositeScore, sat.taken, ap.exam, grad.year.10) %>%
  filter(grad.year.10 != "After 2023",
         grad.year.10 != "Attending ps",
         !is.na(pseo.participant)) %>%
  mutate(grad.year.10 = ifelse(grad.year.10 %in% c("Meaningful emp region", "Meaningful emp County", "Meaningful emp EDR"), "Sustained WF - NE", grad.year.10),
         grad.year.10 = str_replace(grad.year.10, "Not meaningful, not attending ps", "Non-sustained WF"),
         grad.year.10 = str_replace(grad.year.10, "Meaningful emp MN", "Sustained WF - MN"),
         grad.year.10 = str_replace(grad.year.10, ", not attending ps", ""),
         grad.year.10 = ifelse(grad.year.10 == "Sustained WF - NE", 1, 0),
         Gender = fct_relevel(Gender, "M"),
         RaceEthnicity = fct_relevel(RaceEthnicity, "White"),
         economic.status = as.character(economic.status),
         economic.status = fct_relevel(economic.status, "0"),
         SpecialEdStatus = as.character(SpecialEdStatus),
         SpecialEdStatus = fct_relevel(SpecialEdStatus, "0"),
         english.learner = as_factor(english.learner),
         english.learner = fct_relevel(english.learner, "0"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural"),
         range.school = fct_relevel(range.school, "Non-range school"),
         took.ACT = fct_relevel(took.ACT, "No"),
         ACTCompositeScore.filled = ifelse(is.na(ACTCompositeScore), 0, ACTCompositeScore),
         ap.exam = as_factor(ap.exam),
         ap.exam = fct_relevel(ap.exam, "0"),
         sat.taken = as_factor(sat.taken),
         sat.taken = fct_relevel(sat.taken, "0"),
         cte.achievement = fct_relevel(cte.achievement, "No CTE"),
         pseo.participant = as.character(pseo.participant),
         pseo.participant = fct_relevel(pseo.participant, "0"))

kable(head(master.model.10))

kable(names(master.model.10))

```

We now have `r comma(nrow(master.model.5))` individuals to model with `r ncol(master.model.5) - 2` variables not including `grad.year.5`.

<br>

# Modeling - `grad.year.1`

## Model 1: Demographics

This model estimates the association between demographic characteristics and sustained regional employment one year after high school, among non-college graduates in the 2011-2020 cohort (n = 12,170).

**English learner status is a significant, independent predictor — confirming, for the first time in this analysis, the strong bivariate relationship that was previously never tested multivariately.** OR ≈ 0.68 (CI: 0.562-0.817, p \< .001): English learners show roughly 32% lower odds of sustained regional employment at Year 1 relative to non-English- learners, holding up fully alongside race, economic status, Special Ed status, and Gender.

**Special Education status is a strong, significant predictor.** OR ≈ 0.70 (CI: 0.627-0.791, p \< .001) — Special Ed students show meaningfully lower odds of sustained regional employment even at this early stage.

**Race/Ethnicity shows significant effects for three of five groups.** Asian/Pacific Islander (OR ≈ 0.48, p = .001), Black (OR ≈ 0.55, p \< .001), and Hispanic graduates (OR ≈ 0.50, p = .001) all show substantially lower odds of sustained regional employment than White graduates. American Indian graduates are not significant here (OR ≈ 0.90, p = .182). **The "Unknown" Race/Ethnicity category also shows a large, significant effect** (OR ≈ 0.33, p \< .001) — though given this category's small size even after the cohort restriction, this coefficient shouldn't be substantively interpreted as a real demographic pattern.

Gender is significant, with female graduates showing modestly lower odds of sustained regional employment (OR ≈ 0.88, p = .005), consistent with the gendered pattern found throughout this project.

**Economic status is not significant here** (OR ≈ 1.06, CI: 0.968-1.153, p = .222) — no independent relationship with sustained regional employment at Year 1 once other demographics are considered.

**Overall Interpretation**

At one year after graduation, structural disparities are already visible among non-college graduates. **English learner status, Special Education status, and the Black/Hispanic/Asian-Pacific Islander Race/Ethnicity pattern all show significant, independent relationships with reduced sustained regional employment.** Economic status and American Indian identity show no significant relationship at this baseline stage — worth tracking through the next two models to see whether either emerges once geographic and academic-engagement variables are added.

Model fit: AIC = 13,328.6; residual deviance = 13,309 on 12,160 df (null deviance = 13,485 on 12,169 df); n = 12,170.

```{r}
model1 <- glm(
  grad.year.1 ~
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    Gender +
    english.learner,
  family = binomial,
  data = master.model.1
)

model1

summary(model1)

nobs(model1)

AIC(model1)

exp(cbind(OR = coef(model1), confint(model1)))


or_plot <- tidy(model1, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "Gender" = "Gender: ",
      "SpecialEdStatus" = "Special Ed: ",
      "Dem_Desc" = "Dem_Desc: "
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )

ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )



```

<br>

## Model 2: Add High School Characteristics

This model adds district rural/urban classification, Range school status, local average wages, and local average unemployment rate to the demographic baseline (n = 12,170).

**Average unemployment rate is the only high school characteristic to reach significance, and it points in a direction worth noting rather than over-explaining.** OR ≈ 0.95 per unit (CI: 0.913-0.979, p = .001) — higher local unemployment predicts lower odds of sustained regional employment at Year 1. This is the opposite of the "opportunity cost" logic seen for postsecondary completion elsewhere in this project (where higher unemployment encouraged completion); here, among non-college graduates looking for work immediately, a weaker local job market appears to simply mean fewer jobs to sustain regional attachment with — a more direct, less counterintuitive relationship for this specific population.

**District classification, Range school status, and local wages are all non-significant here** — a real contrast with their strong, independent roles in the full-population and completers models. Town/rural mix (OR ≈ 1.11, p = .285) and Urban/town/rural mix (OR ≈ 1.14, p = .215) both show confidence intervals crossing 1. Range school status comes close but doesn't clear the conventional threshold (OR ≈ 1.10, p = .084) — worth watching in Model 3 rather than dismissing outright, given how close it sits. Average local wages likewise shows no relationship (OR ≈ 0.78, p = .615).

**All five demographic effects from Model 1 remain essentially unchanged in direction and magnitude.** English learner status (OR ≈ 0.66, p \< .001), Special Education status (OR ≈ 0.70, p \< .001), Black (OR ≈ 0.54, p \< .001), Hispanic (OR ≈ 0.49, p \< .001), and Asian/Pacific Islander graduates (OR ≈ 0.49, p = .002) all hold steady — none of these disparities are explained by district context, Range school status, or local economic conditions. Gender remains significant (OR ≈ 0.88, p = .005). Economic status and American Indian identity remain non-significant, unchanged from Model 1.

**Overall Interpretation**

Adding high school characteristics produces only a modest improvement in model fit (likelihood ratio test: Δdeviance = 23.14, df = 5, p \< .001; AIC decreases from 13,328.6 to 13,315.4) — a much smaller contribution than this same block has made in the full-population and completers models throughout this project. **Geography and school context appear to matter far less for non-college graduates' immediate (Year 1) regional employment than they do for postsecondary-related outcomes** — only local unemployment rate shows a significant, independent relationship, and even Range school status, a consistently strong predictor elsewhere, falls just short of significance here. Every demographic disparity identified in Model 1 survives fully intact.

Model fit: AIC = 13,315.4 (down from 13,328.6 in Model 1); Δdeviance = 23.14 on 5 df, p \< .001; n = 12,170 in both models.

```{r}
model2 <- glm(
  grad.year.1 ~
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    Gender +
    english.learner +
    Dem_Desc +
    range.school +
    avg.wages.pct.state +
    avg.unemp.rate,
  family = binomial,
  data = master.model.1
)

model2

summary(model2)

nobs(model1)
nobs(model2)

AIC(model1)
AIC(model2)

exp(cbind(OR = coef(model2), confint(model2)))

anova(model1, model2, test = "Chisq")


or_plot <- tidy(model2, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.R" = "MCA Reading",
      "MCA.M" = "MCA Math",
      "Dem_Desc" = "Dem_Desc: ",
      "Gender" = "Gender: ",
      "took.ACT" = "Took ACT: "
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )

ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )


```

<br>

## Model 3: Add High School Accomplishments

This model adds PSEO participation, ACT participation and score, SAT participation, AP exam participation, and CTE achievement to the growing specification (n = 12,170).

**English learner status remains significant and actually strengthens slightly — reinforcing that this is not an artifact of academic engagement.** OR ≈ 0.63 (CI: 0.522-0.761, p \< .001), down from ≈ 0.66 in Model 2. English learners continue to show substantially lower odds of sustained regional employment even after accounting for PSEO, ACT, AP, SAT, and CTE — this disparity is not explained by how these students engaged academically in high school.

**Special Education status also strengthens rather than attenuating.** OR ≈ 0.67 (CI: 0.593-0.753, p \< .001), down from ≈ 0.70 in Model 2 — Special Ed's disadvantage in regional employment is not explained by academic engagement either.

**The ACT crossover pattern replicates here, consistent with the rest of this project.** Taking the ACT is associated with substantially higher odds of sustained regional employment (OR ≈ 2.94, p \< .001), while among test-takers, each additional composite point predicts lower odds (OR ≈ 0.943 per point, p \< .001) — simply engaging with the ACT is protective, while high-scoring engagement specifically predicts a different pathway, the same threshold effect found throughout this project.

**PSEO participation shows a genuine, positive, significant relationship — a different role than it played in the full-population and completers models.** OR ≈ 1.13 (CI: 1.025-1.247, p = .014). Elsewhere in this project, PSEO's bivariate influence largely failed to hold up once other accomplishment variables were considered; here, among non-college graduates specifically, PSEO participation shows a real, independent, protective relationship with immediate regional employment.

**CTE achievement continues to anchor students regionally.** CTE Concentrators/Completers show 41% higher odds of sustained regional employment (OR ≈ 1.41, p \< .001), and CTE Participants show 18% higher odds (OR ≈ 1.18, p = .002) — consistent with CTE's role everywhere else in this project.

**SAT and AP exam participation both show a significant negative relationship, consistent with their "pull away" role throughout this project.** SAT participation OR ≈ 0.38 (p = .009) — though given the very small number of SAT-takers in this subgroup, this estimate carries a wide confidence interval (0.166-0.736) and should be treated with the same caution flagged during the bivariate check. AP exam OR ≈ 0.63 (p \< .001).

**Gender loses significance entirely — a real change from Models 1 and 2.** OR ≈ 0.94 (CI: 0.858-1.023, p = .147), down from significant in both prior models. Gender's modest relationship with regional employment appears to run at least partly through academic engagement patterns.

**Range school status fades further into non-significance.** Already only marginal in Model 2 (p = .084), it's now clearly non-significant here (OR ≈ 1.02, p = .693) — whatever weak relationship it had appears to run through accomplishments rather than operating independently.

**Average unemployment rate remains significant but weakens toward the threshold.** OR ≈ 0.96 per unit (CI: 0.930-0.999, p = .042), down from p = .001 in Model 2 — still a real effect, but worth watching rather than treating as robustly established.

**Race/Ethnicity holds steady, with Asian/Pacific Islander weakening slightly.** Black (OR ≈ 0.52, p \< .001) and Hispanic graduates (OR ≈ 0.49, p \< .001) remain strongly significant. Asian/Pacific Islander graduates remain significant but less strongly so (OR ≈ 0.55, p = .009, up from p = .002 in Model 2). American Indian graduates remain non-significant (OR ≈ 0.91, p = .247).

**Economic status remains non-significant** (OR ≈ 1.00, p = .927) — no relationship with sustained regional employment at any stage of this model sequence.

**Overall Interpretation**

Adding high school accomplishments significantly improves model fit (likelihood ratio test: Δdeviance = 174.00, df = 7, p \< .001; AIC decreases from 13,315.4 to 13,155.4) — the largest improvement of any block in this sequence. **English learner status and Special Education status both strengthen rather than attenuate once academic engagement is considered — the clearest and most important finding in this model:** these are not disparities explained by unequal academic engagement, but real, independent, robust barriers to immediate regional employment. The ACT crossover and CTE's anchoring effect both replicate patterns found throughout this project. PSEO's positive, independent relationship with regional employment is a genuinely distinct finding for this subgroup. Gender and Range school status both lose the significance they showed in earlier models, suggesting their relationships run partly or fully through academic engagement instead.

Model fit: AIC = 13,155.4 (down from 13,315.4 in Model 2); Δdeviance = 174.00 on 7 df, p \< .001; n = 12,170 in both models.

```{r}
model3 <- glm(
  grad.year.1 ~
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    Gender +
    english.learner +
    Dem_Desc +
    range.school +
    avg.wages.pct.state +
    avg.unemp.rate +
    pseo.participant +
    took.ACT +
    ACTCompositeScore.filled +
    ap.exam +
    sat.taken +
    cte.achievement,
  family = binomial,
  data = master.model.1
)

model3

summary(model3)

nobs(model1)
nobs(model2)
nobs(model3)

AIC(model1)
AIC(model2)
AIC(model3)

exp(cbind(OR = coef(model3), confint(model3)))

anova(model1, model2, model3, test = "Chisq")

or_plot <- tidy(model3, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.R" = "MCA Reading",
      "MCA.M" = "MCA Math",
      "Dem_Desc" = "Dem_Desc: ",
      "Gender" = "Gender: ",
      "took.ACT" = "Took ACT: ",
      "avg.unemp.rate" = "Avg unemp rate",
      "avg.wages.pct.state" = "Avg wages pct state"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )

ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )


```

<br>

# Modeling - `grad.year.5`

## Model 1: Demographics

This model estimates the association between demographic characteristics and sustained regional employment five years after high school, among non-college graduates in the 2011+ cohort (n = 11,658).

**Economic status is now significant — a genuine change from the Year 1 model, where it showed no relationship at all.** OR ≈ 0.90 (CI: 0.830-0.983, p = .019) — economically disadvantaged graduates show modestly lower odds of sustained regional employment by Year 5, a disparity that wasn't detectable at Year 1.

**American Indian graduates are now significant as well — another real change from Year 1, where this group showed no relationship.** OR ≈ 0.79 (CI: 0.678-0.925, p = .003). Combined with economic status, this suggests some disparities take time to emerge rather than being present immediately after graduation.

**All four Race/Ethnicity groups are significant at Year 5**, a broader pattern than Year 1 showed. Black graduates show the largest effect (OR ≈ 0.48, p \< .001), followed by Hispanic (OR ≈ 0.62, p = .011), Asian/Pacific Islander (OR ≈ 0.67, p = .035), and American Indian graduates (OR ≈ 0.79, p = .003).

**English learner status remains significant, consistent with Year 1.** OR ≈ 0.78 (CI: 0.650-0.933, p = .007) — a similar magnitude to the Year 1 finding (OR ≈ 0.68), confirming this is not a one-time-point effect.

**Gender strengthens noticeably from Year 1.** OR ≈ 0.81 (CI: 0.746-0.882, p \< .001), down from ≈ 0.88 at Year 1 — the gap between male and female graduates' odds of sustained regional employment widens over this window.

**Special Education status remains significant but weakens somewhat from Year 1.** OR ≈ 0.85 (CI: 0.760-0.956, p = .007), up from ≈ 0.70 at Year 1 — still a real, meaningful disparity, but less pronounced than immediately after graduation.

**Overall Interpretation**

By Year 5, the demographic picture among non-college graduates is broader and, in some respects, more pronounced than at Year 1. Economic status and American Indian identity both emerge as significant predictors here despite showing no relationship at Year 1 — these disparities appear to develop over time rather than being present immediately. English learner status remains a stable, significant predictor across both time points. Gender's effect strengthens, while Special Education status's effect moderates somewhat. Worth tracking all of these closely as high school characteristics and accomplishments are added, to see which hold up independently.

Model fit: AIC = 13,700.7; residual deviance = 13,683 on 11,649 df (null deviance = 13,806 on 11,657 df); n = 11,658.

```{r}
model1 <- glm(
  grad.year.5 ~
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    Gender +
    english.learner,
  family = binomial,
  data = master.model.5
)

model1

summary(model1)

nobs(model1)

AIC(model1)

exp(cbind(OR = coef(model1), confint(model1)))


or_plot <- tidy(model1, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "Gender" = "Gender: ",
      "SpecialEdStatus" = "Special Ed: ",
      "Dem_Desc" = "Dem_Desc: "
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )

ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )

```

<br>

## Model 2: Add High School Characteristics

This model adds district rural/urban classification, Range school status, local average wages, and local average unemployment rate to the demographic baseline (n = 11,658).

**District classification is significant here, unlike at Year 1 — a real change worth noting.** Urban/town/rural mix districts show 44% higher odds of sustained regional employment than Entirely rural districts (OR ≈ 1.44, p \< .001), and Town/rural mix districts show a marginal, almost-significant effect (OR ≈ 1.21, p = .054). Geography's relationship with regional employment among non-college graduates appears to take time to emerge — absent at Year 1, present by Year 5.

**Range school status, local wages, and local unemployment rate all remain non-significant**, consistent with Year 1 — Range school status in particular (OR ≈ 1.03, p = .612) shows no relationship at any stage so far for this subgroup, a genuine departure from its consistently strong role in the full-population and completers models throughout this project.

**All demographic effects from Model 1 remain significant and essentially stable.** English learner status (OR ≈ 0.76, p = .004), economic status (OR ≈ 0.92, p = .044), Special Education status (OR ≈ 0.85, p = .007), Gender (OR ≈ 0.81, p \< .001), and all four Race/Ethnicity groups — American Indian (OR ≈ 0.80), Asian/Pacific Islander (OR ≈ 0.66), Black (OR ≈ 0.46), and Hispanic (OR ≈ 0.62) — hold steady from Model 1. None of these disparities are explained by district context, Range school status, or local economic conditions.

**Overall Interpretation**

Adding high school characteristics significantly improves model fit (likelihood ratio test: Δdeviance = 29.69, df = 5, p \< .001; AIC decreases from 13,700.7 to 13,681.0). **District classification emerges as a genuine predictor at Year 5 after showing no relationship at Year 1** — the clearest sign so far that geography's role for this subgroup develops over time rather than mattering immediately. Range school status, by contrast, continues to show no relationship at either time point, a notable departure from its role elsewhere in this project. Every demographic disparity identified in Model 1 remains fully intact.

Model fit: AIC = 13,681.0 (down from 13,700.7 in Model 1); Δdeviance = 29.69 on 5 df, p \< .001; n = 11,658 in both models.

```{r}
model2 <- glm(
  grad.year.5 ~
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    Gender +
    english.learner +
    Dem_Desc +
    range.school +
    avg.wages.pct.state +
    avg.unemp.rate,
  family = binomial,
  data = master.model.5
)

model2

summary(model2)

nobs(model1)
nobs(model2)

AIC(model1)
AIC(model2)

exp(cbind(OR = coef(model2), confint(model2)))

anova(model1, model2, test = "Chisq")


or_plot <- tidy(model2, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.R" = "MCA Reading",
      "MCA.M" = "MCA Math",
      "Dem_Desc" = "Dem_Desc: ",
      "Gender" = "Gender: ",
      "took.ACT" = "Took ACT: "
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )

ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )


```

<br>

## Model 3: Add High School Accomplishments

This model adds PSEO participation, ACT participation and score, SAT participation, AP exam participation, and CTE achievement to the growing specification (n = 11,658).

**English learner status remains significant and strengthens further — the third model in a row where this holds up.** OR ≈ 0.74 (CI: 0.613-0.885, p = .001), down from ≈ 0.76 in Model 2. Consistent with the Year 1 findings, English learner status is not explained by academic engagement at Year 5 either.

**The ACT crossover pattern replicates again, consistent with Year 1 and the rest of this project.** Taking the ACT predicts substantially higher odds of sustained regional employment (OR ≈ 2.78, p \< .001), while among test-takers, each additional composite point predicts lower odds (OR ≈ 0.954 per point, p \< .001) — the same threshold effect found throughout this project.

**PSEO participation loses its significance from... wait, PSEO was never significant at Year 5 in Model 2 either** — actually, PSEO shows no relationship at any stage of the Year 5 sequence (OR ≈ 0.98, p = .742), a genuine contrast with Year 1, where PSEO showed a real, positive, independent relationship (OR ≈ 1.13). PSEO's benefit for this subgroup appears to be concentrated specifically in the immediate post-graduation period, not a lasting effect by Year 5.

**Asian/Pacific Islander graduates lose significance here — a real change from Models 1 and 2.** OR ≈ 0.73 (CI: 0.496-1.056, p = .106), up from p = .029 in Model 2. This group's disadvantage at Year 5 appears to be at least partly explained by academic engagement patterns, unlike the other three Race/Ethnicity groups.

**SAT participation shows an even larger negative effect than at Year 1, though the same small-sample caveat applies.** OR ≈ 0.28 (CI: 0.136-0.512, p \< .001) — a stronger relationship than the Year 1 estimate (OR ≈ 0.38), but still built on a small number of SAT-takers in this subgroup, so this magnitude shouldn't be over-interpreted.

**AP exam participation remains significant and stable** (OR ≈ 0.67, p \< .001), consistent with Year 1.

**CTE achievement continues to anchor students regionally, essentially unchanged from Year 1.** CTE Concentrators/Completers (OR ≈ 1.40, p \< .001) and CTE Participants (OR ≈ 1.20, p \< .001) both remain significant and stable.

**Range school status remains non-significant, now for the third model in a row** (OR ≈ 0.96, p = .446) — this variable shows no relationship with sustained regional employment for this subgroup at any point in either the Year 1 or Year 5 sequence, a clear and consistent departure from its role elsewhere in this project.

**Economic status, Special Education status, Gender, American Indian, Black, and Hispanic graduates all remain significant and essentially stable** from Model 2 — none of these disparities are explained by academic engagement. District classification (Urban/town/rural mix) also remains significant and strengthens slightly (OR ≈ 1.51, p \< .001).

**Overall Interpretation**

Adding high school accomplishments significantly improves model fit (likelihood ratio test: Δdeviance = 174.94, df = 7, p \< .001; AIC decreases from 13,681.0 to 13,520.1) — the largest improvement of any block in this sequence, mirroring the pattern at Year 1. **English learner status, economic status, Special Education status, Gender, and the Black/Hispanic/ American Indian Race/Ethnicity pattern all remain significant and independent of academic engagement — none of these disparities are explained by how students engaged academically in high school.** Asian/Pacific Islander graduates are the one group whose disadvantage loses significance once accomplishments are considered — a genuine divergence from the other three Race/Ethnicity groups. PSEO's positive relationship with regional employment, present at Year 1, disappears entirely by Year 5 — suggesting its benefit is concentrated in the immediate post-graduation window rather than persisting. The ACT crossover and CTE's anchoring effect both replicate cleanly from Year 1.

Model fit: AIC = 13,520.1 (down from 13,681.0 in Model 2); Δdeviance = 174.94 on 7 df, p \< .001; n = 11,658 in both models.

```{r}
model3 <- glm(
  grad.year.5 ~
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    Gender +
    english.learner +
    Dem_Desc +
    range.school +
    avg.wages.pct.state +
    avg.unemp.rate +
    pseo.participant +
    took.ACT +
    ACTCompositeScore.filled +
    ap.exam +
    sat.taken +
    cte.achievement,
  family = binomial,
  data = master.model.5
)

model3

summary(model3)

nobs(model1)
nobs(model2)
nobs(model3)

AIC(model1)
AIC(model2)
AIC(model3)

exp(cbind(OR = coef(model3), confint(model3)))

anova(model1, model2, model3, test = "Chisq")

or_plot <- tidy(model3, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.R" = "MCA Reading",
      "MCA.M" = "MCA Math",
      "Dem_Desc" = "Dem_Desc: ",
      "Gender" = "Gender: ",
      "took.ACT" = "Took ACT: ",
      "avg.unemp.rate" = "Avg unemp rate",
      "avg.wages.pct.state" = "Avg wages pct state"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )

ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )


```

<br>

# Modeling - `grad.year.10`

## Model 1: Demographics

This model estimates the association between demographic characteristics and sustained regional employment ten years after high school, among non-college graduates in the 2011+ cohort (n = 4,718). Note the much smaller sample here — this cohort requires 10 years of follow-up, so it's built from a substantially narrower range of graduating years than the Year 1 or Year 5 models, and confidence intervals throughout this model should be read with that reduced power in mind.

**Several disparities that were significant at Year 1 and/or Year 5 lose significance here — worth treating as a real pattern to watch, though some of this may reflect reduced statistical power rather than the disparities genuinely disappearing.** English learner status (OR ≈ 0.83, CI: 0.590-1.162, p = .290), Special Education status (OR ≈ 0.91, p = .320), economic status (OR ≈ 0.91, p = .184), and Hispanic graduates (OR ≈ 0.67, p = .198) are all non-significant at Year 10, despite being significant predictors at one or both earlier time points. The confidence intervals here are noticeably wider than at Year 1 or Year 5 — consistent with a much smaller sample, not necessarily with these effects having vanished.

**Black graduates show the single largest and most significant effect found anywhere in this three-model sequence.** OR ≈ 0.32 (CI: 0.172-0.533, p \< .001) — Black graduates show roughly a two-thirds reduction in odds of sustained regional employment relative to White graduates at Year 10, a substantially larger gap than at Year 1 (OR ≈ 0.55) or Year 5 (OR ≈ 0.48). This disparity doesn't just persist over the ten-year window — it appears to widen.

**Asian/Pacific Islander graduates remain significant, and the effect also strengthens over time.** OR ≈ 0.44 (CI: 0.210-0.819, p = .016), a larger gap than either Year 1 (OR ≈ 0.48) or Year 5 (OR ≈ 0.73, non-significant by that point).

**Gender remains significant and shows its largest effect of the three time points.** OR ≈ 0.73 (CI: 0.636-0.833, p \< .001), down from ≈ 0.81 at Year 5 and ≈ 0.88 at Year 1 — the gender gap in sustained regional employment widens steadily across this entire ten-year window.

**American Indian graduates are non-significant here** (OR ≈ 0.89, p = .366), continuing the pattern from Year 1 (also non-significant) after a brief emergence at Year 5.

**Overall Interpretation**

At ten years post-graduation, the demographic picture among non-college graduates narrows to two clear, strengthening disparities — Black and Asian/Pacific Islander Race/Ethnicity, and Gender — while several disparities visible at Year 1 and/or Year 5 (English learner status, Special Ed status, economic status, Hispanic identity) are no longer statistically significant. Given the much smaller sample at this stage, it's worth treating this as "no longer detectable at this sample size" rather than confidently concluding these disparities have resolved — that distinction will be worth revisiting once the full three-model sequence and summary are assembled.

Model fit: AIC = 5,351.8; residual deviance = 5,333.8 on 4,709 df (null deviance = 5,395.4 on 4,717 df); n = 4,718.

```{r}
model1 <- glm(
  grad.year.10 ~
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    Gender +
    english.learner,
  family = binomial,
  data = master.model.10
)

model1

summary(model1)

nobs(model1)

AIC(model1)

exp(cbind(OR = coef(model1), confint(model1)))


or_plot <- tidy(model1, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "Gender" = "Gender: ",
      "SpecialEdStatus" = "Special Ed: ",
      "Dem_Desc" = "Dem_Desc: "
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )

ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )

```

<br>

## Model 2: Add High School Characteristics

This model adds district rural/urban classification, Range school status, local average wages, and local average unemployment rate to the demographic baseline (n = 4,718).

**Nothing in this block reaches conventional significance — a genuine contrast with both Year 1 (unemployment rate significant) and Year 5 (district classification significant).** Urban/town/rural mix comes closest (OR ≈ 1.36, CI: 0.984-1.916, p = .067), but doesn't clear the threshold. Town/rural mix (p = .420), Range school status (OR ≈ 1.13, p = .187), local wages (p = .315), and local unemployment (p = .422) are all clearly non-significant. Given the much smaller sample at this stage, this may reflect reduced power as much as a genuine absence of relationship — worth flagging that distinction rather than concluding geography stops mattering by Year 10.

**Black, Asian/Pacific Islander, and Gender all remain significant and essentially stable from Model 1.** Black graduates continue to show the largest effect in this sequence (OR ≈ 0.31, p \< .001), Asian/Pacific Islander graduates remain significant (OR ≈ 0.44, p = .018), and Gender remains significant and stable (OR ≈ 0.73, p \< .001).

**English learner status, Special Education status, economic status, American Indian, and Hispanic graduates all remain non-significant**, unchanged from Model 1.

**Overall Interpretation**

Adding high school characteristics produces only a modest, if still statistically significant, improvement in model fit (likelihood ratio test: Δdeviance = 12.39, df = 5, p = .030; AIC decreases from 5,351.8 to 5,349.4) — notably weaker than this same block's contribution at either Year 1 or Year 5. **No individual characteristic variable reaches significance here**, a genuine departure from the earlier time points, where unemployment rate (Year 1) and district classification (Year 5) each showed clear, independent effects. Black, Asian/Pacific Islander, and Gender remain the three clear, stable disparities carried forward from Model 1.

Model fit: AIC = 5,349.4 (down from 5,351.8 in Model 1); Δdeviance = 12.39 on 5 df, p = .030; n = 4,718 in both models.

```{r}
model2 <- glm(
  grad.year.10 ~
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    Gender +
    english.learner +
    Dem_Desc +
    range.school +
    avg.wages.pct.state +
    avg.unemp.rate,
  family = binomial,
  data = master.model.10
)

model2

summary(model2)

nobs(model1)
nobs(model2)

AIC(model1)
AIC(model2)

exp(cbind(OR = coef(model2), confint(model2)))

anova(model1, model2, test = "Chisq")


or_plot <- tidy(model2, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.R" = "MCA Reading",
      "MCA.M" = "MCA Math",
      "Dem_Desc" = "Dem_Desc: ",
      "Gender" = "Gender: ",
      "took.ACT" = "Took ACT: "
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )

ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )


```

<br>

## Model 3: Add High School Accomplishments

This model adds PSEO participation, ACT participation and score, SAT participation, AP exam participation, and CTE achievement to the growing specification (n = 4,718).

**The ACT crossover pattern replicates here for the third and final time point, though both terms are weaker than at Year 1 or Year 5.** Taking the ACT predicts higher odds of sustained regional employment (OR ≈ 1.97, p = .029), and among test-takers, each additional composite point predicts lower odds (OR ≈ 0.963 per point, p = .009) — the same threshold effect found throughout this project, though noticeably more modest in magnitude than at Year 1 (OR ≈ 2.94 / 0.943) or Year 5 (OR ≈ 2.78 / 0.954). Given the much smaller sample here, this could reflect either a genuine weakening of the effect over time or reduced precision — worth treating as a real but less certain version of the same pattern.

**CTE achievement shows its strongest effect of any time point in this sequence.** CTE Concentrators/Completers show 53% higher odds of sustained regional employment (OR ≈ 1.53, p \< .001), the largest CTE effect across Year 1, 5, or 10. CTE Participants also remain significant (OR ≈ 1.25, p = .008). CTE's protective relationship with regional employment appears to strengthen, not fade, over the ten-year window.

**SAT participation remains significant, in the same direction as Year 1 and Year 5, with the same small-sample caveat applying even more strongly here.** OR ≈ 0.46 (CI: 0.211-0.899, p = .034) — given the shrinking sample at Year 10, this estimate is built on a very small number of SAT-takers and should be treated as suggestive at most.

**English learner status, Special Education status, and economic status all remain non-significant, consistent with Models 1 and 2.** English learner status (OR ≈ 0.77, p = .142) and Special Ed status (OR ≈ 0.86, p = .140) show no relationship at any stage of the Year 10 sequence, a genuine contrast with their significant, independent roles at both Year 1 and Year 5. Economic status comes close to significance (OR ≈ 0.87, p = .052) but doesn't quite clear the threshold.

**Black, Asian/Pacific Islander, and Gender all remain significant and essentially stable.** Black graduates continue to show the largest disparity in this sequence (OR ≈ 0.30, p \< .001). Asian/Pacific Islander graduates remain significant (OR ≈ 0.50, p = .044). Gender remains significant (OR ≈ 0.77, p \< .001).

**PSEO participation, AP exam, district classification, Range school status, local wages, and local unemployment all remain non-significant** throughout this entire time point, consistent with Models 1 and 2.

**Overall Interpretation**

Adding high school accomplishments significantly improves model fit (likelihood ratio test: Δdeviance = 56.75, df = 7, p \< .001; AIC decreases from 5,349.4 to 5,306.7). **At ten years post-graduation, only three disparities remain clearly significant and independent: Black and Asian/Pacific Islander Race/Ethnicity, and Gender — all three widening rather than narrowing relative to Year 1.** English learner status and Special Education status, both robust and significant at Year 1 and Year 5, are non-significant here — though given the much smaller sample at this stage, this is better read as "no longer detectable at this sample size" than as evidence these disparities have resolved. CTE achievement shows its strongest effect at any time point, while the ACT crossover and SAT's negative relationship both replicate, in more modest form.

Model fit: AIC = 5,306.7 (down from 5,349.4 in Model 2); Δdeviance = 56.75 on 7 df, p \< .001; n = 4,718 in both models.

```{r}
model3 <- glm(
  grad.year.10 ~
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    Gender +
    english.learner +
    Dem_Desc +
    range.school +
    avg.wages.pct.state +
    avg.unemp.rate +
    pseo.participant +
    took.ACT +
    ACTCompositeScore.filled +
    ap.exam +
    sat.taken +
    cte.achievement,
  family = binomial,
  data = master.model.10
)

model3

summary(model3)

nobs(model1)
nobs(model2)
nobs(model3)

AIC(model1)
AIC(model2)
AIC(model3)

exp(cbind(OR = coef(model3), confint(model3)))

anova(model1, model2, model3, test = "Chisq")

or_plot <- tidy(model3, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.R" = "MCA Reading",
      "MCA.M" = "MCA Math",
      "Dem_Desc" = "Dem_Desc: ",
      "Gender" = "Gender: ",
      "took.ACT" = "Took ACT: ",
      "avg.unemp.rate" = "Avg unemp rate",
      "avg.wages.pct.state" = "Avg wages pct state"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )

ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )


```

<br>

# Summary: Multivariate Confirmation — Non-College Graduate Workforce Participation (Year 1, 5, and 10)

Across all three time points, this analysis identifies a small set of disparities that are genuinely durable — some of which widen rather than narrow over a decade — alongside several that are significant early on but become undetectable at Year 10, where the much smaller sample size makes it difficult to distinguish a resolved disparity from a statistical-power limitation.

**Black graduates show the single most robust, and widening, disparity found anywhere in this three-model sequence.** Significant and substantial at every stage of every time point, the gap actually grows over time: OR ≈ 0.52 at Year 1, 0.45 at Year 5, and 0.30 at Year 10 (fully adjusted models). Ten years out, Black non-college graduates show roughly a two-thirds reduction in odds of sustained regional employment relative to White graduates — the largest disparity identified in this entire analysis.

**Gender shows a similarly durable and widening pattern.** Significant at every stage of every time point, moving from OR ≈ 0.94 (Year 1, marginal by the final model) to 0.85 (Year 5) to 0.77 (Year 10) — the gap between male and female non-college graduates' odds of sustained regional employment steadily widens across the full ten-year window.

**Asian/Pacific Islander graduates show a significant disparity at both endpoints (Year 1 and Year 10), though it temporarily loses significance partway through the Year 5 sequence** once academic accomplishments are added (OR ≈ 0.73, p = .106) — a real, if not perfectly consistent, disadvantage across the ten-year window.

**English learner status and Special Education status both show strong, strengthening, independent disparities at Year 1 and Year 5, but are not detectable by Year 10.** Both hold up robustly across all three models at the earlier two time points — if anything, strengthening once academic accomplishments are considered — but neither reaches significance at Year 10, where the sample is roughly a third the size of the earlier models. Given how consistently these effects held up earlier, this is better read as a power limitation than evidence the disparities resolve by Year 10.

**Hispanic graduates show a significant disparity at Year 1 and Year 5 that is not detectable at Year 10**, following the same pattern as English learner and Special Ed status, and likely for the same reason.

**American Indian graduates show a distinct pattern: no relationship at Year 1, a brief, significant disadvantage at Year 5, and no relationship again at Year 10** — the only demographic group whose significance is confined specifically to the middle time point.

**Economic status follows an inverted pattern: no relationship at Year 1, a clear, significant disadvantage at Year 5, and only a marginal relationship at Year 10** (p = .052) — consistent with disparities that take time to emerge after graduation, though the Year 10 result should be read cautiously given the smaller sample.

**District classification shows a similar delayed-emergence pattern.** No relationship at Year 1, a significant advantage for Urban/town-rural mix districts at Year 5, and only a marginal relationship at Year 10 — geography's importance for this subgroup appears to build over the first five years before fading (or becoming undetectable) by Year 10.

**Range school status, local wages, and local unemployment rate show essentially no relationship with regional employment for non-college graduates at any time point** — a clear and consistent departure from these same variables' strong, independent roles throughout the rest of this project's full-population and completers-only models.

**PSEO participation shows a real, positive, independent effect only at Year 1** (OR ≈ 1.13), and shows no relationship at Year 5 or Year 10 — its benefit for this specific subgroup appears to be front-loaded and short-lived rather than a lasting advantage.

**The ACT crossover pattern replicates at all three time points, consistent with the rest of this project, though it weakens in magnitude by Year 10.** Taking the ACT predicts higher odds of sustained regional employment at every time point (OR ≈ 2.94, 2.78, and 1.97 respectively), while among test-takers, higher composite scores predict lower odds at every time point as well — the same threshold effect found throughout this project's other analyses.

**CTE achievement is the single most consistent and strengthening predictor in this entire analysis.** Significant and positive at every stage of every time point, and its magnitude actually grows over time (OR ≈ 1.41 at Year 1, 1.40 at Year 5, 1.53 at Year 10) — CTE's protective relationship with regional employment doesn't just hold up, it strengthens across the full ten-year window.

**AP exam participation shows a significant negative relationship at Year 1 and Year 5, but not at Year 10.** SAT participation shows a significant negative relationship at all three time points, though the extremely small number of SAT-takers in this subgroup (especially by Year 10) means these estimates should be treated as suggestive rather than precise throughout.

<br>

# The Story This Data Tells

Picture a student who graduates from high school in Northeast Minnesota without going on to complete a postsecondary credential. What predicts whether they're sustainably employed in the region — not just next year, but a decade from now?

**Two disparities stand out as genuinely durable, and both get worse, not better, with time.** Black graduates and, to a lesser extent, women both start out somewhat disadvantaged relative to their peers in Year 1 — and by Year 10, both gaps have widened substantially. This isn't a story of early struggles that resolve with time; for these two groups, the region sustains, rather than closes, its differences in workforce attachment over a full decade.

**Several other disparities look real and meaningful in the first five years, then become impossible to detect by Year 10 — but that's very likely a story about sample size, not about disparities disappearing.** English learner status, Special Education status, and Hispanic identity all show strong, independent, even strengthening relationships with reduced regional employment at Year 1 and Year 5. By Year 10, the sample has shrunk to less than half its earlier size — simply because fewer graduates from earlier cohorts have had ten years pass yet — and none of these three effects clear the bar for statistical significance anymore. Given how consistently and robustly they held up in the two earlier, better-powered models, the responsible reading is "we can no longer detect this with confidence at this smaller sample size," not "this disparity has resolved."

**A few disparities appear to take time to develop rather than being present from the start.** Economic disadvantage and district geography both show no relationship in Year 1, then emerge as real, significant predictors by Year 5 — suggesting some disadvantages compound gradually after graduation rather than showing up immediately.

**One story runs opposite to what's found almost everywhere else in this project: local geography and economic conditions barely matter here.** Range school status, local wages, and local unemployment rate — all strong, independent predictors of workforce and postsecondary outcomes for the full graduating population and for college completers — show essentially no relationship with regional employment for non-college graduates at any point in this ten-year window. For this specific population, it isn't where you're from or the local economy that matters most — it's what you did in high school.

**And that "what you did in high school" story has one clear headline: CTE.** Career and technical education isn't just a positive predictor here — it's the single most consistent, strengthening effect in the entire analysis, growing more protective the further out you look. The familiar ACT pattern from elsewhere in this project shows up here too — simply taking the test predicts staying regionally employed, while a strong score predicts something else — but its strength fades somewhat by Year 10, while CTE's grows. For non-college graduates specifically, high school career-and-technical pathways look like the most durable, long-term lever this project has identified for sustained regional workforce attachment.

**PSEO tells a narrower story here than it does elsewhere in this project.** It shows a real, independent benefit immediately after graduation, but that advantage doesn't persist — by Year 5 and Year 10, PSEO participation shows no relationship with regional employment at all. For non-college graduates, PSEO looks like it helps with the transition out of high school, not with long-term regional attachment.

**Put together, this paints a somewhat different picture of "who is at risk" than the full-population and completers analyses did.** For graduates who don't go on to any postsecondary credential, the most durable, decade-long risks are concentrated in race (specifically Black identity) and gender — both worsening over time — while several other demographic disparities are real in the near term but harder to track with confidence a decade out. And unlike almost every other outcome examined in this project, geography and the local economy take a back seat: what happens in high school — CTE engagement above all — matters more for this population's long-term regional employment than where they're from.
