---
title: "Post-secondary pathways"
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
library(htmlwidgets)
library(RColorBrewer)
library(readxl)
library(janitor)
library(lubridate)
library(systemfonts)
reset_font_cache()
library(ggtext)
library(gmodels)
library(fastDummies)
library(car)
library(glmnet)
library(glmnetUtils)
library(pscl)
library(sjPlot)
library(ggforce)
library(openxlsx)
```

```{r themes and shapefiles, include=FALSE}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", linewidth  = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.key.size = unit(1, "lines"),
        legend.margin = margin(0,0,0,0),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.text = element_text(margin = margin(l = 2)),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", linewidth = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))


theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
    select(5,6) %>%
    unique() %>%
    mutate(edr = str_replace(edr, "  ", " "),
           planning.region = str_replace(planning.region, " Minnesota", ""),
           planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
           edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

counties.regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
  rename(mif = `MIF Region`) %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Name = str_replace(Name, "Mcleod", "McLeod"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc) ,
         edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"))

counties.regions.1 <- counties.regions %>%
  mutate(statefp = "27",
         project.pr = ifelse(edr %in% c("EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 4 - West Central"), "Northwest",
                             ifelse(edr == "EDR 3 - Arrowhead", "Northeast",
                                    ifelse(edr %in% c("EDR 5 - North Central", "EDR 7E- East Central", "EDR 7W- Central"), "Central",
                                           ifelse(edr %in% c("EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 8 - Southwest"), "Southwest",
                                                  ifelse(edr %in% c("EDR 9 - South Central", "EDR 10 - Southeast"), "Southern", as.character(planning.region)))))),
         project.pr = fct_relevel(project.pr, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southern"))


color.ruca <- c("Entirely rural" = "#009933", "Town/rural mix" = "#99CC33", "Urban/town/rural mix" = "#CC9966", "Entirely urban" = "#754C29", "Minnesota" = "black")

color.pr <- c("Northwest" = 	"#4575b4", "Northeast" = "grey", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.edr.simple <- c("EDR 1" = "#b3cde3", "EDR 2" = "#8c96c6", "EDR 3" = "#fe9929", "EDR 4" = "#8856a7", "EDR 5" = "#810f7c", "EDR 6E" = "#e5f5f9", "EDR 6W" = "#bdc9e1", "EDR 7E" = "#99d8c9", "EDR 7W" = "#2ca25f", "EDR 8" = "#74a9cf", "EDR 9" = "#0570b0", "EDR 10" = "#d7301f", "EDR 11" = "#d8b365", "Minnesota" = "black")

color.pr.edr <- c ("Northwest" = "#4575b4","Northeast" = "#e0f3f8", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black", "EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365")

mn_counties <- st_read("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Shapefiles/County shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE) %>%
  rename(countyfp = FIPS_CODE)
```

```{r master}
master.original <- read_csv("Data/SLEDS/Masters/After analysis/Master-after-ct.csv") 

kable(names(master.original))

kable(head(master.original))
```

# Dependent Variable

The cross-tab and ANOVA analysis exploring relationships between demographic, high school, and post-secondary variables with employment outcomes highlighted the importance of post-secondary variables in an individuals journey to meaningful employment inside and outside the region. Due to this, it makes sense to explore if some of the other variables play a role in post-secondary outcomes - whether an individual attends, graduates, the type of institution they attend and the location.

We will explore this here. We will explore the following post-secondary variables;

1.  `attended.ps` (ever attended postsecondary)
2.  `ps.grad` (graduated from postsecondary)
3.  `highest.cred.level`
4.  `ps.grad.location`

Each dependent variable has the following categories;

```{r dependent variable categories}
data <- master.original %>%
  select(attended.ps, ps.grad, highest.cred.level, ps.grad.location) %>%
  mutate(across(c("attended.ps":"ps.grad.location"), factor))
```

**attended.ps**

```{r echo = FALSE}
levels(data$attended.ps)
```

**ps.grad**

```{r echo = FALSE}
levels(data$ps.grad)
```

**highest.cred.level**

```{r echo = FALSE}
levels(data$highest.cred.level)
```

**ps.grad.location**

```{r echo = FALSE}
levels(data$ps.grad.location)
```

<br>

# Predictors

For all of these, we will use the predictors that were in Tier 1 and Tier 2 of the previous cross-tab analysis.

- Demographics
  - `Gender`
  - `economic.status`
  - `RaceEthnicity`
  - `SpecialEdStatus`
- High school characteristics
  - `Dem_Desc`
  - `avg.unemp.rate`
  - `avg.wage.pct.state`
- High school accomplishments
  - `ACTCompositeScore`
  - `pseo.participant`
  - `cte.achievement`
  - `total.cte.courses.taken`
  - `MCA.M`
  - `MCA.R`
  - `MCA.S`

<br>

# `attended.ps`

## Summary

Across all three variable groups — demographic characteristics, high school characteristics, and high school accomplishments — the factors most strongly associated with post-secondary attendance are individual-level characteristics, not school- or community-level ones.

Among demographic predictors, special education status and economic disadvantage stand out as the strongest barriers to post-secondary attendance, with gaps of 35 and 22 percentage points respectively. Gender and race/ethnicity are meaningful but secondary, and the race/ethnicity association is likely partially explained by overlap with economic disadvantage — particularly for American Indian students in the region.

High school characteristics contribute very little. District type has a negligible association with attendance, and while local unemployment and wage levels show statistically significant differences, the effect sizes are too small to be practically meaningful. Where a student goes to school matters far less than who that student is and what they do while they are there.

High school accomplishments tell the most differentiated story. PSEO participation is the single strongest predictor across all variables examined, followed closely by academic performance as measured by ACT scores and MCA proficiency in math, reading, and science. CTE engagement runs in the opposite direction — students with deeper CTE involvement are somewhat less likely to attend post-secondary — pointing to a distinct workforce-first orientation among that group that post-secondary attendance alone does not capture.

Taken together, these findings suggest that post-secondary attendance is shaped primarily by preparation and opportunity factors established well before graduation. Students who arrive at their senior year academically prepared, economically stable, and without a special education designation attend post-secondary at substantially higher rates. For students without those advantages, the gap in attendance is large and consistent across multiple measures.

<br>

## Demographics

All four demographic variables show a statistically significant relationship with post-secondary attendance, though effect sizes vary meaningfully across the group.

**Special education status** shows the strongest association (φ = 0.273). Students without a special education designation attended post-secondary at a rate of 73%, compared to 38% for students with a special education designation — a gap of 35 percentage points. The interpretation of this finding warrants some care, as post-secondary attendance may carry different meaning for special education students depending on individual transition plans and IEP goals.

**Economic status** is the second strongest predictor (φ = 0.228). Students not identified as economically disadvantaged attended post-secondary at 76%, compared to 54% for those who were — a 22-percentage-point gap. This association is likely intertwined with other demographic factors, particularly race/ethnicity, given the regional concentration of economically disadvantaged students among certain groups.

**Race and ethnicity** shows a statistically significant but more modest association (V = 0.164). White students had the highest post-secondary attendance rate at 72%, while American Indian students had the lowest at 52%. Other racial and ethnic groups clustered between 57% and 66%. A large "Unknown" category (n = 8,298) attended at 56%, and the non-random nature of that classification warrants caution in interpretation.

**Gender** shows the smallest effect in this group (φ = 0.121). Female graduates attended post-secondary at a rate of 74%, compared to 62% for male graduates — an 11-percentage-point gap that, while statistically significant, is modest relative to the other demographic predictors.

Taken together, special education status and economic status are the dominant demographic predictors of post-secondary attendance in this cohort. Gender and race/ethnicity are meaningful but secondary. The race/ethnicity association in particular should be interpreted alongside economic status, as overlap between these characteristics may account for a portion of the observed gap.

::: panel-tabset
### `Gender`

```{r}
tab_xtab(var.row = master.original$attended.ps,
         var.col = master.original$Gender,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/attended-ps/Demographics/Gender.html")
```

### `economic.status`

```{r}
tab_xtab(var.row = master.original$attended.ps,
         var.col = master.original$economic.status,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/attended-ps/Demographics/economic-status.html")
```

### `RaceEthnicity`

```{r}
tab_xtab(var.row = master.original$attended.ps,
         var.col = master.original$RaceEthnicity,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/attended-ps/Demographics/RaceEthnicity.html")
```

### `SpecialEdStatus`

```{r}
tab_xtab(var.row = master.original$attended.ps,
         var.col = master.original$SpecialEdStatus,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/attended-ps/Demographics/SpecialEdStatus.html")
```
:::

<br>

## High School Characteristics

Among the three high school characteristic variables examined, district demographic designation shows a statistically significant but negligible association with post-secondary attendance, while the two school-level economic indicators show small but significant differences across attendance groups.

**District demographic designation** (Cramer's V = 0.029) is technically statistically significant but the effect size is essentially zero — the weakest association in this entire analysis. Post-secondary attendance rates are nearly identical across school types: 67% for entirely rural districts, 66% for town/rural mix districts, and 69% for urban/town/rural mix districts. For practical purposes, the type of district a student graduates from has no meaningful relationship with whether they attend post-secondary education.

**Average local unemployment rate** shows a small but statistically significant difference across attendance groups (F = 428.4, p \< .001). Students who attended post-secondary came from school districts with an average unemployment rate approximately 0.29 percentage points lower than those who did not attend. While the difference is statistically significant given the large sample size, the magnitude is modest and should be interpreted with caution.

**Average local wage as a percentage of the state average** also shows a small but statistically significant difference (F = 65.8, p \< .001). Students who attended post-secondary came from districts where average wages were about 0.4 percentage points higher relative to the state average compared to non-attendees. As with unemployment rate, the effect is real but small.

Taken together, high school characteristics as measured here contribute very little to explaining post-secondary attendance. The near-zero effect for district type is particularly notable given that geography played a strong role in workforce outcomes in the prior analysis — suggesting that the type of community a student comes from matters more for what happens after post-secondary than for whether they attend in the first place.

::: panel-tabset
### `Dem_Desc`

```{r}
tab_xtab(var.row = master.original$attended.ps,
         var.col = master.original$Dem_Desc,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/attended-ps/High school characteristics/Dem_Desc.html")
```

### `avg.unemp.rate`

```{r}
summary <- aov(avg.unemp.rate ~ attended.ps, data = master.original)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/attended-ps/High school characteristics/avg-unemp-rate.xlsx", overwrite = TRUE)


```

### `avg.wages.pct.state`

```{r}
summary <- aov(avg.wages.pct.state ~ attended.ps, data = master.original)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/attended-ps/High school characteristics/avg-wage-pct-state.xlsx", overwrite = TRUE)
```
:::

<br>

## High School Accomplishments

High school accomplishment variables show the widest range of effect sizes in this analysis, with PSEO participation and academic performance measures standing out as the strongest predictors of post-secondary attendance.

**PSEO participation** is the strongest predictor in this group (φ = 0.279). Students who participated in PSEO during high school attended post-secondary at a rate of 80%, compared to 53% for non-participants — a 27-percentage-point gap. This is unsurprising given that PSEO students are already enrolled in college coursework while in high school, but the magnitude of the association confirms that PSEO participation is a robust early signal of post-secondary attendance.

**ACT Composite Score** shows a large and statistically significant difference across attendance groups (F = 739.9, p \< .001). Students who attended post-secondary scored approximately 2.0 points higher on the ACT than those who did not. Given that ACT scores are only available for test-takers — a selected population — this association likely understates the true relationship between academic preparation and post-secondary attendance.

The three **MCA proficiency scores** — math, reading, and science — all show statistically significant differences across attendance groups, with students who attended post-secondary scoring higher on each measure. The estimated differences are 0.35 proficiency levels for math, 0.44 for reading, and 0.39 for science (all p \< .001). These are consistent in direction and meaningful in magnitude, reinforcing the broader pattern that academic preparation in high school is closely linked to post-secondary attendance.

**CTE achievement level** shows a statistically significant but weak association (Cramer's V = 0.071). Notably, the direction of the relationship is not straightforward — CTE concentrators and completers attended post-secondary at a lower rate (63%) than both CTE participants (70%) and students with no CTE involvement (70%). This likely reflects the vocational orientation of CTE concentrators, many of whom may be pursuing workforce entry directly rather than post-secondary education.

**Total CTE courses taken** shows a small but statistically significant difference (F = 565.7, p \< .001). Students who did not attend post-secondary actually took more CTE courses on average — about 0.54 more courses than those who did attend. This is consistent with the CTE achievement finding and reinforces the interpretation that deeper CTE engagement is associated with a workforce-first pathway rather than post-secondary attendance.

Taken together, academic preparation variables — PSEO participation, ACT scores, and MCA proficiency — are the dominant high school accomplishment predictors of post-secondary attendance. CTE engagement tells a different story: greater CTE involvement is modestly associated with lower post-secondary attendance, suggesting these students are on a distinct trajectory that the post-secondary attendance outcome may not fully capture.

::: panel-tabset
### `ACTCompositeScore`

```{r}
summary <- aov(ACTCompositeScore ~ attended.ps, data = master.original)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/attended-ps/High school accomplishments/ACTCompositeScore.xlsx", overwrite = TRUE)
```

### `pseo.participation`

```{r}
tab_xtab(var.row = master.original$attended.ps,
         var.col = master.original$pseo.participant,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/attended-ps/High school accomplishments/pseo-participant.html")
```

### `cte.achievement`

```{r}
tab_xtab(var.row = master.original$attended.ps,
         var.col = master.original$cte.achievement,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/attended-ps/High school accomplishments/cte-achievement.html")
```

### `total.cte.courses.taken`

```{r}
summary <- aov(total.cte.courses.taken ~ attended.ps, data = master.original)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/attended-ps/High school accomplishments/total-cte-courses-taken.xlsx", overwrite = TRUE)
```

### `MCA.M`

```{r}
summary <- aov(MCA.M ~ attended.ps, data = master.original)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/attended-ps/High school accomplishments/MCA-M.xlsx", overwrite = TRUE)
```

### `MCA.R`

```{r}
summary <- aov(MCA.R ~ attended.ps, data = master.original)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/attended-ps/High school accomplishments/MCA-R.xlsx", overwrite = TRUE)

```

### `MCA.S`

```{r}
summary <- aov(MCA.S ~ attended.ps, data = master.original)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/attended-ps/High school accomplishments/MCA-S.xlsx", overwrite = TRUE)
```
:::

<br>

# `ps.grad`

## Summary

The predictors of post-secondary graduation closely parallel those for attendance, but with a notable pattern: effect sizes are consistently larger for graduation than for attendance across nearly every variable examined. The factors that shape whether students enroll also shape whether they complete — and in most cases, they shape completion even more strongly.

Among demographic predictors, special education status and economic disadvantage remain the dominant barriers, with graduation gaps of 35 and 27 percentage points respectively. The gender gap widens slightly at graduation relative to attendance, suggesting that male students who do enroll face additional headwinds in completing a credential. Race and ethnicity disparities are particularly stark at the graduation stage — American Indian and Black students graduate at roughly 21% and 22% respectively, compared to 49% for White students. These are not small differences, and they persist even after accounting for the fact that all students in this analysis had already graduated high school.

High school characteristics again contribute very little. District type is essentially unrelated to graduation, and while local labor market conditions show statistically significant associations, the effect sizes are negligible. The community context in which a student attended high school does not meaningfully predict whether they complete post-secondary education.

High school accomplishments tell the clearest and most consistent story. PSEO participation shows the strongest association of any variable in this analysis (φ = 0.308), with a 30-percentage-point graduation gap between participants and non-participants. ACT scores and MCA proficiency measures — particularly reading — show meaningful differences that are somewhat larger at graduation than at attendance, pointing to the sustained role of academic preparation in degree completion. CTE engagement again runs counter to the post-secondary completion pattern, with deeper CTE involvement associated with lower graduation rates — a finding consistent with a workforce-first orientation established well before students leave high school.

Taken together, post-secondary graduation appears to be shaped by many of the same forces as attendance, but those forces exert even greater influence at the completion stage. Students who arrive at post-secondary without strong academic preparation, without economic stability, or without early college experience face compounding disadvantages that extend from enrollment all the way through to credential attainment.

<br>

## Demographics

All four demographic variables show a statistically significant relationship with post-secondary graduation, and the pattern of effect sizes closely mirrors what was observed for attendance — with special education status and economic disadvantage again emerging as the strongest predictors.

**Special education status** shows the strongest association in this group (φ = 0.262). Among students without a special education designation, 48% graduated from a post-secondary institution, compared to just 13% of students with a special education designation — a 35-percentage-point gap. As with attendance, this finding should be interpreted with care given that post-secondary graduation may not align with the goals established in many students' transition plans.

**Economic status** is the second strongest predictor (φ = 0.270). Students not identified as economically disadvantaged graduated at a rate of 53%, compared to 26% for those who were — a 27-percentage-point gap. Notably, this gap is somewhat wider than the one observed for attendance, suggesting that economic disadvantage may present ongoing barriers to degree completion beyond the initial decision to enroll.

**Race and ethnicity** shows a moderate association (Cramer's V = 0.169). White students graduated at a rate of 49%, while American Indian and Black students had the lowest graduation rates at 21% and 22% respectively. Hispanic students graduated at 31%, and Asian/Pacific Islander students at 45%. The gaps between White students and American Indian and Black students are particularly pronounced and warrant careful attention in the narrative.

**Gender** shows a meaningful and widened association compared to attendance (φ = 0.134). Female graduates completed post-secondary at a rate of 49%, compared to 36% for male graduates — a 13-percentage-point gap. The gender gap in graduation is somewhat larger than the gap in attendance, suggesting that male students who enroll are less likely to complete.

Taken together, the demographic predictors of post-secondary graduation closely track those for attendance, but the gaps in several cases are wider. Economic disadvantage, special education status, and race/ethnicity in particular appear to present compounding barriers — not just to entering post-secondary education, but to completing it.

::: panel-tabset
### `Gender`

```{r}
data <- master.original %>%
  filter(ps.grad != "Attending ps")

tab_xtab(var.row = data$ps.grad,
         var.col = data$Gender,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/ps-grad/Demographics/Gender.html")
```

### `economic.status`

```{r}
data <- master.original %>%
  filter(ps.grad != "Attending ps")

tab_xtab(var.row = data$ps.grad,
         var.col = data$economic.status,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/ps-grad/Demographics/economic-status.html")
```

### `RaceEthnicity`

```{r}
data <- master.original %>%
  filter(ps.grad != "Attending ps",
         RaceEthnicity != "Unknown")

tab_xtab(var.row = data$ps.grad,
         var.col = data$RaceEthnicity,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/ps-grad/Demographics/RaceEthnicity.html")
```

### `SpecialEdStatus`

```{r}
data <- master.original %>%
  filter(ps.grad != "Attending ps")

tab_xtab(var.row = data$ps.grad,
         var.col = data$SpecialEdStatus,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/ps-grad/Demographics/SpecialEdStatus.html")
```
:::

<br>

## High School Characteristics

The pattern for high school characteristics and post-secondary graduation closely mirrors what was observed for attendance — school-level variables contribute very little explanatory power, with effect sizes that are statistically significant but negligible in practical terms.

**District demographic designation** again shows the weakest association in this group (Cramer's V = 0.022). Graduation rates are nearly identical across district types: 39% for entirely rural, 42% for town/rural mix, and 43% for urban/town/rural mix. As with attendance, the type of district a student graduates from has no meaningful relationship with post-secondary graduation.

**Average local unemployment rate** shows a small but statistically significant difference across graduation groups (F = 572.0, p \< .001). Students who graduated from a post-secondary institution came from districts with average unemployment rates approximately 0.33 percentage points lower than non-graduates. The effect is consistent in direction with the attendance finding but remains modest in magnitude.

**Average local wage as a percentage of the state average** also shows a small but statistically significant difference (F = 43.4, p \< .001). Students who graduated post-secondary came from districts where wages were about 0.003 percentage points higher relative to the state average — an effect so small it is essentially negligible despite reaching statistical significance.

Taken together, high school characteristics remain weak predictors of post-secondary graduation, just as they were for attendance. The community a student comes from does not meaningfully determine whether they complete a post-secondary credential, reinforcing that individual-level factors carry the explanatory weight in this analysis.

::: panel-tabset
### `Dem_Desc`

```{r}
tab_xtab(var.row = data$ps.grad,
         var.col = data$Dem_Desc,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/ps-grad/High school characteristics/Dem_Desc.html")
```

### `avg.unemp.rate`

```{r}
summary <- aov(avg.unemp.rate ~ ps.grad, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/ps-grad/High school characteristics/avg-unemp-rate.xlsx", overwrite = TRUE)

```

### `avg.wages.pct.state`

```{r}
summary <- aov(avg.wages.pct.state ~ ps.grad, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/ps-grad/High school characteristics/avg-wage-pct-state.xlsx", overwrite = TRUE)
```
:::

<br>

## High School Accomplishments

::: panel-tabset
### `ACTCompositeScore`

```{r}
summary <- aov(ACTCompositeScore ~ ps.grad, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/ps-grad/High school accomplishments/ACTCompositeScore.xlsx", overwrite = TRUE)
```

### `pseo.participant`

```{r}
tab_xtab(var.row = data$ps.grad,
         var.col = data$pseo.participant,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/ps-grad/High school accomplishments/pseo-participant.html")
```

### `cte.achievement`

```{r}
tab_xtab(var.row = data$ps.grad,
         var.col = data$cte.achievement,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/ps-grad/High school accomplishments/cte-achievement.html")
```

### `total.cte.courses.taken`

```{r}
summary <- aov(total.cte.courses.taken ~ ps.grad, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/ps-grad/High school accomplishments/total-cte-courses-taken.xlsx", overwrite = TRUE)
```

### `MCA.M`

```{r}
summary <- aov(MCA.M ~ ps.grad, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/ps-grad/High school accomplishments/MCA-M.xlsx", overwrite = TRUE)
```

### `MCA.R`

```{r}
summary <- aov(MCA.R ~ ps.grad, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/ps-grad/High school accomplishments/MCA-R.xlsx", overwrite = TRUE)

```

### `MCA.S`

```{r}
summary <- aov(MCA.S ~ ps.grad, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/ps-grad/High school accomplishments/MCA-S.xlsx", overwrite = TRUE)
```
:::

<br>

# `highest.cred.level`

## Summary

The predictors of highest credential level tell a richer story than attendance or graduation alone, because the outcome captures not just whether students complete but how far they go. The patterns established in earlier outcomes persist here, but the Tukey HSD results add meaningful texture about where in the credential spectrum the key distinctions lie.

Demographic factors continue to show strong associations, with economic status and special education status as the dominant predictors. Economically disadvantaged students who attend post-secondary are heavily concentrated in the "some college" category — nearly half stop short of any credential. Special education students show an even more extreme pattern, with the vast majority exiting without completing and very few reaching a bachelor's degree. The gender gap widens at the credential level, with women substantially more likely to attain a bachelor's degree or higher and men more likely to stop at some college or exit with less than an associate degree. Race and ethnicity shows a weaker overall association at this stage, but the concentration of American Indian and Black students in the "some college" category — at 55% and 62% respectively — is a finding that deserves attention in its own right.

High school characteristics remain negligible. District type, local unemployment, and local wages contribute nothing meaningful to explaining credential level, consistent with every other outcome examined in this analysis.

High school accomplishments again provide the strongest and most differentiated signal. PSEO participation stands out as the single strongest predictor, with its primary influence concentrated at the bachelor's degree level rather than shorter credentials — PSEO participants and non-participants attain associate degrees at nearly identical rates, but diverge sharply at the bachelor's degree level. ACT scores produce the clearest ordered gradient across all four credential categories, distinguishing not just completers from non-completers but bachelor's degree completers from associate degree completers from those who exit without a credential. MCA proficiency scores in math, reading, and science follow a consistent pattern — bachelor's degree completers are distinguishable from all other groups, but the some college and less-than-associate-degree groups are largely indistinguishable from each other, suggesting that incomplete participation is not primarily a function of academic preparation.

CTE engagement continues to tell a distinctly different story. Deeper CTE involvement is consistently associated with shorter credential pathways and lower rates of bachelor's degree attainment — a pattern now confirmed across attendance, graduation, and credential level. This is not a negative finding about CTE; it reflects a coherent workforce-first orientation among students with deeper CTE engagement, whose post-secondary trajectories may be better evaluated against labor market outcomes than credential attainment alone.

<br>

## Demographics

This analysis examines highest credential level attained among post-secondary attendees who are no longer enrolled, including students who stopped short of completing a credential. The four credential categories — bachelor's degree or higher, associate degree, less than associate degree, and some college — capture the full range of post-secondary attainment, from completion to partial participation.

**Economic status** shows the strongest association in this group (Cramer's V = 0.264). Among students not identified as economically disadvantaged, 46% attained a bachelor's degree or higher, compared to just 20% for economically disadvantaged students. The "some college" category tells the complementary story — 49% of economically disadvantaged students stopped at some college without completing a credential, compared to 27% of non-disadvantaged students. Economic disadvantage is strongly associated not just with whether students complete, but with how far they get when they do attend.

**Special education status** shows a similarly strong association (Cramer's V = 0.233). Among non-special-ed students, 41% attained a bachelor's degree or higher and 31% stopped at some college. Among special education students, the distribution shifts dramatically — just 7% attained a bachelor's degree or higher, while 62% stopped at some college and 21% reached less than an associate degree. Special education students who attend post-secondary are far more likely to exit without a credential than their non-special-ed peers.

**Gender** shows a moderate association (Cramer's V = 0.148). Female students were more likely to attain a bachelor's degree or higher (44% vs. 32% for males) and less likely to stop at some college (29% vs. 38%). Males were also more likely to exit with less than an associate degree (15% vs. 10%). The gender gap in credential attainment is consistent in direction with the gaps observed for attendance and graduation, and adds further detail — men who attend post-secondary are more likely to leave without completing at any level.

**Race and ethnicity** shows a weaker but still statistically significant association (Cramer's V = 0.088). Asian/Pacific Islander students had the highest rate of bachelor's degree attainment at 53%, followed by White students at 41%. American Indian and Black students had the lowest bachelor's degree attainment rates at 20% and 20% respectively, and the highest rates of stopping at some college — 55% for American Indian students and 62% for Black students. These patterns are consistent with the graduation disparities observed earlier and reinforce that racial and ethnic gaps in post-secondary outcomes extend beyond enrollment to credential attainment.

Taken together, economic status and special education status again dominate as the strongest demographic predictors, and the credential-level detail adds an important layer to the story told by graduation rates alone. A substantial share of economically disadvantaged and special education students do reach post-secondary — but they are far more likely to leave without a credential, concentrated in the "some college" category. This suggests that the barriers these students face are not only at the point of entry, but persist throughout their post-secondary experience.

::: panel-tabset
### `Gender`

```{r}
data <- master.original %>%
  filter(!highest.cred.level %in% c("Attending ps", "No college"))

tab_xtab(var.row = data$highest.cred.level,
         var.col = data$Gender,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/highest-cred-level/Demographics/Gender.html")
```

### `economic.status`

```{r}
data <- master.original %>%
  filter(!highest.cred.level %in% c("Attending ps", "No college"))

tab_xtab(var.row = data$highest.cred.level,
         var.col = data$economic.status,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/highest-cred-level/Demographics/economic-status.html")
```

### `RaceEthnicity`

```{r}
data <- master.original %>%
  filter(!highest.cred.level %in% c("Attending ps", "No college"),
         RaceEthnicity != "Unknown")

tab_xtab(var.row = data$highest.cred.level,
         var.col = data$RaceEthnicity,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/highest-cred-level/Demographics/RaceEthnicity.html")
```

### `SpecialEdStatus`

```{r}
data <- master.original %>%
  filter(!highest.cred.level %in% c("Attending ps", "No college"))

tab_xtab(var.row = data$highest.cred.level,
         var.col = data$SpecialEdStatus,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/highest-cred-level/Demographics/SpecialEdStatus.html")
```
:::

<br>

## High School Characteristics

As with the previous outcomes, high school characteristics show negligible associations with highest credential level attained, and the Tukey HSD results add important nuance to what the overall ANOVA tests detect.

**District demographic designation** again produces a statistically significant but practically meaningless association (Cramer's V = 0.027). Credential distributions are nearly identical across district types. Entirely rural districts show a slightly higher share of students stopping at some college (38%) compared to town/rural mix (32%) and urban/town/rural mix (33%) districts, and a lower share reaching a bachelor's degree or higher (34% vs. 39% and 40%). While these differences are real, the effect size is too small to carry meaningful interpretive weight.

**Average local unemployment rate** shows a statistically significant overall ANOVA result (F = 7.69, p \< .001), but the Tukey HSD results reveal that the only meaningful pairwise differences involve the "some college" group. Students who stopped at some college came from districts with modestly higher average unemployment rates than bachelor's degree completers (estimate = 0.077, p \< .001) and less-than-associate-degree completers (estimate = 0.116, p \< .001). No significant differences were found among the credential-completing groups themselves. Local unemployment rate is essentially unrelated to the level of credential attained among completers.

**Average local wage as a percentage of the state average** also reaches statistical significance overall (F = 5.64, p \< .001), but again the Tukey HSD results show limited practical differentiation. Bachelor's degree completers and less-than-associate-degree completers came from districts with modestly higher wages relative to the state average compared to associate degree completers, but the differences are very small and the "some college" group does not differ significantly from bachelor's degree completers. The wage associations here do not form a clear or interpretable gradient across credential levels.

Taken together, the community and school context in which a student attended high school has no meaningful relationship with the credential level they ultimately attain. This finding is consistent across all three outcomes examined so far — attendance, graduation, and credential level — and reinforces that individual-level factors, not school or community characteristics, are the primary drivers of post-secondary attainment in this region.

::: panel-tabset
### `Dem_Desc`

```{r}
tab_xtab(var.row = data$highest.cred.level,
         var.col = data$Dem_Desc,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/highest-cred-level/High school characteristics/Dem_Desc.html")
```

### `avg.unemp.rate`

```{r}
summary <- aov(avg.unemp.rate ~ highest.cred.level, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/highest-cred-level/High school characteristics/avg.unemp.rate.xlsx", overwrite = TRUE)

```

### `avg.wages.pct.state`

```{r}
summary <- aov(avg.wages.pct.state ~ highest.cred.level, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/highest-cred-level/High school characteristics/avg.wage.pct.state.xlsx", overwrite = TRUE)
```
:::

<br>

## High School Accomplishments

High school accomplishment variables show strong and highly differentiated associations with highest credential level, with clear gradients across credential categories for nearly every predictor. The Tukey HSD results add important texture, revealing where the meaningful distinctions lie.

**PSEO participation** is again the strongest predictor in this group (Cramer's V = 0.308). Among PSEO participants, 51% attained a bachelor's degree or higher, compared to just 22% of non-participants. Non-participants were far more likely to stop at some college (45% vs. 24%) or exit with less than an associate degree (17% vs. 10%). PSEO participation shows virtually no difference in associate degree attainment between the two groups (16% vs. 16%), suggesting its primary influence is on bachelor's degree completion rather than shorter-term credential attainment.

**ACT Composite Score** shows the largest and most clearly ordered effect among the continuous predictors (F = 1138.1, p \< .001). Tukey HSD results reveal a sharp and statistically significant gradient: bachelor's degree completers scored approximately 3.4 points higher than associate degree completers, who in turn scored about 0.9 points higher than less-than-associate-degree completers. The "some college" group scored similarly to associate degree completers — consistent with the idea that this group had the academic preparation to attend but not to complete. Every pairwise comparison except associate degree vs. some college is statistically significant, indicating that ACT score differentiates credential levels across the full attainment spectrum.

**MCA Math proficiency** shows a similarly ordered gradient (F = 977.1, p \< .001). Bachelor's degree completers scored approximately 0.48 proficiency levels higher than associate degree completers, who scored about 0.23 levels higher than less-than-associate-degree completers. The some college group scored similarly to the less-than-associate-degree group, with no significant difference between them — suggesting that among students who attended without completing, math proficiency does not distinguish those who got close from those who stopped short.

**MCA Reading proficiency** follows the same pattern (F = 1062.4, p \< .001). Bachelor's degree completers scored approximately 0.33 proficiency levels higher than associate degree completers, with statistically significant differences across all pairwise comparisons except some college vs. less than associate degree. Reading proficiency distinguishes completers from non-completers more than it distinguishes among non-completers.

**MCA Science proficiency** shows a nearly identical pattern (F = 887.0, p \< .001). Bachelor's degree completers scored approximately 0.43 proficiency levels higher than associate degree completers, with significant differences across credential levels. The some college and less-than-associate-degree groups do not differ significantly from each other in science proficiency, consistent with the math and reading findings.

**CTE achievement level** shows a moderate and substantively interesting association (Cramer's V = 0.135). CTE concentrators and completers are most heavily concentrated in the some college (39%) and less-than-associate-degree (18%) categories, and least likely to attain a bachelor's degree or higher (27%). Students with no CTE involvement show the inverse pattern — 48% attained a bachelor's degree or higher, and only 30% stopped at some college. CTE participants fall between the two extremes. This reinforces the now-consistent finding that deeper CTE engagement is associated with a pathway oriented toward shorter-term or workforce credentials rather than four-year degree completion.

**Total CTE courses taken** shows a significant and clearly ordered association with credential level (F = 442.6, p \< .001). Bachelor's degree completers took approximately 0.73 fewer CTE courses than associate degree completers, while less-than-associate-degree completers took about 0.59 more courses than associate degree completers, and the some college group took about 0.84 more courses than bachelor's degree completers. Every pairwise comparison is statistically significant, and the gradient is consistent — more CTE coursework is associated with lower credential attainment, while less CTE coursework is associated with higher attainment.

Taken together, high school accomplishments show strong, ordered, and consistent associations with highest credential level. Academic preparation variables — particularly PSEO participation and ACT scores — produce the clearest gradients, with bachelor's degree completers distinguishable from all other groups. CTE engagement again points in the opposite direction, with deeper involvement associated with shorter credential pathways. The "some college" group consistently resembles the less-than-associate-degree group on academic preparation measures, suggesting that incomplete post-secondary participation is not primarily explained by a lack of academic readiness but by other factors not captured in these variables.

::: panel-tabset
### `ACTCompositeScore`

```{r}
summary <- aov(ACTCompositeScore ~ highest.cred.level, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/highest-cred-level/High school accomplishments/ACTCompositeScore.xlsx", overwrite = TRUE)
```

### `pseo.participant`

```{r}
tab_xtab(var.row = data$highest.cred.level,
         var.col = data$pseo.participant,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/highest-cred-level/High school accomplishments/pseo-participant.html")
```

### `cte.achievement`

```{r}
tab_xtab(var.row = data$highest.cred.level,
         var.col = data$cte.achievement,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/highest-cred-level/High school accomplishments/cte-achievement.html")
```

### `total.cte.courses.taken`

```{r}
summary <- aov(total.cte.courses.taken ~ highest.cred.level, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/highest-cred-level/High school accomplishments/total-cte-courses-taken.xlsx", overwrite = TRUE)
```

### `MCA.M`

```{r}
summary <- aov(MCA.M ~ highest.cred.level, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/highest-cred-level/High school accomplishments/MCA-M.xlsx", overwrite = TRUE)
```

### `MCA.R`

```{r}
summary <- aov(MCA.R ~ highest.cred.level, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/highest-cred-level/High school accomplishments/MCA-R.xlsx", overwrite = TRUE)

```

### `MCA.S`

```{r}
summary <- aov(MCA.S ~ highest.cred.level, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/highest-cred-level/High school accomplishments/MCA-S.xlsx", overwrite = TRUE)
```
:::

<br>

# `ps.grad.location`

## Summary

Post-secondary graduation location tells a fundamentally different story than the previous three outcomes. Where graduates complete their credential is less about whether they face barriers and more about the pull of opportunity — academic preparation, early college exposure, and community economic context all shape the geographic direction students take, rather than simply whether they succeed.

The most consistent finding across all predictor groups is that regional completion is associated with lower academic preparation, greater economic constraint, and deeper CTE engagement — while completing outside the region, whether elsewhere in Minnesota or out of state, is associated with stronger academic profiles and fewer financial barriers. Importantly, the distinction between completing in Minnesota versus leaving the state is largely unrelated to academic preparation; what separates regional completers from everyone else is the dominant divide.

Among demographic predictors, economic status shows the most meaningful association. Economically disadvantaged graduates complete within the region at substantially higher rates, most likely reflecting financial constraints on geographic mobility rather than preference. Special education graduates show a similar pattern. Gender has essentially no relationship with where graduates complete, and race and ethnicity differences are negligible once the Unknown category is removed.

High school characteristics matter more for this outcome than for any other examined in this analysis. Entirely rural graduates leave the region at striking rates — nearly half complete elsewhere in Minnesota — suggesting that rural students who do succeed in reaching and completing post-secondary are doing so largely outside their home region. Local wage levels and unemployment rates also show interpretable associations, with students from relatively stronger local economies more likely to leave the state entirely for post-secondary.

High school accomplishments again provide the sharpest signal. PSEO participation and stronger academic preparation — higher ACT scores and MCA proficiency across all three subjects — are consistently associated with completing outside the region, with no meaningful difference between MN and out-of-state completers. CTE engagement runs in the opposite direction, with deeper involvement strongly associated with regional completion. The academic and CTE pathways point in opposite geographic directions, and that divergence has direct implications for regional workforce pipeline planning — the students most academically prepared are the most likely to leave, while those with the deepest vocational orientation are the most likely to stay.

<br>

## Demographics

This outcome examines where post-secondary graduates completed their credential — within Northeast Minnesota (the region), elsewhere in Minnesota, or outside Minnesota — and is limited to students who graduated from a post-secondary institution. It is a fundamentally different question from the previous outcomes: not whether students complete, but where they complete and what that implies for regional workforce supply.

Overall, 59% of post-secondary graduates completed within the region, 24% completed elsewhere in Minnesota, and 17% completed outside Minnesota entirely. These baseline proportions provide important context for interpreting the demographic differences below.

**Economic status** shows the most meaningful association in this group (Cramer's V = 0.104). Economically disadvantaged graduates completed within the region at a substantially higher rate (68%) than non-disadvantaged graduates (56%). Conversely, non-disadvantaged graduates were more likely to complete elsewhere in Minnesota (26% vs. 20%) or outside Minnesota (19% vs. 12%). This pattern is consistent with financial constraints limiting how far students are able or willing to travel for post-secondary education — and has direct implications for regional workforce retention, as economically disadvantaged graduates who complete locally may be more likely to remain in the region afterward.

**Special education status** shows a small but statistically significant association (Cramer's V = 0.072). Special education graduates completed within the region at a notably higher rate (74%) than non-special-ed graduates (58%), and were less likely to complete elsewhere in Minnesota (15% vs. 17%) or outside Minnesota (11% vs. 17%). As with economic status, this likely reflects geographic constraints on post-secondary choice rather than a preference for regional institutions per se.

**Race and ethnicity** shows a negligible association (Cramer's V = 0.033). While statistically significant, the differences across racial and ethnic groups are small and do not form a clear or interpretable pattern. American Indian graduates complete within the region at the highest rate (68%), while Asian/Pacific Islander graduates do so at the lowest (49%). The small cell sizes for non-White groups warrant caution in drawing strong conclusions from these differences.

**Gender** shows essentially no association with graduation location (Cramer's V = 0.029). Female and male graduates complete within the region, elsewhere in Minnesota, and outside Minnesota at nearly identical rates. Where students complete their post-secondary credential is not meaningfully related to gender.

Taken together, graduation location is most strongly shaped by economic status, with disadvantaged students substantially more likely to complete within the region. Special education status follows a similar pattern. These findings suggest that regional post-secondary completion is driven in part by access and proximity constraints rather than preference — a distinction worth naming carefully in the narrative, as it has implications for how regional institutions serve and retain students.

::: panel-tabset
### `Gender`

```{r}
data <- master.original %>%
  filter(!ps.grad.location %in% c("Attending ps", "Did not graduate ps"))

tab_xtab(var.row = data$ps.grad.location,
         var.col = data$Gender,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/ps-grad-location/Demographics/Gender.html")
```

### `economic.status`

```{r}
data <- master.original %>%
  filter(!ps.grad.location %in% c("Attending ps", "Did not graduate ps"))

tab_xtab(var.row = data$ps.grad.location,
         var.col = data$economic.status,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/ps-grad-location/Demographics/economic-status.html")
```

### `RaceEthnicity`

```{r}
data <- master.original %>%
  filter(!ps.grad.location %in% c("Attending ps", "Did not graduate ps"),
         RaceEthnicity != "Unknown")

tab_xtab(var.row = data$ps.grad.location,
         var.col = data$RaceEthnicity,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/ps-grad-location/Demographics/RaceEthnicity.html")
```

### `SpecialEdStatus`

```{r}
data <- master.original %>%
  filter(!ps.grad.location %in% c("Attending ps", "Did not graduate ps"))
tab_xtab(var.row = data$ps.grad.location,
         var.col = data$SpecialEdStatus,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/ps-grad-location/Demographics/SpecialEdStatus.html")
```
:::

<br>

## High School Characteristics

Unlike the previous outcomes, high school characteristics show more interpretively interesting associations with graduation location — particularly district demographic designation and local wage levels — though the overall effect sizes remain modest.

**District demographic designation** shows the strongest association in this group and its strongest association across all four outcomes (Cramer's V = 0.113). The pattern is striking: graduates from entirely rural districts completed within the region at just 38%, compared to 56% for town/rural mix and 61% for urban/town/rural mix districts. Entirely rural graduates were far more likely to complete elsewhere in Minnesota (49%) — more than double the rate of urban/town/rural mix graduates (21%). Graduates from entirely rural districts appear to leave the region for post-secondary at substantially higher rates, which has direct implications for workforce retention in the most rural parts of Northeast Minnesota.

**Average local wage as a percentage of the state average** shows a statistically significant and clearly ordered association across all three graduation locations (F = 107.6, p \< .001). Graduates who completed outside Minnesota came from districts with the highest wages relative to the state average, followed by regional graduates, with MN graduates (outside the region) coming from the lowest-wage districts. All three pairwise comparisons are statistically significant except the region vs. outside MN contrast. This suggests that students from relatively higher-wage communities are more likely to leave the state entirely for post-secondary — potentially reflecting greater economic resources to do so, or the pull of institutions in neighboring states.

**Average local unemployment rate** also shows a statistically significant association (F = 19.2, p \< .001), though the effect is smaller. Graduates who completed outside Minnesota came from districts with the lowest average unemployment rates, while regional graduates came from districts with the highest. All three pairwise comparisons are statistically significant. The direction is consistent with the wage finding — students from relatively stronger local labor markets are more likely to leave for post-secondary, while those from weaker markets tend to complete closer to home.

Taken together, high school characteristics tell a more meaningful story for graduation location than for any other post-secondary outcome examined. Where graduates complete their credential is associated with the economic character of their home community — students from more prosperous or more rural communities are more likely to leave the region or the state for post-secondary, while students from lower-wage or more urbanized communities tend to complete closer to home. These findings connect directly to regional workforce pipeline concerns and are worth highlighting for audiences focused on talent retention.

::: panel-tabset
### `Dem_Desc`

```{r}
tab_xtab(var.row = data$ps.grad.location,
         var.col = data$Dem_Desc,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/ps-grad-location/High school characteristics/Dem_Desc.html")
```

### `avg.unemp.rate`

```{r}
summary <- aov(avg.unemp.rate ~ ps.grad.location, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/ps-grad-location/High school characteristics/avg-unemp.rate.xlsx", overwrite = TRUE)


```

### `avg.wages.pct.state`

```{r}
summary <- aov(avg.wages.pct.state ~ ps.grad.location, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/ps-grad-location/High school characteristics/avg-wage-pct-state.xlsx", overwrite = TRUE)
```
:::

<br>

## High School Accomplishments

High school accomplishment variables show a distinctive pattern for graduation location: the key distinction is not between Minnesota and out-of-state graduates, but between regional graduates and everyone else. Across nearly every predictor, regional graduates score lower on academic preparation measures than both other groups, while MN and out-of-state graduates are largely indistinguishable from each other.

**PSEO participation** is the strongest predictor in this group (Cramer's V = 0.117). PSEO participants graduated within the region at a lower rate (54%) than non-participants (66%), and were more likely to complete elsewhere in Minnesota (27% vs. 21%) or outside Minnesota (19% vs. 13%). PSEO participation — which connects students to college coursework and institutions during high school — appears to expand the geographic range of where students ultimately complete, drawing more graduates away from regional institutions.

**ACT Composite Score** shows a large and clearly ordered association (F = 536.9, p \< .001), but the Tukey HSD results reveal an important asymmetry. Regional graduates scored approximately 2.4 points lower than both MN graduates and out-of-state graduates, but MN and out-of-state graduates did not differ significantly from each other. Higher ACT scores are associated with leaving the region for post-secondary completion, but are not associated with whether students leave the state or stay within Minnesota.

**MCA Math, Reading, and Science proficiency** scores all follow the same pattern across all three measures. Regional graduates score significantly lower than both MN graduates (outside the region) and out-of-state graduates, but MN and out-of-state graduates do not differ significantly from each other on any of the three MCA measures. The consistent finding across all academic preparation variables is that regional completion is associated with lower academic preparation, while the decision to leave the state versus stay in Minnesota is essentially unrelated to academic preparation.

**CTE achievement level** shows a small but statistically significant association (Cramer's V = 0.068). CTE concentrators and completers graduated within the region at the highest rate (66%), compared to 59% for CTE participants and 54% for students with no CTE involvement. This is consistent with the broader CTE pattern observed across all outcomes — deeper CTE engagement is associated with staying closer to home for post-secondary completion, reinforcing its connection to regionally-oriented workforce pathways.

**Total CTE courses taken** also shows a statistically significant association (F = 63.3, p \< .001). Regional graduates took more CTE courses on average than both MN graduates and out-of-state graduates, and all three pairwise comparisons are significant. More CTE coursework is associated with completing post-secondary closer to home, while fewer CTE courses is associated with leaving — the mirror image of what is seen with academic preparation variables.

Taken together, the high school accomplishment predictors of graduation location tell a coherent story. Academic preparation expands geographic range — students with stronger academic profiles are more likely to leave the region for post-secondary, though equally likely to stay in Minnesota or leave the state entirely. CTE engagement narrows geographic range — students with deeper CTE involvement are more likely to complete within the region. These findings reinforce the distinct trajectories associated with academic and CTE pathways, and have direct implications for understanding regional post-secondary pipeline dynamics.

::: panel-tabset
### `ACTCompositeScore`

```{r}
summary <- aov(ACTCompositeScore ~ ps.grad.location, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/ps-grad-location/High school accomplishments/ACTCompositeScore.xlsx", overwrite = TRUE)
```

### `pseo.participant`

```{r}
tab_xtab(var.row = data$ps.grad.location,
         var.col = data$pseo.participant,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/ps-grad-location/High school accomplishments/pseo-participant.html")
```

### `cte.achievement`

```{r}
tab_xtab(var.row = data$ps.grad.location,
         var.col = data$cte.achievement,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/ps-grad-location/High school accomplishments/cte-achievement.html")
```

### `total.cte.courses.taken`

```{r}
summary <- aov(total.cte.courses.taken ~ ps.grad.location, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/ps-grad-location/High school accomplishments/total-cte-courses-taken.xlsx", overwrite = TRUE)

```

### `MCA.M`

```{r}
summary <- aov(MCA.M ~ ps.grad.location, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/ps-grad-location/High school accomplishments/MCA-M.xlsx", overwrite = TRUE)
```

### `MCA.R`

```{r}
summary <- aov(MCA.R ~ ps.grad.location, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/ps-grad-location/High school accomplishments/MCA-R.xlsx", overwrite = TRUE)

```

### `MCA.S`

```{r}
summary <- aov(MCA.S ~ ps.grad.location, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/ps-grad-location/High school accomplishments/MCA-S.xlsx", overwrite = TRUE)
```
:::

<br>

# Summary: Post-Secondary Pathways Among Northeast Minnesota Graduates

This analysis examined four post-secondary outcomes — attendance, graduation, highest credential level, and graduation location — across three groups of predictors: demographic characteristics, high school characteristics, and high school accomplishments. Taken together, the findings tell a coherent and interpretively rich story about who pursues post-secondary education in Northeast Minnesota, who completes it, how far they go, and where they end up.

<br>

## The Dominant Finding: Individual Characteristics, Not School or Community Context, Drive Post-Secondary Outcomes

Across all four outcomes, high school characteristics — district type, local unemployment, and local wages — contribute almost nothing to explaining post-secondary attainment. Where a student went to school matters far less than who that student is and what they did while they were there. This finding is consistent and robust, holding across attendance, graduation, and credential level. The one exception is graduation location, where district type and local economic conditions show meaningful associations — but even there, the individual-level predictors are stronger.

This has an important implication for how the region thinks about post-secondary attainment gaps: the levers are not primarily structural or geographic. They are individual, relational, and experiential.

<br>

## Two Distinct Pathways Emerge — and They Point in Different Directions

The data reveal two coherent and divergent trajectories among Northeast Minnesota graduates.

The first is an academically-oriented pathway, characterized by PSEO participation, strong ACT scores, and high MCA proficiency. Students on this pathway attend post-secondary at high rates, complete at high rates, and are most likely to attain a bachelor's degree or higher. They are also the most likely to leave the region — and the state — for post-secondary. Academic preparation expands geographic range at every stage.

The second is a CTE-oriented pathway, characterized by deeper CTE engagement and more CTE coursework. Students on this pathway are somewhat less likely to attend post-secondary and substantially less likely to complete a four-year credential. But those who do complete are more likely to do so within the region. CTE engagement narrows geographic range and is consistently associated with shorter-credential and workforce-first outcomes.

These are not simply better and worse trajectories — they are different ones, oriented toward different destinations. The risk is in evaluating both pathways against post-secondary completion benchmarks designed primarily for the academic pathway. CTE-oriented students who enter the workforce directly or complete shorter credentials are not failing against the post-secondary completion standard; they are succeeding against a different one.

<br>

## Barriers Are Concentrated Among Specific Populations — and They Compound Across Outcomes

The demographic findings tell a story of compounding disadvantage. Economic disadvantage and special education status are the strongest demographic predictors at every stage — attendance, graduation, and credential level — and the gaps widen as the outcome becomes more demanding. Students who face these barriers are not just less likely to attend; they are less likely to complete, less likely to reach a four-year credential, and more likely to stop at some college without finishing anything.

The racial and ethnic gaps, particularly for American Indian and Black students, are stark at the graduation and credential level stages. These students attend post-secondary at rates that, while lower than White students, are not dramatically so — but they complete at far lower rates. The barrier for these students appears to intensify after enrollment, not primarily at the point of entry.

Gender gaps are consistent but secondary. Female graduates outperform male graduates at every post-secondary stage, and the gap widens at credential attainment — men who enroll are more likely to leave without completing.

<br>

## Regional Retention Is Structurally Tilted Against the Most Prepared Students

The graduation location findings surface a tension that is directly relevant to regional workforce planning. The students most likely to complete post-secondary within the region are those who face the most constraints: economically disadvantaged students, special education students, students from lower-wage communities, and students with deeper CTE engagement. The students most likely to leave — and not return — are the most academically prepared.

This is not a new dynamic in rural regions, but the data make it concrete. The regional post-secondary pipeline is shaped by a sorting process in which preparation and ambition tend to pull students outward, while constraint and vocational orientation tend to keep them local. Strategies aimed at strengthening regional post-secondary attainment need to grapple with this tension directly — because improving academic preparation without improving regional opportunity may simply accelerate outmigration.

<br>

## Key Takeaways

- Academic preparation — particularly PSEO participation, ACT performance, and MCA proficiency — is the strongest and most consistent predictor of post-secondary attendance and completion across all outcomes.
- Economic disadvantage and special education status are the dominant demographic barriers, and their effects compound at every successive stage of post-secondary attainment.
- CTE engagement is associated with a coherent workforce-first orientation — not failure — and is the strongest predictor of regional post-secondary completion among accomplishment variables.
- High school and community characteristics are largely irrelevant to whether students attend or complete post-secondary; they matter most for where graduates complete, with entirely rural students leaving the region at disproportionately high rates.
- The students most likely to complete post-secondary within the region are those with the fewest options, not the most. Regional workforce pipeline strategies need to account for this structural reality.
