---
title: "Multivariate Confirmation Analysis - Logistic Regression"
editor: visual
---

```{r include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
library(htmlwidgets)
library(RColorBrewer)
library(readxl)
library(janitor)
library(lubridate)
library(systemfonts)
reset_font_cache()
library(ggtext)
library(gmodels)
library(fastDummies)
library(car)
library(glmnet)
library(glmnetUtils)
library(pscl)
library(sjPlot)
library(ggforce)
```

```{r themes and shapefiles, include=FALSE}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", linewidth  = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.key.size = unit(1, "lines"),
        legend.margin = margin(0,0,0,0),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.text = element_text(margin = margin(l = 2)),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", linewidth = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))


theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
    select(5,6) %>%
    unique() %>%
    mutate(edr = str_replace(edr, "  ", " "),
           planning.region = str_replace(planning.region, " Minnesota", ""),
           planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
           edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

counties.regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
  rename(mif = `MIF Region`) %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Name = str_replace(Name, "Mcleod", "McLeod"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc) ,
         edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"))

counties.regions.1 <- counties.regions %>%
  mutate(statefp = "27",
         project.pr = ifelse(edr %in% c("EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 4 - West Central"), "Northwest",
                             ifelse(edr == "EDR 3 - Arrowhead", "Northeast",
                                    ifelse(edr %in% c("EDR 5 - North Central", "EDR 7E- East Central", "EDR 7W- Central"), "Central",
                                           ifelse(edr %in% c("EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 8 - Southwest"), "Southwest",
                                                  ifelse(edr %in% c("EDR 9 - South Central", "EDR 10 - Southeast"), "Southern", as.character(planning.region)))))),
         project.pr = fct_relevel(project.pr, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southern"))


color.ruca <- c("Entirely rural" = "#009933", "Town/rural mix" = "#99CC33", "Urban/town/rural mix" = "#CC9966", "Entirely urban" = "#754C29", "Minnesota" = "black")

color.pr <- c("Northwest" = 	"#4575b4", "Northeast" = "grey", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.edr.simple <- c("EDR 1" = "#b3cde3", "EDR 2" = "#8c96c6", "EDR 3" = "#fe9929", "EDR 4" = "#8856a7", "EDR 5" = "#810f7c", "EDR 6E" = "#e5f5f9", "EDR 6W" = "#bdc9e1", "EDR 7E" = "#99d8c9", "EDR 7W" = "#2ca25f", "EDR 8" = "#74a9cf", "EDR 9" = "#0570b0", "EDR 10" = "#d7301f", "EDR 11" = "#d8b365", "Minnesota" = "black")

color.pr.edr <- c ("Northwest" = "#4575b4","Northeast" = "#e0f3f8", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black", "EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365")

mn_counties <- st_read("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Shapefiles/County shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE) %>%
  rename(countyfp = FIPS_CODE)

```

```{r master dataset}
master.original <- read_csv("Data/SLEDS/Masters/After analysis/Master-final-after-mca.csv")

names(master.original)
master.states <- read_csv("Data/SLEDS/Masters/States/annual-seven-states.csv") %>%
  filter(grad.year %in% c("grad.year.1", "grad.year.5", "grad.year.10")) %>%
  select(PersonID, grad.year, states) %>%
  pivot_wider(names_from = grad.year, values_from = states) 

master <- master.original %>%
  left_join(master.states, by = "PersonID") %>%
  mutate(grad.year.5 = ifelse(grad.year.5 %in% c("Meaningful emp County", "Meaningful emp EDR"), "Sustained WF - NE",
                              ifelse(grad.year.5 == "Meaningful emp MN", "Sustained WF - MN",
                                     ifelse(grad.year.5 == "No MN emp record, not attending ps", "No MN emp record", 
                                            ifelse(grad.year.5 == "Not meaningful, not attending ps", "Non-sustained WF", as.character(grad.year.5))))),
         grad.year.5 = as_factor(grad.year.5))

kable(head(master))

kable(names(master))
```

# Multivariate Confirmation Analysis: Logistic Regression

The purpose of this project is to explain which characteristics are associated with entry into distinct workforce participation pathways following high school graduation in Northeast Minnesota (EDR 3). These pathways include sustained employment in the region, sustained employment elsewhere in Minnesota, non-sustained employment, and having no Minnesota employment record.

The policy question of greatest interest is what factors are associated with **sustained employment in the region**. Earlier analysis relied on bivariate analyses (cross-tabulations, Cramér’s V, and ANOVA) to identify patterns of stratification across workforce participation outcomes. A consistent finding was that postsecondary experiences—particularly credential attainment and where education was completed—are strongly associated with long-term workforce attachment. Academic preparation, early college exposure (PSEO), career and technical education (CTE), and structural demographic factors also demonstrated meaningful relationships.

To strengthen confidence in these findings, a limited multivariate analysis is conducted using logistic regression. The purpose of this modeling is not to establish causality or build a predictive tool, but to examine whether the strongest bivariate relationships remain directionally consistent when key variables are considered simultaneously. This approach helps clarify whether preparation variables retain independent associations after accounting for credential attainment, and whether regional postsecondary completion remains influential when controlling for background and readiness.

Logistic regression is used here as a confirmatory tool rather than the primary analytic framework.

<br>

## Methodology

### **Dependent Variables**

This analysis models four distinct outcomes, each restricted to the appropriate subset of the cohort:

1.  **Sustained regional workforce attachment** (`grad.year.5`) — binary (1 = meaningful employment within the region five years after high school graduation, 0 = otherwise), modeled across the full graduating cohort.
2.  **Postsecondary completion** (`ps.grad`) — binary (1 = completed a postsecondary credential, 0 = did not), modeled across **all high school graduates**, regardless of whether they ever attended postsecondary education.
3.  **Postsecondary completion within the region** — binary (1 = completed a postsecondary credential within the region, 0 = completed elsewhere in Minnesota or out of state), restricted to **postsecondary completers only**.
4.  **Sub-baccalaureate credential attainment** — binary (1 = highest credential earned is an associate degree or less-than-associate credential, 0 = bachelor's degree or higher), restricted to **postsecondary completers only**. *\[Coding pending final confirmation against the exact factor levels in the data.\]*

### **Sample**

All four models are restricted to the **2011–2019 graduating cohorts**. Earlier cohorts (2008–2010) are excluded because PSEO participation was not consistently tracked in the data system prior to 2011 — missingness in that variable is concentrated almost entirely in those years rather than distributed randomly, and including them would risk misclassifying genuine PSEO participants as non-participants. And PSEO participation was identified as having a very strong relationship to our dependent variables in the bivariate analysis.

### Independent Variables

Variables were selected based on (1) a demonstrated bivariate relationship with the relevant outcome(s) in the cross-tabulation and ANOVA analysis, and (2) practical usability in a multivariate model, with closely related variables consolidated to avoid multicollinearity. The same variable set and coding is used across all four outcomes.

**Demographics**

- **`Gender`**
- **`RaceEthnicity`**
- **`economic.status`**
- **`SpecialEdStatus`**
- **`english.learner`**

**High school characteristics**

- **`Dem_Desc`** (district rural/urban classification)
- **`avg.wages.pct.state`** (local average wages as a percentage of the state average)
- **`avg.unemp.rate`** (local average unemployment rate)

**High school accomplishments**

- **`pseo.participant`** (Post-Secondary Enrollment Options participation)
- **`took.ACT`** (whether the student took the ACT)
- **`ACTCompositeScore`** (ACT composite score, recoded to 0 for non-test-takers — see note below)
- **`ap.exam`** (AP exam participation)
- **`cte.achievement`** (CTE concentration/completion level)
- **`MCA.M`** (MCA Math proficiency)

**Postsecondary variables**

- `ps.grad` - used only for the Sustained WF - Region as the dependent variable
- **`highest.cred.level`** - these will be tested on the dependent variables dedicated to
- **`ps.grad.location`** (location of postsecondary completion)

### Notes on Variable Selection

**Academic achievement (MCA).** MCA Math, Reading, and Science scores are highly correlated with one another (r = 0.63–0.75) and with ACT performance. Including all three MCA subject scores in the same model produces unstable, difficult-to-interpret coefficients without adding meaningful explanatory power. MCA Math was retained as the single representative of academic proficiency; it has the lowest missingness of the three subjects and a comparable or stronger bivariate relationship with the outcomes examined.

**CTE engagement.** `cte.achievement` (concentrator/completer/participant/no CTE) was retained over `total.cte.courses.taken`, a continuous measure of the same underlying construct. The two are redundant, and the categorical version is more directly interpretable for a non-technical audience.

**ACT participation and score.** Students who did not take the ACT have no composite score by definition — this is a structural absence tied entirely to an observed variable (`took.ACT`), not missing data in the conventional sense. To retain the full sample rather than dropping non-test-takers, `ACTCompositeScore` was recoded to 0 for non-test-takers and entered into the model alongside `took.ACT`. With both terms present, `took.ACT` captures the effect of test participation itself, while `ACTCompositeScore` captures the achievement gradient specifically among test-takers, without the two effects being confounded.

**Variables dropped.** `range.school`, `total.cte.courses.taken`, `sat.taken`, `MCA.R`, and `MCA.S` were excluded due to a combination of weak bivariate association, redundancy with a retained variable, or both.

### Modeling Approach

Each outcome is modeled using a sequential block-entry approach, adding one group of predictors at a time to assess the relative contribution of each. Each block is compared to the prior one using a likelihood ratio test to confirm it adds significant explanatory power, with AIC and pseudo-R² tracked across the sequence to summarize overall model improvement.

**Sustained regional workforce attachment:**

- Model 1: Demographics
- Model 2: + High school characteristics
- Model 3: + High school accomplishments
- Model 4a: + Highest credential attained
- Model 4b: + Postsecondary graduation location

Model 4 is entered in two stages rather than as a single combined block. This preserves the ability to observe whether credential attainment and graduation geography operate as independent mechanisms or whether one substantially attenuates the other once both are in the model — a distinction that is central to this analysis's findings on regional workforce retention.

**Postsecondary completion, all high school graduates:**

- Model 1: Demographics
- Model 2: + High school characteristics
- Model 3: + High school accomplishments

**Postsecondary completion within the region, completers only:**

- Model 1: Demographics
- Model 2: + High school characteristics
- Model 3: + High school accomplishments

**Sub-baccalaureate credential attainment, completers only:**

- Model 1: Demographics
- Model 2: + High school characteristics
- Model 3: + High school accomplishments

Following these base model sequences, additional modeling will be considered only if the results indicate a need for further specification (e.g., interaction effects, additional mediation checks).

<br>

# Analysis Approach

This multivariate analysis is structured as a series of theoretically guided logistic regression models designed to evaluate mechanisms associated with sustained regional employment. The goal is confirmatory testing of previously observed patterns rather than prediction or causal inference.

The dependent variable is a binary indicator of **Sustained employment – region** (1 = sustained employment in the region; 0 = all other workforce outcomes).

<br>

## Modeling Strategy

Predictors are introduced sequentially to evaluate distinct mechanisms and to examine coefficient attenuation across specifications. This staged approach strengthens interpretability and reduces redundancy among highly related postsecondary variables.

### Model 1: Structural and Preparation Baseline

Includes:

- `RaceEthnicity`
- `economic.status`
- `SpecialEdStatus`
- `englisher.learner`
- `Gender`
- `MCA.M`
- `pseo.participant`
- `cte.achievement`
- `avg.unemp.rate`

This model evaluates whether structural background and preparation factors are independently associated with sustained regional employment.

### Model 2: Credential Stratification Mechanism

Adds:

- `highest.cred.level`

This model tests whether credential depth explains regional workforce attachment beyond preparation and demographic factors. Changes in coefficients from Model 1 to Model 2 are examined to assess attenuation and potential pathway mediation.

### Model 3: Regional Anchoring Mechanism

In a separate specification, the model introduces:

- `ps.grad.location`

This model evaluates whether completing postsecondary education within Northeast Minnesota is associated with sustained regional employment, holding structural and preparation factors constant.

### Model 4: Combined Attainment and Geography

Includes both:

- `highest.cred.level`
- `ps.grad.location`

This specification assesses whether regional completion retains influence after accounting for credential depth, and whether geographic anchoring operates independently of attainment level.

<br>

## Evaluation Criteria

Results will be reported using: - Odds ratios - 95% confidence intervals - Statistical significance levels - Directional interpretation

Model comparison metrics (AIC and pseudo R²) will be used to assess relative improvement across specifications within each outcome's model sequence, but emphasis will remain on coefficient stability, attenuation patterns, and substantive interpretation rather than overall fit. <br>

## Interpretation Framework

Interpretation will focus on: - Direction and relative magnitude of associations - Changes in coefficients across blocks within each model sequence - Evidence of attenuation as additional blocks are introduced, with particular attention to the sustained-regional-employment model, where credential attainment and graduation location are entered in separate stages to assess whether one mechanism absorbs the other

Findings will be described as associations holding other included variables constant. No causal claims will be made.

This analysis evaluates three central mechanisms underlying postsecondary and regional workforce outcomes in Northeast Minnesota, each anchored to a specific model:

1.  **Preparation Sorting Mechanism** (academic readiness, early college exposure, technical pathways) — evaluated primarily through the postsecondary completion model (all high school graduates).
2.  **Credential Stratification Mechanism** (attainment depth) — evaluated primarily through the sub-baccalaureate credential model (completers only).
3.  **Regional Anchoring Mechanism** (location of postsecondary completion) — evaluated primarily through the regional completion model (completers only).

The sustained regional employment model synthesizes these mechanisms, testing whether credential depth and graduation location operate as independent predictors of long-term regional workforce attachment or whether one substantially explains the other once both are accounted for.

Together, these models provide multivariate confirmation of patterns identified in earlier descriptive analyses while preserving conceptual clarity and interpretability.

# Prep data

The first thing I need to do is setup my dataset for modeling. I'm going to create the workforce participation into a binary and will use grad.year.5 as my modeling year.

```{r}

model_vars_final <- c("Gender", "RaceEthnicity", "economic.status", "SpecialEdStatus",
                      "english.learner", "Dem_Desc", "avg.wages.pct.state", "avg.unemp.rate",
                      "pseo.participant", "took.ACT", "ACTCompositeScore.filled", "ap.exam",
                      "cte.achievement", "MCA.M", "ps.grad",
                      "grad.year.5")

master.model <- master %>%
  filter(grad.year > 2010) %>%
  select(PersonID, Gender, RaceEthnicity, economic.status, SpecialEdStatus, english.learner, Dem_Desc, avg.wages.pct.state, avg.unemp.rate, pseo.participant, took.ACT, ACTCompositeScore, ap.exam, cte.achievement, MCA.M, ps.grad, grad.year.5) %>%
  filter(grad.year.5 != "After 2023",
         grad.year.5 != "Attending ps",
         ps.grad != "Attending ps") %>%
  mutate(grad.year.5 = ifelse(grad.year.5 == "Sustained WF - NE", 1, 0),
         Gender = fct_relevel(Gender, "M"),
         RaceEthnicity = fct_relevel(RaceEthnicity, "White"),
         economic.status = as_factor(economic.status),
         economic.status = fct_relevel(economic.status, "0"),
         SpecialEdStatus = as.character(SpecialEdStatus),
         SpecialEdStatus = fct_relevel(SpecialEdStatus, "0"),
         english.learner = as_factor(english.learner),
         english.learner = fct_relevel(english.learner, "0"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural"),
         pseo.participant = as.character(pseo.participant),
         pseo.participant = fct_relevel(pseo.participant, "0"),
         took.ACT = fct_relevel(took.ACT, "No"),
         ACTCompositeScore.filled = ifelse(is.na(ACTCompositeScore), 0, ACTCompositeScore),
         MCA.M.missing = ifelse(is.na(MCA.M), 1, 0),
         MCA.M.filled = ifelse(is.na(MCA.M), 0, MCA.M),
         ap.exam = as_factor(ap.exam),
         ap.exam = fct_relevel(ap.exam, "0"),
         cte.achievement = fct_relevel(cte.achievement, "No CTE"),
         ps.grad = ifelse(ps.grad == "Y", "Yes", ps.grad),
         ps.grad = fct_relevel(ps.grad, "No")) 

nrow(master.model)  # should now be back up near 20,386, since MCA.M no longer forces complete-case loss

kable(head(master.model))

kable(names(master.model))

model_vars_final <- c("Gender", "RaceEthnicity", "economic.status", "SpecialEdStatus",
                      "english.learner", "Dem_Desc", "avg.wages.pct.state", "avg.unemp.rate",
                      "pseo.participant", "took.ACT", "ACTCompositeScore.filled", "ap.exam",
                      "cte.achievement", "MCA.M.filled", "MCA.M.missing",
                      "ps.grad", "grad.year.5")

colSums(is.na(master.model[, model_vars_final]))

nrow(master.model)
```

We now have `r comma(nrow(master.model))` individuals to model with `r ncol(master.model) - 2` variables not including `grad.year.5`.

<br>

# Modeling - Sustained WF - Region

## Model 1: Demographics (Sustained Regional Workforce Attachment)

This model estimates the association between demographic characteristics and sustained regional workforce attachment five years after high school graduation, using the full analytic sample (n = 20,386). The outcome is coded 1 = sustained workforce attachment within the region ("Sustained WF - NE"), 0 = all other outcomes (non-sustained, sustained elsewhere in Minnesota, or no Minnesota employment record).

Demographic factors show a consistent pattern: nearly every group examined is *less* likely than its reference category to achieve sustained regional attachment.

**Race and ethnicity** shows the widest range of effects in this model.

- **Black graduates:** approximately 52% lower odds of sustained regional attachment relative to White graduates (OR ≈ 0.48).
- **Asian/Pacific Islander graduates:** approximately 43% lower odds (OR ≈ 0.57).
- **Hispanic graduates:** approximately 37% lower odds (OR ≈ 0.63).
- **American Indian graduates:** approximately 15% lower odds (OR ≈ 0.85), the smallest gap among the four groups but still statistically significant.

**Other demographic predictors** show smaller but still significant effects.

- **English learner status:** approximately 24% lower odds of sustained regional attachment (OR ≈ 0.76).
- **Female graduates:** approximately 11% lower odds relative to male graduates (OR ≈ 0.89).
- **Special education status:** approximately 15% lower odds (OR ≈ 0.85).
- **Economic disadvantage:** approximately 7% lower odds (OR ≈ 0.93) — the weakest effect in the model, significant at p \< .05 but only just.

**Interpretation**

At the demographics-only stage, every predictor points in the same direction: relative to the reference group (male, White, not economically disadvantaged, not special education, not an English learner), each demographic category examined shows *lower* odds of sustained regional workforce attachment. Race and ethnicity produce the largest disparities, with Black and Asian/Pacific Islander graduates showing the steepest gaps. Economic status has the smallest effect of the five predictors, sitting just inside statistical significance.

This is a baseline model only — it does not yet account for high school context, academic preparation, or postsecondary pathways, all of which are added in subsequent blocks. Whether these demographic gaps persist, widen, or attenuate once those factors are introduced is the central question the block-sequenced comparison is designed to answer.

**Model fit:** AIC = 25,272.6; residual deviance = 25,255 on 20,377 df (null deviance = 25,416 on 20,385 df).

**Note on missing data:** MCA Math scores are missing at a substantially higher rate for English learner students (26.1%) than for non-English-learners (2.4%), consistent with the likelihood that English learners are more frequently eligible for testing exemptions or accommodations. To prevent this from disproportionately excluding English learners from the analytic sample once academic achievement enters the model (Model 3), a missing-data indicator (`MCA.M.missing`) was created alongside a filled version of the score (`MCA.M.filled`, coded 0 for missing cases), and both are entered as separate terms starting in Model 3. This preserves the full sample across all four model blocks.

```{r}
model1_dv1 <- glm(
  grad.year.5 ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner,
  family = binomial,
  data = master.model
)

model1_dv1

summary(model1_dv1)
nobs(model1_dv1)
AIC(model1_dv1)

# odds ratios + 95% CIs, per your evaluation criteria
exp(cbind(OR = coef(model1_dv1), confint(model1_dv1)))

or_plot <- tidy(model1_dv1, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )

ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )



```

<br>

## Model 2: Demographics + High School Characteristics

This model adds high school characteristics — district rural/urban classification, local wage levels, and local unemployment rates — to the demographic baseline established in Model 1. The analytic sample remains n = 20,386.

**District classification (Dem_Desc)** emerges as the strongest new predictor in this block, with a clear and interpretable gradient.

- **Town/rural mix districts:** approximately 31% higher odds of sustained regional attachment relative to entirely rural districts (OR ≈ 1.31).
- **Urban/town/rural mix districts:** approximately 60% higher odds relative to entirely rural districts (OR ≈ 1.60).

The pattern is consistent and increasing: the more urban a graduate's home district, the higher the odds of sustained attachment to the regional workforce.

**Local economic conditions do not show an independent effect at this stage.**

- **Average local wages (% of state average)** is not statistically significant (OR ≈ 1.34, CI: 0.72–2.51, crosses 1).
- **Average local unemployment rate** is not statistically significant (OR ≈ 1.00, CI: 0.976–1.027, crosses 1).

Despite both showing meaningful associations in the bivariate analysis, neither holds up as an independent predictor once entered alongside demographics and district type. Consistent with this project's approach to weak or non-significant predictors, these variables are retained in the model but not built into the narrative interpretation.

**Demographic effects from Model 1 largely persist, with one notable exception.**

- Gender, Race/Ethnicity (all four categories), Special Education status, and English learner status remain statistically significant, in the same direction, and at similar magnitudes to Model 1.
- **Economic status is no longer statistically significant** (OR ≈ 0.947, CI: 0.888–1.009, crosses 1) — a change from Model 1, where it was significant, if modestly so (OR ≈ 0.928, p = .022). This suggests that some of the economic-status effect on regional attachment observed in Model 1 may be explained by geographic/community context (district type, local labor market conditions) rather than economic status operating independently. Whether this attenuation persists or reverses will be tracked through Models 3 and 4.

**Overall Interpretation**

Adding high school characteristics significantly improves model fit over the demographics-only baseline (likelihood ratio test: Δdeviance = 80.07, df = 4, p \< .001; AIC decreases from 25,272.6 to 25,200.6). District rural/urban classification is the one high-school-characteristic variable that shows a clear, significant, and interpretable relationship with sustained regional attachment, while local wage and unemployment levels do not hold up independently despite their bivariate significance. The attenuation of the economic-status effect is the most substantively interesting development in this block, and points toward geography playing a role in explaining part of what looked like a purely economic effect in Model 1.

**Model fit:** AIC = 25,200.6 (down from 25,272.6 in Model 1); Δdeviance = 80.07 on 4 df, p \< .001; n = 20,386 in both models.

```{r}
model2_dv1 <- glm(
  grad.year.5 ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    avg.wages.pct.state +
    avg.unemp.rate,
  family = binomial,
  data = master.model
)

nobs(model1_dv1)
nobs(model2_dv1)


AIC(model1_dv1)
AIC(model2_dv1)

exp(cbind(OR = coef(model2_dv1), confint(model2_dv1)))

anova(model1_dv1, model2_dv1, test = "Chisq")

or_plot <- tidy(model2_dv1, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )

ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )

```

<br>

## Model 3: Demographics + High School Characteristics + High School Accomplishments

This model adds high school accomplishments — PSEO participation, ACT participation and score, AP exam participation, CTE achievement, and MCA Math (with the missing-data adjustment) — to the growing specification. The analytic sample remains n = 20,386.

**CTE achievement continues to anchor students regionally**, consistent with the pattern established across this entire project.

- **CTE Concentrators/Completers:** approximately 37% higher odds of sustained regional attachment relative to no CTE involvement (OR ≈ 1.37).
- **CTE Participants:** approximately 19% higher odds (OR ≈ 1.19).

**AP exam participation is associated with lower odds of regional attachment** (OR ≈ 0.68) — consistent with the broader pattern that college-preparatory engagement pulls graduates away from the region.

**PSEO participation is no longer statistically significant** in this specification (OR ≈ 1.01, CI: 0.94–1.08, crosses 1). Despite being one of the strongest bivariate predictors examined throughout this project, PSEO does not show an independent effect once ACT participation, ACT score, AP exam, and CTE are accounted for together — its apparent influence in isolation appears to substantially overlap with these other college-preparatory measures.

**ACT shows a genuinely more complex pattern than a single coefficient can convey, and deserves a careful read rather than a headline number.** Both ACT terms are highly significant, but they point in different directions:

- **Took the ACT (vs. did not):** OR ≈ 3.91 — a large, positive association with sustained regional attachment.
- **ACT composite score (among test-takers):** OR ≈ 0.93 per point — each additional point is associated with *lower* odds of regional attachment.

Combined, these two terms describe a crossover: relative to students who never took the ACT, students who took it and scored **below approximately 19–20** (just under the national median of 21) show *higher* odds of sustained regional attachment, while students who took it and scored **above that threshold** show *lower* odds. In other words, simply engaging with the ACT is not, on its own, a signal of leaving the region — it is specifically *high-scoring* ACT engagement that predicts departure, while modest ACT engagement looks more similar to (or even stronger than) not testing at all. Reporting the "took ACT" coefficient alone, without this qualification, would overstate and mischaracterize the relationship.

**MCA Math shows a similar pattern worth naming carefully.** The filled score itself is not significant (OR ≈ 1.03, CI: 0.982–1.072, crosses 1) — once the missingness is properly accounted for, the raw math proficiency score does not independently predict this outcome. However, **the missing-data flag itself is strongly significant and substantively large** (OR ≈ 0.186): students with a missing MCA Math score have dramatically lower odds of sustained regional attachment than students with an observed score, holding everything else constant.

This result connects directly to the English learner finding below, and is worth flagging as a genuine, if uncomfortable, insight rather than a nuisance parameter. **English learner status, which was significant in Models 1 and 2, is no longer significant in Model 3** (OR ≈ 0.91, CI: 0.769–1.076, crosses 1). Since English learners are missing MCA Math at a much higher rate (26.1% vs. 2.4%), a substantial share of what looked like an "English learner effect" in the earlier models appears to actually be driven by whatever unobserved circumstances are associated with a missing MCA score — likely testing exemptions, transfers, or other structural factors correlated with both EL status and reduced regional workforce attachment. The missing-data indicator is absorbing this signal rather than the English learner variable itself. This is exactly the kind of masking effect the missing-indicator approach was designed to detect, and it changes the interpretation meaningfully: English learner status alone may not be the operative factor — something correlated with missing academic records is.

**Several other coefficients shift notably from Model 2, in both directions.**

- **Economic status returns to significance**, and more strongly than in Model 1 (OR ≈ 0.844, CI: 0.789–0.902) — after appearing attenuated in Model 2, its effect reemerges once academic and college-prep measures are accounted for.
- **Average local wages becomes significant** for the first time (OR ≈ 2.34, CI: 1.228–4.452), having shown no independent effect in Model 2.
- **Gender is no longer significant** (OR ≈ 0.950, CI: 0.892–1.011, crosses 1), a change from Models 1 and 2 where it was a modest but reliable predictor.
- **Special Education status strengthens** (OR ≈ 0.761, compared to OR ≈ 0.847 in Model 2).
- Race/Ethnicity effects persist across all four categories, largely stable in direction and magnitude relative to prior models.
- **District classification (Dem_Desc)** remains significant and essentially unchanged from Model 2.

**Overall Interpretation**

Adding high school accomplishments substantially improves model fit (likelihood ratio test: Δdeviance = 694.52, df = 8, p \< .001; AIC decreases from 25,200.6 to 24,522.0) — the largest improvement of any block so far. Academic and college-preparatory engagement (ACT, AP, and by extension the pattern typically associated with PSEO) continues to point away from regional attachment, while CTE engagement continues to anchor students to the region, replicating the "dual pathways" pattern found throughout this project's descriptive work.

The most important development in this block is analytical, not just predictive: the attenuation of the English learner effect, paired with the strong and significant MCA-missingness flag, suggests that at least part of what appeared to be an English-learner-specific disparity in earlier models is better explained by whatever underlying circumstances are associated with missing academic records for that population. This distinction matters for how the finding should be communicated — the disparity is real, but its likely mechanism is more specific than "English learner status" alone.

**Model fit:** AIC = 24,522.0 (down from 25,200.6 in Model 2); Δdeviance = 694.52 on 8 df, p \< .001; n = 20,386 across all three models.

```{r}
model3_dv1 <- glm(
  grad.year.5 ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    avg.wages.pct.state +
    avg.unemp.rate +
    pseo.participant +
    took.ACT +
    ACTCompositeScore.filled +
    ap.exam + 
    cte.achievement +
    MCA.M.filled +
    MCA.M.missing,
  family = binomial,
  data = master.model
)

nobs(model1_dv1)
nobs(model2_dv1)
nobs(model3_dv1)


AIC(model1_dv1)
AIC(model2_dv1)
AIC(model3_dv1)

exp(cbind(OR = coef(model3_dv1), confint(model3_dv1)))

anova(model1_dv1, model2_dv1, model3_dv1, test = "Chisq")

or_plot <- tidy(model3_dv1, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )

ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )


```

<br>

## Model 4a: Demographics + High School Characteristics + High School Accomplishments + Credential Attainment

This model adds highest credential level attained (relative to "Did not graduate ps") to the specification. The analytic sample remains n = 20,386.

**Credential attainment is strongly associated with sustained regional attachment, but the two credential levels show a very different magnitude of effect.**

- **Associate degree or less:** approximately 118% higher odds of sustained regional attachment relative to non-completers (OR ≈ 2.18).
- **Bachelor's degree or higher:** approximately 16% higher odds relative to non-completers (OR ≈ 1.16) — significant, but far more modest.

Both credential levels point in the same direction — completing *any* postsecondary credential is associated with higher odds of regional attachment than not completing one at all — but sub-baccalaureate credentials show an effect roughly 13 times larger than bachelor's degrees. This is a substantively important finding: credential depth alone does not straightforwardly predict regional attachment. Shorter, more technical credentials anchor graduates to the region far more strongly than four-year degrees do.

**Several accomplishments-block coefficients attenuate somewhat once credential level is introduced**, consistent with the idea that some of their apparent effect operates through eventual credential attainment rather than independently of it:

- **Took the ACT:** OR drops from 3.91 (Model 3) to 3.27 — still large and highly significant, but noticeably smaller.
- **MCA Math missingness:** OR moves from 0.186 to 0.201 — still a strong effect, modestly attenuated.
- **PSEO participation** flips from a (non-significant) positive association to a (still non-significant) negative one — remains uninformative either way.

**Most other predictors remain stable in direction and significance:**

- Race/Ethnicity effects persist, though American Indian and Asian/Pacific Islander move to the edge of significance (CIs now narrowly include or approach 1).
- CTE achievement, AP exam, economic status, Special Ed status, and Dem_Desc all remain significant with similar magnitudes to Model 3.
- **Gender returns to significance** (OR ≈ 0.928), reversing its non-significant status in Model 3.
- English learner status remains non-significant, consistent with Model 3's finding that the MCA-missingness flag — not English learner status itself — appears to carry this signal.

**Overall Interpretation**

Adding credential attainment significantly improves model fit (likelihood ratio test: Δdeviance = 375.05, df = 2, p \< .001; AIC decreases from 24,522.0 to 24,151.0). The central finding of this block is the sharp asymmetry between credential types: sub-baccalaureate credentials show a dramatically stronger association with regional workforce attachment than bachelor's degrees do. This raises the question this project has flagged before — whether this reflects the nature of the credential itself, or where those credentials tend to be earned (a question the next model, adding graduation location, is specifically designed to test).

**Model fit:** AIC = 24,151.0 (down from 24,522.0 in Model 3); Δdeviance = 375.05 on 2 df, p \< .001; n = 20,386 across all four models.

```{r}
model4a_dv1 <- glm(
  grad.year.5 ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    avg.wages.pct.state +
    avg.unemp.rate +
    pseo.participant +
    took.ACT +
    ACTCompositeScore.filled +
    ap.exam + 
    cte.achievement +
    MCA.M.filled +
    MCA.M.missing +
    highest.cred.level,
  family = binomial,
  data = master.model
)

nobs(model1_dv1)
nobs(model2_dv1)
nobs(model3_dv1)
nobs(model4a_dv1)

AIC(model1_dv1)
AIC(model2_dv1)
AIC(model3_dv1)
AIC(model4a_dv1)

exp(cbind(OR = coef(model4a_dv1), confint(model4a_dv1)))

anova(model1_dv1, model2_dv1, model3_dv1, model4a_dv1, test = "Chisq")

or_plot <- tidy(model4a_dv1, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )


ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )


```

<br>

## Model 4b: Full Specification — Demographics + High School Characteristics + High School Accomplishments + Credential Attainment + Graduation Location

This model adds postsecondary graduation location (relative to "Did not graduate ps") to the full specification, completing the staged test of whether credential depth or graduation geography drives sustained regional workforce attachment. The analytic sample remains n = 20,386.

**This is the key result of the entire model sequence: once graduation location is added, credential attainment loses its independent effect entirely.**

- **Associate degree or less:** OR drops from 2.18 (Model 4a) to 1.73, and the confidence interval now crosses 1 (CI: 0.874–3.316) — **no longer statistically significant.**
- **Bachelor's degree or higher:** OR moves from 1.16 to 1.21, also no longer significant (CI: 0.608–2.325).

Both credential categories, which were strong and significant predictors in Model 4a, become statistically indistinguishable from non-completers once graduation location is accounted for. This confirms that the large sub-baccalaureate effect observed in Model 4a was substantially — perhaps entirely — a reflection of *where* those credentials tend to be earned, not something inherent to shorter credentials themselves.

**Graduation location itself shows a more nuanced pattern than a single "region vs. elsewhere" story.**

- **Graduated elsewhere in Minnesota:** approximately 52% lower odds of sustained regional attachment relative to non-completers (OR ≈ 0.48, statistically significant).
- **Graduated outside Minnesota:** approximately 45% lower odds, though this does not reach significance (OR ≈ 0.55, CI: 0.283–1.101).
- **Graduated within the region:** OR ≈ 1.65, in the expected positive direction, but **not statistically significant** (CI: 0.859–3.262).

The lack of significance on the "graduated within the region" category is notable given how central this variable has been throughout this project's descriptive analysis, and is worth a caveat: `highest.cred.level` and `ps.grad.location` are correlated with each other (sub-baccalaureate credentials are disproportionately regional, as this project's earlier work established), and entering both together in the same model has widened the confidence intervals on each — a sign the two variables are now competing to explain overlapping variance rather than each cleanly isolating its own effect. The overall pattern (regional completion trending positive, MN-elsewhere and out-of-state trending negative) is consistent with expectations, even where individual categories fall short of significance.

**Nearly all other coefficients remain stable, with one previously-significant predictor now falling out.**

- Gender, all four Race/Ethnicity categories, economic status, Special Ed status, Dem_Desc, took.ACT, ACT score, AP exam, CTE achievement, and the MCA missingness flag all remain significant, in the same direction and similar magnitude to Model 4a.
- **Average local wages is no longer significant** (OR ≈ 1.51, CI: 0.779–2.909), having been significant in Models 3 and 4a. This is now the second high-school-characteristic economic variable (alongside unemployment rate) to show no reliable independent effect in the fully specified model.
- English learner status and PSEO participation remain non-significant, consistent with every prior model in this sequence.

**Overall Interpretation**

Adding graduation location produces the largest single-block improvement in model fit of the entire sequence (likelihood ratio test: Δdeviance = 514.19, df = 3, p \< .001; AIC decreases from 24,151.0 to 23,642.8). More importantly, it resolves the central analytical question this staged model sequence was built to answer: **the strong sub-baccalaureate credential effect observed in Model 4a is substantially explained by graduation geography, not credential depth itself.** Once location is accounted for, credential level no longer independently predicts sustained regional attachment.

This is consistent with — and provides direct multivariate confirmation of — the locked-in finding from this project's earlier descriptive and confirmatory work: geography dominates credential type in predicting regional workforce participation. Preparation and academic engagement (ACT, AP, CTE, MCA-missingness) continue to operate as independent, stable predictors throughout every block, while the postsecondary story is fundamentally about *where* graduates complete their education rather than *what* they complete.

**A variance inflation factor (VIF) check confirms the collinearity implied by the coefficient changes above.** `highest.cred.level` and `ps.grad.location` show elevated VIFs (11.75 and 4.96 respectively, on a scale where values above 5–10 indicate meaningful collinearity), confirming that these two variables substantially overlap in the variance they explain. This is consistent with — and provides a formal statistical basis for — the interpretation that graduation location, not credential depth, is the more fundamental driver of sustained regional workforce attachment. (The elevated VIFs on `took.ACT` and `ACTCompositeScore.filled`, by contrast, are an expected and accepted consequence of the missing-indicator method used to handle ACT non-participation, not a new concern.)x

**Model fit:** AIC = 23,642.8 (down from 24,151.0 in Model 4a); Δdeviance = 514.19 on 3 df, p \< .001; n = 20,386 across all five models.

```{r}
model4b_dv1 <- glm(
  grad.year.5 ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    avg.wages.pct.state +
    avg.unemp.rate +
    pseo.participant +
    took.ACT +
    ACTCompositeScore.filled +
    ap.exam + 
    cte.achievement +
    MCA.M.filled +
    MCA.M.missing +
    highest.cred.level +
    ps.grad.location,
  family = binomial,
  data = master.model
)

nobs(model1_dv1)
nobs(model2_dv1)
nobs(model3_dv1)
nobs(model4a_dv1)
nobs(model4b_dv1)

AIC(model1_dv1)
AIC(model2_dv1)
AIC(model3_dv1)
AIC(model4a_dv1)
AIC(model4b_dv1)

exp(cbind(OR = coef(model4b_dv1), confint(model4b_dv1)))

anova(model1_dv1, model2_dv1, model3_dv1, model4a_dv1, model4b_dv1, test = "Chisq")

library(car)
vif(model4b_dv1)

or_plot <- tidy(model4, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )


ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )

```

<br>

```{r}
model4c_dv1 <- glm(
  grad.year.5 ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    avg.wages.pct.state +
    avg.unemp.rate +
    pseo.participant +
    took.ACT +
    ACTCompositeScore.filled +
    ap.exam + 
    cte.achievement +
    MCA.M.filled +
    MCA.M.missing +
    ps.grad.location,
  family = binomial,
  data = master.model
)

nobs(model1_dv1)
nobs(model2_dv1)
nobs(model3_dv1)
nobs(model4a_dv1)
nobs(model4b_dv1)
nobs(model4c_dv1)

AIC(model1_dv1)
AIC(model2_dv1)
AIC(model3_dv1)
AIC(model4a_dv1)
AIC(model4b_dv1)
AIC(model4c_dv1)

exp(cbind(OR = coef(model4c_dv1), confint(model4c_dv1)))

anova(model1_dv1, model2_dv1, model3_dv1, model4a_dv1, model4b_dv1, model4c_dv1, test = "Chisq")

library(car)
vif(model4c_dv1)

or_plot <- tidy(model4, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )


ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )

test <- master.model %>%
  mutate(
    cred_location_combined = case_when(
      highest.cred.level == "Did not graduate ps" ~ "Did not graduate ps",
      TRUE ~ paste(highest.cred.level, ps.grad.location, sep = " - ")
    ),
    cred_location_combined = fct_relevel(cred_location_combined, "Did not graduate ps")
  )

table(test$cred_location_combined)
```

<br>

## Multivariate Confirmation Summary

Across progressively specified logistic regression models, sustained regional employment in Northeast Minnesota is driven most strongly by **postsecondary geography**, with credential depth playing a secondary role once location is considered. High school CTE participation remains a consistent independent pathway, while academic preparation measures align modestly with outward mobility. Structural disparities persist across all specifications.

Model fit improves meaningfully as postsecondary variables are introduced:

- Model 1 (Preparation + Structure): AIC ≈ 34,286\
- Model 2 (+ Credential Depth): AIC ≈ 33,676\
- Model 3 (+ Geography): AIC ≈ 32,922\
- Model 4 (+ Credential + Geography): AIC ≈ 32,771

The largest improvement occurs when postsecondary geography is introduced, indicating that location of completion explains more variation than credential depth alone.

[**Key Findings**]{.underline}

- **Graduating within the region** is the strongest predictor across all specifications:

  - Model 3: \~142% higher odds (OR ≈ 2.42).
  - Model 4: \~197% higher odds (OR ≈ 2.97).

  Regional postsecondary completion is the single most powerful anchoring mechanism in the analysis.

- Graduating **elsewhere in Minnesota** is associated with substantially lower regional attachment (\~44% lower odds in Model 3).

- Graduating **outside Minnesota** is also associated with lower regional attachment (\~36% lower odds in Model 3).

- Credential depth alone appears highly influential in Model 2:

  - Associate degree: \~120% higher odds.
  - Less than Associate degree: \~118% higher odds.

  However, once graduation location is included (Model 4), these effects attenuate substantially, indicating that the earlier credential effect largely reflects where the credential was earned rather than credential depth itself.

- **Bachelor’s degree or higher** is consistently mobility-aligned in the fully specified model (\~40% lower odds), suggesting four-year attainment is associated with outward workforce trajectories.

- **CTE participation remains independently associated with regional retention across all models:**

  - CTE concentrators/completers: \~29–38% higher odds.
  - CTE participants: \~16–20% higher odds.

  These effects persist even after accounting for both credential depth and graduation location, indicating that CTE functions as a direct workforce-alignment pathway rather than merely channeling students into regional colleges.

- Academic mobility indicators show modest but consistent outward orientation:

  - Higher MCA Math scores: \~4–7% lower odds per unit increase.
  - PSEO participation: \~6–8% lower odds.

- Structural disparities remain substantial across models:

  - Black students: \~43–48% lower odds.
  - Asian/PI students: \~32–38% lower odds.
  - Hispanic students: \~22–24% lower odds (marginal in later models).
  - Economic disadvantage: \~9–12% lower odds.
  - Special Education status: \~17–22% lower odds.
  - Female students: \~9–11% lower odds.
  - Local unemployment rates: modest negative association (\~4–5% lower odds per percentage-point increase).

[**Overall Interpretation**]{.underline}

Two primary mechanisms shape sustained regional employment in Northeast Minnesota.

First, **regional postsecondary embeddedness is the dominant driver**. Completing postsecondary education within the region dramatically increases the likelihood of long-term regional workforce attachment. The substantial improvement in model fit when geography is introduced confirms that where education is completed matters more than credential level alone.

Second, **CTE participation operates as a stable and independent alignment pathway**. While smaller in magnitude than the geography effect, CTE remains statistically robust across all specifications. This suggests that structured career pathways in high school directly connect students to regional workforce opportunities.

Credential depth initially appears highly influential, but much of that effect is absorbed by geography. Sub-baccalaureate credentials anchor students primarily when earned within the region. Bachelor’s attainment, particularly when earned outside the region, aligns with mobility rather than retention.

Academic proficiency and early college exposure are modestly mobility-oriented, indicating that stronger academic preparation may facilitate broader opportunity structures beyond Northeast Minnesota.

Importantly, racial and structural disparities persist even after accounting for preparation, credential attainment, and graduation location. Educational pathways alone do not fully explain differences in sustained regional workforce attachment.

Overall, the Northeast Minnesota results point to a clear anchoring pipeline: high school career alignment (CTE) and, most decisively, regional postsecondary completion drive long-term regional employment. Out-of-region completion and mobility-aligned academic pathways are associated with outward workforce trajectories.

<br>

# Modeling - college completion

Prior models identified two dominant predictors of sustained regional employment: graduating from a postsecondary institution within the region and participation in high school CTE pathways. To further clarify the mechanism linking education to workforce retention, this supplemental analysis examines which demographic and high school characteristics are associated with completing college, no matter where, and then the modeling the sub-population of college completers on whether they completed college in the region or somewhere else.

This model does not replace the workforce participation analysis. Rather, it evaluates whether the same preparation and structural factors that differentiate workforce outcomes also predict entry into the regional postsecondary pathway. If variables such as CTE participation strongly predict in-region graduation, this would suggest that their influence on sustained regional employment operates, at least in part, through geographic educational alignment. Conversely, if certain predictors do not differentiate regional graduation, this would indicate a more direct association with workforce retention independent of postsecondary geography.

By modeling regional postsecondary graduation as the outcome, this analysis tests whether regional workforce retention reflects a two-step pipeline—high school pathway → regional postsecondary completion → regional employment—or whether these mechanisms operate in parallel.

The dependent variable is coded as:

1 = Graduated from a postsecondary institution within the region

0 = All others (including graduates elsewhere and non-graduates)

The full analytic sample is retained to evaluate the entire educational pipeline rather than restricting to postsecondary completers.

```{r}
master.model.ps.grad <- master.model %>%
  mutate(
    ps_grad_binary =
      ifelse(ps.grad == "Y", 1, 0)
  )

table(master.model.ps.grad$ps_grad_binary)
prop.table(table(master.model.ps.grad$ps_grad_binary))

colSums(is.na(master.model.ps.grad[, c(
  "RaceEthnicity",
  "economic.status",
  "SpecialEdStatus",
  "english.learner",
  "Gender",
  "MCA.M",
  "pseo.participant",
  "cte.achievement",
  "avg.unemp.rate"
)]))

levels(master.model.ps.grad$RaceEthnicity)
levels(master.model.ps.grad$economic.status)
levels(master.model.ps.grad$SpecialEdStatus)
levels(master.model.ps.grad$english.learner)
levels(master.model.ps.grad$Gender)
levels(master.model.ps.grad$cte.achievement)
levels(master.model.ps.grad$pseo.participant)

prop.table(table(master.model.ps.grad$ps_grad_binary,
                 master.model.ps.grad$cte.achievement), 2)

prop.table(table(master.model.ps.grad$ps_grad_binary,
                 master.model.ps.grad$RaceEthnicity), 2)

```

Everything looks sound so let's begin modeling.

## Model 1: College completion, full population

This model estimates the association between demographic characteristics, academic preparation, and high school pathways with postsecondary completion, regardless of institution location. The outcome is coded 1 = completed postsecondary and 0 = did not complete. The full analytic sample is retained (n = 27,364).

The strongest negative predictors of completion are structural factors.

- **Special Education status:** approximately 62% lower odds of completing postsecondary education (OR ≈ 0.38).
- **Economic disadvantage:** roughly 59% lower odds (OR ≈ 0.41).
- **American Indian students:** about 51% lower odds relative to White students (OR ≈ 0.49).
- **Black students:** approximately 46% lower odds (OR ≈ 0.54).
- **Hispanic students:** about 27% lower odds (OR ≈ 0.73).
- **Asian/PI students:** not statistically distinct from White students once controls are included.

These effects are large and consistent, indicating that the most substantial sorting in the education-to-workforce pipeline occurs at the point of postsecondary completion.

Academic preparation variables show strong positive associations.

- **Female students:** approximately 66% higher odds of completing postsecondary education (OR ≈ 1.66).
- **PSEO participation:** about 44% higher odds (OR ≈ 1.44).
- **Higher MCA Math scores:** roughly 29% higher odds per unit increase (OR ≈ 1.29).

These findings confirm that academic readiness and early college exposure substantially increase the likelihood of postsecondary completion.

Career and Technical Education (CTE) participation does not independently predict completion once other variables are included. Both CTE concentrators/completers and CTE participants are not statistically distinct from non-CTE students in this specification. This indicates that CTE functions differently in the broader pipeline, aligning more directly with workforce outcomes rather than increasing college completion rates.

Local labor market conditions show a very modest association, with slightly higher unemployment rates associated with marginally higher completion odds, though the magnitude is small.

[**Overall Interpretation**]{.underline}

Postsecondary completion represents the most structurally stratified stage of the regional education-to-workforce pathway. Race, socioeconomic status, and Special Education status strongly differentiate who completes college, while academic preparation and early college exposure substantially increase completion likelihood.

These findings clarify earlier results. Large disparities observed in regional postsecondary graduation are driven primarily by overall completion gaps rather than geographic sorting among graduates. In contrast, CTE participation does not significantly affect completion, reinforcing the conclusion that its association with sustained regional employment operates through a more direct workforce-alignment mechanism rather than through increased college attainment.

```{r}
model_ps_grad <- glm(
  ps_grad_binary ~
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Gender +
    pseo.participant +
    MCA.M +
    cte.achievement +
    avg.unemp.rate,
  family = binomial,
  data = master.model.ps.grad
)

nobs(model_ps_grad)
AIC(model_ps_grad)

exp(cbind(
  OR = coef(model_ps_grad),
  confint(model_ps_grad)
))

or_plot <- tidy(model_ps_grad, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )


ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )

```

<br>

## Model 2: College completion region, college-grads only

```{r}
master.model.ps.grad.region <- master.model %>%
  filter(ps.grad == "Y") %>%
  mutate(ps.grad.location.binary = ifelse(ps.grad.location == "PS grad Region", 1, 0))

table(master.model.ps.grad.region$ps.grad.location.binary)
prop.table(table(master.model.ps.grad.region$ps.grad.location.binary))

colSums(is.na(master.model.ps.grad.region[, c(
  "RaceEthnicity",
  "economic.status",
  "SpecialEdStatus",
  "english.learner",
  "Gender",
  "MCA.M",
  "pseo.participant",
  "cte.achievement",
  "avg.unemp.rate"
)]))

levels(master.model.ps.grad.region$RaceEthnicity)
levels(master.model.ps.grad.region$economic.status)
levels(master.model.ps.grad.region$SpecialEdStatus)
levels(master.model.ps.grad.region$english.learner)
levels(master.model.ps.grad.region$Gender)
levels(master.model.ps.grad.region$cte.achievement)
levels(master.model.ps.grad.region$pseo.participant)

prop.table(table(master.model.ps.grad.region$ps.grad.location.binary,
                 master.model.ps.grad.region$cte.achievement), 2)

prop.table(table(master.model.ps.grad.region$ps.grad.location.binary,
                 master.model.ps.grad.region$RaceEthnicity), 2)


```

Looks good.

This model estimates the association between demographic characteristics, academic preparation, and high school pathways with graduating from a postsecondary institution within the region, restricted to individuals who completed postsecondary education (n = 11,987). The outcome is coded 1 = graduated in-region and 0 = graduated outside the region.

The strongest positive predictors of regional graduation reflect local alignment pathways.

- **CTE concentrators/completers:** approximately 43% higher odds of graduating from a regional institution (OR ≈ 1.43).
- **CTE participants:** about 18% higher odds (OR ≈ 1.18).
- **Economic disadvantage:** roughly 57% higher odds of graduating in-region (OR ≈ 1.57).
- **American Indian students:** approximately 45% higher odds relative to White students (OR ≈ 1.45).
- **Special Education status:** about 32% higher odds (OR ≈ 1.32).

These effects suggest that students with stronger local ties or structural constraints are more likely to complete postsecondary education within the region rather than elsewhere.

Academic mobility indicators show the opposite pattern.

- **Higher MCA Math scores:** approximately 21% lower odds of graduating in-region per unit increase (OR ≈ 0.79).
- **PSEO participation:** about 11% lower odds (OR ≈ 0.89).
- **Female students:** roughly 9% lower odds (OR ≈ 0.91).

These patterns indicate that academically stronger and early-college students are more likely to complete degrees outside the region once they enroll in postsecondary education.

Racial differences shift substantially once restricting to completers.

- **Black students:** approximately 44% lower odds of graduating in-region (OR ≈ 0.56).
- **Asian/PI students:** about 39% lower odds (OR ≈ 0.61).
- **Hispanic students:** not statistically distinct from White students in this specification.

Local unemployment rates show a small negative association, though the magnitude is modest.

[**Overall Interpretation**]{.underline}

Once college completion is held constant, regional graduation patterns look very different from overall completion disparities. The strongest predictors of graduating within the region are CTE participation and indicators of local structural constraint, while higher academic achievement and early college exposure are associated with graduating outside the region.

This confirms that large disparities in regional graduation observed earlier are driven primarily by completion gaps rather than geography alone. Among those who complete college, regional sorting reflects a combination of local alignment pathways (CTE) and mobility gradients associated with academic strength.

Together with prior models, the pipeline becomes clear: structural factors strongly shape who completes college; among completers, CTE and local attachment increase the likelihood of graduating within the region; and regional graduation, in turn, is the dominant predictor of sustained regional employment.

```{r}

model_ps_grad_region <- glm(
  ps.grad.location.binary ~
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Gender +
    pseo.participant +
    MCA.M +
    cte.achievement +
    avg.unemp.rate,
  family = binomial,
  data = master.model.ps.grad.region
)

nobs(model_ps_grad_region)
AIC(model_ps_grad_region)

exp(cbind(
  OR = coef(model_ps_grad_region),
  confint(model_ps_grad_region)
))

or_plot <- tidy(model_ps_grad_region, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )


ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )

```

<br>

## **Model 3: An Academic Readiness Index**

The Model 2 specification above draws on a fixed set of predictors carried through this entire modeling sequence. Reviewing it against the bivariate analysis of graduation location revealed several variables that show a meaningful relationship with this outcome but weren't yet part of the model: district type, local wage levels, ACT scores, and MCA reading and science proficiency (only MCA math was included).

Adding all of these separately creates a problem. ACT scores and the three MCA subject scores are all measuring closely related things — a student's academic preparation — and are highly correlated with one another. Entering them as five separate variables in the same model makes the results unstable and hard to interpret: the model can't cleanly tell which of five overlapping measures is doing the work.

To address this, we constructed a single **Academic Readiness Index** to summarize academic engagement and achievement during high school. Each graduate received one point for each of the following they met, for a possible score of 0 to 6:

- Scored at or above proficiency on the MCA Math assessment
- Scored at or above proficiency on the MCA Reading assessment
- Scored at or above proficiency on the MCA Science assessment
- Scored 21 or higher on the ACT (the national median composite score)
- Participated in at least one AP exam
- Participated in Post-Secondary Enrollment Options (PSEO)

This index is not a measure of academic ability. It is a straightforward count of how much a student engaged with college-preparatory academics in high school.

**Deciding whether the index is the right approach**

Before adopting the index in place of its six individual components, we tested whether collapsing them into a single equally-weighted score was statistically defensible, using three checks:

1.  **A likelihood ratio test** comparing a model with the six components entered separately against a model using the single index. This test directly evaluates whether treating all six components as equally important is a reasonable simplification, or whether it masks meaningful differences between them.
2.  **A multicollinearity check** (variance inflation factors) on the model with all six components entered separately, to confirm whether those variables are in fact too closely related to reliably separate in the model.
3.  **An internal consistency check** among the six components themselves, to confirm they move together as a single underlying construct rather than capturing unrelated things.

Based on these results, we retained the individual components rather than adopting the composite index. A likelihood ratio test found that the disaggregated model fit significantly better than the index model (Δdeviance = 53.8, df = 5, p \< .001), indicating that equal weighting across the six components was not statistically supported. Variance inflation factors for the separate model were all low (\< 1.5), indicating that multicollinearity — the primary motivation for considering an index — was not in fact a meaningful concern. Internal consistency across the six items was also modest (Cronbach's α = 0.61), reflecting the fact that test-based measures (MCA scores, ACT) and participation-based measures (AP exam, PSEO) do not move together as a single underlying construct.

```{r}
# Build master base dataset

master.test <- master %>%
  filter(ps.grad == "Y") %>%
  select(PersonID,Gender, RaceEthnicity, economic.status, SpecialEdStatus, Dem_Desc, avg.wages.pct.state, avg.unemp.rate, pseo.participant, ACTCompositeScore, MCA.M, MCA.R, MCA.S, took.ACT, cte.achievement, ap.exam, ps.grad.location, grad.year.5) %>%
  filter(grad.year.5 != "After 2023",
         grad.year.5 != "Attending ps",
         ps.grad.location != "Did not graduate ps") %>%
  mutate(grad.year.5 = ifelse(grad.year.5 %in% c("Meaningful emp region", "Meaningful emp County", "Meaningful emp EDR"), "Meaningful emp region", grad.year.5),
         grad.year.5 = str_replace(grad.year.5, ", not attending ps", ""),
         grad.year.5 = ifelse(grad.year.5 == "Meaningful emp region", 1, 0),
         ps.grad.location = ifelse(ps.grad.location == "PS grad Region", 1, 0))


# ------------------------------------------------------------------
# 0. Build the six binary readiness components
# ------------------------------------------------------------------
# NOTE ON MCA THRESHOLD: assuming the standard MN MCA achievement
# scale (1 = does not meet, 2 = partially meets, 3 = meets, 4 = exceeds),
# "at or above proficiency" = level 3 or 4. Confirm this matches your
# coding of MCA.M/.R/.S before trusting the output.

master.test.index <- master.test %>%
  mutate(
    mca.m.prof = ifelse(MCA.M >= 3, 1, 0),
    mca.r.prof = ifelse(MCA.R >= 3, 1, 0),
    mca.s.prof = ifelse(MCA.S >= 3, 1, 0),
    act.prof   = ifelse(ACTCompositeScore >= 21, 1, 0),
    ap.prof    = ifelse(ap.exam >= 1, 1, 0),
    pseo.prof  = ifelse(pseo.participant == "1", 1, 0)
  )
# ------------------------------------------------------------------
# 1. Missingness decision point (still open from earlier)
# ------------------------------------------------------------------
# ACT and AP exam are selected populations - not everyone takes them.
# This code takes the COMPLETE-CASE approach: only students with all six
# components observed are included in the comparison below. This keeps
# Model A and Model B on identical samples, which the LRT requires.
# If you'd rather code missing-as-0, swap this filter out - but flag
# that choice explicitly in the write-up if you do, since it changes
# what the index is measuring for those students.

readiness_vars <- c("mca.m.prof", "mca.r.prof", "mca.s.prof",
                    "act.prof", "ap.prof", "pseo.prof")

model_data <- master.test.index %>%
  filter(if_all(all_of(readiness_vars), ~ !is.na(.))) %>%
  mutate(
    academic.readiness.index = mca.m.prof + mca.r.prof + mca.s.prof +
      act.prof + ap.prof + pseo.prof
  )

nrow(model_data)  # compare this to nobs() from the original Model 2 (n = 11,987) - report the drop

# ------------------------------------------------------------------
# 2. Model A: six components entered separately
# ------------------------------------------------------------------
model_a_separate <- glm(
  ps.grad.location ~
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    Gender +
    Dem_Desc +
    avg.wages.pct.state +
    avg.unemp.rate +
    cte.achievement +
    mca.m.prof +
    mca.r.prof +
    mca.s.prof +
    act.prof +
    ap.prof +
    pseo.prof,
  family = binomial,
  data = model_data
)

summary(model_a_separate)

# ------------------------------------------------------------------
# 3. Model B: single Academic Readiness Index
# ------------------------------------------------------------------
model_b_index <- glm(
  ps.grad.location ~
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    Gender +
    Dem_Desc +
    avg.wages.pct.state +
    avg.unemp.rate +
    cte.achievement +
    academic.readiness.index,
  family = binomial,
  data = model_data
)

summary(model_b_index)

# ------------------------------------------------------------------
# 4. Likelihood ratio test - the key decision test
# ------------------------------------------------------------------
anova(model_b_index, model_a_separate, test = "Chisq")

# Non-significant p-value -> equal weighting is defensible, use the index
# Significant p-value -> the six components diverge, consider keeping separate

# ------------------------------------------------------------------
# 5. Multicollinearity check on Model A
# ------------------------------------------------------------------
library(car)
vif(model_a_separate)

# VIFs above ~5 on the readiness components support using the index
# regardless of what the LRT shows

# ------------------------------------------------------------------
# 6. Internal consistency of the six components
# ------------------------------------------------------------------
library(psych)

readiness_matrix <- model_data %>%
  select(all_of(readiness_vars))

tetrachoric(readiness_matrix)   # correlation matrix for binary items
alpha(readiness_matrix)         # Cronbach's alpha

```

<br>

## Model 4 - reworking college completion region

```{r}
master.model.ps.grad.region.2 <- master.test 

test <- master.model.ps.grad.region.2 %>%
  filter(is.na(pseo.participant))
names(master.model.ps.grad.region.2)
```

<br>

# Synthesis of Analysis – Northeast Minnesota

Across a series of logistic regression models—progressively specifying high school preparation, postsecondary credential attainment, postsecondary geography, college completion, and regional graduation among completers—a clear and layered mechanism emerges. Sustained regional employment five years after high school is driven primarily by postsecondary geographic embeddedness, reinforced by CTE participation, and shaped throughout by strong structural disparities that begin at college completion.

Academic readiness increases college completion. Regional graduation anchors workforce retention. These are related but distinct mechanisms.

<br>

## Big Takeaways

**1. Regional postsecondary graduation is the dominant workforce retention mechanism.**\
Graduating from a postsecondary institution within the region is the single strongest predictor of sustained regional employment.

- **Regional graduation (any credential):** \~142% to \~197% higher odds of sustained regional employment (OR ≈ 2.42–2.97 across models).
- Graduating elsewhere in Minnesota: \~44% lower odds.
- Graduating outside Minnesota: \~36% lower odds.

Once graduation location is included, credential level largely loses its independent effect. Associate degrees and sub-baccalaureate credentials no longer predict retention once geography is accounted for. Where the degree is earned matters far more than what degree is earned.

Geography—not credential depth—is the decisive workforce anchor.

<br>

**2. College completion is the most structurally stratified stage of the pipeline.**\
Before geography even enters the picture, completion itself is highly unequal.

In the postsecondary completion model:

- **Special Education status:** \~62% lower odds of completion (OR ≈ 0.38).
- **Economic disadvantage:** \~59% lower odds (OR ≈ 0.41).
- **American Indian students:** \~51% lower odds.
- **Black students:** \~46% lower odds.
- **Hispanic students:** \~27% lower odds.

In contrast:

- **Female students:** \~66% higher odds of completion.
- **PSEO participation:** \~44% higher odds.
- **Higher MCA Math:** \~29% higher odds per unit increase.

These results show that the largest structural sorting in the education-to-workforce pipeline occurs at college completion—not at regional sorting among graduates.

Regional graduation disparities observed earlier are largely downstream of completion gaps.

<br>

**3. Among completers, regional sorting reflects local alignment versus mobility gradients.**\
When restricting to students who completed postsecondary education, the geography mechanism becomes clearer.

In the regional graduation model among completers:

- **CTE concentrators/completers:** \~43% higher odds of graduating in-region.
- **CTE participants:** \~18% higher odds.
- **Economic disadvantage:** \~57% higher odds of graduating in-region.
- **American Indian students:** \~45% higher odds.
- **Special Education status:** \~32% higher odds.

In contrast, academic mobility indicators reduce regional graduation:

- **Higher MCA Math:** \~21% lower odds per unit increase.
- **PSEO participation:** \~11% lower odds.
- **Female students:** \~9% lower odds.

This pattern suggests that academically stronger students and early-college participants are more likely to complete degrees outside the region, while locally aligned or structurally constrained students are more likely to remain within the regional postsecondary system.

Regional graduation is therefore shaped by both local attachment and mobility expansion.

<br>

**4. CTE operates through both educational geography and direct workforce alignment.**\
CTE does not significantly increase overall college completion. However, among completers, CTE increases the likelihood of graduating in-region. In workforce models, CTE remains positively associated with sustained regional employment even after controlling for graduation location.

- **CTE concentrators/completers:** \~29–38% higher odds of sustained regional employment in workforce models.
- **CTE participants:** \~16–20% higher odds.

This indicates that CTE operates partly through regional educational pathways and partly as a direct workforce-alignment mechanism.

CTE matters—but its effect is modest relative to postsecondary geography.

<br>

**5. Academic readiness aligns with opportunity expansion rather than regional anchoring.**\
Higher MCA scores and PSEO participation increase the likelihood of college completion but reduce the likelihood of regional graduation among completers. In workforce models, higher MCA scores are modestly associated with lower regional retention.

This pattern is consistent:

- Academic readiness increases postsecondary progression.
- But it expands geographic mobility rather than reinforcing regional attachment.

Academic strength facilitates opportunity beyond the region.

<br>

**6. Structural disparities persist into workforce outcomes.**\
Even after accounting for preparation, completion, and graduation location, disparities remain in sustained regional employment.

In workforce models:

- **Black students:** \~43–48% lower odds.
- **Asian/PI students:** \~32–38% lower odds.
- **Hispanic students:** \~22–24% lower odds (weaker in fully specified models).
- **Economic disadvantage:** \~9–12% lower odds.
- **Special Education status:** \~17–22% lower odds.
- **Female students:** \~9–11% lower odds.

These gaps are not eliminated by controlling for credential level or graduation location. Structural inequality remains embedded throughout the pipeline.

<br>

## Relative Strength of Mechanisms

Across all models, the ordering of influence is clear:

1.  **Regional postsecondary graduation** → strongest predictor of sustained regional employment.
2.  **College completion** → most structurally stratified stage.
3.  **CTE participation** → consistent but secondary alignment mechanism.
4.  **Academic readiness** → promotes completion but aligns with mobility rather than retention.
5.  **Structural background factors** → shape completion and persist into workforce outcomes.

<br>

## Overall Conclusion

Sustained regional employment in Northeast Minnesota is shaped by a layered pipeline:

- Structural factors strongly influence who completes college.
- Among completers, local alignment pathways—especially CTE—shape who graduates within the region.
- Regional graduation is the dominant anchor for long-term regional workforce attachment.

Credential attainment alone does not drive retention. Geography does.

Academic readiness expands opportunity but does not anchor students regionally. Four-year and outward pathways align with mobility rather than local retention. CTE contributes independently, both through regional educational pathways and through more direct workforce alignment.

Most importantly, structural disparities emerge at college completion and persist through regional graduation and into workforce outcomes. Educational pathways explain part of regional retention, but they do not fully account for persistent racial and socioeconomic gaps.

In short, Northeast Minnesota’s workforce retention pipeline is geographically anchored, institutionally embedded, partially reinforced by career-aligned high school preparation, and constrained by enduring structural inequality that begins well before workforce entry.

<br>

# What This Analysis Means for Northeast Minnesota Economic & Workforce Developers

These findings reflect strong statistical associations rather than proven causal effects. However, the consistency and magnitude of the relationships provide clear directional insight for regional workforce strategy.

If the goal is to keep more young people in Northeast Minnesota to fill workforce shortages, the data point to one dominant pattern: Students who graduate from a postsecondary institution within the region are far more likely to be working in the region five years later.

Graduating from a regional institution is associated with roughly **2.4 to 3 times higher odds** of sustained regional employment compared to students who do not graduate in-region. This is the single strongest relationship observed in the analysis.

In practical terms: Students who earn their degree here are much more likely to be employed here later.

<br>

## 1. Regional Colleges Are Strongly Linked to Workforce Retention

Graduating from a regional postsecondary institution is associated with approximately **142% to 197% higher odds** of sustained regional employment, even after accounting for academic preparation, demographics, and economic background.

By contrast:

- Graduating elsewhere in Minnesota is associated with roughly **44% lower odds** of regional employment.
- Graduating outside Minnesota is associated with roughly **36% lower odds**.

These results suggest that workforce retention is closely linked to **institutional embeddedness and local ties**. The location of postsecondary completion is far more strongly associated with retention than credential level alone.

For workforce developers, this implies that strengthening partnerships with regional colleges—especially 2-year public institutions—may be one of the most effective levers available.

<br>

## 2. The Largest Sorting Occurs at College Completion

Before geography even enters the picture, the most substantial differences appear at college completion.

Students who are:

- Economically disadvantaged have about **59% lower odds of completing college**.
- In Special Education have about **62% lower odds**.
- American Indian students have about **51% lower odds**.
- Black students have about **46% lower odds**.

This indicates that many potential regional workers are lost earlier in the pipeline.

Improving college completion—particularly for low-income and historically underrepresented students—appears strongly linked to increasing the pool of individuals who could later contribute to the regional workforce.

Retention is associated with geography, but supply begins with completion.

<br>

## 3. CTE Is a Consistent but Secondary Lever

Career and Technical Education (CTE) participation shows stable positive associations:

- CTE concentrators/completers have about **29–38% higher odds** of sustained regional employment.
- Among students who complete college, CTE concentrators are about **43% more likely** to graduate from a regional institution.

CTE is not strongly associated with overall college completion, but it is associated with both regional graduation and regional employment. This suggests that CTE pathways are linked to local workforce alignment in ways that extend beyond postsecondary attainment alone.

Expanding high-quality CTE programs aligned with regional industry needs may therefore support workforce retention, even if their effect is more modest than that of regional college graduation.

<br>

## 4. Academic Readiness Expands Opportunity—Often Beyond the Region

Higher academic performance increases college completion:

- Higher MCA Math scores are associated with about **29% higher odds of completion per unit increase**.
- PSEO participation is associated with about **44% higher odds of completion**.

However, among students who complete college:

- Higher MCA scores are associated with about **21% lower odds of graduating in-region**.
- PSEO participants are about **11% less likely** to graduate in-region.

These patterns suggest that academically strong students are more geographically mobile. Academic readiness increases progression into postsecondary education, but it is not strongly associated with regional retention.

For workforce developers, this may mean that retaining high-achieving students requires earlier employer engagement, internship pipelines, and visible regional career opportunities.

<br>

## 5. Structural Disparities Persist Into Workforce Outcomes

Even after accounting for preparation, completion, and graduation location:

- Black students have about **43–48% lower odds** of sustained regional employment.
- Asian/PI students have about **32–38% lower odds**.
- Economically disadvantaged students have about **9–12% lower odds**.
- Students in Special Education have about **17–22% lower odds**.

Educational pathways explain part of the variation in workforce retention, but not all of it. Structural disparities remain associated with employment outcomes even after controlling for key educational factors.

This suggests that workforce strategies may also need to address hiring practices, advancement pathways, and inclusive workplace supports—not just educational alignment.

<br>

## Strategic Implications for the Region

The analysis points to a layered system:

1.  **College completion is the most structurally unequal stage.**
2.  **Regional postsecondary graduation is the strongest predictor of regional workforce attachment.**
3.  **CTE provides a consistent, though secondary, alignment pathway.**
4.  **Academic strength increases completion but is associated with greater geographic mobility.**

Taken together, the results suggest that strengthening regional colleges, improving completion rates, expanding aligned CTE pathways, and building clearer bridges from education to local employment are all strongly associated with higher regional workforce attachment.

Geography is the strongest anchor observed in the data.\
Completion is the largest structural barrier.\
CTE is a meaningful alignment tool.

While these findings do not establish causation, the magnitude and consistency of the associations provide clear guidance for regional workforce development strategy.
