---
title: "Post-secondary pathways"
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
library(htmlwidgets)
library(RColorBrewer)
library(readxl)
library(janitor)
library(lubridate)
library(systemfonts)
reset_font_cache()
library(ggtext)
library(gmodels)
library(fastDummies)
library(car)
library(glmnet)
library(glmnetUtils)
library(pscl)
library(sjPlot)
library(ggforce)
library(openxlsx)
```

```{r themes and shapefiles, include=FALSE}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", linewidth  = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.key.size = unit(1, "lines"),
        legend.margin = margin(0,0,0,0),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.text = element_text(margin = margin(l = 2)),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", linewidth = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))


theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
    select(5,6) %>%
    unique() %>%
    mutate(edr = str_replace(edr, "  ", " "),
           planning.region = str_replace(planning.region, " Minnesota", ""),
           planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
           edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

counties.regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
  rename(mif = `MIF Region`) %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Name = str_replace(Name, "Mcleod", "McLeod"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc) ,
         edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"))

counties.regions.1 <- counties.regions %>%
  mutate(statefp = "27",
         project.pr = ifelse(edr %in% c("EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 4 - West Central"), "Northwest",
                             ifelse(edr == "EDR 3 - Arrowhead", "Northeast",
                                    ifelse(edr %in% c("EDR 5 - North Central", "EDR 7E- East Central", "EDR 7W- Central"), "Central",
                                           ifelse(edr %in% c("EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 8 - Southwest"), "Southwest",
                                                  ifelse(edr %in% c("EDR 9 - South Central", "EDR 10 - Southeast"), "Southern", as.character(planning.region)))))),
         project.pr = fct_relevel(project.pr, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southern"))


color.ruca <- c("Entirely rural" = "#009933", "Town/rural mix" = "#99CC33", "Urban/town/rural mix" = "#CC9966", "Entirely urban" = "#754C29", "Minnesota" = "black")

color.pr <- c("Northwest" = 	"#4575b4", "Northeast" = "grey", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.edr.simple <- c("EDR 1" = "#b3cde3", "EDR 2" = "#8c96c6", "EDR 3" = "#fe9929", "EDR 4" = "#8856a7", "EDR 5" = "#810f7c", "EDR 6E" = "#e5f5f9", "EDR 6W" = "#bdc9e1", "EDR 7E" = "#99d8c9", "EDR 7W" = "#2ca25f", "EDR 8" = "#74a9cf", "EDR 9" = "#0570b0", "EDR 10" = "#d7301f", "EDR 11" = "#d8b365", "Minnesota" = "black")

color.pr.edr <- c ("Northwest" = "#4575b4","Northeast" = "#e0f3f8", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black", "EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365")

mn_counties <- st_read("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Shapefiles/County shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE) %>%
  rename(countyfp = FIPS_CODE)
```

```{r master}
master.original <- read_csv("Data/SLEDS/Masters/After analysis/Master-after-ct.csv") 

kable(names(master.original))

kable(head(master.original))
```

# Dependent Variable

The cross-tab and ANOVA analysis exploring relationships between demographic, high school, and post-secondary variables with employment outcomes highlighted the importance of post-secondary variables in an individuals journey to meaningful employment inside and outside the region. Due to this, it makes sense to explore if some of the other variables play a role in post-secondary outcomes - whether an individual attends, graduates, the type of institution they attend and the location.

We will explore this here. We will explore this in three sections;

[**Tier 1: Core Postsecondary Access and Attainment**]{.underline}

These outcomes are most foundational and closely tied to employment.

1\. `attended.ps.within.first.year.hsgrad`

**Question:** Who transitions directly to postsecondary? **Why it matters:** Strongly linked to long-term success and minimizes "disconnection."

2\. `attended.ps` (ever attended postsecondary)

**Question:** Who attends postsecondary at all? **Why it matters:** Broader view than immediate enrollment.

3\. `ps.grad` (graduated from postsecondary)

**Question:** Who completes a credential? **Why it matters:** Credential completion is often the real driver of employment outcomes.

[**🔹 Tier 2: Type of Postsecondary Experience**]{.underline}

This outcome differentiate the **nature** of the experience, which may shape career pathways and economic outcomes.

4\. `highest.cred.level`

**Question:** Who earns certificates, associate degrees, bachelor’s degrees? **Why it matters:** Different credentials yield different labor market returns.

[**Tier 3: Location of Postsecondary Experience**]{.underline}

These outcomes reveal **mobility**, local engagement, or place-based equity patterns.

5\. `ps.grad.location`

**Question:** Who stays in the region for school? **Why it matters:** Tied to affordability and regional workforce planning.

Each dependent variable has the following categories;

```{r dependent variable categories}
data <- master.original %>%
  select(attended.ps.within.first.year.hsgrad, attended.ps, ps.grad, ps.grad.InstitutionSector, highest.cred.level, ps.grad.location) %>%
  mutate(across(c("attended.ps.within.first.year.hsgrad":"ps.grad.location"), factor))
```

**attended.ps.within.first.yer.hsgrad**

```{r echo = FALSE}
levels(data$attended.ps.within.first.year.hsgrad)

```

**attended.ps**

```{r echo = FALSE}
levels(data$attended.ps)
```

**ps.grad**

```{r echo = FALSE}
levels(data$ps.grad)
```

**highest.cred.level**

```{r echo = FALSE}
levels(data$highest.cred.level)
```

**ps.grad.location**

```{r echo = FALSE}
levels(data$ps.grad.location)
```

<br>

# Predictors

For all of these, we will use the predictors that were in Tier 1 and Tier 2 of the previous cross-tab analysis.

-   Demographics
    -   `economic.status`
    -   `RaceEthnicity`
    -   `SpecialEdStatus`
-   High school characteristics
    -   `Dem_Desc`
    -   `avg.unemp.rate`
    -   `avg.wage.pct.state`
-   High school accomplishments
    -   `ACTCompositeScore`
    -   `pseo.participant`
    -   `cte.achievement`
    -   `total.cte.courses.taken`
    -   `MCA.M`
    -   `MCA.R`
    -   `MCA.S`

<br>

# [**Tier 1: Core Postsecondary Access and Attainment**]{.underline}

Tier 1 examined how demographic background, high school context, and high school accomplishments predict three key milestones: (1) **immediate enrollment** (“attended.ps.within.first.year.hsgrad”), (2) **ever attending postsecondary** (“attended.ps”), and (3) **completing a credential** (“ps.grad”).

------------------------------------------------------------------------

## **Key Findings**

1.  **Demographics are the strongest and most consistent predictors across all three outcomes.**
    -   **Economic Status**: Economically disadvantaged students lagged behind at every stage—less likely to enroll immediately, less likely to attend at all, and much less likely to graduate.
    -   **Race/Ethnicity**: White and Asian/PI students consistently had higher enrollment and graduation rates, while American Indian, Black, and Hispanic students faced persistent barriers at all stages, with the largest disparities at completion.
    -   **Special Education Status**: Students with special education status showed markedly lower rates of enrollment, attendance, and especially graduation.
2.  **High School Characteristics showed weak associations.**
    -   **School Locale**: Rural, mixed, and urban settings had minimal differences in postsecondary pathways.
    -   **Community Economic Conditions**: Students from lower-unemployment, higher-wage areas were modestly more likely to enroll, attend, and graduate, but these effects were smaller than individual-level demographics or accomplishments.
3.  **High School Accomplishments were powerful predictors, especially for enrollment and graduation.**
    -   **Academic Achievement**: ACT and MCA scores (particularly Reading and Science) were strongly linked to both immediate enrollment and graduation, showing that academic readiness is a key driver of persistence and attainment.
    -   **PSEO Participation**: The single strongest programmatic predictor across all outcomes. PSEO students were much more likely to enroll immediately, attend at all, and eventually graduate.
    -   **CTE Involvement**: Higher CTE course-taking and concentrator/completor status were associated with lower likelihood of both enrollment and graduation, suggesting that intensive CTE pathways align more with direct workforce entry than with postsecondary progression.

------------------------------------------------------------------------

## **Overall Conclusion**:

Tier 1 analysis reveals that **postsecondary access and attainment are shaped by both structural inequities and academic preparation**. Demographic barriers—economic disadvantage, special education status, and race/ethnicity—consistently depress outcomes at every stage. Academic readiness (ACT and MCA scores) and program participation (especially PSEO) are the most powerful positive predictors of successful transitions and completion. In contrast, CTE intensity is inversely associated with postsecondary outcomes, reflecting its alignment with alternative workforce pathways.

Together, these findings suggest that strategies to improve postsecondary success must focus on **closing equity gaps, bolstering academic readiness, and expanding access to PSEO and similar dual-enrollment programs**, while also balancing the role of CTE in supporting both workforce entry and potential academic progression.

<br>

## `attended.ps.within.first.year.hsgrad`

### Summary

**Key Findings by Variable Group**

**Demographics**

-   **Economic Status**: Cramer’s V = 0.246. Non-disadvantaged students had much higher immediate enrollment (72.1%) than disadvantaged students (47.6%). Economic disadvantage is a major barrier to direct enrollment.\
-   **Race/Ethnicity**: Cramer’s V = 0.096. White (67%) and Asian/PI (58.7%) students enrolled at higher rates, while American Indian (45.8%), Black (51.4%), and Hispanic (51.6%) students enrolled at lower rates.\
-   **Special Education Status**: Cramer’s V = 0.278. Students without special education status were far more likely to enroll immediately (68.6%) than those with special education status (32.6%).

**High School Characteristics**

-   **School Locale (Dem_Desc)**: Cramer’s V = 0.022. Very weak differences across rural, mixed, and urban contexts (all \~62%).\
-   **Average Unemployment Rate**: F(2, 49813) = 302.20, p \< .001. Students from higher-unemployment areas were less likely to enroll immediately.\
-   **Average Wages (as % of State Average)**: F(2, 49813) = 33.08, p \< .001. Students from higher-wage areas were slightly more likely to attend postsecondary (immediately or later).

**High School Accomplishments**

-   **ACT Composite Score**: F(2, 28493) = 396.75, p \< .001. Higher ACT scores strongly predicted immediate enrollment, with immediate enrollees scoring about 2 points higher than non-attendees.\
-   **PSEO Participation**: Cramer’s V = 0.295. PSEO participants enrolled immediately at very high rates (76.8%) compared to non-participants (48.2%).\
-   **CTE Achievement**: Cramer’s V = 0.052. CTE concentrators/completors were less likely to enroll immediately (57.9%) than participants (64.9%) or non-CTE students (65.7%).\
-   **Total CTE Courses Taken**: F(2, 49813) = 301.90, p \< .001. More CTE courses were associated with lower immediate enrollment.\
-   **MCA Math**: F(2, 47621) = 885.58, p \< .001. Immediate enrollees scored much higher (0.50 points above delayed enrollees).\
-   **MCA Reading**: F(2, 47319) = 1475.87, p \< .001. Immediate enrollees scored substantially higher (0.46 points above non-attendees).\
-   **MCA Science**: F(2, 40378) = 1025.37, p \< .001. Immediate enrollees scored much higher (0.41 points above non-attendees); no difference between delayed and non-attendees.

------------------------------------------------------------------------

### **Relative Importance of Predictors (Ranked by Effect Size/Strength of Association):**

1.  **Special Education Status (Cramer’s V = 0.278)** – strong demographic predictor.
2.  **PSEO Participation (Cramer’s V = 0.295)** – strongest programmatic predictor.
3.  **Economic Status (Cramer’s V = 0.246)** – major structural barrier.
4.  **MCA Reading (F = 1475.87)** – strongest academic achievement predictor.
5.  **MCA Science (F = 1025.37)** – strong academic predictor.
6.  **MCA Math (F = 885.58)** – strong academic predictor.
7.  **ACT Composite Score (F = 396.75)** – important test-based predictor.
8.  **Average Unemployment Rate (F = 302.20)** – moderate community context effect.
9.  **Total CTE Courses Taken (F = 301.90)** – more courses linked to weaker transition.
10. **Race/Ethnicity (Cramer’s V = 0.096)** – modest but meaningful disparities.
11. **Average Wages (F = 33.08)** – small community effect.
12. **CTE Achievement (Cramer’s V = 0.052)** – weak association.
13. **School Locale (Cramer’s V = 0.022)** – negligible differences.

------------------------------------------------------------------------

**Conclusion**:\
Immediate postsecondary enrollment is shaped most strongly by **student characteristics (economic status, special education status, race/ethnicity)**, **program participation (PSEO)**, and **academic achievement (MCA and ACT scores)**. Community economic conditions (unemployment and wages) also matter but to a lesser extent, while school locale and CTE achievement have minimal influence. This suggests that interventions targeting equity gaps, strengthening academic preparation, and expanding access to PSEO could have the greatest impact on improving direct transitions to postsecondary education.

### Demographics

**Key Findings**:

-   **Economic Status**: Cramer’s V = 0.246, p \< .001. Non-disadvantaged students enrolled immediately at much higher rates (72.1%) than economically disadvantaged students (47.6%). Economically disadvantaged students were more likely to never attend (46.1% vs. 24.0%).
-   **Race/Ethnicity**: Cramer’s V = 0.096, p \< .001. White (67%) and Asian/PI (58.7%) students had the highest immediate enrollment. American Indian students had the lowest (45.8%), while Black (51.4%) and Hispanic (51.6%) students also enrolled at below-average rates.
-   **Special Education Status**: Cramer’s V = 0.278, p \< .001. Students without special education status were much more likely to enroll immediately (68.6%) than those with special education status (32.6%). Special education students were more likely to never attend (62.2% vs. 26.7%).

**Conclusion**:\
Demographic factors play a significant role in shaping immediate postsecondary enrollment. Economic disadvantage and special education status are linked to substantially lower rates of transition, with both groups far more likely to never enroll. Race and ethnicity also matter, with White and Asian/PI students showing stronger transitions and American Indian, Black, and Hispanic students facing greater barriers. Overall, these patterns highlight persistent equity gaps in postsecondary access immediately after high school.

::: panel-tabset
#### Economic Status → Attended Postsecondary Within First Year of High School Graduation

**Key Findings**:\
Chi-square = 3003.42, Cramer’s V = 0.246, p \< .001. Immediate enrollment (“Yes”) was much higher among students not economically disadvantaged (72.1%) than among economically disadvantaged students (47.6%). Economically disadvantaged students were more likely to never attend postsecondary (“Did not attend ps”: 46.1% vs. 24.0%) and somewhat more likely to delay enrollment beyond the first year (“No”: 6.3% vs. 3.9%).

**Conclusion**:\
Economic disadvantage is moderately and significantly associated with lower immediate postsecondary enrollment and higher non-enrollment. Students not economically disadvantaged transition directly to college at much higher rates, while economically disadvantaged students are more likely to delay or never attend.

```{r}

tab_xtab(var.row = master.original$attended.ps.within.first.year.hsgrad,
         var.col = master.original$economic.status,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/attended-ps-within-first-year-hsgrad/Demographics/economic-status.html")
```

#### Race/Ethnicity → Attended Postsecondary Within First Year of High School Graduation (Unknown Removed)

**Key Findings**:\
Chi-square = 759.71, Cramer’s V = 0.096, p \< .001. Immediate postsecondary enrollment (“Yes”) was highest among White students (67%), followed by Asian/PI (58.7%), Hispanic (51.6%), and Black students (51.4%). American Indian students had the lowest immediate enrollment rate (45.8%). Non-enrollment (“Did not attend ps”) was most common among American Indian students (48.2%) and least common among White students (27.6%).

**Conclusion**:\
Race and ethnicity remain significantly associated with immediate postsecondary enrollment, though the effect size is modest. White and Asian/PI students transition more directly to postsecondary, while American Indian students face the greatest barriers, with Black and Hispanic students also enrolling at lower-than-average rates.

```{r}
data <- master.original %>%
  filter(RaceEthnicity != "Unknown")

tab_xtab(var.row = data$attended.ps.within.first.year.hsgrad,
         var.col = data$RaceEthnicity,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/attended-ps-within-first-year-hsgrad/Demographics/RaceEthnicity.html")
```

#### Special Education Status → Attended Postsecondary Within First Year of High School Graduation

**Key Findings**:\
Chi-square = 3844.40, Cramer’s V = 0.278, p \< .001. Students without special education status had a much higher immediate postsecondary enrollment rate (68.6%) compared to students with special education status (32.6%). Students in special education were far more likely to never attend postsecondary (62.2% vs. 26.7%). Rates of delayed enrollment (“No”) were similar between groups (about 5%).

**Conclusion**:\
Special education status is strongly associated with immediate postsecondary enrollment. Students with special education status are substantially less likely to transition directly into postsecondary and much more likely to never enroll at all.

```{r}
tab_xtab(var.row = master.original$attended.ps.within.first.year.hsgrad,
         var.col = master.original$SpecialEdStatus,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/attended-ps-within-first-year-hsgrad/Demographics/SpecialEdStatus.html")
```
:::

<br>

### High School Characteristics

**Key Findings**:

-   **School Locale (Dem_Desc)**: Cramer’s V = 0.022, p \< .001. Enrollment patterns were nearly identical across locales: rural (61.8%), town/rural mix (61.3%), and urban/town/rural mix (63.8%). The association is statistically significant but negligible in size.
-   **Average Unemployment Rate**: F(2, 49813) = 302.20, p \< .001. Students from higher-unemployment areas were less likely to enroll immediately, while immediate enrollees came from communities with relatively lower unemployment.
-   **Average Wages (as % of State Average)**: F(2, 49813) = 33.08, p \< .001. Students from higher-wage areas were somewhat more likely to attend postsecondary (either immediately or after a delay). Differences between immediate and delayed enrollment were not significant.

**Conclusion**:\
High school characteristics show weaker associations with immediate postsecondary enrollment than demographics. School locale has almost no impact, while local economic context matters somewhat—students in areas with higher unemployment and lower wages are less likely to transition into postsecondary. These results suggest that while community economic conditions shape opportunity, school setting itself does not strongly influence immediate enrollment.

<br>

::: panel-tabset
#### School Locale (Dem_Desc) → Attended Postsecondary Within First Year of High School Graduation

**Key Findings**:\
Chi-square = 49.99, Cramer’s V = 0.022, p \< .001. Immediate enrollment (“Yes”) rates were similar across locales: entirely rural (61.8%), town/rural mix (61.3%), and urban/town/rural mix (63.8%). Non-enrollment (“Did not attend ps”) was slightly higher in town/rural mix schools (34.5%) compared to urban/town/rural mix (31.3%) and entirely rural (32.8%).

**Conclusion**:\
School locale shows a statistically significant but very weak association with immediate postsecondary enrollment. Differences across rural, mixed, and urban/town/rural contexts are small, suggesting that school setting itself is not a major driver of postsecondary transition.

```{r}
tab_xtab(var.row = master.original$attended.ps.within.first.year.hsgrad,
         var.col = master.original$Dem_Desc,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/attended-ps-within-first-year-hsgrad/High school characteristics/Dem_Desc.html")
```

#### Average Unemployment Rate → Attended Postsecondary Within First Year of High School Graduation

**Key Findings**:\
ANOVA: F(2, 49813) = 302.20, p \< .001.\
Mean unemployment rates differed significantly across groups:

-   Did not attend postsecondary: highest average unemployment rate.
-   Delayed enrollment (“No”): about 0.68 percentage points lower.
-   Immediate enrollment (“Yes”): about 0.26 percentage points lower than those who never attended, and 0.41 points lower than those who delayed enrollment.

**Conclusion**:\
Students from areas with higher average unemployment rates are less likely to enroll in postsecondary immediately after high school. Immediate enrollees tend to come from communities with relatively lower unemployment, suggesting local labor market conditions may influence the timing of postsecondary transition.

```{r}
summary <- aov(avg.unemp.rate ~ attended.ps.within.first.year.hsgrad, data = master.original)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/attended-ps-within-first-year-hsgrad/High school characteristics/avg-unemp-rate.xlsx", overwrite = TRUE)


```

#### Average Wages (as % of Minnesota State Average) → Attended Postsecondary Within First Year of High School Graduation

**Key Findings**:\
ANOVA: F(2, 49813) = 33.08, p \< .001.

-   Students who never attended postsecondary were from areas with slightly lower average wages.
-   Immediate enrollees (“Yes”) came from areas with wages about 0.39 percentage points higher than those who never attended.
-   Delayed enrollees (“No”) also came from areas with slightly higher wages than non-attendees (0.45 points higher).
-   No significant difference was found between immediate and delayed enrollees.

**Conclusion**:\
Average local wages are modestly associated with postsecondary enrollment patterns. Students from communities with somewhat higher wages are more likely to pursue postsecondary education—whether immediately or after a delay—while students from lower-wage areas are more likely to never attend.

```{r}
summary <- aov(avg.wages.pct.state ~ attended.ps.within.first.year.hsgrad, data = master.original)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/attended-ps-within-first-year-hsgrad/High school characteristics/avg-wage-pct-state.xlsx", overwrite = TRUE)
```
:::

<br>

### High School Accomplishments

::: panel-tabset
#### ACT Composite Score → Attended Postsecondary Within First Year of High School Graduation

**Key Findings**:\
ANOVA: F(2, 28493) = 396.75, p \< .001.

-   Students who never attended postsecondary had the lowest ACT Composite Scores.
-   Delayed enrollees scored about 0.9 points higher on average than non-attendees.
-   Immediate enrollees scored about 2.0 points higher than non-attendees, and about 1.1 points higher than delayed enrollees.

**Conclusion**:\
ACT performance is strongly associated with immediate postsecondary enrollment. Higher ACT scores are linked to a greater likelihood of transitioning directly to postsecondary, while lower scores are tied to delaying or never attending.

```{r}
summary <- aov(ACTCompositeScore ~ attended.ps.within.first.year.hsgrad, data = master.original)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/attended-ps-within-first-year-hsgrad/High school accomplishments/ACTCompositeScore.xlsx", overwrite = TRUE)

```

#### PSEO Participation → Attended Postsecondary Within First Year of High School Graduation

**Key Findings**:\
Chi-square = 3454.84, Cramer’s V = 0.295, p \< .001. Immediate enrollment (“Yes”) was much higher among PSEO participants (76.8%) compared to non-participants (48.2%). Non-participants were more likely to never attend postsecondary (46.7% vs. 20.2%). Delayed enrollment was low for both groups, but slightly higher among non-participants (5.1% vs. 3.0%).

**Conclusion**:\
PSEO participation is strongly associated with immediate postsecondary enrollment. Students who participated in PSEO were far more likely to transition directly to postsecondary and much less likely to never attend, highlighting the program’s role in supporting successful transitions.

```{r}
tab_xtab(var.row = master.original$attended.ps.within.first.year.hsgrad,
         var.col = master.original$pseo.participant,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/attended-ps-within-first-year-hsgrad/High school accomplishments/pseo-parcipant.html")
```

#### CTE Achievement → Attended Postsecondary Within First Year of High School Graduation

**Key Findings**:\
Chi-square = 268.13, Cramer’s V = 0.052, p \< .001. Immediate enrollment (“Yes”) was highest among students with no CTE involvement (65.7%) and CTE participants (64.9%), while concentrators/completors had lower immediate enrollment (57.9%). Conversely, concentrators/completors were more likely to never attend postsecondary (37.1%) compared to participants (30.4%) or students with no CTE (29.6%).

**Conclusion**:\
CTE achievement shows a small but significant association with immediate postsecondary enrollment. Students deeply involved in CTE (concentrators/completors) are less likely to transition directly to postsecondary and more likely to never attend, while students with lighter or no CTE engagement transition at higher rates.

```{r}
tab_xtab(var.row = master.original$attended.ps.within.first.year.hsgrad,
         var.col = master.original$cte.achievement,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/attended-ps-within-first-year-hsgrad/High school accomplishments/cte-achievement.html")
```

#### Total CTE Courses Taken → Attended Postsecondary Within First Year of High School Graduation

**Key Findings**:\
ANOVA: F(2, 49813) = 301.90, p \< .001.

-   Students who never attended postsecondary had taken the most CTE courses on average.
-   Delayed enrollees took about 0.25 fewer CTE courses than non-attendees.
-   Immediate enrollees took the fewest, averaging about 0.56 fewer courses than non-attendees and 0.31 fewer than delayed enrollees.

**Conclusion**:\
Greater CTE course-taking is associated with a lower likelihood of immediate postsecondary enrollment. Students who transition directly into postsecondary tend to have completed fewer CTE courses, while heavier CTE course loads are linked to delaying or forgoing postsecondary education.

```{r}
summary <- aov(total.cte.courses.taken ~ attended.ps.within.first.year.hsgrad, data = master.original)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/attended-ps-within-first-year-hsgrad/High school accomplishments/total-cte-courses-taken.xlsx", overwrite = TRUE)
```

#### MCA Math Scores → Attended Postsecondary Within First Year of High School Graduation

**Key Findings**:\
ANOVA: F(2, 47621) = 885.58, p \< .001.

-   Students who never attended postsecondary had the lowest MCA Math scores.
-   Delayed enrollees scored slightly higher (0.12 points above non-attendees).
-   Immediate enrollees scored much higher—0.38 points above non-attendees and 0.50 points above delayed enrollees.

**Conclusion**:\
Math achievement is strongly associated with immediate postsecondary enrollment. Higher MCA Math scores are linked to a greater likelihood of enrolling right after high school, while lower scores are associated with delaying or never attending.

```{r}
summary <- aov(MCA.M ~ attended.ps.within.first.year.hsgrad, data = master.original)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/attended-ps-within-first-year-hsgrad/High school accomplishments/MCA-M.xlsx", overwrite = TRUE)
```

#### MCA Reading Scores → Attended Postsecondary Within First Year of High School Graduation

**Key Findings**:\
ANOVA: F(2, 47319) = 1475.87, p \< .001.

-   Students who never attended postsecondary had the lowest MCA Reading scores.
-   Delayed enrollees scored about 0.15 points higher than non-attendees.
-   Immediate enrollees scored substantially higher—0.46 points above non-attendees and 0.31 points above delayed enrollees.

**Conclusion**:\
Reading achievement is strongly associated with immediate postsecondary enrollment. Higher MCA Reading scores are linked to a greater likelihood of transitioning directly into postsecondary, while lower scores are tied to delaying or never attending.

```{r}
  summary <- aov(MCA.R ~ attended.ps.within.first.year.hsgrad, data = master.original)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/attended-ps-within-first-year-hsgrad/High school accomplishments/MCA-R.xlsx", overwrite = TRUE)

```

#### MCA Science Scores → Attended Postsecondary Within First Year of High School Graduation

**Key Findings**:\
ANOVA: F(2, 40378) = 1025.37, p \< .001.

-   Students who never attended postsecondary had the lowest MCA Science scores.
-   Delayed enrollees scored about the same as non-attendees (no significant difference).
-   Immediate enrollees scored much higher—0.41 points above non-attendees and 0.38 points above delayed enrollees.

**Conclusion**:\
Science achievement is strongly associated with immediate postsecondary enrollment. Higher MCA Science scores are linked to a greater likelihood of enrolling right after high school, while low scores are tied to non-enrollment or delay.

```{r}
summary <- aov(MCA.S ~ attended.ps.within.first.year.hsgrad, data = master.original)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/attended-ps-within-first-year-hsgrad/High school accomplishments/MCA-S.xlsx", overwrite = TRUE)
```
:::

<br>

## `Attended.ps`

### Summary

**Key Findings Across Variable Groups**

Demographics

-   **Economic Status**: φ = 0.228, p \< .001. Non-disadvantaged students were far more likely to attend (76%) than disadvantaged students (53.9%).
-   **Race/Ethnicity**: Cramer’s V = 0.164, p \< .001. White (72.4%) and Asian/PI (65.7%) students had the highest attendance, while American Indian students (51.8%) had the lowest, with Black (57.7%) and Hispanic (58.7%) also below average.
-   **Special Education Status**: φ = 0.273, p \< .001. Attendance was much higher for students without special education status (73.3%) compared to those with it (37.8%).

High School Characteristics

-   **School Locale (Dem_Desc)**: Cramer’s V = 0.029, p \< .001. Differences across rural, town/rural mix, and urban/town/rural mix were minimal (65–69%).
-   **Average Unemployment Rate**: F(1, 49814) = 428.40, p \< .001. Students from lower-unemployment areas were more likely to attend.
-   **Average Wages (% of State Average)**: F(1, 49814) = 65.84, p \< .001. Attendees came from areas with slightly higher average wages.

High School Accomplishments

-   **ACT Composite Score**: F(1, 28494) = 739.92, p \< .001. Attendees scored about 2 points higher than non-attendees.
-   **PSEO Participation**: φ = 0.279, p \< .001. Participation strongly boosted attendance (79.8% vs. 53.3%).
-   **CTE Achievement**: Cramer’s V = 0.071, p \< .001. Concentrators/completors had lower attendance (62.9%) than participants (69.6%) and students with no CTE (70.4%).
-   **Total CTE Courses Taken**: F(1, 49814) = 565.70, p \< .001. Non-attendees had taken more CTE courses.
-   **MCA Math**: F(1, 47622) = 1220.58, p \< .001. Attendees scored 0.35 points higher.
-   **MCA Reading**: F(1, 47320) = 2654.13, p \< .001. Attendees scored 0.44 points higher, the strongest MCA predictor.
-   **MCA Science**: F(1, 40379) = 1721.96, p \< .001. Attendees scored 0.39 points higher.

------------------------------------------------------------------------

**Overall Conclusion**:\
Postsecondary attendance is most strongly influenced by **demographic barriers** (economic disadvantage, special education status, and race/ethnicity), **program participation (PSEO)**, and **academic achievement (ACT and MCA scores, especially Reading)**. Community economic conditions (unemployment and wages) also play a role, though weaker, while **CTE involvement** shows an inverse relationship—students taking more CTE courses or reaching concentrator/completor status were less likely to attend.

Taken together, the results suggest that promoting equitable access to rigorous academics, reducing economic and special education barriers, and expanding opportunities like **PSEO** may be the most effective levers for improving overall postsecondary participation.

<br>

### Demographics

**Key Findings**:

-   **Economic Status**: φ = 0.228, p \< .001. Postsecondary attendance was higher for non-disadvantaged students (76%) than for economically disadvantaged students (53.9%). Disadvantaged students were almost twice as likely to never attend (46.1% vs. 24%).
-   **Race/Ethnicity**: Cramer’s V = 0.164, p \< .001. Attendance was highest for White (72.4%) and Asian/PI (65.7%) students. Attendance was lower among Hispanic (58.7%) and Black (57.7%) students, and lowest among American Indian students (51.8%). Students classified as “Unknown” also showed below-average rates (55.8%).
-   **Special Education Status**: φ = 0.273, p \< .001. Students without special education status were far more likely to attend postsecondary (73.3%) compared to students with special education status (37.8%). Special education students were more than twice as likely to never attend (62.2% vs. 26.7%).

**Conclusion**:\
Demographic factors are strongly associated with whether students ever attend postsecondary. Economic disadvantage and special education status present the largest barriers, while racial disparities are also significant, with American Indian, Black, and Hispanic students less likely to attend. These findings highlight that systemic inequities in access persist even when considering postsecondary attendance at any point after high school.

::: panel-tabset
#### Economic Status → Ever Attended Postsecondary (Attended.ps)

**Key Findings**:\
Chi-square = 2593.35, φ = 0.228, p \< .001. Postsecondary attendance was much higher among students not identified as economically disadvantaged (76%) compared to economically disadvantaged students (53.9%). Conversely, economically disadvantaged students were far more likely to never attend postsecondary (46.1% vs. 24%).

**Conclusion**:\
Economic status is strongly associated with whether a student ever attends postsecondary. Economically disadvantaged students are substantially less likely to attend, highlighting persistent inequities in long-term access to higher education.

```{r}
tab_xtab(var.row = master.original$attended.ps,
         var.col = master.original$economic.status,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/attended-ps/Demographics/economic-status.html")
```

#### Race/Ethnicity → Ever Attended Postsecondary (Attended.ps)

**Key Findings**:\
Chi-square = 1345.33, Cramer’s V = 0.164, p \< .001. Postsecondary attendance was highest among White students (72.4%) and Asian/PI students (65.7%). Attendance was lower among Hispanic (58.7%) and Black students (57.7%), and lowest among American Indian students (51.8%). Students classified as “Unknown” were also below average (55.8%).

**Conclusion**:\
Race and ethnicity are significantly associated with whether students ever attend postsecondary. White and Asian/PI students are more likely to attend, while American Indian, Black, and Hispanic students are less likely to do so, highlighting persistent racial disparities in postsecondary access.

```{r}
tab_xtab(var.row = master.original$attended.ps,
         var.col = master.original$RaceEthnicity,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/attended-ps/Demographics/RaceEthnicity.html")
```

#### Special Education Status → Ever Attended Postsecondary (Attended.ps)

**Key Findings**:\
Chi-square = 3707.63, φ = 0.273, p \< .001. Postsecondary attendance was much higher among students without special education status (73.3%) compared to students with special education status (37.8%). Conversely, special education students were much more likely to never attend postsecondary (62.2% vs. 26.7%).

**Conclusion**:\
Special education status is strongly associated with postsecondary attendance. Students with special education status are far less likely to ever attend, highlighting a major access gap that persists even when considering lifetime enrollment.

```{r}
tab_xtab(var.row = master.original$attended.ps,
         var.col = master.original$SpecialEdStatus,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/attended-ps/Demographics/SpecialEdStatus.html")
```
:::

<br>

## High School Characteristics

**Key Findings**:

-   **School Locale (Dem_Desc)**: Cramer’s V = 0.029, p \< .001. Postsecondary attendance rates were nearly identical across locales: entirely rural (67.2%), town/rural mix (65.5%), and urban/town/rural mix (68.7%). Differences are negligible.
-   **Average Unemployment Rate**: F(1, 49814) = 428.40, p \< .001. Students from lower-unemployment areas were significantly more likely to attend postsecondary, with attendees coming from communities averaging 0.29 percentage points lower unemployment.
-   **Average Wages (as % of State Average)**: F(1, 49814) = 65.84, p \< .001. Postsecondary attendees came from areas with wages about 0.40 percentage points higher than those who never attended.

**Conclusion**:\
High school contextual factors show mixed influence on postsecondary attendance. School locale itself has virtually no impact, but broader community economic conditions matter: students from lower-unemployment, higher-wage areas are more likely to attend postsecondary. This suggests that local labor market opportunities and economic vitality influence long-term educational access more than school setting.

::: panel-tabset
#### School Locale (Dem_Desc) → Ever Attended Postsecondary (Attended.ps)

**Key Findings**:\
Chi-square = 42.93, Cramer’s V = 0.029, p \< .001. Postsecondary attendance rates were very similar across locales: entirely rural (67.2%), town/rural mix (65.5%), and urban/town/rural mix (68.7%).

**Conclusion**:\
School locale shows a statistically significant but negligible association with whether students ever attend postsecondary. Differences across rural, mixed, and urban/town/rural contexts are minimal, suggesting that school setting itself does not meaningfully influence overall postsecondary access.

```{r}
tab_xtab(var.row = master.original$attended.ps,
         var.col = master.original$Dem_Desc,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/attended-ps/High school characteristics/Dem_Desc.html")
```

#### Average Unemployment Rate → Ever Attended Postsecondary (Attended.ps)

**Key Findings**:\
ANOVA: F(1, 49814) = 428.40, p \< .001. Students who attended postsecondary came from areas with unemployment rates about 0.29 percentage points lower on average than those who never attended.

**Conclusion**:\
Community unemployment levels are significantly associated with postsecondary attendance. Students from higher-unemployment areas are less likely to ever attend postsecondary, suggesting that local labor market conditions play a role in shaping long-term educational access.

```{r}
summary <- aov(avg.unemp.rate ~ attended.ps, data = master.original)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/attended-ps/High school characteristics/avg-unemp-rate.xlsx", overwrite = TRUE)


```

#### Average Wages (as % of Minnesota State Average) → Ever Attended Postsecondary (Attended.ps)

**Key Findings**:\
ANOVA: F(1, 49814) = 65.84, p \< .001. Students who attended postsecondary came from areas where average wages were about 0.40 percentage points higher than those who never attended.

**Conclusion**:\
Local wage levels show a modest but significant association with postsecondary attendance. Students from communities with higher relative wages are somewhat more likely to pursue postsecondary education, while those from lower-wage areas are less likely to attend.

```{r}
summary <- aov(avg.wages.pct.state ~ attended.ps, data = master.original)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/attended-ps/High school characteristics/avg-wage-pct-state.xlsx", overwrite = TRUE)
```
:::

<br>

## High School Accomplishments

**Key Findings**:

-   **ACT Composite Score**: F(1, 28494) = 739.92, p \< .001. Postsecondary attendees scored nearly 2 points higher on average than non-attendees.
-   **PSEO Participation**: φ = 0.279, p \< .001. Attendance was much higher among PSEO participants (79.8%) compared to non-participants (53.3%).
-   **CTE Achievement**: Cramer’s V = 0.071, p \< .001. Attendance was lowest among concentrators/completors (62.9%) compared to participants (69.6%) and students with no CTE (70.4%).
-   **Total CTE Courses Taken**: F(1, 49814) = 565.70, p \< .001. Non-attendees had taken more CTE courses, with attendees averaging 0.54 fewer courses.
-   **MCA Math**: F(1, 47622) = 1220.58, p \< .001. Attendees scored about 0.35 points higher than non-attendees.
-   **MCA Reading**: F(1, 47320) = 2654.13, p \< .001. Attendees scored about 0.44 points higher than non-attendees, the strongest MCA predictor.
-   **MCA Science**: F(1, 40379) = 1721.96, p \< .001. Attendees scored about 0.39 points higher than non-attendees.

**Conclusion**:\
High school accomplishments are strongly tied to postsecondary attendance. **PSEO participation** stands out as the most powerful programmatic predictor, while **academic achievement (ACT and MCA scores)** is consistently higher among attendees, especially reading scores. In contrast, heavier involvement in **CTE (more courses and concentrator/completor status)** is associated with lower attendance rates, suggesting a tradeoff between workforce-oriented preparation and pursuing postsecondary education.

::: panel-tabset
#### ACT Composite Score → Ever Attended Postsecondary (Attended.ps)

**Key Findings**:\
ANOVA: F(1, 28494) = 739.92, p \< .001. Students who attended postsecondary had ACT Composite Scores nearly 2 points higher on average than those who never attended.

**Conclusion**:\
ACT performance is a strong predictor of postsecondary attendance. Higher ACT scores are strongly associated with a greater likelihood of attending postsecondary at some point after high school.

```{r}
summary <- aov(ACTCompositeScore ~ attended.ps, data = master.original)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/attended-ps/High school accomplishments/ACTCompositeScore.xlsx", overwrite = TRUE)
```

#### PSEO Participation → Ever Attended Postsecondary (Attended.ps)

**Key Findings**:\
Chi-square = 3096.86, φ = 0.279, p \< .001. Postsecondary attendance was much higher among PSEO participants (79.8%) compared to non-participants (53.3%). Non-participants were far more likely to never attend postsecondary (46.7% vs. 20.2%).

**Conclusion**:\
PSEO participation is strongly associated with postsecondary attendance. Students who participated in PSEO were far more likely to attend postsecondary at some point, highlighting the program’s effectiveness in expanding long-term access to higher education.

```{r}
tab_xtab(var.row = master.original$attended.ps,
         var.col = master.original$pseo.participant,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/attended-ps/High school accomplishments/pseo-participant.html")
```

#### CTE Achievement → Ever Attended Postsecondary (Attended.ps)

**Key Findings**:\
Chi-square = 254.60, Cramer’s V = 0.071, p \< .001. Postsecondary attendance was lowest among CTE concentrators/completors (62.9%) and higher among CTE participants (69.6%) and students with no CTE involvement (70.4%). Non-attendance was most common among concentrators/completors (37.1%) compared to participants (30.4%) and non-CTE students (29.6%).

**Conclusion**:\
CTE achievement shows a small but significant association with postsecondary attendance. Students deeply involved in CTE (concentrators/completors) are less likely to attend postsecondary compared to peers who were only participants or had no CTE courses.

```{r}
tab_xtab(var.row = master.original$attended.ps,
         var.col = master.original$cte.achievement,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/attended-ps/High school accomplishments/cte-achievement.html")
```

#### Total CTE Courses Taken → Ever Attended Postsecondary (Attended.ps)

**Key Findings**:\
ANOVA: F(1, 49814) = 565.70, p \< .001. Students who attended postsecondary had taken fewer CTE courses on average, with a mean about 0.54 courses lower than those who never attended.

**Conclusion**:\
Greater CTE course-taking is negatively associated with postsecondary attendance. Students who took more CTE courses were less likely to pursue postsecondary education, suggesting that heavier CTE engagement may steer students toward workforce pathways rather than higher education.

```{r}
summary <- aov(total.cte.courses.taken ~ attended.ps, data = master.original)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/attended-ps/High school accomplishments/total-cte-courses-taken.xlsx", overwrite = TRUE)
```

#### MCA Math Scores → Ever Attended Postsecondary (Attended.ps)

**Key Findings**:\
ANOVA: F(1, 47622) = 1220.58, p \< .001. Students who attended postsecondary had MCA Math scores about 0.35 points higher on average than those who never attended.

**Conclusion**:\
Math achievement is strongly associated with postsecondary attendance. Higher MCA Math scores are linked to a greater likelihood of pursuing postsecondary education.

```{r}
summary <- aov(MCA.M ~ attended.ps, data = master.original)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/attended-ps/High school accomplishments/MCA-M.xlsx", overwrite = TRUE)
```

#### MCA Reading Scores → Ever Attended Postsecondary (Attended.ps)

**Key Findings**:\
ANOVA: F(1, 47320) = 2654.13, p \< .001. Students who attended postsecondary had MCA Reading scores about 0.44 points higher on average than those who never attended.

**Conclusion**:\
Reading achievement is very strongly associated with postsecondary attendance. Students with higher MCA Reading scores are substantially more likely to attend postsecondary at some point.

```{r}
summary <- aov(MCA.R ~ attended.ps, data = master.original)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/attended-ps/High school accomplishments/MCA-R.xlsx", overwrite = TRUE)

```

#### MCA Science Scores → Ever Attended Postsecondary (Attended.ps)

**Key Findings**:\
ANOVA: F(1, 40379) = 1721.96, p \< .001. Students who attended postsecondary had MCA Science scores about 0.39 points higher on average than those who never attended.

**Conclusion**:\
Science achievement is strongly associated with postsecondary attendance. Students with higher MCA Science scores are much more likely to attend postsecondary at some point compared to students with lower scores.

```{r}
summary <- aov(MCA.S ~ attended.ps, data = master.original)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/attended-ps/High school accomplishments/MCA-S.xlsx", overwrite = TRUE)
```
:::

<br>

## `ps.grad`

### Summary

**Key Findings Across Variable Groups**

Demographics

-   **Economic Status**: φ = 0.270, p \< .001. Graduation was much higher among non-disadvantaged students (53.1%) compared to economically disadvantaged students (25.7%). Disadvantaged students were nearly three times as likely not to graduate.
-   **Race/Ethnicity**: Cramer’s V = 0.169, p \< .001 (Unknown removed). White (49.2%) and Asian/PI (44.7%) students had the highest graduation rates, while Hispanic (31.3%), Black (21.6%), and American Indian (20.8%) students had much lower rates.
-   **Special Education Status**: φ = 0.262, p \< .001. Students without special education status graduated at far higher rates (48.2%) than those with it (13.2%).

High School Characteristics

-   **School Locale (Dem_Desc)**: Cramer’s V = 0.022, p \< .001. Graduation rates were very similar across rural (39.2%), town/rural mix (41.6%), and urban/town/rural mix (43.2%).
-   **Average Unemployment Rate**: F(1, 44033) = 572.00, p \< .001. Graduates came from areas with unemployment rates about 0.33 percentage points lower than non-graduates.
-   **Average Wages (% of State Average)**: F(1, 44033) = 43.37, p \< .001. Graduates came from areas with wages about 0.33 percentage points higher than non-graduates.

High School Accomplishments

-   **ACT Composite Score**: F(1, 24476) = 1619.08, p \< .001. Graduates scored about 2.41 points higher than non-graduates.
-   **PSEO Participation**: φ = 0.308, p \< .001. Graduation was much more common among PSEO participants (57%) than non-participants (26.7%).
-   **CTE Achievement**: Cramer’s V = 0.103, p \< .001. Concentrators/completors had the lowest graduation rates (35.5%), while participants (42.7%) and non-CTE students (47.5%) graduated at higher rates.
-   **Total CTE Courses Taken**: F(1, 44033) = 693.05, p \< .001. Non-graduates had taken about 0.60 more CTE courses than graduates.
-   **MCA Math**: F(1, 41955) = 1748.50, p \< .001. Graduates scored 0.41 points higher than non-graduates.
-   **MCA Reading**: F(1, 41687) = 3087.39, p \< .001. Graduates scored 0.47 points higher, the strongest MCA predictor.
-   **MCA Science**: F(1, 34774) = 2188.12, p \< .001. Graduates scored 0.44 points higher.

------------------------------------------------------------------------

**Overall Conclusion**:\
Postsecondary graduation is shaped most strongly by **student-level factors**. Economic disadvantage and special education status are powerful barriers, while race and ethnicity reveal deep inequities in attainment. Among school-related experiences, **PSEO participation** stands out as the most influential positive predictor of graduation. Academic preparation (ACT and MCA scores, especially Reading and Science) also plays a major role. In contrast, **greater CTE involvement**—both in number of courses and concentrator/completor status—is linked to lower graduation rates, suggesting that heavy CTE pathways align more with workforce entry than degree attainment. Community economic conditions (lower unemployment and higher wages) show modest but consistent associations.

Taken together, these findings highlight that **supporting equity, strengthening academic readiness, and expanding dual-enrollment opportunities like PSEO** are critical levers for increasing postsecondary completion rates.

<br>

### Demographics

**Key Findings**:

-   **Economic Status**: φ = 0.270, p \< .001. Graduation was much more common among non-disadvantaged students (53.1%) compared to economically disadvantaged students (25.7%). Disadvantaged students were almost three times as likely not to graduate (74.3% vs. 46.9%).
-   **Race/Ethnicity**: Cramer’s V = 0.169, p \< .001 (Unknown removed). White (49.2%) and Asian/PI (44.7%) students had the highest graduation rates, while Hispanic (31.3%), Black (21.6%), and American Indian (20.8%) students graduated at much lower rates.
-   **Special Education Status**: φ = 0.262, p \< .001. Graduation rates were substantially higher for students without special education status (48.2%) compared to those with it (13.2%). Special education students were far more likely not to graduate (86.8% vs. 51.8%).

**Conclusion**:\
Demographic characteristics strongly influence postsecondary completion. Economic disadvantage and special education status are linked to particularly low graduation rates, while race and ethnicity reveal pronounced disparities, with White and Asian/PI students most likely to graduate and American Indian, Black, and Hispanic students least likely. These patterns indicate that structural barriers extend beyond access and deeply affect attainment.

::: panel-tabset
#### Economic Status → Postsecondary Graduation (Ps.grad)

**Key Findings**:\
Chi-square = 3204.41, φ = 0.270, p \< .001. Postsecondary graduation was much more common among students not economically disadvantaged (53.1%) compared to disadvantaged students (25.7%). Conversely, economically disadvantaged students were far more likely not to graduate (74.3% vs. 46.9%).

**Conclusion**:\
Economic status is strongly associated with postsecondary completion. Economically disadvantaged students are much less likely to graduate once they enroll, highlighting major inequities not only in access but also in degree attainment.

```{r}
data <- master.original %>%
  filter(ps.grad != "Attending ps")

tab_xtab(var.row = data$ps.grad,
         var.col = data$economic.status,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/ps-grad/Demographics/economic-status.html")
```

#### Race/Ethnicity → Postsecondary Graduation (Ps.grad) (Unknown Removed)

**Key Findings**:\
Chi-square = 1130.51, Cramer’s V = 0.169, p \< .001. Graduation rates varied by group:\
- White: 49.2% graduated, 50.8% did not.\
- Asian/PI: 44.7% graduated, 55.3% did not.\
- Hispanic: 31.3% graduated, 68.7% did not.\
- Black: 21.6% graduated, 78.4% did not.\
- American Indian: 20.8% graduated, 79.2% did not.

**Conclusion**:\
Race and ethnicity are significantly associated with postsecondary completion. White and Asian/PI students are the most likely to graduate, while Black, Hispanic, and American Indian students have substantially lower completion rates. This highlights persistent racial disparities in achieving postsecondary credentials.

```{r}
data <- master.original %>%
  filter(ps.grad != "Attending ps",
         RaceEthnicity != "Unknown")

tab_xtab(var.row = data$ps.grad,
         var.col = data$RaceEthnicity,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/ps-grad/Demographics/RaceEthnicity.html")
```

#### Special Education Status → Postsecondary Graduation (Ps.grad)

**Key Findings**:\
Chi-square = 3018.74, φ = 0.262, p \< .001. Postsecondary graduation was much higher among students without special education status (48.2%) compared to students with special education status (13.2%). Special education students were far more likely not to graduate (86.8% vs. 51.8%).

**Conclusion**:\
Special education status is strongly associated with postsecondary completion. Students with special education status are far less likely to graduate once they enroll, highlighting a significant attainment gap.

```{r}
data <- master.original %>%
  filter(ps.grad != "Attending ps")

tab_xtab(var.row = data$ps.grad,
         var.col = data$SpecialEdStatus,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/ps-grad/Demographics/SpecialEdStatus.html")
```
:::

<br>

### High School Characteristics

**Key Findings**:

-   **School Locale (Dem_Desc)**: Cramer’s V = 0.022, p \< .001. Graduation rates were similar across locales: entirely rural (39.2%), town/rural mix (41.6%), and urban/town/rural mix (43.2%). Differences are negligible.
-   **Average Unemployment Rate**: F(1, 44033) = 572.00, p \< .001. Graduates came from areas with unemployment rates about 0.33 percentage points lower than non-graduates.
-   **Average Wages (as % of State Average)**: F(1, 44033) = 43.37, p \< .001. Graduates came from areas with average wages about 0.33 percentage points higher than non-graduates.

**Conclusion**:\
High school contextual characteristics show limited but consistent influence on postsecondary graduation. Locale itself has virtually no impact, while broader community economic conditions—lower unemployment and higher wages—are modestly associated with greater likelihood of completing a credential. This suggests that local economic vitality provides some support for persistence through postsecondary education, though it is not a dominant factor compared to student-level characteristics.

::: panel-tabset
#### School Locale (Dem_Desc) → Postsecondary Graduation (Ps.grad)

**Key Findings**:\
Chi-square = 20.89, Cramer’s V = 0.022, p \< .001. Graduation rates were similar across locales: entirely rural (39.2%), town/rural mix (41.6%), and urban/town/rural mix (43.2%).

**Conclusion**:\
School locale has a statistically significant but negligible association with postsecondary graduation. Differences between rural, mixed, and urban/town/rural schools are very small, suggesting that locale has little impact on ultimate completion rates.

```{r}
tab_xtab(var.row = data$ps.grad,
         var.col = data$Dem_Desc,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/ps-grad/High school characteristics/Dem_Desc.html")
```

#### Average Unemployment Rate → Postsecondary Graduation (Ps.grad)

**Key Findings**:\
ANOVA: F(1, 44033) = 572.00, p \< .001. Students who graduated from postsecondary came from areas with unemployment rates about 0.33 percentage points lower than those who did not graduate.

**Conclusion**:\
Community unemployment levels are significantly associated with postsecondary completion. Students from lower-unemployment areas are more likely to graduate, suggesting that stronger local labor market conditions may support persistence and credential attainment.

```{r}
summary <- aov(avg.unemp.rate ~ ps.grad, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/ps-grad/High school characteristics/avg-unemp-rate.xlsx", overwrite = TRUE)

```

#### Average Wages (as % of Minnesota State Average) → Postsecondary Graduation (Ps.grad)

**Key Findings**:\
ANOVA: F(1, 44033) = 43.37, p \< .001. Students who graduated from postsecondary came from areas with average wages about 0.33 percentage points higher than those who did not graduate.

**Conclusion**:\
Local wage levels are modestly associated with postsecondary graduation. Students from somewhat higher-wage communities are more likely to complete a credential, suggesting that stronger local economic conditions may support persistence through postsecondary programs.

```{r}
summary <- aov(avg.wages.pct.state ~ ps.grad, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/ps-grad/High school characteristics/avg-wage-pct-state.xlsx", overwrite = TRUE)
```
:::

<br>

### High School Accomplishments

**Key Findings**:

-   **ACT Composite Score**: F(1, 24476) = 1619.08, p \< .001. Graduates scored about 2.41 points higher on average than non-graduates.
-   **PSEO Participation**: φ = 0.308, p \< .001. Graduation was much more common among PSEO participants (57%) than non-participants (26.7%).
-   **CTE Achievement**: Cramer’s V = 0.103, p \< .001. Graduation was lowest among CTE concentrators/completors (35.5%), compared to participants (42.7%) and students with no CTE involvement (47.5%).
-   **Total CTE Courses Taken**: F(1, 44033) = 693.05, p \< .001. Non-graduates had taken about 0.60 more CTE courses on average than graduates.
-   **MCA Math**: F(1, 41955) = 1748.50, p \< .001. Graduates scored 0.41 points higher than non-graduates.
-   **MCA Reading**: F(1, 41687) = 3087.39, p \< .001. Graduates scored 0.47 points higher than non-graduates, the strongest MCA predictor.
-   **MCA Science**: F(1, 34774) = 2188.12, p \< .001. Graduates scored 0.44 points higher than non-graduates.

**Conclusion**:\
High school accomplishments are strongly tied to postsecondary completion. **PSEO participation** is the single strongest programmatic predictor, while **academic achievement (ACT and MCA scores, particularly Reading and Science)** consistently predicts higher graduation rates. In contrast, **greater CTE engagement (more courses and concentrator/completor status)** is linked to lower graduation rates, suggesting a tradeoff between workforce preparation and degree attainment. Overall, academic preparation and dual-enrollment opportunities play critical roles in supporting successful postsecondary completion.

::: panel-tabset
#### ACT Composite Score → Postsecondary Graduation (Ps.grad)

**Key Findings**:\
ANOVA: F(1, 24476) = 1619.08, p \< .001. Students who graduated from postsecondary had ACT Composite Scores about 2.41 points higher on average than those who did not graduate.

**Conclusion**:\
ACT performance is strongly associated with postsecondary completion. Higher ACT scores substantially increase the likelihood of graduating, underscoring the role of academic preparation in degree attainment.

```{r}
summary <- aov(ACTCompositeScore ~ ps.grad, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/ps-grad/High school accomplishments/ACTCompositeScore.xlsx", overwrite = TRUE)
```

#### PSEO Participation → Postsecondary Graduation (Ps.grad)

**Key Findings**:\
Chi-square = 3210.71, φ = 0.308, p \< .001. Postsecondary graduation was much more common among PSEO participants (57%) compared to non-participants (26.7%). Non-participants were far more likely not to graduate (73.3% vs. 43%).

**Conclusion**:\
PSEO participation is strongly associated with postsecondary completion. Students who participated in PSEO were more than twice as likely to graduate, underscoring the program’s powerful role in promoting credential attainment.

```{r}
tab_xtab(var.row = data$ps.grad,
         var.col = data$pseo.participant,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/ps-grad/High school accomplishments/pseo-participant.html")
```

#### CTE Achievement → Postsecondary Graduation (Ps.grad)

**Key Findings**:\
Chi-square = 465.86, Cramer’s V = 0.103, p \< .001. Graduation rates were lowest among CTE concentrators/completors (35.5%), higher among CTE participants (42.7%), and highest among students with no CTE involvement (47.5%). Non-graduation was most common among concentrators/completors (64.5%) compared to participants (57.3%) and non-CTE students (52.5%).

**Conclusion**:\
CTE achievement is significantly associated with postsecondary graduation. Students who were CTE concentrators or completors were less likely to graduate, while those with limited or no CTE participation had higher completion rates. This suggests that deeper CTE engagement may orient students more toward workforce pathways than degree attainment.

```{r}
tab_xtab(var.row = data$ps.grad,
         var.col = data$cte.achievement,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/ps-grad/High school accomplishments/cte-achievement.html")
```

#### Total CTE Courses Taken → Postsecondary Graduation (Ps.grad)

**Key Findings**:\
ANOVA: F(1, 44033) = 693.05, p \< .001. Students who graduated from postsecondary had taken about 0.60 fewer CTE courses on average compared to those who did not graduate.

**Conclusion**:\
Greater CTE course-taking is negatively associated with postsecondary completion. Students with heavier CTE course loads were less likely to graduate, reinforcing the pattern that strong engagement in CTE is linked to lower rates of credential attainment.

```{r}
summary <- aov(total.cte.courses.taken ~ ps.grad, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/ps-grad/High school accomplishments/total-cte-courses-taken.xlsx", overwrite = TRUE)
```

#### MCA Math Scores → Postsecondary Graduation (Ps.grad)

**Key Findings**:\
ANOVA: F(1, 41955) = 1748.50, p \< .001. Students who graduated from postsecondary had MCA Math scores about 0.41 points higher on average than those who did not graduate.

**Conclusion**:\
Math achievement is strongly associated with postsecondary completion. Higher MCA Math scores are linked to a greater likelihood of graduating, suggesting that math proficiency supports persistence and success through postsecondary programs.

```{r}
summary <- aov(MCA.M ~ ps.grad, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/ps-grad/High school accomplishments/MCA-M.xlsx", overwrite = TRUE)
```

#### MCA Reading Scores → Postsecondary Graduation (Ps.grad)

**Key Findings**:\
ANOVA: F(1, 41687) = 3087.39, p \< .001. Students who graduated from postsecondary had MCA Reading scores about 0.47 points higher on average than those who did not graduate.

**Conclusion**:\
Reading achievement is very strongly associated with postsecondary completion. Higher MCA Reading scores are linked to substantially greater likelihood of graduating, making reading proficiency one of the strongest academic predictors of credential attainment.

```{r}
summary <- aov(MCA.R ~ ps.grad, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/ps-grad/High school accomplishments/MCA-R.xlsx", overwrite = TRUE)

```

#### MCA Science Scores → Postsecondary Graduation (Ps.grad)

**Key Findings**:\
ANOVA: F(1, 34774) = 2188.12, p \< .001. Students who graduated from postsecondary had MCA Science scores about 0.44 points higher on average than those who did not graduate.

**Conclusion**:\
Science achievement is strongly associated with postsecondary completion. Higher MCA Science scores are linked to a significantly greater likelihood of graduating, reinforcing the importance of academic preparation in science for persistence and attainment.

```{r}
summary <- aov(MCA.S ~ ps.grad, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/ps-grad/High school accomplishments/MCA-S.xlsx", overwrite = TRUE)
```
:::

<br>

# [**Tier 2: Type of Postsecondary Experience**]{.underline}

<br>

## `highest.cred.level`

### **Key Findings Across Variable Groups**

Demographics

-   **Economic Status**: Cramer’s V = 0.264, p \< .001. Non-disadvantaged students were far more likely to earn a bachelor’s or higher (46.2% vs. 20.4%), while disadvantaged students disproportionately stopped at “Some college” (49.2% vs. 26.6%).
-   **Race/Ethnicity**: Cramer’s V = 0.088, p \< .001. White (40.9%) and Asian/PI (52.5%) students had the highest bachelor’s+ attainment. Black (20.1%), Hispanic (26.7%), and American Indian (19.5%) students were far more likely to leave with “Some college” or sub-associate credentials.
-   **Special Education Status**: Cramer’s V = 0.233, p \< .001. Students with special education status were overwhelmingly concentrated in “Some college” (61.6%), with very few earning a bachelor’s+ (6.9%) compared to non-special education students (41.4%).

High School Characteristics

-   **School Locale (Dem_Desc)**: Cramer’s V = 0.028, p \< .001. Credential attainment was nearly identical across rural, mixed, and urban/town/rural schools.
-   **Average Unemployment Rate**: F(3, 28012) = 9.11, p \< .001. Students from higher-unemployment areas were slightly more likely to stop at sub-associate levels, while those from lower-unemployment areas were somewhat more likely to reach bachelor’s+.
-   **Average Wages (% of State Average)**: F(3, 28012) = 6.70, p \< .001. Students from higher-wage areas were modestly more likely to earn bachelor’s+, while those from lower-wage areas were slightly more likely to stop short.

High School Accomplishments

-   **ACT Composite Score**: F(3, 19493) = 1165.82, p \< .001. Higher ACT scores strongly predicted bachelor’s+ attainment. “Some college” students scored similarly to associate earners, indicating persistence rather than readiness drove non-completion.
-   **PSEO Participation**: Cramer’s V = 0.308, p \< .001. PSEO participants were more than twice as likely to earn a bachelor’s+ (50.5% vs. 22.2%) and much less likely to stop at “Some college.”
-   **CTE Achievement**: Cramer’s V = 0.140, p \< .001. CTE concentrators/completors were more likely to stop at associate degrees (28.2%) or “Some college” (38.9%) and less likely to reach bachelor’s+ (25.6%). Non-CTE students had the highest bachelor’s+ attainment (47%).
-   **Total CTE Courses Taken**: F(3, 28012) = 471.36, p \< .001. More CTE courses were associated with lower attainment; bachelor’s+ earners took the fewest CTE courses, while sub-associate and “Some college” groups took the most.
-   **MCA Math, Reading, Science**: All strongly associated with credential level (F \> 850, p \< .001). Bachelor’s+ earners consistently scored 0.3–0.6 points higher across subjects, with Reading and Science showing the strongest predictive value.

------------------------------------------------------------------------

### **Overall Conclusion**

The level of postsecondary credential earned is shaped most powerfully by **demographic factors** and **academic accomplishments**. Economic disadvantage, special education status, and race/ethnicity all create steep barriers, concentrating students in sub-associate outcomes or “Some college” without completion. In contrast, strong academic preparation (ACT, MCA scores) and **PSEO participation** sharply increase the likelihood of earning a bachelor’s or higher degree. Intensive CTE engagement, while supporting workforce readiness, is consistently associated with lower attainment levels, especially associate degrees or non-completion.

**Community context (unemployment and wages)** exerts a modest effect, but school locale has virtually no impact. Overall, Tier 2 results highlight that **rigorous academics, equitable supports, and access to dual-enrollment opportunities** are critical levers for ensuring students not only graduate but also achieve higher levels of credential attainment.

<br>

### Demographics

**Key Findings**:

-   **Economic Status**: Cramer’s V = 0.264, p \< .001. Non-disadvantaged students were far more likely to complete a bachelor’s or higher degree (46.2% vs. 20.4%) while disadvantaged students disproportionately left with “Some college” and no degree (49.2% vs. 26.6%).
-   **Race/Ethnicity**: Cramer’s V = 0.088, p \< .001. White (40.9%) and Asian/PI (52.5%) students were most likely to earn bachelor’s or higher degrees, while American Indian (19.5%), Black (20.1%), and Hispanic (26.7%) students disproportionately stopped short, often leaving with no degree.
-   **Special Education Status**: Cramer’s V = 0.233, p \< .001. Students with special education status were overwhelmingly concentrated in “Some college” (61.6%), with only 6.9% earning a bachelor’s or higher, compared to 41.4% of non-special education students.

**Conclusion**:\
Demographic factors play a major role in shaping the highest credential attained. **Economic disadvantage** and **special education status** are the strongest barriers, sharply reducing the likelihood of earning a bachelor’s or higher degree. **Racial disparities** also persist, with White and Asian/PI students attaining higher levels of education, while American Indian, Black, and Hispanic students are much more likely to leave postsecondary without completing a credential. These patterns highlight that inequities extend well beyond access and graduation, influencing the ultimate level of educational attainment students are able to achieve.

::: panel-tabset
#### Economic Status → Highest Credential Level (Highest.cred.level)

**Key Findings**:\
Chi-square = 1959.32, Cramer’s V = 0.264, p \< .001.

-   **Non-disadvantaged students** were much more likely to earn a bachelor’s degree or higher (46.2%) compared to disadvantaged students (20.4%).
-   **Disadvantaged students** were far more likely to leave with “Some college” and no degree (49.2% vs. 26.6%).
-   Associate’s and sub-associate credentials showed smaller gaps, with disadvantaged students slightly more likely to stop at these levels.

**Conclusion**:\
Economic status is strongly associated with the highest credential earned. Economically disadvantaged students are far less likely to reach a bachelor’s degree or higher and far more likely to leave postsecondary without completing a credential. This points to economic status as a key driver not only of access but also of persistence and attainment level.

```{r}
data <- master.original %>%
  filter(!highest.cred.level %in% c("Attending ps", "No college"))

tab_xtab(var.row = data$highest.cred.level,
         var.col = data$economic.status,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/highest-cred-level/Demographics/economic-status.html")
```

#### Race/Ethnicity → Highest Credential Level (Highest.cred.level)

**Key Findings**:\
Chi-square = 627.58, Cramer’s V = 0.088, p \< .001. Patterns of attainment varied:

-   **White students** were the most likely to earn a bachelor’s or higher (40.9%) and less likely to stop at “Some college” (31.2%).
-   **Asian/PI students** also had high bachelor’s/graduate attainment (52.5%), with relatively fewer leaving with no degree (30.2%).
-   **Black students** were the least likely to reach a bachelor’s or higher (20.1%) and most likely to stop at “Some college” (62.1%).
-   **Hispanic students** had lower bachelor’s/graduate attainment (26.7%) and higher “Some college” outcomes (45.4%).
-   **American Indian students** showed a similar pattern, with only 19.5% completing bachelor’s or higher and more than half (55.2%) leaving with no degree.

**Conclusion**:\
Race and ethnicity are significantly associated with the highest credential earned, though the effect size is modest. White and Asian/PI students are far more likely to achieve bachelor’s or higher degrees, while American Indian, Black, and Hispanic students are disproportionately likely to leave postsecondary without completing a degree. This reflects persistent racial disparities not only in access and graduation, but also in the ultimate level of attainment.

```{r}
data <- master.original %>%
  filter(!highest.cred.level %in% c("Attending ps", "No college"),
         RaceEthnicity != "Unknown")

tab_xtab(var.row = data$highest.cred.level,
         var.col = data$RaceEthnicity,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/highest-cred-level/Demographics/RaceEthnicity.html")
```

#### Special Education Status → Highest Credential Level (Highest.cred.level)

**Key Findings**:\
Chi-square = 1522.15, Cramer’s V = 0.233, p \< .001.

-   **Students without special education status**: 41.4% earned a bachelor’s degree or higher, 30.7% left with “Some college,” 16.2% stopped at an associate’s, and 11.7% earned less than an associate credential.
-   **Students with special education status**: Only 6.9% earned a bachelor’s or higher. The vast majority either left with “Some college” and no degree (61.6%) or stopped at an associate’s degree (10.8%).

**Conclusion**:\
Special education status is strongly associated with the highest credential attained. Students with special education status are overwhelmingly concentrated in the “Some college” category, with very limited progression to bachelor’s or higher degrees. This reflects persistent barriers to full attainment among students with special education needs.

```{r}
data <- master.original %>%
  filter(!highest.cred.level %in% c("Attending ps", "No college"))

tab_xtab(var.row = data$highest.cred.level,
         var.col = data$SpecialEdStatus,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/highest-cred-level/Demographics/SpecialEdStatus.html")
```
:::

<br>

## High School Characteristics

**Key Findings**:

-   **School Locale (Dem_Desc)**: Cramer’s V = 0.028, p \< .001. Credential attainment was nearly identical across rural, mixed, and urban/town/rural schools, with differences of only a few percentage points.
-   **Average Unemployment Rate**: F(3, 28012) = 9.11, p \< .001. Students from higher-unemployment areas were modestly more likely to stop at sub-associate credentials, while those from lower-unemployment areas were somewhat more likely to reach higher credentials.
-   **Average Wages (as % of State Average)**: F(3, 28012) = 6.70, p \< .001. Students from higher-wage areas were slightly more likely to earn bachelor’s or higher degrees, while students from lower-wage areas were somewhat more likely to stop short of an associate degree.

**Conclusion**:\
High school contextual characteristics have only limited influence on the highest credential earned. **School locale shows virtually no effect**, while **local economic conditions**—lower unemployment and higher wages—are modestly associated with higher attainment levels. These community effects are real but small, and far less impactful than demographic or academic factors.

::: panel-tabset
#### School Locale (Dem_Desc) → Highest Credential Level (Highest.cred.level)

**Key Findings**:\
Chi-square = 45.47, Cramer’s V = 0.028, p \< .001. Credential attainment patterns were very similar across locales:

-   **Entirely rural**: 38.5% stopped at “Some college,” 32.9% earned a bachelor’s or higher.
-   **Town/rural mix**: 32.8% “Some college,” 37.7% bachelor’s or higher.
-   **Urban/town/rural mix**: 33.2% “Some college,” 39.1% bachelor’s or higher.

Differences across locales were minimal, generally within 5 percentage points.

**Conclusion**:\
School locale has a statistically significant but negligible association with the highest credential earned. Students from rural, mixed, and urban/town/rural schools attain similar levels of postsecondary credentials, suggesting locale plays little role in shaping ultimate attainment.

```{r}
tab_xtab(var.row = data$highest.cred.level,
         var.col = data$Dem_Desc,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/highest-cred-level/High school characteristics/Dem_Desc.html")
```

#### Average Unemployment Rate → Highest Credential Level (Highest.cred.level)

**Key Findings**:\
ANOVA: F(3, 28012) = 9.11, p \< .001.

-   Students whose highest attainment was **less than an associate degree** came from areas with slightly higher unemployment rates compared to both associate and bachelor’s+ earners (difference ≈ 0.11 percentage points, p \< .01).
-   No significant differences emerged between bachelor’s and associate degree earners, or between bachelor’s and “Some college.”

**Conclusion**:\
Community unemployment levels show a small but significant association with the highest credential attained. Students from higher-unemployment areas are modestly more likely to stop at sub-associate credentials, while those from lower-unemployment areas are somewhat more likely to advance to higher levels. Overall, the effect is statistically real but substantively limited.

```{r}
summary <- aov(avg.unemp.rate ~ highest.cred.level, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/highest-cred-level/High school characteristics/avg.unemp.rate.xlsx", overwrite = TRUE)

```

#### Average Wages (as % of Minnesota State Average) → Highest Credential Level (Highest.cred.level)

**Key Findings**:\
ANOVA: F(3, 28012) = 6.70, p \< .001.

-   Students who attained a **bachelor’s degree or higher** came from areas with slightly higher relative wages than associate earners (difference ≈ 0.42 percentage points, p \< .001).
-   Students with **less than an associate degree** also came from somewhat lower-wage areas than associate earners (difference ≈ 0.38 percentage points, p \< .05).
-   No significant differences emerged between “Some college” and either associate or bachelor’s groups.

**Conclusion**:\
Local wage levels show a modest association with the highest credential earned. Students from higher-wage areas are somewhat more likely to complete bachelor’s or higher degrees, while those from lower-wage areas are slightly more likely to stop short of an associate degree. These effects are real but small in magnitude compared to student-level demographic and academic factors.

```{r}
summary <- aov(avg.wages.pct.state ~ highest.cred.level, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/highest-cred-level/High school characteristics/avg.wage.pct.state.xlsx", overwrite = TRUE)
```
:::

<br>

## High School Accomplishments

**Key Findings**:

-   **ACT Composite Score**: F(3, 19493) = 1165.82, p \< .001. Students earning a bachelor’s or higher scored 3–4 points above those stopping at associate or sub-associate levels. “Some college” students had ACT scores similar to associate earners, suggesting non-completion is tied more to persistence than readiness.
-   **PSEO Participation**: Cramer’s V = 0.308, p \< .001. PSEO participants were more than twice as likely to earn a bachelor’s or higher (50.5% vs. 22.2%) and far less likely to leave with “Some college” (23.9% vs. 44.9%). Associate and sub-associate outcomes were nearly identical between groups.
-   **CTE Achievement**: Cramer’s V = 0.140, p \< .001. Concentrators/completors were less likely to earn bachelor’s+ (25.6%) and more likely to stop at associate (28.2%) or “Some college” (38.9%). Non-CTE students had the highest bachelor’s+ attainment (47%).
-   **Total CTE Courses Taken**: F(3, 28012) = 471.36, p \< .001. Students who earned bachelor’s+ took fewer CTE courses, while those with sub-associate or “Some college” outcomes had taken more. Largest gap: bachelor’s+ vs. sub-associate, with \>1.3 course difference.
-   **MCA Math**: F(3, 27568) = 925.88, p \< .001. Bachelor’s+ earners scored 0.45–0.66 points higher than other groups. Lower scores characterized sub-associate and “Some college” outcomes.
-   **MCA Reading**: F(3, 27410) = 1053.96, p \< .001. Bachelor’s+ earners scored 0.32–0.59 points higher than other groups, making reading one of the strongest predictors of higher attainment.
-   **MCA Science**: F(3, 22192) = 854.01, p \< .001. Bachelor’s+ earners scored 0.41–0.60 points higher than other groups, while “Some college” and sub-associate students scored notably lower.

**Conclusion**:\
High school accomplishments are among the most powerful predictors of the highest credential earned. **PSEO participation** stands out as the strongest programmatic factor, dramatically increasing the likelihood of bachelor’s or higher attainment. **Academic performance (ACT, MCA Reading, and MCA Science especially)** is tightly linked to ultimate attainment, differentiating bachelor’s+ earners from those who stop earlier. In contrast, **greater CTE engagement** (more courses and concentrator/completor status) is associated with lower levels of attainment, with students more likely to finish at the associate level or leave with “Some college.” Overall, rigorous academics and early postsecondary experiences strongly support higher credential attainment, while intensive CTE pathways align more with sub-baccalaureate outcomes.

::: panel-tabset
#### ACT Composite Score → Highest Credential Level (Highest.cred.level)

**Key Findings**:\
ANOVA: F(3, 19493) = 1165.82, p \< .001.

-   Students who earned a **bachelor’s degree or higher** had ACT scores about 3.39 points higher than associate degree earners, and about 4.22 points higher than those who stopped short of an associate degree.
-   Students with **less than an associate degree** had significantly lower ACT scores than associate earners (–0.83).
-   Students in the **“Some college”** group scored similarly to associate earners (no significant difference).

**Conclusion**:\
ACT performance is strongly associated with the highest credential attained. Higher ACT scores are linked to completing a bachelor’s or higher degree, while lower scores increase the likelihood of stopping short of an associate degree. Students who attended college without finishing a credential had ACT scores similar to associate earners, suggesting that non-completion is driven more by persistence factors than academic readiness alone.

```{r}
summary <- aov(ACTCompositeScore ~ highest.cred.level, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/highest-cred-level/High school accomplishments/ACTCompositeScore.xlsx", overwrite = TRUE)
```

#### PSEO Participation → Highest Credential Level (Highest.cred.level)

**Key Findings**:\
Chi-square = 1933.77, Cramer’s V = 0.308, p \< .001.

-   **PSEO participants** were far more likely to complete a bachelor’s degree or higher (50.5%) compared to non-participants (22.2%).
-   Participants were also less likely to leave with “Some college” and no degree (23.9% vs. 44.9%).
-   Rates of associate degree and sub-associate credential attainment were nearly identical between participants and non-participants (\~16%).

**Conclusion**:\
PSEO participation is strongly associated with higher credential attainment. Students who participated in PSEO were more than twice as likely to earn a bachelor’s or higher degree and far less likely to leave postsecondary without a credential. This highlights PSEO as one of the most effective levers for boosting ultimate educational attainment.

```{r}
tab_xtab(var.row = data$highest.cred.level,
         var.col = data$pseo.participant,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/highest-cred-level/High school accomplishments/pseo-participant.html")
```

#### CTE Achievement → Highest Credential Level (Highest.cred.level)

**Key Findings**:\
Chi-square = 1097.17, Cramer’s V = 0.140, p \< .001. Credential attainment patterns differed by CTE involvement:

-   **CTE concentrators/completors** were the least likely to earn a bachelor’s or higher (25.6%) and the most likely to leave with “Some college” (38.9%). They were also more likely than other groups to stop at the associate degree level (28.2%).
-   **CTE participants** showed more balanced outcomes, with 36.9% earning a bachelor’s or higher, 34% stopping at “Some college,” and 20.3% earning an associate’s degree.
-   **Non-CTE students** were the most likely to reach a bachelor’s or higher (47%) and the least likely to stop at associate’s (15.7%) or “Some college” (29.8%).

**Conclusion**:\
CTE achievement is significantly associated with the highest credential attained. Students with deeper CTE engagement (concentrators/completors) are less likely to earn a bachelor’s or higher degree and more likely to stop at an associate degree or leave with “Some college.” In contrast, students with no CTE experience have the highest rates of bachelor’s and graduate-level attainment. This highlights how intensive CTE pathways tend to align with sub-baccalaureate outcomes and workforce preparation rather than advanced academic credentials.

```{r}
tab_xtab(var.row = data$highest.cred.level,
         var.col = data$cte.achievement,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/highest-cred-level/High school accomplishments/cte-achievement.html")
```

#### Total CTE Courses Taken → Highest Credential Level (Highest.cred.level)

**Key Findings**:\
ANOVA: F(3, 28012) = 471.36, p \< .001.

-   Students who earned a **bachelor’s degree or higher** had taken fewer CTE courses on average than associate degree earners (–0.72 courses, p \< .001).
-   Students with **less than an associate degree** had taken significantly more CTE courses than associate earners (+0.63, p \< .001).
-   Students who left with **“Some college”** had also taken slightly more CTE courses than associate earners (+0.13, p \< .01).
-   The largest gap was between **bachelor’s+ earners and those with less than an associate degree**, with the latter taking over 1.35 more CTE courses on average.

**Conclusion**:\
Greater CTE course-taking is negatively associated with ultimate attainment. Students who take more CTE courses are less likely to earn a bachelor’s or higher degree and more likely to stop short—either with “Some college” or with credentials below the associate level. In contrast, students who take fewer CTE courses are disproportionately represented among bachelor’s and graduate degree earners, reinforcing the strong alignment of intensive CTE pathways with sub-baccalaureate outcomes.

```{r}
summary <- aov(total.cte.courses.taken ~ highest.cred.level, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/highest-cred-level/High school accomplishments/total-cte-courses-taken.xlsx", overwrite = TRUE)
```

#### MCA Math Scores → Highest Credential Level (Highest.cred.level)

**Key Findings**:\
ANOVA: F(3, 27568) = 925.88, p \< .001.

-   Students with a **bachelor’s degree or higher** scored about 0.45 points higher than associate degree earners, and about 0.66 points higher than those with less than an associate degree.
-   Students in the **“Some college”** group scored 0.20 points lower than associate earners, and about 0.66 points lower than bachelor’s+ earners.
-   Students with **less than an associate degree** scored 0.21 points lower than associate earners.

**Conclusion**:\
Math achievement is strongly associated with the highest credential earned. Students with higher MCA Math scores are far more likely to complete bachelor’s or higher degrees, while those with lower scores are more likely to stop at associate’s, sub-associate credentials, or leave with “Some college.”

```{r}
summary <- aov(MCA.M ~ highest.cred.level, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/highest-cred-level/High school accomplishments/MCA-M.xlsx", overwrite = TRUE)
```

#### MCA Reading Scores → Highest Credential Level (Highest.cred.level)

**Key Findings**:\
ANOVA: F(3, 27410) = 1053.96, p \< .001.

-   Students with a **bachelor’s degree or higher** scored about 0.32 points higher than associate degree earners, 0.59 points higher than those with less than an associate degree, and 0.54 points higher than those who left with “Some college.”
-   Students with **less than an associate degree** and those with **“Some college”** scored significantly lower than both associate earners and bachelor’s+ earners.

**Conclusion**:\
Reading achievement is very strongly associated with the highest credential earned. Higher MCA Reading scores sharply increase the likelihood of attaining a bachelor’s or higher degree, while lower scores are linked to stopping at sub-associate levels or leaving with no degree. This makes reading proficiency one of the clearest academic differentiators of ultimate educational attainment.

```{r}
summary <- aov(MCA.R ~ highest.cred.level, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/highest-cred-level/High school accomplishments/MCA-R.xlsx", overwrite = TRUE)

```

#### MCA Science Scores → Highest Credential Level (Highest.cred.level)

**Key Findings**:\
ANOVA: F(3, 22192) = 854.01, p \< .001.

-   Students with a **bachelor’s degree or higher** scored about 0.41 points higher than associate earners, 0.57 points higher than sub-associate earners, and 0.60 points higher than those with “Some college.”
-   Students with **less than an associate degree** scored about 0.16 points lower than associate earners.
-   Students with **“Some college”** scored about 0.18 points lower than associate earners.

**Conclusion**:\
Science achievement is strongly associated with the highest credential earned. Students with higher MCA Science scores are far more likely to attain bachelor’s or higher degrees, while those with lower scores are concentrated among associate earners, sub-associate credentials, or incomplete “Some college” outcomes.

```{r}
summary <- aov(MCA.S ~ highest.cred.level, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/highest-cred-level/High school accomplishments/MCA-S.xlsx", overwrite = TRUE)
```
:::

<br>

# [**Tier 3: Location of Postsecondary Experience**]{.underline}

## ps.grad.location

Tier 3 examined how demographic factors, high school characteristics, and high school accomplishments shape where students complete their postsecondary credentials: **within their high school region, elsewhere in Minnesota, or outside the state.**

------------------------------------------------------------------------

### **Key Findings**

Demographics

-   **Economic Status**: Economically disadvantaged students were more likely to graduate within their home region (67.8% vs. 55.7%), while non-disadvantaged students showed greater mobility, more often graduating elsewhere in MN or out of state.
-   **Race/Ethnicity**: American Indian (67.8%) and Hispanic (63.6%) graduates were most likely to stay in-region, while Asian/PI graduates were the least likely (48.7%), with higher rates graduating elsewhere in MN or out of state. White graduates showed a balanced pattern.
-   **Special Education Status**: Students with special education status were overwhelmingly more likely to remain in-region (73.8%) than non-special education students (57.7%), who were more mobile across the state and beyond.

High School Characteristics

-   **School Locale**: Rural graduates were less likely to remain in-region (38%) and more likely to graduate elsewhere in MN, while mixed/urban graduates were more likely to stay local (56–61%).
-   **Economic Conditions**: Students from lower-unemployment, higher-wage areas were more likely to graduate in their home region or outside the state, while those from higher-unemployment, lower-wage areas were more likely to complete elsewhere in Minnesota. Effects were modest but consistent.

High School Accomplishments

-   **Academic Achievement (ACT & MCA scores)**: Students with higher ACT and MCA scores were significantly more likely to graduate elsewhere in Minnesota or out of state. Lower-scoring students were more likely to remain in-region.
-   **PSEO Participation**: Participants were more geographically mobile, with higher rates of completing elsewhere in Minnesota or outside the state, while non-participants were more likely to stay local.
-   **CTE Engagement**: Students with deeper CTE involvement (more courses, concentrator/completor status) were far more likely to graduate in-region, while students with little or no CTE experience showed greater mobility.

------------------------------------------------------------------------

### **Overall Conclusion**

Tier 3 results show that **geographic mobility in postsecondary completion is closely tied to academic strength, program participation, and demographics.**

-   Students who are **academically stronger or PSEO participants** are more likely to graduate away from their home region, either elsewhere in Minnesota or out of state.
-   Students who are **economically disadvantaged, American Indian, Hispanic, or in special education**, as well as those with deeper **CTE engagement**, are more likely to remain local.
-   **Community economic conditions** and **school locale** play smaller but notable roles, with rural and lower-wage contexts reinforcing regional completion.

In sum, Tier 3 highlights a divide between **mobile, academically prepared students** who pursue credentials beyond their home region, and **students facing structural or programmatic constraints** who are more likely to complete locally. This underscores the importance of academic preparation, resource access, and program opportunities in shaping not just whether students succeed in postsecondary, but **where** they do so.

<br>

### Demographics

**Key Findings**:

-   **Economic Status**: Cramer’s V = 0.104, p \< .001. Economically disadvantaged students were more likely to graduate within their high school region (67.8% vs. 55.7%), while non-disadvantaged students were more likely to graduate elsewhere in Minnesota (25.8% vs. 19.9%) or outside the state (18.5% vs. 12.3%).
-   **Race/Ethnicity**: Cramer’s V = 0.051, p \< .001. American Indian (67.8%) and Hispanic (63.6%) graduates were most likely to remain in-region, while Asian/PI graduates were the least likely (48.7%), with higher rates completing elsewhere in MN (30.5%) or outside MN (20.8%). White graduates showed balanced outcomes across regions.
-   **Special Education Status**: Cramer’s V = 0.072, p \< .001. Students with special education status were far more likely to remain in their high school region (73.8% vs. 57.7%), while non-special education students were more likely to complete elsewhere in MN (24.9% vs. 14.8%) or outside the state (17.4% vs. 11.4%).

**Conclusion**:\
Demographic characteristics modestly influence the location of postsecondary graduation. **Economically disadvantaged, American Indian, Hispanic, and special education students are more likely to stay within their home region**, reflecting limited geographic mobility. In contrast, **non-disadvantaged, White, and Asian/PI students show greater mobility**, more often completing their credentials elsewhere in Minnesota or out of state. These differences suggest that resources, community ties, and support needs shape students’ ability or decision to pursue postsecondary opportunities beyond their home region.

::: panel-tabset
#### Economic Status → Postsecondary Graduation Location (Ps.grad.location)

**Key Findings**:\
Chi-square = 202.79, Cramer’s V = 0.104, p \< .001.

-   **Economically disadvantaged students** were more likely to graduate within their high school region (67.8%) compared to non-disadvantaged peers (55.7%).
-   **Non-disadvantaged students** were more likely to graduate elsewhere in Minnesota (25.8% vs. 19.9%) or outside the state (18.5% vs. 12.3%).

**Conclusion**:\
Economic status is modestly associated with where students complete postsecondary credentials. Economically disadvantaged students tend to stay closer to home, graduating within their high school region, while non-disadvantaged students are more geographically mobile, more often completing credentials outside their region, both within and outside Minnesota.

```{r}
data <- master.original %>%
  filter(!ps.grad.location %in% c("Attending ps", "Did not graduate ps"))

tab_xtab(var.row = data$ps.grad.location,
         var.col = data$economic.status,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/ps-grad-location/Demographics/economic-status.html")
```

#### Race/Ethnicity → Postsecondary Graduation Location (Ps.grad.location)

**Key Findings**:\
Chi-square = 98.74, Cramer’s V = 0.051, p \< .001. Location of graduation showed modest differences by race/ethnicity:

-   **American Indian (67.8%)** and **Hispanic (63.6%)** graduates were most likely to remain in their high school region.
-   **Black graduates (57.0%)** also tended to remain in-region, but with a somewhat higher share completing outside MN (20.1%).
-   **White graduates (57.9%)** were more balanced, with 24.8% completing elsewhere in Minnesota and 17.3% outside the state.
-   **Asian/PI graduates (48.7%)** were the least likely to remain in-region, with higher rates of completing elsewhere in Minnesota (30.5%) or outside MN (20.8%).

**Conclusion**:\
Race and ethnicity are significantly associated with where students graduate from postsecondary institutions, though the effect size is small. American Indian and Hispanic graduates are more likely to remain in-region, while Asian/PI and White graduates show greater mobility, often completing outside their region or out of state. These patterns suggest varying levels of geographic mobility across racial groups, which may reflect differences in access, resources, or cultural and community ties.

```{r}

tab_xtab(var.row = data$ps.grad.location,
         var.col = data$RaceEthnicity,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/ps-grad-location/Demographics/RaceEthnicity.html")
```

#### Special Education Status → Postsecondary Graduation Location (Ps.grad.location)

**Key Findings**:\
Chi-square = 95.81, Cramer’s V = 0.072, p \< .001.

-   **Students with special education status** were much more likely to graduate within their high school region (73.8%) compared to non-special education students (57.7%).
-   **Non-special education students** were more likely to graduate elsewhere in Minnesota (24.9% vs. 14.8%) or outside the state (17.4% vs. 11.4%).

**Conclusion**:\
Special education status is modestly associated with where students complete postsecondary credentials. Students with special education needs are more likely to stay local and graduate within their high school region, while non-special education students demonstrate greater geographic mobility, completing degrees elsewhere in Minnesota or out of state.

```{r}
tab_xtab(var.row = data$ps.grad.location,
         var.col = data$SpecialEdStatus,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/ps-grad-location/Demographics/SpecialEdStatus.html")
```
:::

<br>

## High School Characteristics

**Key Findings**:

-   **School Locale (Dem_Desc)**: Cramer’s V = 0.113, p \< .001. Graduates from entirely rural schools were less likely to remain in-region (38%) and more likely to graduate elsewhere in Minnesota (49.3%). Graduates from town/rural mix (55.9%) and urban/town/rural mix schools (61%) were more likely to stay in-region.
-   **Average Unemployment Rate**: F(2, 18642) = 19.17, p \< .001. Graduates from lower-unemployment areas were slightly more likely to graduate in their home region or outside Minnesota, while graduates from higher-unemployment areas were more likely to graduate elsewhere in Minnesota.
-   **Average Wages (as % of State Average)**: F(2, 18642) = 107.57, p \< .001. Students from higher-wage areas were more likely to graduate either in their home region or outside the state, while those from lower-wage areas were more likely to graduate elsewhere in Minnesota.

**Conclusion**:\
High school contextual characteristics exert a modest influence on where students complete postsecondary credentials. **Locale matters somewhat**, with rural students more likely to leave their region for other Minnesota institutions, while mixed/urban students are more likely to remain local. **Economic conditions also play a role**: students from stronger labor markets (lower unemployment, higher wages) are more likely to complete either locally or out of state, whereas weaker labor markets are tied to completion elsewhere in Minnesota. These effects are consistent but modest compared to demographic and accomplishment-based factors.

::: panel-tabset
#### School Locale (Dem_Desc) → Postsecondary Graduation Location (Ps.grad.location)

**Key Findings**:\
Chi-square = 472.32, Cramer’s V = 0.113, p \< .001. Graduation location varied modestly by school locale:

-   **Entirely rural graduates** were more likely to complete outside their region (49.3% elsewhere in MN, 12.6% outside MN). Only 38% stayed within their high school region.
-   **Town/rural mix graduates** were much more likely to stay in-region (55.9%) and less likely to go outside the state (15.4%).
-   **Urban/town/rural mix graduates** were the most likely to remain in-region (61%) and the most likely to graduate outside MN (17.9%), with relatively fewer moving elsewhere in MN (21%).

**Conclusion**:\
School locale shows a modest association with where students graduate from postsecondary institutions. Graduates from entirely rural schools are less likely to remain in-region and more likely to move elsewhere in Minnesota, while graduates from mixed or urban/town/rural schools are more likely to complete within their home region. These findings suggest that regional school context modestly shapes geographic mobility after high school graduation.

```{r}
tab_xtab(var.row = data$ps.grad.location,
         var.col = data$Dem_Desc,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/ps-grad-location/High school characteristics/Dem_Desc.html")
```

#### Average Unemployment Rate → Postsecondary Graduation Location (Ps.grad.location)

**Key Findings**:\
ANOVA: F(2, 18642) = 19.17, p \< .001.

-   Graduates from **outside Minnesota** came from areas with unemployment rates about 0.19 percentage points lower than those who graduated elsewhere in Minnesota.
-   Graduates from their **home region** also came from areas with unemployment rates about 0.11 percentage points lower than elsewhere-in-MN graduates.
-   Differences between **Region** and **Outside MN** graduates were smaller (0.08 percentage points) but still significant.

**Conclusion**:\
Community unemployment rates are modestly associated with where students complete their postsecondary credentials. Graduates from lower-unemployment areas are somewhat more likely to finish either within their home region or out of state, while those from higher-unemployment areas are more likely to graduate elsewhere in Minnesota. The effects are statistically reliable but small in magnitude.

```{r}
summary <- aov(avg.unemp.rate ~ ps.grad.location, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/ps-grad-location/High school characteristics/avg-unemp.rate.xlsx", overwrite = TRUE)


```

#### Average Wages (as % of Minnesota State Average) → Postsecondary Graduation Location (Ps.grad.location)

**Key Findings**:\
ANOVA: F(2, 18642) = 107.57, p \< .001.

-   Graduates from **outside Minnesota** came from areas with average wages about 0.012 points higher than those who graduated elsewhere in Minnesota (p \< .001).
-   Graduates from their **home region** came from areas with wages about 0.014 points higher than elsewhere-in-MN graduates (p \< .001).
-   No significant difference emerged between **Region** and **Outside MN** graduates.

**Conclusion**:\
Local wage levels are modestly associated with where students complete postsecondary credentials. Students from higher-wage areas are more likely to graduate either within their home region or outside the state, while students from lower-wage areas are somewhat more likely to graduate elsewhere in Minnesota. The effect, while statistically significant, is relatively small in practical terms.

```{r}
summary <- aov(avg.wages.pct.state ~ ps.grad.location, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/ps-grad-location/High school characteristics/avg-wage-pct-state.xlsx", overwrite = TRUE)
```
:::

<br>

## High School Accomplishments

**Key Findings**:

-   **ACT Composite Score**: F(2, 14470) = 536.91, p \< .001. Students who graduated elsewhere in Minnesota had the highest ACT scores, while those who graduated in-region or outside Minnesota had lower scores.
-   **PSEO Participation**: Cramer’s V = 0.117, p \< .001. PSEO participants were more geographically mobile—less likely to graduate in-region (54% vs. 66%) and more likely to graduate elsewhere in Minnesota or outside the state.
-   **CTE Achievement**: Cramer’s V = 0.068, p \< .001. CTE concentrators/completors were the most likely to remain in their high school region (65.8%) and the least likely to graduate outside MN (13.2%). Non-CTE students were the least likely to stay local and the most likely to graduate out of state.
-   **Total CTE Courses Taken**: F(2, 18642) = 63.26, p \< .001. Students who stayed in-region had taken the most CTE courses, while those who graduated elsewhere in MN or out of state had taken fewer.
-   **MCA Math**: F(2, 18435) = 204.54, p \< .001. Region graduates had significantly lower math scores than those who graduated elsewhere in MN or out of state.
-   **MCA Reading**: F(2, 18346) = 156.66, p \< .001. Region graduates had significantly lower reading scores than elsewhere-in-MN and out-of-state graduates.
-   **MCA Science**: F(2, 14864) = 236.46, p \< .001. Region graduates had significantly lower science scores than both elsewhere-in-MN and out-of-state graduates.

**Conclusion**:\
High school accomplishments are strongly associated with postsecondary graduation location. **Higher academic achievement (ACT and MCA scores)** is linked to greater geographic mobility, with stronger students more likely to graduate elsewhere in Minnesota or out of state. **PSEO participation** also increases mobility, while **CTE engagement (more courses and concentrator/completor status)** anchors students closer to home, with higher rates of in-region graduation. Overall, academic strength and dual-enrollment experiences encourage mobility, while CTE pathways reinforce regional ties.

::: panel-tabset
#### ACT Composite Score → Postsecondary Graduation Location (Ps.grad.location)

**Key Findings**:\
ANOVA: F(2, 14470) = 536.91, p \< .001.

-   Students who graduated in their **home region** had ACT scores about 2.39 points lower than those who graduated elsewhere in Minnesota.
-   Students who graduated **outside Minnesota** also had ACT scores about 2.30 points lower than elsewhere-in-MN graduates.
-   No significant difference was observed between Region and Outside-MN graduates.

**Conclusion**:\
ACT scores are strongly associated with where students complete postsecondary credentials. Students with higher ACT scores are more likely to graduate from institutions elsewhere in Minnesota, while those with lower scores tend to remain in their home region or attend out-of-state institutions. The lack of difference between Region and Outside-MN suggests that both represent pathways for students with relatively lower ACT performance compared to those who stay within the state but outside their home region.

```{r}
summary <- aov(ACTCompositeScore ~ ps.grad.location, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/ps-grad-location/High school accomplishments/ACTCompositeScore.xlsx", overwrite = TRUE)
```

#### PSEO Participation → Postsecondary Graduation Location (Ps.grad.location)

**Key Findings**:\
Chi-square = 187.50, Cramer’s V = 0.117, p \< .001:contentReference[oaicite:0]{index="0"}.

-   **Non-PSEO participants** were much more likely to graduate within their high school region (66%) compared to PSEO participants (54%).
-   **PSEO participants** were more likely to graduate elsewhere in Minnesota (27% vs. 20.6%) or outside the state (19% vs. 13.4%).

**Conclusion**:\
PSEO participation is significantly associated with where students graduate. Students who participated in PSEO show greater geographic mobility, more often completing their degrees outside their home region—in other parts of Minnesota or out of state. Non-PSEO students, by contrast, are more likely to remain in their high school region for postsecondary completion.

```{r}
tab_xtab(var.row = data$ps.grad.location,
         var.col = data$pseo.participant,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/ps-grad-location/High school accomplishments/pseo-participant.html")
```

#### CTE Achievement → Postsecondary Graduation Location (Ps.grad.location)

**Key Findings**:\
Chi-square = 171.97, Cramer’s V = 0.068, p \< .001:contentReference[oaicite:0]{index="0"}.

-   **CTE concentrators/completors** were the most likely to remain in their high school region (65.8%) and the least likely to graduate outside MN (13.2%).
-   **CTE participants** were somewhat less likely to stay in-region (58.9%) and more likely to graduate outside MN (16.9%).
-   **Non-CTE students** were the least likely to remain in-region (54.4%) and the most likely to graduate outside MN (19.2%).

**Conclusion**:\
CTE achievement shows a modest but significant association with graduation location. Students with deeper CTE engagement are more likely to stay local and complete within their high school region, while students with little or no CTE involvement demonstrate greater geographic mobility, graduating elsewhere in Minnesota or out of state. This suggests that CTE pathways reinforce regional ties, while non-CTE students pursue more mobile postsecondary trajectories.

```{r}
tab_xtab(var.row = data$ps.grad.location,
         var.col = data$cte.achievement,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/ChatGPT/Post-secondary pathways ct/ps-grad-location/High school accomplishments/cte-achievement.html")
```

#### Total CTE Courses Taken → Postsecondary Graduation Location (Ps.grad.location)

**Key Findings**:\
ANOVA: F(2, 18642) = 63.26, p \< .001.

-   Graduates who remained in their **home region** had taken significantly more CTE courses than both elsewhere-in-MN (+0.24 courses) and outside-MN graduates (+0.45 courses).
-   Graduates who completed **outside Minnesota** had taken fewer CTE courses than those in-region (–0.45 courses).
-   The smallest number of CTE courses was observed among students graduating **elsewhere in Minnesota**, who took about 0.21 fewer courses than out-of-state graduates.

**Conclusion**:\
The number of CTE courses taken is significantly related to where students graduate. Students with heavier CTE course loads are more likely to stay in their home region, while those with fewer CTE courses are more likely to graduate outside their region, either elsewhere in Minnesota or out of state. This reinforces the pattern that stronger CTE engagement is tied to local completion, while less CTE involvement aligns with greater geographic mobility.

```{r}
summary <- aov(total.cte.courses.taken ~ ps.grad.location, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/ps-grad-location/High school accomplishments/total-cte-courses-taken.xlsx", overwrite = TRUE)

```

#### MCA Math Scores → Postsecondary Graduation Location (Ps.grad.location)

**Key Findings**:\
ANOVA: F(2, 18435) = 204.54, p \< .001.

-   **Region graduates** had significantly lower MCA Math scores than both elsewhere-in-MN graduates (–0.30, p \< .001) and outside-MN graduates (–0.26, p \< .001).
-   No significant difference in Math scores was found between elsewhere-in-MN and outside-MN graduates.

**Conclusion**:\
Math achievement is strongly associated with graduation location. Students with higher MCA Math scores are more likely to graduate outside their home region, whether elsewhere in Minnesota or out of state. Students with lower math scores are more likely to remain in their home region for postsecondary completion, indicating that higher academic achievement is linked to greater geographic mobility.

```{r}
summary <- aov(MCA.M ~ ps.grad.location, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/ps-grad-location/High school accomplishments/MCA-M.xlsx", overwrite = TRUE)
```

#### MCA Reading Scores → Postsecondary Graduation Location (Ps.grad.location)

**Key Findings**:\
ANOVA: F(2, 18346) = 156.66, p \< .001.

-   **Region graduates** had significantly lower MCA Reading scores than both elsewhere-in-MN graduates (–0.19, p \< .001) and outside-MN graduates (–0.18, p \< .001).
-   No significant difference was found between elsewhere-in-MN and outside-MN graduates.

**Conclusion**:\
Reading achievement is strongly associated with graduation location. Students with higher MCA Reading scores are more likely to graduate outside their home region, whether elsewhere in Minnesota or out of state. Students with lower reading scores are more likely to remain in their home region for postsecondary completion, showing a clear link between stronger academic skills and greater geographic mobility.

```{r}
summary <- aov(MCA.R ~ ps.grad.location, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/ps-grad-location/High school accomplishments/MCA-R.xlsx", overwrite = TRUE)

```

#### MCA Science Scores → Postsecondary Graduation Location (Ps.grad.location)

**Key Findings**:\
ANOVA: F(2, 14864) = 236.46, p \< .001.

-   **Region graduates** had significantly lower MCA Science scores than both elsewhere-in-MN (–0.29, p \< .001) and outside-MN graduates (–0.28, p \< .001).
-   No significant difference was found between elsewhere-in-MN and outside-MN graduates.

**Conclusion**:\
Science achievement is strongly associated with graduation location. Students with higher MCA Science scores are more likely to graduate outside their home region, whether elsewhere in Minnesota or out of state. Students with lower science scores are more likely to remain in their home region, reinforcing the pattern that higher academic achievement is linked with greater geographic mobility.

```{r}
summary <- aov(MCA.S ~ ps.grad.location, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/ChatGPT/Post-secondary pathways ct/ps-grad-location/High school accomplishments/MCA-S.xlsx", overwrite = TRUE)
```
:::

<br>

# Summary of Postsecondary Pathways Analysis (All Tiers)

This study examined how **demographics, high school characteristics, and high school accomplishments** shape postsecondary pathways across three tiers of outcomes:

1.  **Tier 1: Core Postsecondary Access and Attainment**
    -   Immediate enrollment (attended.ps.within.first.year.hsgrad)\
    -   Ever attended postsecondary (attended.ps)\
    -   Postsecondary graduation (ps.grad)
2.  **Tier 2: Type of Postsecondary Experience**
    -   Highest credential earned (highest.cred.level)
3.  **Tier 3: Location of Postsecondary Experience**
    -   Location of postsecondary graduation (ps.grad.location)

------------------------------------------------------------------------

# Key Findings by Tier

## Tier 1: Core Access and Attainment

-   **Demographics** were the strongest and most consistent predictors. Economic disadvantage, race/ethnicity (especially American Indian, Black, and Hispanic students), and special education status all reduced likelihood of enrollment, attendance, and completion.
-   **High School Characteristics** (locale, local unemployment, wages) had weak and inconsistent effects.
-   **High School Accomplishments** were powerful predictors. Higher ACT/MCA scores and **PSEO participation** strongly boosted immediate enrollment, persistence, and graduation. In contrast, **greater CTE engagement** was associated with lower enrollment and graduation, aligning with workforce entry rather than postsecondary continuation.

## Tier 2: Type of Postsecondary Experience (Highest Credential Earned)

-   **Economic disadvantage** and **special education status** created the steepest barriers to earning a bachelor’s or higher degree, with these groups concentrated in “Some college” or sub-associate credentials.
-   **Racial disparities** persisted: White and Asian/PI students were far more likely to complete bachelor’s or graduate degrees, while American Indian, Black, and Hispanic students were more likely to stop at associate’s or leave with no credential.
-   **Academic readiness and PSEO participation** were the strongest positive predictors of bachelor’s+ attainment. Students with higher ACT/MCA scores and PSEO experience were much more likely to progress to advanced credentials.
-   **CTE engagement** again tracked with sub-baccalaureate attainment, with concentrators/completors more likely to finish with associate’s or stop at “Some college.”

## Tier 3: Location of Postsecondary Graduation

-   **Demographics shaped mobility**: disadvantaged, American Indian, Hispanic, and special education students were most likely to graduate within their high school region, while non-disadvantaged, White, and Asian/PI students showed greater mobility (elsewhere in MN or out-of-state).
-   **High school characteristics** mattered modestly. Rural students tended to leave their region but stay within Minnesota, while students from stronger labor markets (lower unemployment, higher wages) were more likely to graduate in-region or out-of-state.
-   **Accomplishments mattered most**: Higher ACT/MCA scores and PSEO participation were strongly tied to graduating outside the home region, while heavier CTE engagement (more courses, concentrator/completor status) was associated with staying local.

------------------------------------------------------------------------

# Overall Interpretation

Across all tiers, the analysis reveals three central themes:

1.  **Structural Inequities Persist**\
    Economic disadvantage, race/ethnicity, and special education status consistently constrain postsecondary success. These students are less likely to enroll immediately, less likely to persist, less likely to graduate, and when they do, they are more likely to complete at lower credential levels and closer to home.

2.  **Academic Readiness and PSEO are Game-Changers**\
    ACT and MCA scores—especially in reading and science—are among the strongest predictors of bachelor’s and graduate attainment, while **PSEO participation stands out as the single most effective programmatic lever**, doubling the likelihood of bachelor’s+ completion and encouraging geographic mobility.

3.  **CTE Pathways Support Local, Sub-Baccalaureate Outcomes**\
    Intensive CTE engagement aligns strongly with associate-level attainment and local graduation. While CTE provides clear workforce pathways, it appears less effective in propelling students toward bachelor’s or graduate degrees, reflecting its different role in student trajectories.

------------------------------------------------------------------------

# Conclusion

The analysis underscores that **postsecondary pathways are not just about whether students attend, but how, where, and how far they progress.**

-   **Equity gaps** remain entrenched, especially for disadvantaged, racially marginalized, and special education students.
-   **Academic preparation and dual-enrollment opportunities** (like PSEO) dramatically expand access to higher-level and geographically broader attainment.
-   **CTE pathways** anchor students locally and at sub-baccalaureate levels, serving as important workforce pipelines but not drivers of higher educational attainment.

Ultimately, policies and practices that strengthen academic readiness, expand access to PSEO, and address demographic inequities will be most effective in raising not only participation and completion rates but also the **level and reach of postsecondary success**.
