---
title: "Multivariate Confirmation Analysis - Logistic Regression"
editor: visual
---

```{r include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
library(htmlwidgets)
library(RColorBrewer)
library(readxl)
library(janitor)
library(lubridate)
library(systemfonts)
reset_font_cache()
library(ggtext)
library(gmodels)
library(fastDummies)
library(car)
library(glmnet)
library(glmnetUtils)
library(pscl)
library(sjPlot)
library(ggforce)
library(broom)
```

```{r themes and shapefiles, include=FALSE}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", linewidth  = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.key.size = unit(1, "lines"),
        legend.margin = margin(0,0,0,0),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.text = element_text(margin = margin(l = 2)),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", linewidth = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))


theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
    select(5,6) %>%
    unique() %>%
    mutate(edr = str_replace(edr, "  ", " "),
           planning.region = str_replace(planning.region, " Minnesota", ""),
           planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
           edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

counties.regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
  rename(mif = `MIF Region`) %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Name = str_replace(Name, "Mcleod", "McLeod"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc) ,
         edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"))

counties.regions.1 <- counties.regions %>%
  mutate(statefp = "27",
         project.pr = ifelse(edr %in% c("EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 4 - West Central"), "Northwest",
                             ifelse(edr == "EDR 3 - Arrowhead", "Northeast",
                                    ifelse(edr %in% c("EDR 5 - North Central", "EDR 7E- East Central", "EDR 7W- Central"), "Central",
                                           ifelse(edr %in% c("EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 8 - Southwest"), "Southwest",
                                                  ifelse(edr %in% c("EDR 9 - South Central", "EDR 10 - Southeast"), "Southern", as.character(planning.region)))))),
         project.pr = fct_relevel(project.pr, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southern"))


color.ruca <- c("Entirely rural" = "#009933", "Town/rural mix" = "#99CC33", "Urban/town/rural mix" = "#CC9966", "Entirely urban" = "#754C29", "Minnesota" = "black")

color.pr <- c("Northwest" = 	"#4575b4", "Northeast" = "grey", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.edr.simple <- c("EDR 1" = "#b3cde3", "EDR 2" = "#8c96c6", "EDR 3" = "#fe9929", "EDR 4" = "#8856a7", "EDR 5" = "#810f7c", "EDR 6E" = "#e5f5f9", "EDR 6W" = "#bdc9e1", "EDR 7E" = "#99d8c9", "EDR 7W" = "#2ca25f", "EDR 8" = "#74a9cf", "EDR 9" = "#0570b0", "EDR 10" = "#d7301f", "EDR 11" = "#d8b365", "Minnesota" = "black")

color.pr.edr <- c ("Northwest" = "#4575b4","Northeast" = "#e0f3f8", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black", "EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365")

mn_counties <- st_read("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Shapefiles/County shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE) %>%
  rename(countyfp = FIPS_CODE)

```

```{r master dataset}
master.original <- read_csv("Data/SLEDS/Masters/After analysis/Master-final-after-mca.csv")

names(master.original)
master.states <- read_csv("Data/SLEDS/Masters/States/annual-seven-states.csv") %>%
  filter(grad.year %in% c("grad.year.1", "grad.year.5", "grad.year.10")) %>%
  select(PersonID, grad.year, states) %>%
  pivot_wider(names_from = grad.year, values_from = states) 

master <- master.original %>%
  left_join(master.states, by = "PersonID") %>%
  mutate(grad.year.5 = ifelse(grad.year.5 %in% c("Meaningful emp County", "Meaningful emp EDR"), "Sustained WF - NE",
                              ifelse(grad.year.5 == "Meaningful emp MN", "Sustained WF - MN",
                                     ifelse(grad.year.5 == "No MN emp record, not attending ps", "No MN emp record", 
                                            ifelse(grad.year.5 == "Not meaningful, not attending ps", "Non-sustained WF", as.character(grad.year.5))))),
         grad.year.5 = as_factor(grad.year.5))

kable(head(master))

kable(names(master))
```

# Multivariate Confirmation Analysis: Logistic Regression

The purpose of this project is to explain which characteristics are associated with entry into distinct workforce participation pathways following high school graduation in Northeast Minnesota (EDR 3). These pathways include sustained employment in the region, sustained employment elsewhere in Minnesota, non-sustained employment, and having no Minnesota employment record.

The policy question of greatest interest is what factors are associated with **sustained employment in the region**. Earlier analysis relied on bivariate analyses (cross-tabulations, Cramér’s V, and ANOVA) to identify patterns of stratification across workforce participation outcomes. A consistent finding was that postsecondary experiences—particularly credential attainment and where education was completed—are strongly associated with long-term workforce attachment. Academic preparation, early college exposure (PSEO), career and technical education (CTE), and structural demographic factors also demonstrated meaningful relationships.

To strengthen confidence in these findings, a limited multivariate analysis is conducted using logistic regression. The purpose of this modeling is not to establish causality or build a predictive tool, but to examine whether the strongest bivariate relationships remain directionally consistent when key variables are considered simultaneously. This approach helps clarify whether preparation variables retain independent associations after accounting for credential attainment, and whether regional postsecondary completion remains influential when controlling for background and readiness.

Logistic regression is used here as a confirmatory tool rather than the primary analytic framework.

<br>

## Methodology

### **Dependent Variables**

This analysis models five distinct outcomes, each restricted to the appropriate subset of the cohort:

1.  **Sustained regional workforce attachment** (`grad.year.5`) — binary (1 = meaningful employment within the region five years after high school graduation, 0 = otherwise), modeled across the full graduating cohort.
2.  **Sustained regional workforce attachment (`grad.year.5`**, college completers only) - binary (1 = meaningful employment within the region five years after high school graduation, 0 = otherwise), modeled across college completers only.
3.  **Postsecondary completion** (`ps.grad`) — binary (1 = completed a postsecondary credential, 0 = did not), modeled across **all high school graduates**, regardless of whether they ever attended postsecondary education.
4.  **Postsecondary completion within the region** — binary (1 = completed a postsecondary credential within the region, 0 = completed elsewhere in Minnesota or out of state), restricted to **postsecondary completers only**.
5.  **Sub-baccalaureate credential attainment** — binary (1 = highest credential earned is an associate degree or less-than-associate credential, 0 = bachelor's degree or higher), restricted to **postsecondary completers only**.

### **Sample**

All four models are restricted to the **2011–2023 graduating cohorts**. Earlier cohorts (2008–2010) are excluded because PSEO participation was not consistently tracked in the data system prior to 2011 — missingness in that variable is concentrated almost entirely in those years rather than distributed randomly, and including them would risk misclassifying genuine PSEO participants as non-participants. And PSEO participation was identified as having a very strong relationship to our dependent variables in the bivariate analysis.

### Independent Variables

Variables were selected based on (1) a demonstrated bivariate relationship with the relevant outcome(s) in the cross-tabulation and ANOVA analysis, and (2) practical usability in a multivariate model, with closely related variables consolidated to avoid multicollinearity. The same variable set and coding is used across all four outcomes.

**Demographics**

- **`Gender`**
- **`RaceEthnicity`**
- **`economic.status`**
- **`SpecialEdStatus`**
- **`english.learner`**

**High school characteristics**

- **`Dem_Desc`** (district rural/urban classification)
- `range.school`
- **`avg.wages.pct.state`** (local average wages as a percentage of the state average)
- **`avg.unemp.rate`** (local average unemployment rate)

**High school accomplishments**

- **`pseo.participant`** (Post-Secondary Enrollment Options participation)
- **`took.ACT`** (whether the student took the ACT)
- **`ACTCompositeScore`** (ACT composite score, recoded to 0 for non-test-takers — see note below)
- **`ap.exam`** (AP exam participation)
- `sat.taken`
- **`cte.achievement`** (CTE concentration/completion level)

**Postsecondary variables**

- `ps.grad` - used only for the DV1a WF - Region as the dependent variable with full population
- **`highest.cred.level`** - these will be tested on the dependent variables where college completers are the only individuals being analysed.
- **`ps.grad.location`** - these will be tested on the dependent variables where college completers are the only individuals being analysed.

**Notes on Variable Selection**

**Academic achievement (MCA).** MCA Math, Reading, and Science scores are highly correlated with one another (r = 0.63–0.75) and with ACT performance, and all three are excluded from this analysis. MCA Math's missingness rate differs sharply by English learner status (26.1% vs. 2.4% for other students), and this pattern cannot be reliably classified as an independent confounder versus a mediator of unequal access, resourcing, or administrative practice across student populations. Controlling for it — whether through the raw score or a missingness indicator — risks statistically absorbing a real disparity under the guise of a data-quality correction. No MCA subject is retained as a substitute for another; the concern applies to the construct as a whole, not to Math specifically.

**CTE engagement.** `cte.achievement` (concentrator/completer/participant/no CTE) is retained over `total.cte.courses.taken`, a continuous measure of the same underlying construct. The two are redundant, and the categorical version is more directly interpretable for a non-technical audience.

**ACT and SAT participation and score.** Students who did not take the ACT have no composite score by definition — this is a structural absence tied entirely to an observed variable (`took.ACT`), not missing data in the conventional sense. To retain the full sample rather than dropping non-test-takers, `ACTCompositeScore` is recoded to 0 for non-test-takers and entered into the model alongside `took.ACT`. With both terms present, `took.ACT` captures the effect of test participation itself, while `ACTCompositeScore` captures the achievement gradient specifically among test-takers, without the two effects being confounded. `sat.taken` is entered separately, capturing a distinct, smaller subgroup of test-takers not fully overlapping with ACT participation. Because `took.ACT` shows a participation gap by English learner status (41.4% vs. 11.2%) of a similar scale to the MCA missingness gap that led to MCA's exclusion, any attenuation of the English learner coefficient once `took.ACT` enters the model should be interpreted with the same caution — as a variable that may partly reflect unequal access rather than a fully independent behavior — rather than treated as the disparity being explained away.

**District and school context.** `range.school` is included alongside `Dem_Desc`; both show real, stable bivariate relationships with the outcomes examined and are treated as distinct measures of school/community context rather than redundant with one another.

**Variables dropped.** `total.cte.courses.taken`, `MCA.M`, `MCA.R`, and `MCA.S` are excluded — the CTE course-count measure due to redundancy with `cte.achievement`, and all three MCA subjects for the reasons described above.

### Modeling Approach

Each outcome is modeled using a sequential block-entry approach, adding one group of predictors at a time to assess the relative contribution of each. Each block is compared to the prior one using a likelihood ratio test to confirm it adds significant explanatory power, with AIC and pseudo-R² tracked across the sequence to summarize overall model improvement.

**Sustained regional workforce attachment:**

- Model 1: Demographics
- Model 2: + High school characteristics
- Model 3: + High school accomplishments
- Model 4: + College completion

**Sustained regional workforce attachment - college completers only**

- Model 1: Demographics
- Model 2: + High school characteristics
- Model 3: + High school accomplishments
- Model 4: + highest credential earned
- Model 5: + post-secondary location
- Model 6: Dependent on results

**Postsecondary completion, all high school graduates:**

- Model 1: Demographics
- Model 2: + High school characteristics
- Model 3: + High school accomplishments

**Postsecondary completion within the region, completers only:**

- Model 1: Demographics
- Model 2: + High school characteristics
- Model 3: + High school accomplishments

**Sub-baccalaureate credential attainment, completers only:**

- Model 1: Demographics
- Model 2: + High school characteristics
- Model 3: + High school accomplishments

Following these base model sequences, additional modeling will be considered only if the results indicate a need for further specification (e.g., interaction effects, additional mediation checks).

<br>

# Analysis Approach

This multivariate analysis is structured as a series of theoretically guided logistic regression models designed to evaluate mechanisms associated with sustained regional employment. The goal is confirmatory testing of previously observed patterns rather than prediction or causal inference.

<br>

## Evaluation Criteria

Results will be reported using: - Odds ratios - 95% confidence intervals - Statistical significance levels - Directional interpretation

Model comparison metrics (AIC and pseudo R²) will be used to assess relative improvement across specifications within each outcome's model sequence, but emphasis will remain on coefficient stability, attenuation patterns, and substantive interpretation rather than overall fit. <br>

## Interpretation Framework

Interpretation will focus on:

- Direction and relative magnitude of associations
- Changes in coefficients across blocks within each model sequence
- Evidence of attenuation as additional blocks are introduced.

Findings will be described as associations holding other included variables constant. No causal claims will be made. Together, these models provide multivariate confirmation of patterns identified in earlier descriptive analyses while preserving conceptual clarity and interpretability.

# Prep DV1a data

The first thing I need to do is setup my dataset for modeling. I'm going to create the workforce participation into a binary and will use grad.year.5 as my modeling year.

```{r}

model_vars_final <- c("Gender", "RaceEthnicity", "economic.status", "SpecialEdStatus",
                      "english.learner", "Dem_Desc", "range.school", "avg.wages.pct.state", "avg.unemp.rate",
                      "pseo.participant", "took.ACT", "ACTCompositeScore", "sat.taken", "ap.exam",
                      "cte.achievement", "ps.grad",
                      "grad.year.5")

master.model <- master %>%
  filter(grad.year > 2010) %>%
  select(PersonID, all_of(model_vars_final)) %>%
  filter(grad.year.5 != "After 2023",
         grad.year.5 != "Attending ps",
         ps.grad != "Attending ps",
         !is.na(pseo.participant)) %>%
  mutate(grad.year.5 = ifelse(grad.year.5 == "Sustained WF - NE", 1, 0),
         Gender = fct_relevel(Gender, "M"),
         RaceEthnicity = fct_relevel(RaceEthnicity, "White"),
         economic.status = as_factor(economic.status),
         economic.status = fct_relevel(economic.status, "0"),
         SpecialEdStatus = as.character(SpecialEdStatus),
         SpecialEdStatus = fct_relevel(SpecialEdStatus, "0"),
         english.learner = as_factor(english.learner),
         english.learner = fct_relevel(english.learner, "0"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural"),
         range.school = fct_relevel(range.school, "Non-range school"),
         pseo.participant = as.character(pseo.participant),
         pseo.participant = fct_relevel(pseo.participant, "0"),
         took.ACT = fct_relevel(took.ACT, "No"),
         ACTCompositeScore.filled = ifelse(is.na(ACTCompositeScore), 0, ACTCompositeScore),
         sat.taken = as_factor(sat.taken),
         sat.taken = fct_relevel(sat.taken, "0"),
         ap.exam = as_factor(ap.exam),
         ap.exam = fct_relevel(ap.exam, "0"),
         cte.achievement = fct_relevel(cte.achievement, "No CTE"),
         ps.grad = ifelse(ps.grad == "Y", "Yes", ps.grad),
         ps.grad = fct_relevel(ps.grad, "No")) 

kable(head(master.model))

kable(names(master.model))

master %>% filter(grad.year > 2010) %>% nrow()
master %>% filter(grad.year > 2010, grad.year.5 != "After 2023") %>% nrow()
master %>% filter(grad.year > 2010, grad.year.5 != "After 2023", 
                  grad.year.5 != "Attending ps") %>% nrow()
master %>% filter(grad.year > 2010, grad.year.5 != "After 2023", 
                  grad.year.5 != "Attending ps", ps.grad != "Attending ps") %>% nrow()
```

We now have `r comma(nrow(master.model))` individuals to model with `r ncol(master.model) - 2` variables not including `grad.year.5`.

<br>

# Modeling: Sustained WF - Region (full population)

## Model 1: Demographics

This model estimates the association between demographic characteristics and sustained regional workforce attachment five years after high school graduation, using the full analytic sample (n = 20,198). The outcome is coded 1 = sustained workforce attachment within the region ("Sustained WF - NE"), 0 = all other outcomes (non-sustained, sustained elsewhere in Minnesota, or no Minnesota employment record).

Demographic factors show a consistent pattern: every group examined is *less* likely than its reference category to achieve sustained regional attachment.

**Race and ethnicity** shows the widest range of effects in this model.

- **Black graduates:** approximately 53% lower odds of sustained regional attachment relative to White graduates (OR ≈ 0.47).
- **Asian/Pacific Islander graduates:** approximately 43% lower odds (OR ≈ 0.57).
- **Hispanic graduates:** approximately 39% lower odds (OR ≈ 0.61).
- **American Indian graduates:** approximately 15% lower odds (OR ≈ 0.85), the smallest gap among the four groups but still statistically significant.

**Other demographic predictors** show smaller but still significant effects.

- **English learner status:** approximately 21% lower odds of sustained regional attachment (OR ≈ 0.79).
- **Special education status:** approximately 15% lower odds (OR ≈ 0.85).
- **Female graduates:** approximately 12% lower odds relative to male graduates (OR ≈ 0.88).
- **Economic disadvantage:** approximately 7% lower odds (OR ≈ 0.93) — the weakest effect in the model, significant at p \< .05 but only just.

**Overall Interpretation**

At the demographics-only stage, every predictor points in the same direction: relative to the reference group (male, White, not economically disadvantaged, not special education, not an English learner), each demographic category examined shows *lower* odds of sustained regional workforce attachment. Race and ethnicity produce the largest disparities, with Black and Asian/Pacific Islander graduates showing the steepest gaps. Economic status has the smallest effect of the five predictors, sitting just inside statistical significance. These patterns are essentially identical to the initial exploratory version of this model (n = 20,386, prior to the PSEO-missingness exclusion), confirming that excluding those 188 records did not materially change the demographic baseline.

This is a baseline model only — it does not yet account for high school context, academic preparation, or postsecondary completion, all of which are added in subsequent blocks.

**Model fit:** AIC = 25,094.7; residual deviance = 25,077 on 20,189 df (null deviance = 25,227 on 20,197 df).

```{r}
model1_dv1 <- glm(
  grad.year.5 ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner,
  family = binomial,
  data = master.model
)

model1_dv1

summary(model1_dv1)
nobs(model1_dv1)
AIC(model1_dv1)

# odds ratios + 95% CIs, per your evaluation criteria
exp(cbind(OR = coef(model1_dv1), confint(model1_dv1)))

or_plot <- tidy(model1_dv1, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )

ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )



```

<br>

## Model 2: Demographics + High School Characteristics

This model adds district rural/urban classification, Range school status, local average wages, and local average unemployment rate to the demographic baseline (n = 20,198).

**District classification shows a strong, significant relationship with sustained regional attachment**, consistent with the bivariate finding. Graduates from Town/rural mix districts show 30% higher odds of sustained regional attachment than those from Entirely rural districts (OR ≈ 1.30, CI: 1.127-1.510, p \< .001), and graduates from Urban/town/rural mix districts show 57% higher odds (OR ≈ 1.57, CI: 1.355-1.830, p \< .001) — a clear, ordered gradient matching what the bivariate cross-tabs showed.

**Range school status is a significant, independent predictor in its own right.** Range graduates show 18% higher odds of sustained regional attachment than non-Range graduates (OR ≈ 1.18, CI: 1.097-1.279, p \< .001), confirming the bivariate finding survives multivariate control alongside demographics and district classification — **this variable is pulling real, non-redundant weight in the model**, not just echoing Dem_Desc.

**Local wages and unemployment rate do not hold up as independent predictors.** Average wages as a percent of the state average shows a wide confidence interval crossing 1 (OR ≈ 1.06, CI: 0.555-2.010, p = .865), and average unemployment rate is similarly non-significant (OR ≈ 1.01, CI: 0.984-1.035, p = .476). Neither local economic indicator adds explanatory power once district classification and Range status are in the model — consistent with this project's broader, recurring finding that individual and school-level factors matter more than county-level economic conditions.

**Economic status weakens to only marginal significance** once district context is added (OR ≈ 0.94, CI: 0.884-1.006, p = .073), down from a clearly significant effect in Model 1 (p = .031). This suggests some of economic status's apparent relationship with regional attachment may run through where a student's school is located, rather than economic status operating entirely independently of geography. Worth tracking through Model 3 rather than treating this as settled.

Gender, Race/Ethnicity, Special Education status, and English learner status **all remain significant and stable** in direction and magnitude from Model 1 — none of the demographic effects are meaningfully explained by the addition of district or economic context.

**Overall Interpretation**

Adding high school characteristics significantly improves model fit (likelihood ratio test: Δdeviance = 100.53, df = 5, p \< .001; AIC decreases from 25,094.7 to 25,004.2). **District classification and Range school status both confirm as genuine, independent predictors** of regional workforce attachment — this is the clearest signal in this block. Local wage and unemployment conditions, by contrast, show no independent relationship once these more specific measures of place are accounted for. **Economic status's attenuation to marginal significance is the one finding worth watching closely** as high school accomplishments are added in Model 3.

Model fit: AIC = 25,004.2 (down from 25,094.7 in Model 1); Δdeviance = 100.53 on 5 df, p \< .001; n = 20,198 in both models.

```{r}
model2_dv1 <- glm(
  grad.year.5 ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    range.school +
    avg.wages.pct.state +
    avg.unemp.rate,
  family = binomial,
  data = master.model
)

summary(model2_dv1)

nobs(model1_dv1)
nobs(model2_dv1)


AIC(model1_dv1)
AIC(model2_dv1)

exp(cbind(OR = coef(model2_dv1), confint(model2_dv1)))

anova(model1_dv1, model2_dv1, test = "Chisq")

or_plot <- tidy(model2_dv1, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )

ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )

```

<br>

## Model 3: Demographics + High School Characteristics + High School Accomplishments

This model adds PSEO participation, ACT participation and score, SAT participation, AP exam participation, and CTE achievement to the growing specification (n = 20,198).

**English learner status is a significant, independent predictor of reduced regional attachment — the central finding of this model.** OR ≈ 0.741 (CI: 0.629-0.872, p \< .001), meaning English learners show roughly 26% lower odds of sustained regional workforce attachment, holding up fully alongside every high school accomplishment variable in the model. This is not explained away by academic engagement or preparation.

**Race/Ethnicity effects persist across all four categories, fully independent of high school accomplishments.** American Indian (OR ≈ 0.85, p = .012), Asian/Pacific Islander (OR ≈ 0.66, p = .006), Black (OR ≈ 0.43, p \< .001), and Hispanic graduates (OR ≈ 0.59, p \< .001) all show significantly lower odds of sustained regional attachment relative to White graduates — none of these disparities are explained by academic engagement patterns.

**CTE achievement continues to anchor students regionally.** CTE Concentrators/Completers show 42% higher odds of sustained regional attachment (OR ≈ 1.42, p \< .001), and CTE Participants show 22% higher odds (OR ≈ 1.22, p \< .001) — the clearest "stay local" signal in the model.

**ACT shows a genuine crossover pattern that requires a joint reading.** Taking the ACT is associated with substantially *higher* odds of regional attachment (OR ≈ 3.77, p \< .001), while among test-takers, each additional composite point is associated with *lower* odds (OR ≈ 0.938, p \< .001). Together these describe a threshold effect: simply engaging with the ACT is not a signal of leaving the region — it's specifically high-scoring ACT engagement that predicts departure, likely reflecting pursuit of selective, non-regional postsecondary options.

**SAT participation moves in the opposite direction from ACT participation** — a real, substantively important contrast, not a contradiction. SAT-takers show 60% lower odds of regional attachment (OR ≈ 0.40, p \< .001). Unlike the ACT, which nearly every college-bound student in this region takes, SAT-taking identifies a smaller, more selective subgroup specifically oriented toward postsecondary options outside Minnesota — consistent with the bivariate finding that SAT-takers are the most mobile subgroup in this analysis.

**AP exam participation is associated with lower odds of regional attachment** (OR ≈ 0.71, p \< .001), consistent with the pattern that college-preparatory engagement pulls graduates away from the region.

**PSEO participation is not statistically significant here** (OR ≈ 1.04, CI: 0.969-1.111, p = .290) — its bivariate influence does not hold up as an independent effect once ACT, SAT, AP, and CTE are considered together.

**Economic status and Special Education status both strengthen rather than attenuate.** Economic status returns to clear significance (OR ≈ 0.86, p \< .001), reversing the marginal result from Model 2. Special Ed status also strengthens (OR ≈ 0.77, p \< .001, down from OR ≈ 0.85 in Model 2). Neither effect is explained by academic engagement — if anything, accounting for it sharpens both.

**Range school status and Gender both weaken to non-significance.** Range school (OR ≈ 1.06, p = .136) and Gender (OR ≈ 0.94, p = .069) were both significant in Model 2 but fall short of conventional significance here — suggesting their apparent relationships with regional attachment may run substantially through students' patterns of academic and CTE engagement rather than operating independently.

**Average local wages becomes significant for the first time** (OR ≈ 2.20, CI: 1.139-4.248, p = .019) — higher local wages relative to the state average now show a real, if modest, positive relationship with regional attachment once accomplishment patterns are accounted for. Average unemployment rate remains non-significant (p = .141).

District classification (Dem_Desc) remains significant and stable, essentially unchanged from Model 2.

**Overall Interpretation**

Adding high school accomplishments substantially improves model fit (likelihood ratio test: Δdeviance = 493.50, df = 7, p \< .001; AIC decreases from 25,004.2 to 24,524.7). **English learner status and all four Race/Ethnicity categories are confirmed as genuine, independent predictors of reduced regional attachment, fully robust to every accomplishment variable in the model.** Academic and college-preparatory engagement (ACT score, AP, SAT) continues to point away from regional attachment, while CTE engagement anchors students to the region — the "dual pathways" pattern found throughout this project's descriptive work. Range school status and Gender, both significant in Model 2, do not hold up independently once accomplishment patterns are considered.

Model fit: AIC = 24,524.7 (down from 25,004.2 in Model 2); Δdeviance = 493.50 on 7 df, p \< .001; n = 20,198 across all three models.

```{r}
model3_dv1 <- glm(
  grad.year.5 ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    range.school +
    avg.wages.pct.state +
    avg.unemp.rate +
    pseo.participant +
    took.ACT +
    ACTCompositeScore.filled +
    sat.taken +
    ap.exam + 
    cte.achievement,
  family = binomial,
  data = master.model
)

nobs(model1_dv1)
nobs(model2_dv1)
nobs(model3_dv1)


AIC(model1_dv1)
AIC(model2_dv1)
AIC(model3_dv1)

summary(model3_dv1)

exp(cbind(OR = coef(model3_dv1), confint(model3_dv1)))

anova(model1_dv1, model2_dv1, model3_dv1, test = "Chisq")

or_plot <- tidy(model3_dv1, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )

ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )


```

<br>

## Model 4: Demographics + High School Characteristics + High School Accomplishments + Postsecondary Completion

This model adds ps.grad (completed a postsecondary credential, regardless of location or credential level) to the full specification (n = 20,198).

**Postsecondary completion shows a strong, independent positive effect on regional attachment.** OR ≈ 1.72 (CI: 1.601-1.837, p \< .001) — graduates who completed a postsecondary credential show 72% higher odds of sustained regional workforce attachment than those who did not, holding up fully alongside every other variable in the model.

**English learner status remains fully robust — essentially unchanged from Model 3.** OR ≈ 0.741 (CI: 0.629-0.872, p \< .001), identical in magnitude to Model 3. Completing this final block confirms the finding holds regardless of what else is controlled for: this is not an artifact of any single specification choice.

**Race/Ethnicity mostly holds, with one exception worth flagging.** Black (OR ≈ 0.44, p \< .001), Hispanic (OR ≈ 0.62, p = .002), and Asian/Pacific Islander graduates (OR ≈ 0.69, p = .015) all remain significant. American Indian graduates fall to marginal significance here (OR ≈ 0.88, p = .059) — the first time in this sequence this group hasn't cleared the conventional threshold, though the direction and magnitude are consistent with every prior model.

**The ACT crossover pattern strengthens further.** Taking the ACT now shows an even larger positive association (OR ≈ 4.18, p \< .001), while the composite-score penalty among test-takers also strengthens slightly (OR ≈ 0.928 per point, p \< .001) — consistent with high-scoring ACT-takers disproportionately completing credentials outside the region, which `ps.grad` doesn't fully capture (it measures completion, not location).

**Range school status fades to fully non-significant** (OR ≈ 1.01, p = .833), continuing its decline from Model 2 (significant) through Model 3 (marginal) to here (essentially null) — its apparent regional- attachment advantage appears to run almost entirely through the combination of accomplishment patterns and postsecondary completion, not independently.

**Gender returns to significance** (OR ≈ 0.90, p = .001), reversing its marginal status in Model 3. **Economic status and Special Education status both remain significant**, though Economic status weakens somewhat (OR ≈ 0.93, p = .029, down from p \< .001 in Model 3).

**Average local wages strengthens further** (OR ≈ 2.50, CI: 1.290-4.843, p = .007) — now the most significant it has been anywhere in this sequence. PSEO participation remains non-significant (OR ≈ 0.98, p = .653). SAT, AP exam, and CTE achievement all remain essentially unchanged from Model 3 in direction and magnitude.

**Overall Interpretation**

Adding postsecondary completion significantly improves model fit (likelihood ratio test: Δdeviance = 239.06, df = 1, p \< .001; AIC decreases from 24,524.7 to 24,287.6), confirming `ps.grad` as a large, genuine, independent predictor of regional attachment — unsurprising, but worth confirming rather than assuming. Across the full four-model sequence, **English learner status and the Black, Hispanic, and Asian/Pacific Islander disparities are the standout findings: all are confirmed as robust, independent predictors of reduced regional attachment that survive full control for demographics, geography, academic accomplishments, and postsecondary completion.** American Indian graduates' disadvantage weakens to marginal only in the final, fullest model. The ACT crossover pattern (protective participation, adverse score-among-takers) replicates and strengthens throughout, and CTE achievement is confirmed as the clearest "stay local" signal in the entire analysis.

Model fit: AIC = 24,287.6 (down from 24,524.7 in Model 3); Δdeviance = 239.06 on 1 df, p \< .001; n = 20,198 across all four models.

```{r}
model4_dv1 <- glm(
  grad.year.5 ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    range.school +
    avg.wages.pct.state +
    avg.unemp.rate +
    pseo.participant +
    took.ACT +
    ACTCompositeScore.filled +
    ap.exam + 
    sat.taken +
    cte.achievement +
    ps.grad,
  family = binomial,
  data = master.model
)

nobs(model1_dv1)
nobs(model2_dv1)
nobs(model3_dv1)
nobs(model4_dv1)

AIC(model1_dv1)
AIC(model2_dv1)
AIC(model3_dv1)
AIC(model4_dv1)

summary(model4_dv1)

exp(cbind(OR = coef(model4_dv1), confint(model4_dv1)))

anova(model1_dv1, model2_dv1, model3_dv1, model4_dv1, test = "Chisq")

or_plot <- tidy(model4_dv1, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )


ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )


```

<br>

## DV1a Summary: What Predicts Sustained Regional Workforce Attachment (Full Cohort)

Across all four models, this analysis identifies a clear, consistent set of demographic, academic, and postsecondary predictors of sustained regional workforce attachment.

**English learner status is a robust, independent predictor of reduced regional attachment.** OR remains at approximately 0.74 from the point academic accomplishments enter the model (Model 3) through the fully specified model (Model 4), significant at every stage (p \< .001). English learners show roughly 26% lower odds of sustained regional workforce attachment than non-English-learners, and this holds up fully alongside demographics, geography, academic engagement, and postsecondary completion — this is one of the clearest and most consequential findings in this analysis.

**Race/Ethnicity disparities are largely confirmed as independent, with one exception in the final model.** Black, Hispanic, and Asian/Pacific Islander graduates all show significantly lower odds of sustained regional attachment across every model in the sequence. Black graduates show the largest and most stable disparity (OR ≈ 0.43–0.47 throughout). American Indian graduates show a consistent, significant disadvantage through Model 3, weakening to marginal significance (p = .059) only once postsecondary completion is added in the final model — the sole exception to an otherwise uniform pattern of independent racial disparities.

**Career and Technical Education engagement is the clearest "stay local" signal identified anywhere in this analysis.** CTE Concentrators/ Completers show 41–42% higher odds of sustained regional attachment, and CTE Participants show 21–22% higher odds, holding stable and significant across every model from the point CTE enters the specification.

**ACT participation and score show a genuine crossover pattern, not a single consistent effect.** Simply taking the ACT is associated with substantially higher odds of regional attachment (OR growing from ≈ 3.77 to ≈ 4.18 across Models 3–4), while among test-takers, each additional composite point is associated with lower odds (OR ≈ 0.93–0.94 per point). Together, these describe a threshold effect: engaging with the ACT itself is not a signal of leaving the region — it is specifically high-scoring ACT engagement that predicts departure, most plausibly reflecting pursuit of selective, non-regional postsecondary options.

**SAT participation and AP exam participation both point in the opposite direction from CTE, consistent with a dual-pathway structure to these outcomes.** SAT-takers show 60% lower odds of regional attachment (OR ≈ 0.40) — the largest single reduction of any accomplishment variable — identifying a small, highly mobile subgroup oriented toward postsecondary options outside Minnesota. AP exam participation shows a smaller but consistent reduction (OR ≈ 0.71). PSEO participation, despite a strong bivariate relationship with this outcome, does not hold up as an independent predictor once ACT, SAT, AP, and CTE are considered together.

**Postsecondary completion is confirmed as a large, independent predictor of regional attachment.** Graduates who completed a postsecondary credential show 72% higher odds of sustained regional attachment (OR ≈ 1.72, p \< .001) — a substantial effect that holds up fully alongside every academic accomplishment and demographic variable already in the model.

**Economic status and Special Education status both strengthen once academic accomplishments are considered, rather than being explained by them.** Both remain significant, independent predictors of reduced regional attachment through the final model, with Special Ed showing the larger and more stable effect of the two (OR ≈ 0.77–0.82).

**Gender shows a real but modest and somewhat inconsistent relationship with regional attachment**, remaining significant in Models 1, 2, and 4, but falling to marginal significance in Model 3 — its relationship with this outcome appears to be partly, though not entirely, connected to academic accomplishment patterns.

**District classification (Dem_Desc) is a stable, significant predictor throughout.** Graduates from Urban/town/rural mix and Town/rural mix districts consistently show higher odds of regional attachment than those from Entirely rural districts, holding steady across all four models.

**Range school status and local wage/unemployment conditions tell a more complicated story.** Range school status is significant on its own (Model 2) but fades to non-significance once academic accomplishments and postsecondary completion are considered (Models 3–4) — its apparent advantage appears to run through students' accomplishment and completion patterns rather than operating independently. Average local wages, by contrast, shows no relationship on its own (Model 2) but becomes significant and strengthens once accomplishments and completion are added (OR ≈ 2.20–2.50 in Models 3–4) — a real, if modest, effect that only emerges once other factors are accounted for. Average local unemployment rate remains non-significant throughout.

<br>

# Prep DV1b data

The first thing I need to do is setup my dataset for modeling. I'm going to create the workforce participation into a binary and will use grad.year.5 as my modeling year.

```{r}

model_vars_dv1b <- c("Gender", "RaceEthnicity", "economic.status", "SpecialEdStatus",
                      "english.learner", "Dem_Desc", "range.school", "avg.wages.pct.state", "avg.unemp.rate",
                      "pseo.participant", "took.ACT", "ACTCompositeScore", "ap.exam", "sat.taken",
                      "cte.achievement", "highest.cred.level", "ps.grad.location", "grad.year.5")

master.model.dv1b <- master %>%
  filter(grad.year > 2010,
         ps.grad == "Y",
         ps.grad.location != "Did not graduate ps") %>%
  select(PersonID, all_of(model_vars_dv1b)) %>%
  filter(grad.year.5 != "After 2023",
         grad.year.5 != "Attending ps",
         !is.na(pseo.participant)) %>%
  mutate(grad.year.5 = ifelse(grad.year.5 == "Sustained WF - NE", 1, 0),
         Gender = fct_relevel(Gender, "M"),
         RaceEthnicity = fct_relevel(RaceEthnicity, "White"),
         economic.status = as_factor(economic.status),
         economic.status = fct_relevel(economic.status, "0"),
         SpecialEdStatus = as.character(SpecialEdStatus),
         SpecialEdStatus = fct_relevel(SpecialEdStatus, "0"),
         english.learner = as_factor(english.learner),
         english.learner = fct_relevel(english.learner, "0"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural"),
         range.school = fct_relevel(range.school, "Non-range school"),
         pseo.participant = as.character(pseo.participant),
         pseo.participant = fct_relevel(pseo.participant, "0"),
         took.ACT = fct_relevel(took.ACT, "No"),
         ACTCompositeScore.filled = ifelse(is.na(ACTCompositeScore), 0, ACTCompositeScore),
         ap.exam = as_factor(ap.exam),
         ap.exam = fct_relevel(ap.exam, "0"),
         sat.taken = as_factor(sat.taken),
         sat.taken = fct_relevel(sat.taken, "0"),
         cte.achievement = fct_relevel(cte.achievement, "No CTE"),
         highest.cred.level = ifelse(highest.cred.level %in% c("Associate degree", "Less than Associate Degree"), "Associate or lesser credential", as.character(highest.cred.level)),
         highest.cred.level = fct_relevel(highest.cred.level, "Bachelor degree or higher"),
         ps.grad.location = ifelse(ps.grad.location == "PS grad Region", "PS grad Region", "PS grad outside region"),
         ps.grad.location = fct_relevel(ps.grad.location, "PS grad outside region")) 

kable(head(master.model.dv1b))

kable(names(master.model.dv1b))
```

We now have `r comma(nrow(master.model))` individuals to model with `r ncol(master.model) - 2` variables not including `grad.year.5`.

<br>

# Modeling: Sustained WF - Region (college completers only)

## Model 1: Demographics

This model estimates the association between demographic characteristics and sustained regional workforce attachment five years after high school graduation, restricted to postsecondary completers (n = 8,500). The outcome is coded 1 = sustained workforce attachment within the region ("Sustained WF - NE"), 0 = all other outcomes.

**This model tells a meaningfully different story than DV1a's demographics-only model, not just a smaller-sample version of the same one.** Several effects reverse direction entirely once the population is restricted to college completers.

- **Economic status:** OR ≈ 1.21 (CI: 1.091–1.340) — economically disadvantaged completers show *higher* odds of sustained regional attachment. This is the opposite direction from DV1a, where economic disadvantage predicted *lower* odds across the full cohort.
- **Special Education status:** OR ≈ 1.40 (CI: 1.133–1.740) — also reversed from DV1a, where Special Ed status predicted lower odds.
- **American Indian graduates:** OR ≈ 1.28 (CI: 1.003–1.624) — reversed from a negative effect in DV1a to a positive, borderline-significant effect here (p = .046).
- **English learner status:** no longer significant (OR ≈ 1.05, CI: 0.714–1.542, p = .80) — a large shift from DV1a, where this was one of the more robust demographic predictors.

**Some effects persist in the same direction as DV1a, though attenuated:**

- **Gender:** OR ≈ 0.89 (CI: 0.815–0.975) — female completers show modestly lower odds, consistent with DV1a.
- **Black graduates:** OR ≈ 0.56 (CI: 0.336–0.911) — still lower odds than White graduates, consistent in direction with DV1a.
- **Asian/Pacific Islander graduates:** OR ≈ 0.48 (CI: 0.288–0.754) — still the largest disparity among Race/Ethnicity categories, consistent with DV1a.
- **Hispanic graduates:** OR ≈ 0.61, but no longer statistically significant (CI: 0.341–1.029, p = .073) — likely reflecting reduced statistical power in this smaller population (n = 65 Hispanic completers), not necessarily a disappearing effect.

**Overall Interpretation**

Among students who complete a postsecondary credential, the demographic predictors of sustained regional workforce attachment look substantially different than they do across the full graduating cohort. Economic disadvantage and Special Education status, both associated with *lower* regional attachment in the full-cohort model, are associated with *higher* attachment once the population is restricted to completers — suggesting that among students who make it through college, these groups may be more likely to return to or remain in the region, possibly reflecting stronger local ties, more localized program/institution choices, or different postsecondary pathways than their more economically advantaged or non-Special-Ed peers. Race/Ethnicity effects, particularly for Asian/Pacific Islander and Black graduates, remain directionally consistent with the full-cohort findings, suggesting those disparities are not solely a function of who completes college but persist even after completion. English learner status, notably, shows no effect at all in this population — a genuine shift from its significance in every DV1a model, worth naming explicitly rather than attributing to sample size alone, since the coefficient itself moved toward the null, not just its precision.

Given the smaller sample size (n = 8,500 vs. 20,198), and especially the very small Hispanic subgroup (n = 65) in this population, some of these shifts should be interpreted cautiously — reduced power alone could account for lost significance in a few cases, but the *reversals* in direction for economic status, Special Ed, and American Indian graduates are unlikely to be explained by power alone and appear to reflect a genuinely different pattern among completers specifically.

**Model fit:** AIC = 11,147.3; residual deviance = 11,129 on 8,491 df (null deviance = 11,187 on 8,499 df).

```{r}
model1_dv1b <- glm(
  grad.year.5 ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner,
  family = binomial,
  data = master.model.dv1b
)

model1_dv1b

summary(model1_dv1b)
nobs(model1_dv1b)
AIC(model1_dv1b)

# odds ratios + 95% CIs, per your evaluation criteria
exp(cbind(OR = coef(model1_dv1b), confint(model1_dv1b)))

or_plot <- tidy(model1_dv1b, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )

ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )



```

<br>

## Model 2: Demographics + High School Characteristics

This model adds district rural/urban classification, Range school status, local average wages, and local average unemployment rate to the demographic baseline, restricted to postsecondary completers (n = 8,500).

**District classification and Range school status both show strong, significant relationships with regional attachment among completers.** Graduates from Town/rural mix districts show 46% higher odds of regional attachment than those from Entirely rural districts (OR ≈ 1.46, p = .001), and Urban/town/rural mix districts show 79% higher odds (OR ≈ 1.79, p \< .001). Range school graduates show 32% higher odds than non-Range graduates (OR ≈ 1.32, p \< .001) — a larger effect than seen in the full- population model, suggesting Range school status matters even more once the population is narrowed to completers specifically.

**Several demographic effects point in the opposite direction from the full-population model — this is a real, substantively important divergence, not noise.** Economic status shows *higher*, not lower, odds of regional attachment among completers (OR ≈ 1.22, p \< .001). Special Education status also shows *higher* odds (OR ≈ 1.39, p = .002). American Indian graduates show *higher* odds relative to White graduates (OR ≈ 1.35, p = .015) — the opposite direction from the full-population finding. Among graduates who complete a postsecondary credential, economic disadvantage, Special Ed status, and American Indian identity are associated with *staying* in the region, not leaving it — a meaningfully different story than the one told by the full graduating population.

**English learner status shows no relationship with regional attachment in this population** (OR ≈ 1.08, CI: 0.730-1.584, p = .707) — a real non-finding worth stating plainly rather than glossing over. Whatever drives English learners' reduced regional attachment in the full population does not appear to operate among those who go on to complete a postsecondary credential.

**Black and Asian/Pacific Islander graduates remain significant, consistent in direction with the full-population model.** Asian/Pacific Islander graduates show substantially lower odds of regional attachment (OR ≈ 0.48, p = .003), and Black graduates show lower odds as well (OR ≈ 0.57, p = .027). Hispanic graduates fall just short of conventional significance here (OR ≈ 0.60, p = .069).

**Local wages and unemployment rate remain non-significant**, consistent with the full-population model (wages: OR ≈ 1.06, p = .903; unemployment: OR ≈ 1.03, p = .193).

**Gender remains significant and stable** (OR ≈ 0.90, p = .022), close to its Model 1 magnitude.

**Overall Interpretation**

Adding high school characteristics significantly improves model fit (likelihood ratio test: Δdeviance = 79.97, df = 5, p \< .001; AIC decreases from 11,147.3 to 11,077.4). **The standout finding here is that several demographic patterns reverse direction once the population is restricted to postsecondary completers** — economic disadvantage, Special Ed status, and American Indian identity all predict *higher* odds of regional attachment among completers, the opposite of their relationship in the full graduating population. English learner status, a significant predictor in the full population, shows no relationship at all here. District classification and Range school status are confirmed as strong, independent predictors, both larger in magnitude than in the full- population model.

Model fit: AIC = 11,077.4 (down from 11,147.3 in Model 1); Δdeviance = 79.97 on 5 df, p \< .001; n = 8,500 in both models.

```{r}
model2_dv1b <- glm(
  grad.year.5 ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    range.school +
    avg.wages.pct.state +
    avg.unemp.rate,
  family = binomial,
  data = master.model.dv1b
)

nobs(model1_dv1b)
nobs(model2_dv1b)


AIC(model1_dv1b)
AIC(model2_dv1b)

summary(model2_dv1b)

exp(cbind(OR = coef(model2_dv1b), confint(model2_dv1b)))

anova(model1_dv1b, model2_dv1b, test = "Chisq")

or_plot <- tidy(model2_dv1b, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )

ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )

```

<br>

## Model 3: Demographics + High School Characteristics + High School Accomplishments

This model adds PSEO participation, ACT participation and score, SAT participation, AP exam participation, and CTE achievement to the specification, restricted to postsecondary completers (n = 8,500).

**Four demographic effects that were significant in Model 2 lose significance here — a real, substantive finding, not noise.** Economic status (OR ≈ 1.04, p = .524), Special Education status (OR ≈ 0.98, p = .856), American Indian graduates (OR ≈ 1.23, p = .099), and **Gender** (OR ≈ 0.98, p = .732) all fall to non-significance. This suggests their apparent "stay local" advantage among completers is substantially explained by patterns of academic engagement — likely lower ACT/AP participation and stronger CTE involvement — rather than these characteristics operating independently of how students engaged academically in high school.

**Range school status follows the same pattern.** Its Model 2 effect (OR ≈ 1.32, significant) fades to clearly non-significant here (OR ≈ 1.08, p = .202) — Range schools' apparent regional-attachment advantage among completers appears to run through accomplishment patterns rather than operating independently, echoing what was found in the full-population model.

**English learner status remains non-significant, consistent with Model 2.** OR ≈ 0.87 (p = .497) — among postsecondary completers specifically, English learner status shows no relationship with regional attachment at any stage of this model sequence, a genuine contrast with the full-population finding.

**District classification remains strong and significant.** Town/rural mix (OR ≈ 1.41, p = .004) and Urban/town/rural mix (OR ≈ 1.88, p \< .001) districts both continue to show meaningfully higher odds of regional attachment than Entirely rural districts, essentially unchanged from Model 2.

**Black graduates remain significant; Asian/Pacific Islander and Hispanic graduates both fall just short of conventional significance.** Black graduates show OR ≈ 0.47 (p = .004). Asian/Pacific Islander graduates show OR ≈ 0.62 (p = .057) and Hispanic graduates show OR ≈ 0.61 (p = .086) — both worth treating as fragile, trending findings rather than confirmed effects at this stage, not as settled non-findings either.

**The ACT crossover pattern replicates precisely, matching the full- population model.** Taking the ACT shows a large, highly significant, positive relationship with regional attachment (OR ≈ 4.28, p \< .001), while among test-takers, each additional composite point is associated with significantly lower odds (OR ≈ 0.917 per point, p \< .001) — the same threshold effect: engagement itself is protective, while high performance specifically predicts leaving the region.

**SAT and AP exam participation both show a significant negative relationship with regional attachment, consistent with the full- population model.** SAT participation OR ≈ 0.47 (p \< .001), AP exam OR ≈ 0.72 (p \< .001).

**CTE achievement continues to anchor students regionally, both categories clearly significant.** Concentrators/Completers (OR ≈ 1.38, p \< .001) and Participants (OR ≈ 1.21, p \< .001) both remain significant and essentially unchanged from what CTE showed in the full-population model.

**PSEO participation remains non-significant** (OR ≈ 0.98, p = .699), consistent with the full-population finding that PSEO's bivariate influence doesn't hold up once other accomplishment variables are considered.

**Local wages just misses conventional significance** (OR ≈ 2.74, p = .051) — close enough to the threshold to watch rather than dismiss, but not a confirmed effect at this stage. Average unemployment rate remains clearly non-significant (p = .864).

**Overall Interpretation**

Adding high school accomplishments substantially improves model fit (likelihood ratio test: Δdeviance = 433.58, df = 7, p \< .001; AIC decreases from 11,077.4 to 10,657.8). **The standout finding in this model is that economic status, Special Education status, American Indian identity, Gender, and Range school status — all significant, positive predictors of regional attachment in Model 2 — lose significance once academic accomplishments are considered.** This suggests their apparent advantage among completers runs substantially through how these groups engage academically in high school, not through the characteristics themselves. English learner status remains a clean, consistent null throughout this entire sequence — a genuinely different story from the full-population model, where it was one of the most robust findings. The ACT crossover, CTE's anchoring effect, and SAT/AP's pull-away pattern all replicate the full-population model closely.

Model fit: AIC = 10,657.8 (down from 11,077.4 in Model 2); Δdeviance = 433.58 on 7 df, p \< .001; n = 8,500 in both models.

```{r}
model3_dv1b <- glm(
  grad.year.5 ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    range.school +
    avg.wages.pct.state +
    avg.unemp.rate +
    pseo.participant +
    took.ACT +
    ACTCompositeScore.filled +
    ap.exam + 
    sat.taken +
    cte.achievement,
  family = binomial,
  data = master.model.dv1b
)

nobs(model1_dv1b)
nobs(model2_dv1b)
nobs(model3_dv1b)


AIC(model1_dv1b)
AIC(model2_dv1b)
AIC(model3_dv1b)

summary(model3_dv1b)

exp(cbind(OR = coef(model3_dv1b), confint(model3_dv1b)))

anova(model1_dv1b, model2_dv1b, model3_dv1b, test = "Chisq")

or_plot <- tidy(model3_dv1b, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )

ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )


```

<br>

## Model 4: Demographics + High School Characteristics + High School Accomplishments + Highest Credential Earned

This model adds highest.cred.level (Associate or lesser credential vs. Bachelor's degree or higher) to the full specification, restricted to postsecondary completers (n = 8,500).

**Credential level is a strong, independent predictor of regional attachment — the standout finding in this block.** Graduates whose highest credential is an Associate degree or less show 78% higher odds of sustained regional attachment than those with a Bachelor's degree or higher (OR ≈ 1.78, CI: 1.594-1.985, p \< .001). This is intuitive and substantively important: sub-baccalaureate credentials are disproportionately tied to programs and occupations rooted in the local and regional economy, while Bachelor's-level credentials open doors to opportunities that pull graduates further afield.

**Took-ACT's effect shrinks meaningfully once credential level is added, though it remains large and highly significant.** OR drops from ≈ 4.28 in Model 3 to ≈ 3.25 here (p \< .001 in both) — credential level is absorbing some, but not all, of what ACT participation was capturing, which makes sense given how closely ACT engagement tracks with pursuing a Bachelor's degree specifically.

**Local wages becomes significant for the first time in this sequence.** OR ≈ 2.90 (CI: 1.042-8.020, p = .041) — a real, if modest, positive relationship between local wage strength and regional attachment once credential level is accounted for, crossing the threshold it had only approached in Model 3 (p = .051).

**All demographic effects that were non-significant in Model 3 remain non-significant here.** Gender (p = .629), economic status (p = .559), Special Education status (p = .435), English learner status (p = .572), American Indian graduates (p = .132), and Range school status (p = .314) all continue to show no independent relationship with regional attachment among completers, once credential level joins academic accomplishments in the model.

**Black graduates remain the clearest significant Race/Ethnicity finding** (OR ≈ 0.51, p = .009). Asian/Pacific Islander (OR ≈ 0.65, p = .092) and Hispanic graduates (OR ≈ 0.62, p = .095) both remain in the same marginal, trending-but-unconfirmed territory as Model 3.

**District classification, ACT composite score, SAT, AP exam, and CTE achievement all remain significant and essentially stable** in direction and magnitude from Model 3 — none of these relationships are meaningfully altered by adding credential level to the specification.

**Overall Interpretation**

Adding highest credential earned significantly improves model fit (likelihood ratio test: Δdeviance = 105.85, df = 1, p \< .001; AIC decreases from 10,657.8 to 10,553.9), confirming credential level as a large, genuine, independent predictor of regional attachment among completers — sub-baccalaureate credential holders are substantially more likely to stay regionally attached than Bachelor's-degree holders. English learner status remains a clean null throughout this entire sequence, consistent with every prior model in the completers-only population. The demographic effects that lost significance in Model 3 — economic status, Special Ed, American Indian identity, Gender, and Range school status — remain non-significant here as well, reinforcing that their earlier relationships with regional attachment ran through academic engagement and credential level, not through the characteristics themselves.

Model fit: AIC = 10,553.9 (down from 10,657.8 in Model 3); Δdeviance = 105.85 on 1 df, p \< .001; n = 8,500 in both models.

```{r}
model4_dv1b <- glm(
  grad.year.5 ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    range.school +
    avg.wages.pct.state +
    avg.unemp.rate +
    pseo.participant +
    took.ACT +
    ACTCompositeScore.filled +
    ap.exam + 
    sat.taken +
    cte.achievement +
    highest.cred.level,
  family = binomial,
  data = master.model.dv1b
)

nobs(model1_dv1b)
nobs(model2_dv1b)
nobs(model3_dv1b)
nobs(model4_dv1b)

AIC(model1_dv1b)
AIC(model2_dv1b)
AIC(model3_dv1b)
AIC(model4_dv1b)

summary(model4_dv1b)

exp(cbind(OR = coef(model4_dv1b), confint(model4_dv1b)))

anova(model1_dv1b, model2_dv1b, model3_dv1b, model4_dv1b, test = "Chisq")

or_plot <- tidy(model4_dv1b, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )


ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )


```

<br>

## Model 5: Full Specification — Demographics + High School Characteristics + High School Accomplishments + Highest Credential Earned + Graduation Location

Model 5: Full Specification — Add Postsecondary Graduation Location

This model adds ps.grad.location (graduated regionally vs. graduated elsewhere in Minnesota) to the full specification — the final model in this sequence (n = 8,500).

**Graduating from a postsecondary institution within the region is by far the single largest predictor in this entire analysis.** OR ≈ 3.22 (CI: 2.893-3.582, p \< .001) — graduates who completed their credential regionally show more than triple the odds of sustained regional workforce attachment compared to those who completed elsewhere in Minnesota. This dwarfs every other effect in the model, demographic or otherwise.

**Several accomplishment-block and credential effects shrink substantially once graduation location is added — a real, expected consequence, not a data problem.** Highest credential level's effect nearly halves (OR ≈ 1.40, down from ≈ 1.78 in Model 4, still p \< .001) — much of the "sub-baccalaureate credentials predict regional attachment" story turns out to run through *where* those credentials were earned, since associate-level and technical programs are disproportionately regional. Took-ACT's effect continues shrinking as well (OR ≈ 2.40, down from ≈ 3.25 in Model 4, still p \< .001), and SAT participation's negative effect weakens too (OR ≈ 0.61, up from ≈ 0.48, p = .013) — some of what these variables were capturing was proximity to a regional postsecondary option, now measured directly.

**Hispanic graduates become significant here — the first time in this entire completers-only sequence.** OR ≈ 0.55 (CI: 0.304-0.970, p = .044), crossing into significance from marginal status in every prior model. Black graduates remain significant (OR ≈ 0.51, p = .012). Asian/Pacific Islander graduates remain just short of significance (OR ≈ 0.61, p = .061). American Indian graduates remain clearly non-significant (OR ≈ 1.11, p = .410).

**Dem_Desc's Town/rural mix effect disappears once graduation location is controlled — a genuine reversal worth flagging plainly.** OR ≈ 1.18 (p = .185), down from significant in Models 2 through 4. Urban/town/rural mix remains significant (OR ≈ 1.53, p \< .001), but the Town/rural mix advantage appears to run largely through where its graduates attend postsecondary, not through district classification independently.

**Local wages loses significance entirely.** OR ≈ 1.43 (CI: 0.500-4.076, p = .504), down from significant in Model 4 (p = .041) — the apparent wage effect there appears to have been substantially explained by graduation location rather than reflecting an independent local-economy relationship.

**Range school status, Gender, economic status, Special Education status, and English learner status all remain non-significant**, consistent with every model from Model 3 onward. English learner status in particular remains a clean, stable null across this entire five-model sequence — the clearest contrast with the full-population model, where it was one of the most robust findings.

**CTE achievement, ACT composite score, and AP exam all remain significant**, each somewhat attenuated in magnitude from Model 4 but still clearly holding up in the same direction.

**Overall Interpretation**

Adding postsecondary graduation location produces the largest single improvement in model fit anywhere in this sequence (likelihood ratio test: Δdeviance = 487.14, df = 1, p \< .001; AIC decreases from 10,553.9 to 10,068.8). **Graduation location is confirmed as the dominant predictor of regional workforce attachment among completers** — larger than demographics, academic accomplishments, or credential level combined. Its addition substantially explains away the apparent effects of credential level, Town/rural mix district classification, and local wages, confirming those were partly proxies for where students completed their education rather than fully independent relationships. Black and, newly, Hispanic graduates are confirmed as facing significantly lower odds of regional attachment even after full control for where and what they studied. English learner status remains a robust null throughout the entire completers-only sequence — a fundamentally different finding from the full-population model, where it was among the strongest and most consequential predictors identified.

Model fit: AIC = 10,068.8 (down from 10,553.9 in Model 4); Δdeviance = 487.14 on 1 df, p \< .001; n = 8,500 across all five models.

```{r}
model5_dv1b <- glm(
  grad.year.5 ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    range.school +
    avg.wages.pct.state +
    avg.unemp.rate +
    pseo.participant +
    took.ACT +
    ACTCompositeScore.filled +
    ap.exam + 
    sat.taken +
    cte.achievement +
    highest.cred.level +
    ps.grad.location,
  family = binomial,
  data = master.model.dv1b
)

nobs(model1_dv1b)
nobs(model2_dv1b)
nobs(model3_dv1b)
nobs(model4_dv1b)
nobs(model5_dv1b)

AIC(model1_dv1b)
AIC(model2_dv1b)
AIC(model3_dv1b)
AIC(model4_dv1b)
AIC(model5_dv1b)

summary(model5_dv1b)

exp(cbind(OR = coef(model5_dv1b), confint(model5_dv1b)))

anova(model1_dv1b, model2_dv1b, model3_dv1b, model4_dv1b, model5_dv1b, test = "Chisq")

or_plot <- tidy(model5_dv1b, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )


ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )

vif(model5_dv1b)
```

<br>

## DV1b Summary: What Predicts Sustained Regional Workforce Attachment (College Completers Only)

Across all five models, this analysis identifies a fundamentally different pattern of predictors than the full-population model — the same outcome, but a genuinely distinct story once the population is restricted to students who completed a postsecondary credential.

**Postsecondary graduation location is the single largest predictor identified in this entire analysis, full population included.** Graduates who completed their credential within the region show more than triple the odds of sustained regional workforce attachment compared to those who completed elsewhere in Minnesota (OR ≈ 3.22, p \< .001) — an effect larger than any demographic, academic, or credential-level variable in either model sequence.

**English learner status shows no relationship with regional attachment at any stage of this sequence — a clean, stable null across all five models.** This stands in sharp contrast to the full-population model, where English learner status was one of the most robust and consequential findings identified. Among students who go on to complete a postsecondary credential, whatever drives reduced regional attachment for English learners in the broader population does not appear to operate.

**Highest credential level is a strong, independent predictor, though much of its apparent strength runs through where that credential was earned.** Sub-baccalaureate credential holders show substantially higher odds of regional attachment than Bachelor's-degree holders throughout (OR ≈ 1.78 before graduation location is considered, OR ≈ 1.40 once it is) — a real, independent effect that survives full specification, but one that is partly explained by the fact that associate-level and technical credentials are disproportionately regional.

**Race/Ethnicity findings differ meaningfully by specification, and one group's disparity only emerges in the fully specified model.** Black graduates show a consistent, significant disadvantage across every model (OR ≈ 0.47–0.57). Hispanic graduates remain marginal through four models before becoming significant only in the full specification (OR ≈ 0.55, p = .044) — their disadvantage among completers is real but only detectable once location, credential level, and full academic engagement are jointly accounted for. Asian/Pacific Islander graduates remain consistently at the edge of significance throughout (never quite crossing the threshold in the fully controlled models). American Indian graduates tell the most distinct story of any group: they show significantly *higher* odds of regional attachment in the demographics-only and characteristics models — the opposite direction from the full-population finding — before fading to non-significance once academic accomplishments enter the specification.

**Economic status, Special Education status, and Gender all show a positive relationship with regional attachment early in the sequence, then lose significance once academic accomplishments are considered.** All three are significant, protective predictors in the demographics-only and characteristics models, but none hold up independently once ACT, SAT, AP, and CTE engagement are in the model — their apparent advantage among completers appears to run through how these groups engage academically in high school, not through the characteristics themselves.

**The ACT crossover pattern replicates precisely, and mirrors the full- population finding exactly.** Taking the ACT is associated with substantially higher odds of regional attachment throughout the sequence, while among test-takers, each additional composite point predicts lower odds — the same threshold effect found in the full population: engagement itself is protective, while high-scoring engagement specifically predicts leaving the region. Both effects shrink somewhat once credential level and graduation location are added, as those variables absorb some of what ACT participation and score were capturing.

**SAT participation and AP exam participation both show a consistent negative relationship with regional attachment**, weakening somewhat by the full specification but remaining significant throughout — consistent with the full-population finding that these measures identify a subgroup oriented toward postsecondary and career options outside the region.

**CTE achievement continues to anchor students regionally across every model in this sequence**, holding significant and stable in both category levels from the point it enters the specification.

**PSEO participation shows no relationship with regional attachment at any stage** — consistent with the full-population model, where PSEO's strong bivariate influence likewise failed to hold up once other academic accomplishment variables were considered.

**District classification and local economic conditions are substantially explained by graduation location.** Urban/town/rural mix district classification remains significant throughout, but Town/rural mix loses significance once graduation location is added — its apparent advantage runs largely through where its graduates attend postsecondary. Local wages briefly becomes significant once credential level is added, then loses significance entirely once graduation location is controlled for — the wage effect was substantially a proxy for regional postsecondary attendance, not an independent local-economy relationship. Local unemployment rate and Range school status remain non-significant from the point academic accomplishments enter the model onward.

<br>

# Prep DV2 Data

This dataset is going to be the entire master set not filtered by post-secondary completion. This analysis attempts to find relationships with whether someone completes college or not.

```{r}
model_vars_dv2<- c("Gender", "RaceEthnicity", "economic.status", "SpecialEdStatus",
                     "english.learner", "Dem_Desc", "range.school", "avg.wages.pct.state", "avg.unemp.rate",
                     "pseo.participant", "took.ACT", "ACTCompositeScore", "ap.exam", "sat.taken",
                     "cte.achievement", "ps.grad")

master.model.dv2 <- master %>%
  filter(grad.year > 2010,
         grad.year <= 2019,
         ps.grad != "Attending ps") %>%
  select(PersonID, all_of(model_vars_dv2)) %>%
  filter(!is.na(pseo.participant)) %>%
  mutate(ps.grad = ifelse(ps.grad == "Y", "Yes", "No"),
         ps.grad = fct_relevel(ps.grad, "No"),
         Gender = fct_relevel(Gender, "M"),
         RaceEthnicity = fct_relevel(RaceEthnicity, "White"),
         economic.status = as_factor(economic.status),
         economic.status = fct_relevel(economic.status, "0"),
         SpecialEdStatus = as.character(SpecialEdStatus),
         SpecialEdStatus = fct_relevel(SpecialEdStatus, "0"),
         english.learner = as_factor(english.learner),
         english.learner = fct_relevel(english.learner, "0"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural"),
         range.school = fct_relevel(range.school, "Non-range school"),
         pseo.participant = as.character(pseo.participant),
         pseo.participant = fct_relevel(pseo.participant, "0"),
         took.ACT = fct_relevel(took.ACT, "No"),
         ACTCompositeScore.filled = ifelse(is.na(ACTCompositeScore), 0, ACTCompositeScore),
         sat.taken = as_factor(sat.taken),
         sat.taken = fct_relevel(sat.taken, "0"),
         ap.exam = as_factor(ap.exam),
         ap.exam = fct_relevel(ap.exam, "0"),
         cte.achievement = fct_relevel(cte.achievement, "No CTE"))


```

# Modeling - college completion, full population

## Model 1: Demographics

This model estimates the association between demographic characteristics and postsecondary completion, modeled across all high school graduates regardless of whether they ever attended postsecondary education (n = 26,327). The outcome is coded 1 = completed a postsecondary credential, 0 = did not.

**Economic status and Special Education status show the largest effects of any demographic predictor examined in this project so far.**

- **Economic disadvantage:** approximately 64% lower odds of postsecondary completion relative to non-disadvantaged students (OR ≈ 0.36).
- **Special Education status:** approximately 73% lower odds of completion (OR ≈ 0.27).

Both are dramatically larger effects than anything seen in the DV1a/DV1b sequences, where these same variables showed effects in the 10–40% range. This makes sense given what the variable is measuring — completion itself is a much more direct, proximate outcome of the barriers these statuses represent than downstream regional employment is.

**Race and ethnicity shows substantial disparities across every category.**

- **American Indian graduates:** approximately 44% lower odds of completion relative to White graduates (OR ≈ 0.56).
- **Black graduates:** approximately 48% lower odds (OR ≈ 0.52).
- **Hispanic graduates:** approximately 43% lower odds (OR ≈ 0.57).
- **Asian/Pacific Islander graduates:** not statistically significant (OR ≈ 0.86, CI: 0.682–1.072, crosses 1) — the only Race/Ethnicity category without a significant gap in this model.
- **Unknown race/ethnicity:** approximately 84% lower odds (OR ≈ 0.16) — given the very small residual size of this category (n ≈ 40) and the fact that "Unknown" doesn't correspond to any real demographic group, this coefficient should not be substantively interpreted; it is retained in the model for completeness but carries no meaningful narrative value.

**English learner status is associated with substantially lower odds of completion** (OR ≈ 0.66) — a much larger effect here than in any DV1a/DV1b model, consistent with postsecondary completion being a more proximate outcome for this predictor as well.

**Gender shows a notably different pattern than in any regional-employment model.**

- **Female graduates:** approximately 74% *higher* odds of postsecondary completion relative to male graduates (OR ≈ 1.74) — a clear reversal from DV1a/DV1b, where female graduates showed modestly *lower* odds of regional workforce attachment. This is a substantively important finding: female graduates complete college at meaningfully higher rates, even though (per DV1a) they are somewhat less likely to remain regionally employed afterward.

**Overall Interpretation**

At the demographics-only stage, economic status and Special Education status emerge as the two strongest predictors of postsecondary completion examined anywhere in this project, far exceeding their effect sizes in the regional-employment models. Race/Ethnicity gaps are present across nearly every category. The gender reversal — women completing college at notably higher rates while showing modestly lower regional workforce attachment — is a pattern worth carrying forward as high school characteristics and accomplishments are added, since it suggests completion and regional attachment may be driven by partially distinct mechanisms for this demographic factor specifically.

This is a baseline model only — it does not yet account for high school context or academic preparation, both added in subsequent blocks.

**Model fit:** AIC = 32,334.4; n = 26,327.

```{r}
model1_dv2 <- glm(
  ps.grad ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner,
    family = binomial,
  data = master.model.dv2
)

nobs(model1_dv2)
AIC(model1_dv2)

summary(model1_dv2)

exp(cbind(
  OR = coef(model1_dv2),
  confint(model1_dv2)
))

or_plot <- tidy(model1_dv2, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )


ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )

```

<br>

## Model 2: Demographics + High School Characteristics

This model adds district rural/urban classification, Range school status, local average wages, and local average unemployment rate to the demographic baseline (n = 26,327).

**Range school status shows a strong, significant positive relationship with postsecondary completion — a genuinely different pattern than seen in the workforce-attachment models.** Range graduates show 34% higher odds of completing a postsecondary credential than non-Range graduates (OR ≈ 1.34, CI: 1.254-1.438, p \< .001). In the regional-attachment sequences, Range status was about staying local; here, it's associated with a genuinely different outcome — completing postsecondary education at all, regardless of where.

**Local wages and unemployment rate both become significant here, and in directions that tell a coherent "opportunity cost" story.** Higher local wages relative to the state average are associated with *lower* odds of postsecondary completion (OR ≈ 0.53, CI: 0.299-0.930, p = .027) — where local jobs pay comparatively well, the incentive to pursue further education appears to weaken. Higher local unemployment is associated with *higher* odds of completion (OR ≈ 1.03 per unit, CI: 1.005-1.048, p = .015) — where local job prospects are weaker, students appear more likely to pursue a credential instead. This is the first model in this entire project where both local economic indicators show a real, intuitive, significant relationship simultaneously.

**District classification shows only a marginal relationship here.** Urban/town/rural mix districts show higher odds of completion that just misses conventional significance (OR ≈ 1.13, p = .053), and Town/rural mix is clearly non-significant (OR ≈ 1.10, p = .129) — a much weaker relationship than Dem_Desc showed for regional workforce attachment.

**All five demographic effects from Model 1 remain essentially unchanged in direction and magnitude.** Economic status (OR ≈ 0.36), Special Education status (OR ≈ 0.26), English learner status (OR ≈ 0.67), and the Race/Ethnicity pattern (American Indian, Black, and Hispanic all significantly lower than White; Asian/Pacific Islander non-significant) all hold steady from Model 1 — none of the demographic disparities in postsecondary completion are explained by district context, Range school status, or local economic conditions.

**Overall Interpretation**

Adding high school characteristics significantly improves model fit (likelihood ratio test: Δdeviance = 81.68, df = 5, p \< .001; AIC decreases from 32,334.4 to 32,262.7). **Range school status and local economic conditions are the standout findings in this block** — Range status is associated with substantially higher odds of completing a credential at all (a different relationship than its "stay local" role in the workforce- attachment models), and local wages and unemployment both show coherent, significant relationships with completion for the first time in this project, consistent with an opportunity-cost interpretation. District classification, by contrast, shows only a weak, largely non-significant relationship with completion — geography matters far less here than it does for where graduates end up working.

Model fit: AIC = 32,262.7 (down from 32,334.4 in Model 1); Δdeviance = 81.68 on 5 df, p \< .001; n = 26,327 in both models.

```{r}
model2_dv2 <- glm(
  ps.grad ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    range.school +
    avg.wages.pct.state +
    avg.unemp.rate,
  family = binomial,
  data = master.model.dv2
)

nobs(model1_dv2)
nobs(model2_dv2)

AIC(model1_dv2)
AIC(model2_dv2)

summary(model2_dv2)

exp(cbind(
  OR = coef(model2_dv2),
  confint(model2_dv2)
))

anova(model1_dv2, model2_dv2, test = "Chisq")


or_plot <- tidy(model2_dv2, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )


ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )

```

<br>

## Model 3: Demographics + High School Characteristics + High School Accomplishments

This model adds PSEO participation, ACT participation and score, SAT participation, AP exam participation, and CTE achievement to the growing specification (n = 26,327).

**English learner status loses significance entirely — a clean, real null, not an artifact of any missing-data control.** OR ≈ 0.90 (CI: 0.749-1.080, p = .264), down from significant in Models 1 and 2. Unlike the regional-attachment models, where English learner status held up as one of the most robust findings, its relationship with overall postsecondary completion appears to run substantially through academic engagement and preparation — this is a genuinely different mechanism for a genuinely different outcome, not a repeat of any prior masking concern, since no missingness-based control is anywhere in this specification.

**PSEO participation shows the strongest accomplishment-block effect in this model, and it makes clear mechanical sense.** OR ≈ 1.65 (CI: 1.557-1.757, p \< .001) — PSEO participants show 65% higher odds of completing a postsecondary credential, unsurprising given PSEO directly provides college credit while still in high school.

**ACT shows a crossover pattern, but in the opposite configuration from the regional-attachment models — a genuinely different, DV-specific pattern worth stating on its own terms.** Simply taking the ACT is associated with *lower* odds of completion (OR ≈ 0.56, p \< .001), while among test-takers, each additional composite point is associated with *higher* odds (OR ≈ 1.08 per point, p \< .001). Read together: students who take the ACT but don't perform strongly on it appear less likely to complete a credential, while strong performers complete at higher rates — a different mechanism than the "engagement is protective, high scores predict leaving" pattern found for regional attachment.

**SAT, AP exam, and CTE Concentrator/Completer status are all non- significant here** — a sharp contrast with the regional-attachment models, where these were some of the strongest predictors. SAT (OR ≈ 1.04, p = .715) and AP exam (OR ≈ 1.04, p = .434) show no relationship with overall completion, and CTE Concentrator/Completer status is likewise non-significant (OR ≈ 1.04, p = .270). CTE Participant status shows a small but significant positive effect (OR ≈ 1.09, p = .010).

**Asian/Pacific Islander graduates become significant here for the first time in this sequence.** OR ≈ 0.77 (CI: 0.605-0.987, p = .039) — this group showed no significant relationship with completion in Models 1 or 2, but a real, if modest, disparity emerges once academic accomplishments are accounted for.

**Economic status, Special Education status, and the other Race/Ethnicity disparities all remain significant, though substantially attenuated.** Economic status (OR ≈ 0.48, down from ≈ 0.36) and Special Ed status (OR ≈ 0.48, down from ≈ 0.26) both weaken considerably but remain among the strongest predictors in the model — academic accomplishments explain a meaningful share of these disparities, but far from all of it. American Indian (OR ≈ 0.69), Black (OR ≈ 0.66), and Hispanic graduates (OR ≈ 0.68) all remain significantly less likely to complete a credential, each somewhat attenuated from Models 1-2 but clearly independent of academic engagement.

**Local wages and unemployment rate both strengthen further**, continuing the pattern from Model 2 (wages: OR ≈ 0.46, p = .012; unemployment: OR ≈ 1.12 per unit, p \< .001) — the opportunity-cost story continues to hold up as academic accomplishments are added.

**Range school status and Gender both remain strong and significant**, Range school strengthening further (OR ≈ 1.56, up from ≈ 1.34) and Gender attenuating somewhat but remaining large (OR ≈ 1.58, down from ≈ 1.75).

**Overall Interpretation**

Adding high school accomplishments produces the largest improvement in model fit of any block in this sequence (likelihood ratio test: Δdeviance = 2,491.21, df = 7, p \< .001; AIC decreases from 32,262.7 to 29,785.5) — substantially larger than the corresponding block in either regional-attachment model. **English learner status is the standout finding here: it emerges as a clean, genuine null once academic accomplishments are considered — a fundamentally different result from its robust, independent role in the regional-attachment models.** PSEO participation is confirmed as the single strongest accomplishment predictor of completion, mechanically consistent with what PSEO actually provides. The ACT crossover reappears but in a different configuration than seen elsewhere in this project, and SAT/AP/CTE-Concentrator status — all strong predictors of regional attachment — show little to no independent relationship with completion itself. Economic status and Special Ed status remain the two largest demographic predictors in the entire model, substantially attenuated but far from explained away.

Model fit: AIC = 29,785.5 (down from 32,262.7 in Model 2); Δdeviance = 2,491.21 on 7 df, p \< .001; n = 26,327 in both models.

```{r}
model3_dv2 <- glm(
  ps.grad ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    range.school +
    avg.wages.pct.state +
    avg.unemp.rate +
    pseo.participant +
    took.ACT +
    ACTCompositeScore.filled +
    ap.exam + 
    sat.taken +
    cte.achievement,
    family = binomial,
  data = master.model.dv2
)

nobs(model1_dv2)
nobs(model2_dv2)
nobs(model3_dv2)

AIC(model1_dv2)
AIC(model2_dv2)
AIC(model3_dv2)

summary(model3_dv2)

exp(cbind(
  OR = coef(model3_dv2),
  confint(model3_dv2)
))

anova(model1_dv2, model2_dv2, model3_dv2, test = "Chisq")


or_plot <- tidy(model2_dv2, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )


ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )

```

<br>

## DV2 Summary: What Predicts Postsecondary Completion (All High School Graduates)

Across all three models, this analysis identifies a distinct pattern of predictors for postsecondary completion — related to, but genuinely different from, what predicts regional workforce attachment.

**English learner status is a clean, genuine null once academic accomplishments are considered — the standout finding of this sequence.** Significant in the demographics-only and characteristics models (OR ≈ 0.66–0.67), English learner status loses significance entirely once PSEO, ACT, SAT, AP, and CTE engagement are added (OR ≈ 0.90, p = .264). This is a fundamentally different result from the regional-attachment models, where English learner status held up as one of the most robust, independent findings in the full-population sequence. For postsecondary completion specifically, whatever reduces English learners' odds appears to run through academic engagement and preparation, not through English learner status independently.

**Economic status and Special Education status are the two largest demographic predictors identified anywhere in this project, and remain substantial even after significant attenuation.** Both show extremely large effects in the demographics-only model (economic status OR ≈ 0.36; Special Ed OR ≈ 0.27) that weaken considerably once academic accomplishments are added (OR ≈ 0.48 for both) — academic engagement explains a meaningful share of these disparities, but far from all of it. Both remain among the strongest predictors in the fully specified model.

**Race/Ethnicity disparities are confirmed as independent and robust, with one group's disparity only emerging once academic accomplishments are considered.** American Indian (OR ≈ 0.69), Black (OR ≈ 0.66), and Hispanic graduates (OR ≈ 0.68) all remain significantly less likely to complete a postsecondary credential across the full model sequence, each somewhat attenuated from the demographics-only model but clearly independent of academic engagement. Asian/Pacific Islander graduates show no significant relationship in the demographics-only or characteristics models, but become significant once accomplishments are added (OR ≈ 0.77, p = .039) — a disparity only detectable in the fully specified model.

**PSEO participation is the strongest single accomplishment predictor of completion, and the relationship is directly mechanical.** OR ≈ 1.65 — PSEO participants show 65% higher odds of completing a postsecondary credential, consistent with PSEO providing actual college credit while students are still in high school.

**ACT shows a crossover pattern, but in a different configuration than found in the regional-attachment models.** Simply taking the ACT is associated with *lower* odds of completion (OR ≈ 0.56), while among test-takers, each additional composite point predicts *higher* odds (OR ≈ 1.08 per point) — students who engage with the ACT but don't score well appear less likely to complete a credential, while strong performers complete at higher rates. This is a genuinely different mechanism from the regional-attachment pattern, where ACT engagement itself was protective and high scores predicted leaving the region.

**SAT participation, AP exam participation, and CTE Concentrator/ Completer status all show little to no independent relationship with completion** — a sharp contrast with the regional-attachment models, where these were some of the strongest predictors. CTE Participant status shows a small but significant positive effect (OR ≈ 1.09).

**Range school status is a strong, independent, and strengthening predictor of completion — a genuinely different role than it plays for regional attachment.** OR grows from ≈ 1.34 (Model 2) to ≈ 1.56 (Model 3) — in the regional-attachment sequences, Range status was about staying local; here, it's associated with completing a credential at all, regardless of location.

**Local wages and local unemployment rate both show a coherent opportunity-cost relationship with completion, strengthening as accomplishments are added.** Higher local wages relative to the state average predict lower odds of completion (OR ≈ 0.46 in the final model) — where local jobs pay well, the incentive to pursue further education appears to weaken. Higher local unemployment predicts higher odds of completion (OR ≈ 1.12 per unit) — where local job prospects are weaker, students appear more likely to pursue a credential instead.

**Gender remains a large, stable, positive predictor throughout**, attenuating somewhat once accomplishments are added (OR ≈ 1.58 in the final model, down from ≈ 1.75) but remaining one of the strongest effects in the entire specification.

**District classification (Dem_Desc) shows only a weak, largely non- significant relationship with completion** — a notably different pattern than seen for regional workforce attachment, where district context was one of the clearer, more consistent findings.

<br>

# Prep DV3 Data

Now we are focusing just on college completers. Since we saw that where someone graduated from college was significantly associated with workforce participation outcomes, this will explore what characteristics are related to that dependent variable.

```{r}
model_vars_dv3<- c("Gender", "RaceEthnicity", "economic.status", "SpecialEdStatus",
                   "english.learner", "Dem_Desc", "range.school", "avg.wages.pct.state", "avg.unemp.rate",
                   "pseo.participant", "took.ACT", "ACTCompositeScore", "ap.exam", "sat.taken",
                   "cte.achievement", "ps.grad.location")

master.model.dv3 <- master %>%
  filter(grad.year > 2010,
         grad.year <= 2019,
         ps.grad == "Y",
         ps.grad.location != "Did not graduate ps",
         RaceEthnicity != "Unknown") %>%
  select(PersonID, all_of(model_vars_dv3)) %>%
  filter(!is.na(pseo.participant)) %>%
  mutate(ps.grad.location = ifelse(ps.grad.location == "PS grad Region", "PS grad Region", "PS grad outside region"),
         ps.grad.location = fct_relevel(ps.grad.location, "PS grad outside region"),
    Gender = fct_relevel(Gender, "M"),
         RaceEthnicity = fct_relevel(RaceEthnicity, "White"),
         economic.status = as_factor(economic.status),
         economic.status = fct_relevel(economic.status, "0"),
         SpecialEdStatus = as.character(SpecialEdStatus),
         SpecialEdStatus = fct_relevel(SpecialEdStatus, "0"),
         english.learner = as_factor(english.learner),
         english.learner = fct_relevel(english.learner, "0"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural"),
    range.school = fct_relevel(range.school, "Non-range school"),
    
         pseo.participant = as.character(pseo.participant),
         pseo.participant = fct_relevel(pseo.participant, "0"),
         took.ACT = fct_relevel(took.ACT, "No"),
         ACTCompositeScore.filled = ifelse(is.na(ACTCompositeScore), 0, ACTCompositeScore),
    sat.taken = as_factor(sat.taken),
    sat.taken = fct_relevel(sat.taken, "0"),
    ap.exam = as_factor(ap.exam),
         ap.exam = fct_relevel(ap.exam, "0"),
         cte.achievement = fct_relevel(cte.achievement, "No CTE"))

```

# Modeling: College location - college completers only

## Model 1: Demographics

This model estimates the association between demographic characteristics and completing a postsecondary credential within the region, restricted to postsecondary completers (n = 12,606). The outcome is coded 1 = completed within the region, 0 = completed elsewhere in Minnesota or out of state. A small number of students with unrecorded ("Unknown") race/ethnicity (n = 6) were excluded due to quasi-complete separation caused by the extremely small cell size.

**Economic status and Special Education status show the strongest effects in this model, and in a notably different direction than in DV2.**

- **Economic disadvantage:** approximately 65% higher odds of completing within the region relative to non-disadvantaged completers (OR ≈ 1.65).
- **Special Education status:** approximately 89% higher odds of completing within the region (OR ≈ 1.89).

Both effects are large, highly significant, and point toward economically disadvantaged and Special Ed completers being considerably more likely to stay in the region for their postsecondary education — plausibly reflecting cost, proximity, or family-tie considerations that favor regional institutions. This is directionally consistent with the reversed demographic pattern found in DV1b's early models (before high school accomplishments attenuated it), suggesting this may be a genuine, recurring feature of how these groups engage with postsecondary geography, not a one-off finding.

**Race/Ethnicity shows a mixed pattern, with only two categories reaching significance.**

- **American Indian graduates:** approximately 37% higher odds of completing within the region relative to White graduates (OR ≈ 1.37).
- **Asian/Pacific Islander graduates:** approximately 34% lower odds (OR ≈ 0.66).
- **Black and Hispanic graduates:** not statistically significant (CIs cross 1), likely reflecting reduced power given their smaller cell sizes in this completers-only population (n = 136 and n = 105, respectively) rather than a definitive absence of a relationship.

**Gender and English learner status show modest or no effects.**

- **Female graduates:** approximately 14% lower odds of completing within the region relative to male graduates (OR ≈ 0.86).
- **English learner status:** no significant effect (OR ≈ 1.00, CI: 0.709–1.427).

**Overall Interpretation**

At the demographics-only stage, economic disadvantage and Special Education status stand out as the strongest predictors of completing postsecondary education within the region — both showing large, positive effects. This is a meaningfully different story than DV2 (postsecondary completion overall), where these same two variables showed the *opposite* direction (substantially lower odds of completing at all). Together, this suggests a nuanced picture: economically disadvantaged and Special Ed students face real barriers to completing postsecondary education in the first place, but among those who do complete, they are considerably more likely to do so within the region — a pattern worth carrying forward as high school characteristics and accomplishments are added, to see whether it holds up or is explained by other factors (similar to what happened with the analogous DV1b reversal).

**Model fit:** AIC = 16,994.1; n = 12,606.

```{r}
model1_dv3 <- glm(
  ps.grad.location ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner,
  family = binomial,
  data = master.model.dv3
)

summary(model1_dv3)
nobs(model1_dv3)
AIC(model1_dv3)
exp(cbind(OR = coef(model1_dv3), confint(model1_dv3)))

or_plot <- tidy(model1_dv3, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )


ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )
```

<br>

## Model 2: Demographics + High School Characteristics

This model adds district rural/urban classification, Range school status, local average wages, and local average unemployment rate to the demographic baseline, restricted to postsecondary completers (n = 12,606).

**District classification shows by far the largest effects found anywhere in this project.** Graduates from Town/rural mix districts show more than double the odds of completing postsecondary education within the region compared to Entirely rural districts (OR ≈ 2.06, p \< .001), and Urban/town/rural mix districts show an even larger effect (OR ≈ 2.36, p \< .001). This makes strong practical sense for this specific outcome — where a student's high school district sits relative to regional postsecondary institutions plausibly shapes where they end up attending far more directly than it shapes broader employment outcomes.

**Range school status is also a strong, independent predictor.** Range graduates show 66% higher odds of completing within the region (OR ≈ 1.66, p \< .001) — consistent with Range communities' documented ties to specific regional institutions and programs.

**Local wages and unemployment rate are both significant here, in the same direction found for DV2 (postsecondary completion overall) — an opportunity-cost-adjacent story, though the wage direction differs from what might be expected.** Higher local wages predict *higher* odds of completing within the region (OR ≈ 2.56, p = .018) — unlike DV2, where higher wages predicted *lower* odds of completing at all. Once a student has decided to pursue postsecondary education, stronger local wages may make staying regional more attractive rather than acting as a disincentive to enroll in the first place. Higher unemployment also predicts higher odds of regional completion (OR ≈ 1.07 per unit, p \< .001), consistent with the same opportunity-cost logic seen in DV2.

**All five demographic effects remain essentially unchanged from Model 1.** Economic status (OR ≈ 1.68), Special Education status (OR ≈ 1.87), and American Indian graduates (OR ≈ 1.45) all remain significant and stable, confirming these groups are more likely to complete regionally once they complete at all. Asian/Pacific Islander graduates remain significantly less likely to complete regionally (OR ≈ 0.66). Black, Hispanic, and English learner status all remain non-significant, unchanged from Model 1.

**Overall Interpretation**

Adding high school characteristics substantially improves model fit (likelihood ratio test: Δdeviance = 316.98, df = 5, p \< .001; AIC decreases from 16,994.1 to 16,687.1) — the largest improvement from this block seen in any model sequence in this project. **District classification is the standout finding, with effects roughly double the size seen for regional workforce attachment** — geography's relationship with *where* a student completes their credential is substantially stronger than its relationship with where they end up working afterward. Range school status and local economic conditions both confirm as genuine, independent predictors, and none of the demographic effects from Model 1 are explained away by adding these variables.

Model fit: AIC = 16,687.1 (down from 16,994.1 in Model 1); Δdeviance = 316.98 on 5 df, p \< .001; n = 12,606 in both models.

```{r}
model2_dv3 <- glm(
  ps.grad.location ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    range.school +
    avg.wages.pct.state +
    avg.unemp.rate,
  family = binomial,
  data = master.model.dv3
)

summary(model2_dv3)

nobs(model1_dv3)
nobs(model2_dv3)

AIC(model1_dv3)
AIC(model2_dv3)

exp(cbind(OR = coef(model2_dv3), confint(model2_dv3)))

anova(model1_dv3, model2_dv3, test = "Chisq")


or_plot <- tidy(model2_dv3, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )


ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )
```

<br>

## Model 3: Demographics + High School Characteristics + High School Accomplishments

This model adds PSEO participation, ACT participation and score, SAT participation, AP exam participation, and CTE achievement to the growing specification, restricted to postsecondary completers (n = 12,606).

**Took-ACT shows the single largest effect found anywhere in this entire project.** OR ≈ 6.58 (CI: 5.111-8.483, p \< .001) — students who took the ACT show more than six times the odds of completing their postsecondary credential within the region, compared to non-test-takers. Among test-takers, the same crossover pattern seen elsewhere reappears: each additional composite point predicts *lower* odds of regional completion (OR ≈ 0.895 per point, p \< .001) — engagement itself is strongly protective for staying regional, while high performance specifically predicts completing elsewhere.

**Local wages strengthens to an extraordinarily large effect.** OR ≈ 7.95 (CI: 3.435-18.405, p \< .001), up sharply from ≈ 2.56 in Model 2. This is a very wide interval alongside a very large point estimate — worth treating as a genuine, strong signal given how far the lower bound sits above 1, but also worth flagging the imprecision rather than leaning too hard on the exact magnitude.

**Special Education status loses significance entirely — a real, substantive change from Models 1 and 2.** OR ≈ 1.19 (CI: 0.964-1.476, p = .109), down from clearly significant (OR ≈ 1.87–1.89) in the prior two models. Special Ed completers' apparent tendency to complete regionally appears to run substantially through their patterns of academic engagement — likely lower ACT/AP participation and stronger CTE involvement — rather than Special Ed status operating independently.

**Race/Ethnicity shows real movement in both directions.** Black graduates become significant here for the first time in this sequence (OR ≈ 0.68, p = .035) — a real, if newly emergent, disparity. Asian/Pacific Islander graduates, by contrast, lose the significance they showed in Model 2 (OR ≈ 0.94, p = .711) — their apparent disadvantage in regional completion appears to run through academic engagement rather than being independent of it. American Indian graduates remain significant but weaken somewhat (OR ≈ 1.25, p = .043). Hispanic graduates remain non-significant.

**Economic status weakens but remains solidly significant** (OR ≈ 1.38, down from ≈ 1.68 in Model 2, still p \< .001) — academic engagement explains some, but far from all, of this relationship.

**SAT and AP exam participation both show a strong negative relationship with regional completion, consistent with their role throughout this project.** SAT participation OR ≈ 0.39 (p \< .001), AP exam OR ≈ 0.72 (p \< .001) — both identify students oriented toward options outside the region, mirroring their role in the workforce-attachment models.

**CTE Concentrator/Completer status remains significant and positive** (OR ≈ 1.24, p \< .001), while CTE Participant status is not (OR ≈ 1.03, p = .528) — only the deeper level of CTE engagement shows an independent relationship with regional completion once other accomplishments are considered.

**PSEO participation is non-significant** (OR ≈ 0.94, p = .167) — consistent with the pattern found in every other model sequence in this project, PSEO's apparent influence doesn't hold up once ACT, SAT, AP, and CTE are jointly considered.

**Gender weakens to only marginal significance** (OR ≈ 0.92, p = .048), right at the conventional threshold — worth treating as a fragile finding at this stage rather than a confirmed one.

**District classification and Range school status both remain strong and significant**, essentially unchanged in direction from Model 2 (Dem_Desc even strengthens slightly), and average unemployment rate remains significant as well (OR ≈ 1.05 per unit, p = .003).

**English learner status remains non-significant throughout this entire sequence** (OR ≈ 0.77, p = .149), consistent with Models 1 and 2 — no relationship with regional completion at any stage.

**Overall Interpretation**

Adding high school accomplishments produces a very large improvement in model fit (likelihood ratio test: Δdeviance = 1,033.23, df = 7, p \< .001; AIC decreases from 16,687.1 to 15,667.9). **Took-ACT's effect here (OR ≈ 6.58) is the largest coefficient found anywhere in this project** — ACT engagement is an extraordinarily strong predictor of completing postsecondary education regionally rather than elsewhere. Special Education status and Asian/Pacific Islander status both lose significance once academic accomplishments are considered, while Black graduates newly emerge as significant — a genuine reshuffling of which demographic disparities are independent versus explained by academic engagement. Economic status, American Indian identity, district classification, and Range school status all remain significant, confirming real, independent relationships with where completers finish their credential.

Model fit: AIC = 15,667.9 (down from 16,687.1 in Model 2); Δdeviance = 1,033.23 on 7 df, p \< .001; n = 12,606 in both models.

```{r}
model3_dv3 <- glm(
  ps.grad.location ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    range.school +
    avg.wages.pct.state +
    avg.unemp.rate +
    pseo.participant +
    took.ACT +
    ACTCompositeScore.filled +
    ap.exam + 
    sat.taken +
    cte.achievement,
  family = binomial,
  data = master.model.dv3
)

summary(model3_dv3)

nobs(model1_dv3)
nobs(model2_dv3)
nobs(model3_dv3)

AIC(model1_dv3)
AIC(model2_dv3)
AIC(model3_dv3)

exp(cbind(OR = coef(model3_dv3), confint(model3_dv3)))

anova(model1_dv3, model2_dv3, model3_dv3, test = "Chisq")


or_plot <- tidy(model3_dv3, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )


ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )
```

<br>

## DV3 Summary: What Predicts Postsecondary Completion Within the Region (College Completers Only)

Across all three models, this analysis identifies district geography and ACT engagement as the two dominant forces shaping where postsecondary completers finish their credential, alongside several demographic patterns that shift meaningfully once academic accomplishments are considered.

**Took-ACT produces the largest single coefficient found anywhere in this project.** Students who took the ACT show more than six times the odds of completing their credential within the region compared to non-test-takers (OR ≈ 6.58, p \< .001). Among test-takers, the familiar crossover reappears: each additional composite point predicts lower odds of regional completion (OR ≈ 0.895 per point, p \< .001) — simply engaging with the ACT is strongly protective for staying regional, while strong performance specifically predicts completing elsewhere.

**District classification shows the largest geographic effects found anywhere in this project, and they strengthen rather than attenuate as other variables are added.** Graduates from Urban/town/rural mix districts show more than double the odds of regional completion relative to Entirely rural districts throughout the sequence (OR ≈ 2.36–2.52), and Town/rural mix districts show a similarly strong, stable effect (OR ≈ 1.93–2.06). Range school status adds a further, independent effect on top of district classification (OR ≈ 1.36–1.66 across models) — geography matters substantially more for *where* a completer finishes their credential than it does for workforce outcomes afterward.

**Special Education status loses significance entirely once academic accomplishments are considered — a real reversal, not a data artifact.** Significant and substantial in the demographics-only and characteristics models (OR ≈ 1.87–1.89), Special Ed status falls to non-significance in the final model (OR ≈ 1.19, p = .109) — its apparent tendency toward regional completion appears to run through academic engagement patterns rather than Special Ed status operating independently.

**Economic status remains significant throughout, though substantially attenuated.** OR weakens from ≈ 1.68 (Model 2) to ≈ 1.38 (Model 3) — academic engagement explains some, but far from all, of economically disadvantaged completers' tendency to complete regionally.

**Race/Ethnicity findings shift meaningfully across the sequence, with no single stable pattern.** American Indian graduates remain significant across all three models (OR ≈ 1.25–1.45), showing a consistent tendency toward regional completion. Asian/Pacific Islander graduates are significant in the demographics-only and characteristics models (OR ≈ 0.66) but lose significance once academic accomplishments are added (OR ≈ 0.94) — their apparent disadvantage runs through academic engagement rather than being independent of it. Black graduates show the opposite trajectory: non-significant in the first two models, then significant in the final model (OR ≈ 0.68, p = .035) — a disparity only detectable once academic accomplishments are fully accounted for. Hispanic graduates and English learner status remain non-significant throughout the entire sequence.

**SAT and AP exam participation both show a strong, stable negative relationship with regional completion**, consistent with their role throughout this project — both identify completers oriented toward postsecondary options outside the region (SAT: OR ≈ 0.39; AP: OR ≈ 0.72 in the final model).

**CTE Concentrator/Completer status shows an independent, positive relationship with regional completion, while CTE Participant status does not** — only the deeper level of CTE engagement holds up once other accomplishment variables are considered.

**PSEO participation shows no relationship with regional completion at any stage** — consistent with the pattern found in every other model sequence in this project.

**Local wages and unemployment rate both show a positive, strengthening relationship with regional completion.** Local wages grows to an extremely large effect in the final model (OR ≈ 7.95, though with a wide confidence interval reflecting real imprecision), and unemployment rate remains significant and stable (OR ≈ 1.05–1.07 per unit) — both consistent with the opportunity-cost pattern found for postsecondary completion overall, though here the relationship concerns *where* students complete rather than *whether* they do.

**Gender shows a weakening relationship across the sequence**, moving from clearly significant in Models 1–2 (OR ≈ 0.86–0.87) to only marginally significant in the final model (OR ≈ 0.92, p = .048) — worth treating as a fragile finding at the fully specified stage rather than a confirmed one.

<br>

# Prep DV4 Data

Another post-secondary independent variable that was strongly associated with workforce participation outcomes was the highest credential earned by an individual. This analysis will explore what characteristics are associated with earned credentials.

```{r}
model_vars_dv4<- c("Gender", "RaceEthnicity", "economic.status", "SpecialEdStatus",
                   "english.learner", "Dem_Desc", "range.school", "avg.wages.pct.state", "avg.unemp.rate",
                   "pseo.participant", "took.ACT", "ACTCompositeScore", "ap.exam", "sat.taken",
                   "cte.achievement","highest.cred.level")

master.model.dv4 <- master %>%
  filter(grad.year > 2010,
         grad.year <= 2019,
         ps.grad == "Y",
         highest.cred.level != "Did not graduate ps",
         RaceEthnicity != "Unknown") %>%
  select(PersonID, all_of(model_vars_dv4)) %>%
  filter(!is.na(pseo.participant)) %>%
  mutate(highest.cred.level = ifelse(highest.cred.level %in% c("Associate degree", "Less than Associate Degree"), "Associate or lesser credential", highest.cred.level),
         highest.cred.level = fct_relevel(highest.cred.level, "Bachelor degree or higher"),
         Gender = fct_relevel(Gender, "M"),
         RaceEthnicity = fct_relevel(RaceEthnicity, "White"),
         economic.status = as_factor(economic.status),
         economic.status = fct_relevel(economic.status, "0"),
         SpecialEdStatus = as.character(SpecialEdStatus),
         SpecialEdStatus = fct_relevel(SpecialEdStatus, "0"),
         english.learner = as_factor(english.learner),
         english.learner = fct_relevel(english.learner, "0"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural"),
         range.school = fct_relevel(range.school, "Non-range school"),
         pseo.participant = as.character(pseo.participant),
         pseo.participant = fct_relevel(pseo.participant, "0"),
         took.ACT = fct_relevel(took.ACT, "No"),
         ACTCompositeScore.filled = ifelse(is.na(ACTCompositeScore), 0, ACTCompositeScore),
         sat.taken = as_factor(sat.taken),
         sat.taken = fct_relevel(sat.taken, "0"),
         ap.exam = as_factor(ap.exam),
         ap.exam = fct_relevel(ap.exam, "0"),
         cte.achievement = fct_relevel(cte.achievement, "No CTE"))

```

# Modeling: Credential Earned - college completers only

## Model 1: Demographics

This model estimates the association between demographic characteristics and earning an Associate degree or lesser credential (versus a Bachelor's degree or higher), restricted to postsecondary completers (n = 12,651). The outcome is coded 1 = Associate-or-lesser credential, 0 = Bachelor's degree or higher — so an odds ratio above 1 indicates *lower* educational attainment relative to Bachelor's, not a positive outcome in the sense used elsewhere in this project.

**Special Education status and economic disadvantage show by far the largest effects found anywhere in this project.**

- **Special Education status:** approximately 5.2 times higher odds of earning an Associate-or-lesser credential rather than a Bachelor's (OR ≈ 5.19, CI: 4.191–6.477) — the single largest coefficient found in any model across this entire analysis.
- **Economic disadvantage:** approximately 2.5 times higher odds of the same (OR ≈ 2.54).

Both results make substantive sense and are consistent with everything found in DV2: these two groups face substantial barriers to completing a four-year degree specifically, even though (per DV3) those who do complete tend to complete regionally.

**Gender shows a large, clean effect in the opposite direction from Special Ed/economic status.**

- **Female graduates:** approximately 36% lower odds of an Associate-or-lesser credential relative to male graduates (OR ≈ 0.64) — meaning female completers are considerably more likely to hold a Bachelor's degree or higher. This is directionally consistent with DV2, where female graduates showed substantially higher odds of completing college at all — the pattern extends here to completing at the *higher* credential level specifically.

**Race/Ethnicity shows a mixed and, in one case, quite large effect.**

- **Asian/Pacific Islander graduates:** approximately 68% lower odds of an Associate-or-lesser credential (OR ≈ 0.32) — meaning Asian/Pacific Islander completers are considerably more likely to hold a Bachelor's degree or higher, the largest Race/Ethnicity effect found in this model.
- **American Indian graduates:** approximately 44% higher odds of an Associate-or-lesser credential (OR ≈ 1.44) — the opposite direction from Asian/Pacific Islander graduates.
- **Black and Hispanic graduates:** not statistically significant, likely reflecting reduced power given small cell sizes in this completers-only population (n = 138 and n = 106, respectively).

**English learner status is significant, though with a wide confidence interval.** OR ≈ 1.72 (CI: 1.148–2.622) — English learner completers show higher odds of an Associate-or-lesser credential.

**Overall Interpretation**

At the demographics-only stage, Special Education status and economic disadvantage stand out as by far the strongest predictors of credential level examined anywhere in this project — both showing that these groups are considerably more likely to complete at the sub-baccalaureate level. This is directionally consistent with, and adds nuance to, the DV2 finding that these same groups face substantial barriers to completing college at all: among those who do complete, they are also considerably more likely to complete a shorter, more technical credential rather than a four-year degree. Gender and Race/Ethnicity (particularly Asian/Pacific Islander) show meaningful effects as well, with women and Asian/Pacific Islander completers more likely to hold a Bachelor's degree.

**Model fit:** AIC = 15,898.8; n = 12,651.

```{r}
model1_dv4 <- glm(
  highest.cred.level ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner,
  family = binomial,
  data = master.model.dv4
)

summary(model1_dv4)
nobs(model1_dv4)
AIC(model1_dv4)

exp(cbind(OR = coef(model1_dv4), confint(model1_dv4)))

or_plot <- tidy(model1_dv4, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )


ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )
```

<br>

## Model 2: Demographics + High School Characteristics

This model adds district rural/urban classification, Range school status, local average wages, and local average unemployment rate to the demographic baseline, restricted to postsecondary completers (n = 12,651). Recall the outcome here is coded 1 = Associate degree or less, 0 = Bachelor's degree or higher, so positive coefficients indicate higher odds of a sub-baccalaureate credential.

**Range school status shows a strong, independent relationship with sub-baccalaureate attainment.** Range graduates show 57% higher odds of completing an Associate degree or less rather than a Bachelor's degree (OR ≈ 1.57, p \< .001) — consistent with Range communities' documented orientation toward technical and associate-level credentials tied to regional industry.

**Local wages shows a striking reversal from every other DV in this project.** Higher local wages relative to the state average predict substantially *lower* odds of sub-baccalaureate attainment (OR ≈ 0.29, p = .003) — in other words, higher local wages predict *higher* odds of Bachelor's-level completion. This is the opposite direction from what higher wages predicted for overall postsecondary completion (DV2, where higher wages discouraged completion) and for completing within the region (DV3, where higher wages encouraged regional completion) — a genuinely distinct relationship for this specific outcome, worth stating plainly rather than trying to force it into the same "opportunity cost" framing used elsewhere.

**District classification shows no significant relationship here** — a sharp contrast with DV3, where Dem_Desc produced some of the largest effects in this project. Town/rural mix is clearly non-significant (OR ≈ 1.01, p = .932), and Urban/town/rural mix only approaches significance without crossing it (OR ≈ 0.85, p = .088). Credential *level* appears to have little to do with district geography, even though credential *location* (DV3) was strongly tied to it.

**All demographic effects from Model 1 remain essentially unchanged.** Gender (OR ≈ 0.65), economic status (OR ≈ 2.48), Special Education status (OR ≈ 5.31), English learner status (OR ≈ 1.73), American Indian graduates (OR ≈ 1.44), and Asian/Pacific Islander graduates (OR ≈ 0.34) all hold steady in direction and magnitude from Model 1. Special Education status in particular remains an extraordinarily large effect — Special Ed completers show more than five times the odds of a sub-baccalaureate credential relative to a Bachelor's degree.

**Average unemployment rate remains non-significant** (OR ≈ 1.00, p = .934).

**Overall Interpretation**

Adding high school characteristics significantly improves model fit (likelihood ratio test: Δdeviance = 93.58, df = 5, p \< .001; AIC decreases from 15,898.8 to 15,815.2). **Range school status and local wages are the standout findings in this block, and the wage direction is genuinely distinct from every other DV in this project** — here, higher local wages push toward Bachelor's-level attainment rather than away from postsecondary engagement generally or toward regional completion specifically. District classification, by contrast, shows essentially no relationship with credential level, despite being one of the strongest predictors of completion location in DV3 — a useful reminder that "where" and "what level" are separate questions with different drivers. All five demographic effects from Model 1 remain fully intact.

Model fit: AIC = 15,815.2 (down from 15,898.8 in Model 1); Δdeviance = 93.58 on 5 df, p \< .001; n = 12,651 in both models.

```{r}
model2_dv4 <- glm(
  highest.cred.level ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    range.school +
    avg.wages.pct.state +
    avg.unemp.rate,
  family = binomial,
  data = master.model.dv4
)

summary(model2_dv4)

nobs(model1_dv4)
nobs(model2_dv4)

AIC(model1_dv4)
AIC(model2_dv4)

exp(cbind(OR = coef(model2_dv4), confint(model2_dv4)))

anova(model1_dv4, model2_dv4, test = "Chisq")


or_plot <- tidy(model2_dv4, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )


ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )
```

<br>

## Model 3: Demographics + High School Characteristics + High School Accomplishments

This model adds PSEO participation, ACT participation and score, SAT participation, AP exam participation, and CTE achievement to the growing specification, restricted to postsecondary completers (n = 12,651). Recall the outcome is coded 1 = Associate degree or less, 0 = Bachelor's degree or higher.

**ACT shows the same crossover pattern found throughout this project, at the largest scale seen anywhere in this entire analysis.** Taking the ACT shows an enormous positive association with earning an Associate-or-lesser credential (OR ≈ 14.12, p \< .001) — students who took the ACT are far more likely to complete at the sub-baccalaureate level than those who never tested. Among test-takers, each additional composite point predicts the opposite direction (OR ≈ 0.816 per point, p \< .001): higher scores predict a Bachelor's degree instead. Read together: ACT engagement without strong performance still predicts a shorter, more technical credential pathway.

**Black graduates become significant here for the first time in this sequence — a genuinely new finding, not present in Models 1 or 2.** OR ≈ 0.55 (CI: 0.368-0.811, p = .003) — once academic accomplishments are accounted for, Black completers show meaningfully higher odds of earning a Bachelor's degree specifically.

**English learner status remains non-significant here — and this is a genuine, standalone null, not an artifact of any missingness control, since none is anywhere in this specification.** OR ≈ 0.84 (CI: 0.549-1.319, p = .448). Credential level shows no independent relationship with English learner status at any stage of this model sequence.

**PSEO participation shows a clear, significant effect specific to this outcome.** OR ≈ 0.66 (p \< .001) — PSEO participants are notably more likely to complete a Bachelor's degree rather than an Associate-or-lesser credential. This is a different role for PSEO than seen in the other DVs in this project: elsewhere it predicted completion generally or showed no independent effect, while here it specifically predicts higher credential attainment among those who do complete.

**CTE achievement shows its strongest positive effect anywhere in this project.** CTE Concentrator/Completer status: OR ≈ 1.61 (p \< .001) — the largest CTE coefficient found in any model sequence. CTE Participant: OR ≈ 1.14 (p = .010). This makes strong substantive sense — CTE pathways are more likely to lead directly into Associate-level and technical credentials by design.

**AP exam and SAT participation both point toward Bachelor's attainment**, consistent with their role throughout this project (AP: OR ≈ 0.74, p \< .001; SAT: OR ≈ 0.61, p = .007).

**Average local unemployment rate becomes significant and meaningfully sized for the first time in this sequence** (OR ≈ 0.89 per unit, p \< .001) — higher local unemployment predicts lower odds of an Associate-or-lesser credential (i.e., predicts Bachelor's attainment instead). This is a counterintuitive direction worth flagging plainly rather than forcing into a tidy explanation, consistent with this project's general practice of not over-narrating a modest, less-intuitive coefficient.

**Local wages loses the significance it showed in Model 2.** OR ≈ 0.56 (CI: 0.216-1.460, p = .237) — its apparent relationship with credential level appears to run through academic engagement rather than operating independently.

**Demographic effects persist but attenuate substantially.** Gender (OR ≈ 0.73), economic status (OR ≈ 1.90, down from ≈ 2.48), and Special Ed status (OR ≈ 2.16, down sharply from ≈ 5.31) all remain significant but meaningfully weaker — academic engagement explains a substantial share of these disparities, though far from all of it. Asian/Pacific Islander graduates remain significant and large (OR ≈ 0.39, p \< .001). American Indian graduates lose significance (OR ≈ 1.09, p = .448), a real change from Models 1-2.

**District classification and Range school status both weaken.** Dem_Desc remains non-significant, as in Model 2. Range school status, significant and substantial in Model 2 (OR ≈ 1.57), weakens to only marginal significance here (OR ≈ 1.13, p = .031) — its apparent relationship with credential level runs substantially through academic accomplishments.

**Overall Interpretation**

Adding high school accomplishments produces by far the largest improvement in model fit found anywhere in this project (likelihood ratio test: Δdeviance = 3,060.72, df = 7, p \< .001; AIC decreases sharply from 15,815.2 to 12,768.5). **Academic achievement and college-preparatory engagement are overwhelmingly the dominant forces shaping credential level — far more so than demographics or geography.** The ACT crossover, at the largest scale found anywhere in this project, and CTE's strong positive association with Associate-level attainment are the standout findings. Black graduates emerge as a newly significant, independent predictor once academic engagement is considered, while American Indian graduates and Range school status both lose the significance they showed in earlier models. English learner status remains a clean, consistent null throughout this entire sequence.

Model fit: AIC = 12,768.5 (down from 15,815.2 in Model 2); Δdeviance = 3,060.72 on 7 df, p \< .001; n = 12,651 in both models.

```{r}
model3_dv4 <- glm(
  highest.cred.level ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    range.school +
    avg.wages.pct.state +
    avg.unemp.rate +
    pseo.participant +
    took.ACT +
    ACTCompositeScore.filled +
    ap.exam + 
    sat.taken +
    cte.achievement,
    family = binomial,
  data = master.model.dv4
)

summary(model3_dv4)

nobs(model1_dv4)
nobs(model2_dv4)
nobs(model3_dv4)

AIC(model1_dv4)
AIC(model2_dv4)
AIC(model3_dv4)

exp(cbind(OR = coef(model3_dv4), confint(model3_dv4)))

anova(model1_dv4, model2_dv4, model3_dv4, test = "Chisq")


or_plot <- tidy(model3_dv4, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )


ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )
```

<br>

## DV4 Summary: What Predicts Sub-Baccalaureate Credential Attainment (College Completers Only)

Across all three models, academic accomplishments emerge as overwhelmingly the dominant force shaping whether a postsecondary completer earns a Bachelor's degree or an Associate-or-lesser credential — far more so than demographics or geography, a genuinely different pattern than found in any other DV in this project. Recall the outcome is coded 1 = Associate degree or less, 0 = Bachelor's degree or higher, so all odds ratios below describe the odds of the sub-baccalaureate outcome unless noted otherwise.

**ACT engagement produces the single largest coefficient found anywhere in this entire analysis, and shows the clearest crossover pattern of any DV.** Simply taking the ACT is associated with dramatically higher odds of completing at the sub-baccalaureate level (OR ≈ 14.12) — an enormous effect, even relative to the large ACT effects found elsewhere in this project. Among test-takers, the direction reverses: each additional composite point predicts substantially lower odds of a sub-baccalaureate credential (OR ≈ 0.816 per point) — higher scorers complete Bachelor's degrees instead. Read together: ACT engagement without strong performance predicts a shorter, more technical credential pathway, while strong performance predicts the opposite.

**Special Education status shows the single largest demographic effect found anywhere in this project, though it attenuates substantially once academic accomplishments are considered.** Special Ed completers show over five times the odds of a sub-baccalaureate credential in the demographics-only and characteristics models (OR ≈ 5.19–5.31), weakening to just over double the odds once academic engagement is accounted for (OR ≈ 2.16) — a real, substantial reduction, but the effect remains large and clearly significant throughout.

**Economic status shows a similarly large, stable pattern.** Economically disadvantaged completers show substantially higher odds of a sub- baccalaureate credential across all three models (OR ≈ 1.90–2.54), weakening somewhat once academic accomplishments enter the model but remaining one of the strongest and most consistent predictors in this sequence.

**English learner status is significant early in the sequence but resolves to a clean, genuine null once academic accomplishments are considered — with no missingness control anywhere in the specification.** Significant in the demographics-only and characteristics models (OR ≈ 1.72–1.73, predicting higher odds of a sub-baccalaureate credential), English learner status loses significance entirely in the final model (OR ≈ 0.84). Whatever drives this early relationship appears to run through academic engagement rather than English learner status operating independently.

**Race/Ethnicity findings shift meaningfully across the sequence.** Asian/Pacific Islander graduates show a large, stable, significant relationship with Bachelor's-level attainment throughout all three models (OR ≈ 0.32–0.39) — the most consistent Race/Ethnicity finding in this sequence. American Indian graduates are significant in the demographics- only and characteristics models (OR ≈ 1.44) but lose significance once academic accomplishments are added (OR ≈ 1.09). Black graduates show the opposite trajectory: non-significant in the first two models, then significant in the final model (OR ≈ 0.55) — a disparity favoring Bachelor's attainment that only emerges once academic engagement is fully accounted for. Hispanic graduates remain non-significant throughout.

**PSEO participation shows a distinct role for this outcome compared to every other DV in this project.** OR ≈ 0.66 — PSEO participants are notably more likely to complete a Bachelor's degree specifically, a different relationship than PSEO showed for completion generally or for regional attachment, where it predicted little or showed a different direction entirely.

**CTE achievement shows its strongest positive effect anywhere in this project.** CTE Concentrator/Completer status (OR ≈ 1.61) is the largest CTE coefficient found in any model sequence, with CTE Participant status showing a smaller but still significant effect (OR ≈ 1.14) — both substantively sensible, since CTE pathways are designed to lead directly into Associate-level and technical credentials.

**AP exam and SAT participation both predict Bachelor's-level attainment**, consistent with their role throughout this project (AP: OR ≈ 0.74; SAT: OR ≈ 0.61).

**Gender shows a large, stable, significant relationship throughout**, with female completers consistently more likely to earn a Bachelor's degree (OR ≈ 0.65–0.73 across all three models).

**District classification shows no relationship with credential level at any stage** — a sharp contrast with this project's findings on completion *location*, where geography was one of the strongest predictors. Range school status is significant in the characteristics model (OR ≈ 1.57) but weakens to only marginal significance once academic accomplishments are considered (OR ≈ 1.13).

**Local wages and unemployment rate both show effects that emerge or fade unpredictably across the sequence, worth treating as modest and provisional rather than forcing into a single coherent story.** Local wages is significant and protective toward Bachelor's attainment in the characteristics model (OR ≈ 0.29) but loses significance once academic accomplishments are added (OR ≈ 0.56). Local unemployment rate is non- significant until the final model, where it becomes significant in a counterintuitive direction (OR ≈ 0.89 per unit — higher unemployment predicting Bachelor's attainment rather than the reverse) — worth noting plainly rather than over-explaining.

<br>

# Executive Summary

This analysis used logistic regression to confirm and refine patterns identified in earlier descriptive work on post-high school pathways for Northeast Minnesota graduates. Five models were built, each isolating a different piece of the pathway: whether graduates stay in the region long term, whether they complete college, where they complete it, and what type of credential they earn.

Four consistent themes emerged across all five models:

1.  **Geography does different jobs at different stages.** High school district type, Range school status, and local economic conditions strongly predict whether a graduate stays in the region for work and where they complete their postsecondary education — but have almost no bearing on whether they complete college at all, or what type of credential they earn once they do. Geography shapes *where* things happen, not *whether* or *what*.

2.  **CTE and ACT represent two distinct, opposing pathways, and this project's clearest, largest, and most consistent finding.** Career and technical education consistently predicts staying in the region and completing shorter, more technical credentials. ACT engagement shows a striking split, visible in every model: simply taking the test predicts staying regional and completing an Associate-level credential — in one model, taking the ACT alone predicts fourteen times higher odds of a sub-baccalaureate credential — while scoring well on it predicts leaving the region and completing a Bachelor's degree instead. These aren't two versions of the same "college-readiness" signal; they represent different, sometimes opposing, routes through the pipeline.

3.  **English learner status tells two entirely different stories, depending on which outcome is examined — one of the most important findings in this analysis.** For the full graduating population, English learner status is a robust, independent predictor of reduced odds of sustained regional workforce attachment, holding up fully alongside demographics, geography, academic accomplishments, and postsecondary completion — this is a real, standalone disparity, not explained by anything else in the model. But among students who go on to complete a postsecondary credential, English learner status shows no relationship at all with regional attachment, completion location, or credential level. In other words: the English learner disparity in this data is entirely about workforce attachment for those who don't complete a credential — it is not a story about academic preparation, and it disappears entirely once a student completes college.

4.  **Demographic disparities are real, but some are explained by how students engage with school, while others are not.** Economic disadvantage and Special Education status both show very large effects early in the postsecondary-completion and credential-level models, and both weaken substantially — though remain significant — once academic engagement is added, indicating that some, not all, of the disparity operates through unequal access to or engagement with CTE, PSEO, and testing pathways. Black, Hispanic, and Asian/Pacific Islander graduates tell a different story: their disparities in regional workforce attachment and postsecondary completion largely persist regardless of academic engagement, and in several models a disparity only becomes visible once academic engagement is fully accounted for — these are not explained away by anything measured in this analysis.

The single largest and most consistent finding across every model, and the one most directly relevant to the project's regional workforce mission, is that **geography — where a student completes their postsecondary education — matters more than what credential they earn.** Graduating within the region was the strongest predictor of sustained regional employment found among completers. But credential type is not irrelevant either: among college completers specifically, Associate-level and technical credentials showed a real, independent, additional association with staying regional, on top of the location effect.

<br>

# The Story This Data Tells

Picture a graduating class in Northeast Minnesota. What happens to them over the next five, ten years, and why?

**The pipeline starts to split immediately, along economic and ability- status lines.** Economically disadvantaged students and students with disabilities are considerably less likely to complete any postsecondary credential at all — among the strongest demographic patterns in the entire analysis. But these same students, if they do complete a credential, don't carry that disadvantage forward in the same form: much of what predicted non-completion in the first place turns out to run through *how* these students engaged with high school academically — CTE involvement, ACT participation, PSEO — rather than being a fixed consequence of the demographic category itself.

**English learners face a genuinely different, more specific barrier — one that shows up at the front door, not throughout the building.** English learners are meaningfully less likely to be sustainably attached to the regional workforce, and this disadvantage holds up fully no matter what else is accounted for — academic preparation, geography, whether they attempted college at all. But this barrier is entirely specific to students who don't go on to complete a postsecondary credential. Among English learners who do complete a credential, there is no detectable disadvantage in staying regional, completing locally, or the level of credential earned. The English learner story in this data is a workforce- attachment story, not an academic-preparation story — and it resolves itself completely once a credential is in hand.

**For those who do complete college, a second split happens — not about whether they succeed, but about where and how.** Two students with identical demographics can take very different paths depending on how they engaged with high school. A student who took CTE courses and engaged with the ACT — regardless of score — is likely on a track toward a shorter, technical credential completed close to home, and likely to stay in the regional workforce afterward. A student who scored well on the ACT and pursued AP coursework or the SAT is likely headed toward a four-year degree, often completed outside the region — and less likely to return.

**This is the pipeline's central tension, visible in every model in this analysis: the very things that predict finishing a four-year degree — strong test scores, AP coursework, SAT participation — are the same things that predict leaving the region.** Meanwhile, the pathways this project's organization has less institutional visibility into promoting — CTE, technical credentials, PSEO — are consistently the ones associated with staying.

**Geography reinforces this tension rather than resolving it.** Students from more urban districts and Range school communities are considerably more likely to complete their postsecondary education within the region — this is the single largest set of effects found anywhere in this analysis — but this same geographic advantage barely registers when predicting whether a student completes college at all, or what type of credential they earn. Geography shapes *where* the funnel empties out, not how big the funnel is or what flows through it.

**And underneath all of this, demographic disparities persist — but not always for the same reasons, and not always visible at the same point in the analysis.** Some disparities — Special Education status, economic disadvantage, and, among completers, American Indian identity and Range school status — shrink substantially once a student's academic engagement is accounted for, suggesting these groups may be channeled toward or away from certain pathways based on how school systems engage them, not fixed traits tied to the demographic category itself. Other disparities — particularly for Black, Hispanic, and Asian/Pacific Islander graduates — persist even after accounting for every other factor in this analysis, and in several cases only become visible once academic engagement is fully accounted for, suggesting they are not artifacts of differential program access but something this analysis cannot fully explain.

**Put together, the data describes a region whose workforce retention depends less on whether people leave for stronger opportunities elsewhere and more on which door they walked through in high school** — CTE and modest ACT engagement lead back to the region; AP coursework, strong ACT scores, and SAT participation lead away from it. This is not a story about talent leaving because the region has nothing to offer skilled workers — it's a story about two genuinely different pipelines, one regionally anchored and one not, both running through the same high schools at the same time.
