---
title: "Disconnected"
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
library(htmlwidgets)
library(RColorBrewer)
library(readxl)
library(janitor)
library(lubridate)
library(systemfonts)
reset_font_cache()
library(ggtext)
library(gmodels)
library(fastDummies)
library(car)
library(glmnet)
library(glmnetUtils)
library(pscl)
library(sjPlot)
library(ggforce)
library(openxlsx)
```

```{r themes and shapefiles, include=FALSE}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", linewidth  = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.key.size = unit(1, "lines"),
        legend.margin = margin(0,0,0,0),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.text = element_text(margin = margin(l = 2)),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", linewidth = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))


theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
    select(5,6) %>%
    unique() %>%
    mutate(edr = str_replace(edr, "  ", " "),
           planning.region = str_replace(planning.region, " Minnesota", ""),
           planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
           edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

counties.regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
  rename(mif = `MIF Region`) %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Name = str_replace(Name, "Mcleod", "McLeod"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc) ,
         edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"))

counties.regions.1 <- counties.regions %>%
  mutate(statefp = "27",
         project.pr = ifelse(edr %in% c("EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 4 - West Central"), "Northwest",
                             ifelse(edr == "EDR 3 - Arrowhead", "Northeast",
                                    ifelse(edr %in% c("EDR 5 - North Central", "EDR 7E- East Central", "EDR 7W- Central"), "Central",
                                           ifelse(edr %in% c("EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 8 - Southwest"), "Southwest",
                                                  ifelse(edr %in% c("EDR 9 - South Central", "EDR 10 - Southeast"), "Southern", as.character(planning.region)))))),
         project.pr = fct_relevel(project.pr, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southern"))


color.ruca <- c("Entirely rural" = "#009933", "Town/rural mix" = "#99CC33", "Urban/town/rural mix" = "#CC9966", "Entirely urban" = "#754C29", "Minnesota" = "black")

color.pr <- c("Northwest" = 	"#4575b4", "Northeast" = "grey", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.edr.simple <- c("EDR 1" = "#b3cde3", "EDR 2" = "#8c96c6", "EDR 3" = "#fe9929", "EDR 4" = "#8856a7", "EDR 5" = "#810f7c", "EDR 6E" = "#e5f5f9", "EDR 6W" = "#bdc9e1", "EDR 7E" = "#99d8c9", "EDR 7W" = "#2ca25f", "EDR 8" = "#74a9cf", "EDR 9" = "#0570b0", "EDR 10" = "#d7301f", "EDR 11" = "#d8b365", "Minnesota" = "black")

color.pr.edr <- c ("Northwest" = "#4575b4","Northeast" = "#e0f3f8", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black", "EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365")

mn_counties <- st_read("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Shapefiles/County shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE) %>%
  rename(countyfp = FIPS_CODE)
```

```{r master}
master.original <- read_csv("Data/SLEDS/Masters/After analysis/Master-after-ct.csv") 

kable(names(master.original))

kable(head(master.original))
```

# Data Prep

In this analysis, I want to dive deeper into whether the disconnected individuals are 5 and 10 years after graduating high school have employment in the region or outside. I think this will be important for people to know. In addition, I want to do some analysis to see if there are any prominent drivers and predictors pushing people into one of the pathways - Not meaningful - regional, or Not meaningful - outside region.

The first step is to identify which individuals were categorized as not having meaningful workforce participation at 5- and 10-years after graduating high school. In our analysis and write-ups, we will focus primarily on the disconnected individuals 5-years after high school.

```{r}
disconnected.grad.year.5 <- master.original %>%
  filter(grad.year.5 == "Not meaningful, not attending ps")

disconnected.grad.year.10 <- master.original %>%
  filter(grad.year.10 == "Not meaningful, not attending ps")

names(disconnected.grad.year.5)

kable(head(disconnected.grad.year.5))
```

At 5-years after high school, there are `r comma(nrow(disconnected.grad.year.5))` individuals that are categorized as having not meaningful workforce participation and not attending post-secondary. Next we have to bring in the employment records and filter them to these individuals.

```{r}
employment.original <- read_csv("Data/SLEDS/Employment/Employment-Northeast.csv") %>%
  drop_na(PersonID)

employment.disconnected.grad.year.5 <- employment.original %>%
  right_join(disconnected.grad.year.5[,c(1)], by = "PersonID")

names(employment.disconnected.grad.year.5)

kable(head(employment.disconnected.grad.year.5))
```

The dataset `employment.disconnected.grad.year.5` provides the following;

-   PersonID: Identifies the graduate from Northeast Minnesota
-   TimeID: Last day of fiscal quarter
-   EmploymentOrganizationID: An internal identification for each employer.
-   EmployeeQuarterlyHoursWorked: The total number of hours worked by an employee at an employer’s location during a specific calendar quarter
-   EmployeeQuarterlyWages: The total wages earned by an employee at an employer’s location during a specific calendar quarter
-   EmployeeTenure: The total number of quarters a worker has been employeed by an organization at a specific location, up to an including the current calendar quarter
-   IndustryID: Auto generated ID code for a record in the Industry dimension.
-   USIndustryCode: Six-digit North American Industrial Classification System (NAICS) code
-   GroupCode: Four-digit North American Industrial Classification System (NAICS) code
-   OwnershipCode: A classification system based upon the controlling interest of the establishment primarily designed to distinguish between governmental and non-governmental establishments
    -   10 = Federal government
    -   20 = State government
    -   30 = Local government
    -   50 = Private
-   EmployerLocationTotalWagesPaid: The total wages paid to all workers employeed by an organization at a specific location during a specific calendar quarter
-   QuarterlyAverageEmployment: The average number of workers employeed by an organization at a specific location during a specific calendar quarter
-   County: Minnesota county code to which this institution belongs. This can be used as a foreign key into the Geography dimension on MNCounty.
    -   [Link to codes](https://sleds.mn.gov/dataDictionary/658)
-   CountyName: The name of the county corresponding to the County and MNCounty Codes for this Location.
-   MNCounty: County Code for this location as assigned by the state of Minnesota.
    -   [Link to codes](https://sleds.mn.gov/dataDictionary/663) City: All of the city data is NA
-   City: All NA so will eliminate.

The next step is to identify what year equals 5-years after completing high school. The original master dataset didn't include their high school graduation year so I will have to bring that in and join it with the `employment.disconnected.grad.year.5`. Then I will filter all employment instances that match that year.

```{r}
hs.grad.year <- read_csv("Data/SLEDS/Masters/Master.csv") %>%
  select(PersonID, grad.year)

employment.disconnected.hs.grad.year.5 <- employment.disconnected.grad.year.5 %>%
  left_join(hs.grad.year, by = "PersonID") %>%
  mutate(hs.grad.year.5 = grad.year + 5,
         emp.year = str_sub(TimeID, 1, 4),
         emp.year = as.integer(emp.year)) %>%
  filter(hs.grad.year.5 == emp.year)


names(employment.disconnected.hs.grad.year.5)

kable(head(employment.disconnected.hs.grad.year.5))
```

Okay, we now have a dataset that contains every individual that was categorized as "Not meaningufl, not attening ps" 5-years after completing high school along with each employment instance that had during the year that matches 5 years after high school. Now we can start to determine if their not meaningful workforce participation took place inside or outside the region, or both. To do this, we will join `employment.disconnected.hs.grad.year.5` with my `county.regions` dataset so that I will be able to identify whether the employer was located in EDR 3 or not. Then I will categorize each individual into the three categories.

```{r}
in.out.disconnected.grad.year.5 <- employment.disconnected.hs.grad.year.5 %>%
  select(PersonID, County, CountyName, hs.grad.year.5) %>%
  mutate(County = formatC(County, width = 3, flag = "0"),
         CountyName = str_replace(CountyName, " County", ""),
         CountyName = str_replace(CountyName, "Saint Louis", "St. Louis"),
         CountyName = str_replace(CountyName, "City of Duluth", "St. Louis"),
         CountyName = str_replace(CountyName, "City of Minneapolis", "Hennepin"),
         CountyName = str_replace(CountyName, "City of St. Paul", "Ramsey"),
         CountyName = str_replace(CountyName, "Shakopee Mdewakanton Community", "Scott"),
         CountyName = str_replace(CountyName, "White Earth Tribal Counci", "Mahnomen"),
         County = ifelse(CountyName == "Mahnomen", "087", County),
         County = ifelse(CountyName == "St. Louis", "137", County),
         County = ifelse(CountyName == "Hennepin", "053", County),
         County = ifelse(CountyName == "Ramsey", "123", County),
         County = ifelse(CountyName == "Scott", "139", County)) %>%
  filter(CountyName != "Statewide (i.e. no fixed location)") %>%
  left_join(counties.regions.1[,c(1, 10)], by = c("County" = "countyfp")) %>%
  filter(CountyName != "Unknown") %>%
  rename(emp.County = County,
         emp.CountyName = CountyName,
         emp.pr = project.pr) %>%
  mutate(not.meaningful.region = ifelse(emp.pr == "Northeast", -1, 1)) %>%
  group_by(PersonID) %>%
  distinct(PersonID, not.meaningful.region) %>%
  ungroup() %>%
  group_by(PersonID) %>%
  mutate(not.meaningful.location = ifelse(sum(not.meaningful.region) == -1, "Employed in region",
                                          ifelse(sum(not.meaningful.region) == 0, "Employed in and out of region", "Employed out of region"))) %>%
  ungroup() %>%
  select(PersonID, not.meaningful.location) %>%
  distinct(PersonID, .keep_all = TRUE)
  
names(in.out.disconnected.grad.year.5)

kable(head(in.out.disconnected.grad.year.5))
```

We now have `r comma(nrow(in.out.disconnected.grad.year.5))` individuals that did not have meaningful workforce participation in Minnesota and are categorized as having their employment in the region, in and out of the region, and only out of the region. We will now merge this with our master dataset and start doing analysis.

```{r}
master <- master.original %>%
  left_join(in.out.disconnected.grad.year.5, by = "PersonID") %>%
    filter(!is.na(not.meaningful.location)) 

names(master)

kable(head(master))
```

<br>

# Distribution

Let's take a look at the distribution of this population. The table below provides the percentage of individuals that have not meaningful workforce participation at 5-years after graduating high school. 53% of them are employed in the region while 27% are employed somewhere else in Minnesota and 20% have employment both in and outside the region.

```{r}
data <- master %>%
  group_by(not.meaningful.location) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  mutate(pct = n / sum(n))

names(data)

datatable(data, class = "cell-border stripe", filter = "top", rownames = FALSE,
          options = list(columnDefs = list(list(className = "dt-center", targets = 1:2)))) %>%
  formatCurrency(2, "", digits = 0) %>%
  formatPercentage(3, digits = 0)

```

Let's see what this looks like in more detail. First, let's take a look at how many different employers they averaged over the course of that particular year. The table below shows that individuals who were employed in and out of the region had a significantly higher number of total employers, wages earned, and total hours worked.

```{r}
employment <- employment.original %>%
  mutate(emp.year = str_sub(TimeID, 1, 4),
         emp.year = as.integer(emp.year),
         emp.month = str_sub(TimeID, 5,6),
         emp.quarter = ifelse(emp.month == "03", 1,
                              ifelse(emp.month == "06", 2,
                                     ifelse(emp.month == "09", 3,
                                            ifelse(emp.month == "12", 4, emp.month))))) %>%
  select(PersonID, EmploymentOrganizationID, EmployeeQuarterlyHoursWorked, EmployeeQuarterlyWages, County, CountyName, emp.year, emp.quarter)


employment.totals <- master %>%
  select(PersonID, not.meaningful.location) %>%
  left_join(hs.grad.year, by = "PersonID")  %>%
  mutate(hs.grad.year.5 = grad.year + 5) %>%
  left_join(employment, by = c("PersonID", "hs.grad.year.5" = "emp.year")) %>%
  group_by(PersonID, EmploymentOrganizationID) %>%
  summarize(n.quarters = sum(n()),
            hours.worked = sum(EmployeeQuarterlyHoursWorked),
            wages.earned = sum(EmployeeQuarterlyWages)) %>%
  ungroup() %>%
  group_by(PersonID) %>%
  mutate(total.hours = sum(hours.worked),
         total.employers = sum(n()),
         total.wages.earned = sum(wages.earned)) %>%
  ungroup() %>%
  filter(total.hours > 0) %>% 
  select(PersonID, total.employers, total.wages.earned, total.hours) %>%
  distinct(PersonID, .keep_all = TRUE)

data <- employment.totals %>%
  left_join(master[,c(1, 59)], by = "PersonID") %>%
  group_by(not.meaningful.location) %>%
  summarize(mean.total.employers = mean(total.employers),
            median.total.employers = median(total.employers),
            mean.wages.earned = mean(total.wages.earned),
            median.wages.earned = median(total.wages.earned),
            mean.total.hours = mean(total.hours),
            median.total.hours = median(total.hours)) %>%
  ungroup() 

datatable(data, class = "cell-border stripe", filter = "top", rownames = FALSE,
          options = list(columnDefs = list(list(className = "dt-center", targets = 1:6)))) %>%
  formatCurrency(2:3, "", digits = 0) %>%
  formatCurrency(4:5, "$", digits = 0) %>%
  formatCurrency(6:7, "", digits = 0)

  
```

<br>

# Analysis

For this analysis I'm going to explore relationships and predictors that play a role in individuals falling into one of the three not meaningful categories - employed in the region, employed inside and outside the region, and employed outside the region. We will use the following methodology;
