---
title: "Multivariate Confirmation Analysis on Persistent Disconnection"
editor: visual
---

```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
library(htmlwidgets)
library(RColorBrewer)
library(readxl)
library(janitor)
library(lubridate)
library(systemfonts)
reset_font_cache()
library(ggtext)
library(gmodels)
library(fastDummies)
library(car)
library(glmnet)
library(glmnetUtils)
library(pscl)
library(sjPlot)
library(ggforce)
library(openxlsx)
```

```{r themes and shapefiles, include=FALSE}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", linewidth  = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.key.size = unit(1, "lines"),
        legend.margin = margin(0,0,0,0),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.text = element_text(margin = margin(l = 2)),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", linewidth = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))


theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
  select(5,6) %>%
  unique() %>%
  mutate(edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

counties.regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
  rename(mif = `MIF Region`) %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Name = str_replace(Name, "Mcleod", "McLeod"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc) ,
         edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"))

counties.regions.1 <- counties.regions %>%
  mutate(statefp = "27",
         project.pr = ifelse(edr %in% c("EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 4 - West Central"), "Northwest",
                             ifelse(edr == "EDR 3 - Arrowhead", "Northeast",
                                    ifelse(edr %in% c("EDR 5 - North Central", "EDR 7E- East Central", "EDR 7W- Central"), "Central",
                                           ifelse(edr %in% c("EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 8 - Southwest"), "Southwest",
                                                  ifelse(edr %in% c("EDR 9 - South Central", "EDR 10 - Southeast"), "Southern", as.character(planning.region)))))),
         project.pr = fct_relevel(project.pr, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southern"))


color.ruca <- c("Entirely rural" = "#009933", "Town/rural mix" = "#99CC33", "Urban/town/rural mix" = "#CC9966", "Entirely urban" = "#754C29", "Minnesota" = "black")

color.pr <- c("Northwest" = 	"#4575b4", "Northeast" = "grey", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.edr.simple <- c("EDR 1" = "#b3cde3", "EDR 2" = "#8c96c6", "EDR 3" = "#fe9929", "EDR 4" = "#8856a7", "EDR 5" = "#810f7c", "EDR 6E" = "#e5f5f9", "EDR 6W" = "#bdc9e1", "EDR 7E" = "#99d8c9", "EDR 7W" = "#2ca25f", "EDR 8" = "#74a9cf", "EDR 9" = "#0570b0", "EDR 10" = "#d7301f", "EDR 11" = "#d8b365", "Minnesota" = "black")

color.pr.edr <- c ("Northwest" = "#4575b4","Northeast" = "#e0f3f8", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black", "EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365")

mn_counties <- st_read("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Shapefiles/County shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE) %>%
  rename(countyfp = FIPS_CODE)
```

```{r master}
master.original <- read_csv("Data/SLEDS/Masters/Master-non-college-grads-disconnection.csv") 

kable(names(master.original))

kable(head(master.original))
```

# Multivariate Confirmation Analysis: Disconnection Persistence Among Non–Postsecondary Graduates

## Introduction

The bivariate analysis of workforce disconnection among non-college graduates identified two distinct disconnection mechanisms operating within a single combined measure. Non-sustained employment (working, but below the sustained-employment threshold) and absence of any Minnesota employment record behave as structurally different phenomena: the former is front-loaded and largely resolves within five years, while the latter is bimodal and behaves as a durable end-state once established. Separating these two mechanisms revealed that different demographic and high-school-accomplishment variables predict each one — most notably, Special Education status and English learner status appeared to drive no-record persistence specifically, while Economic status and Gender appeared to drive non-sustained persistence specifically, and different racial/ethnic groups showed disparities concentrated in one mechanism or the other.

This multivariate analysis exists to confirm whether that mechanism-split pattern holds once all variables are considered simultaneously, rather than one at a time. This step is necessary because earlier work in this project has shown that bivariate relationships — particularly those involving English learner status — can substantially attenuate or shift once other variables, especially missing-data patterns in academic assessment scores, are accounted for in the same model. Before this mechanism-split finding is presented as a confirmed result, it should be tested under the same conditions.

## Methodology

### Study Population

The analytic sample is restricted to individuals who did **not** complete a postsecondary credential, consistent with the non-college-graduate population used throughout this project. The sample is further restricted to individuals who graduated (or exited postsecondary attendance, for those who attended without completing) after 2010, consistent with the cohort restriction applied elsewhere in this project to avoid the PSEO participation tracking gap present in earlier cohorts.

### Dependent Variables

This analysis models two outcomes, each derived from the disconnection-persistence categories established in the bivariate analysis. Each graduate's workforce status was tracked across the first five years after exiting formal education (high school for direct-to-workforce graduates, or postsecondary attendance for those who attended without completing).

- **Persistent non-sustained employment** — binary (1 = non-sustained workforce participation in 4 or 5 of the first 5 years, 0 = otherwise). This isolates the instability mechanism identified in the bivariate work.
- **Persistent no-record status** — binary (1 = no Minnesota employment record in 4 or 5 of the first 5 years, 0 = otherwise). This isolates the disappearance-from-the-system mechanism identified in the bivariate work.

Each outcome is modeled separately, in its own specification, rather than as levels of a single combined variable. This mirrors the logic of the bivariate analysis: combining the two mechanisms into one model would risk re-obscuring the distinction this analysis is specifically designed to confirm.

### Independent Variables

The independent variable set is consistent with the variables used throughout this project's non-college-graduate analyses, with two adjustments made based on findings from this project's earlier work.

**Demographics**

- `Gender`
- `RaceEthnicity`
- `economic.status`
- `SpecialEdStatus`
- `english.learner`

**High school characteristics**

- `Dem_Desc`
- `avg.wages.pct.state`
- `avg.unemp.rate`

**High school accomplishments**

- `pseo.participant`
- `ap.exam`
- `took.ACT`
- `ACTCompositeScore` (recoded to 0 for non-test-takers, entered alongside `took.ACT`, per the missing-indicator approach established earlier in this project)
- `cte.achievement`

**Postsecondary variable**

- `attended.ps` — included specifically because it was the single strongest predictor of overall disconnection in the bivariate analysis, and it is the one variable found to meaningfully affect both disconnection mechanisms rather than only one.

**Adjustments from the bivariate variable list:**

- **MCA tests are excluded.** There is enourmous amount of missing values and strangeness in the data so we are leaving them out entirely.
- **`total.cte.courses.taken` is excluded**, consistent with the redundancy this project has established between the continuous course-count measure and the categorical `cte.achievement` variable.

### Modeling Approach

Each outcome (persistent non-sustained status, persistent no-record status) is modeled using a staged block-entry sequence, consistent with the modeling structure used throughout this project's other multivariate confirmation analyses. Predictors are introduced in three stages to make attenuation of any individual variable's effect visible and attributable to a specific block, rather than only observable in a single final specification.

**Model 1: Demographics and High School Characteristics**

- Demographics: `Gender`, `RaceEthnicity`, `economic.status`, `SpecialEdStatus`, `english.learner`
- High school characteristics: `Dem_Desc`, `avg.wages.pct.state`, `avg.unemp.rate`

This baseline isolates whether structural and demographic positioning alone predicts persistence in either disconnection mechanism, before any measure of academic engagement or preparation is introduced.

**Model 2: Add High School Accomplishments**

- Adds: `pseo.participant`, `took.ACT`, `ACTCompositeScore.filled`, `ap.exam`, `cte.achievement`

This is the model of primary interest for this analysis. Because English learner status has previously been shown to correlate with MCA Math missingness, this block specifically tests whether the English learner effect identified in Model 1 attenuates once high school accomplishments — including the MCA missingness indicator — are introduced. The same logic applies to Special Education status and any other Model 1 effect that might operate through academic engagement rather than independently.

**Model 3: Add Postsecondary Attempt**

- Adds: `attended.ps`

This isolates whether attempting postsecondary education, identified in the bivariate analysis as the one variable meaningfully affecting both disconnection mechanisms, adds independent explanatory power once demographics and high school accomplishments are already accounted for, or whether its bivariate strength was substantially channeled through those same variables (e.g., PSEO participation, CTE achievement, or ACT engagement).

Each model is compared to the prior one using a likelihood ratio test to confirm it adds significant explanatory power, with AIC tracked across the sequence to summarize overall model improvement. This staged structure is applied identically to both outcomes, allowing direct comparison of how each mechanism's predictors evolve across the same three blocks.

### Evaluation Criteria

Results will be reported using odds ratios, 95% confidence intervals, and statistical significance levels. Interpretation will focus specifically on whether the demographic variables identified as mechanism-specific in the bivariate analysis (Special Education and English learner status for no-record persistence; Economic status and Gender for non-sustained persistence; racial/ethnic groups for whichever mechanism their bivariate disparity was concentrated in) remain significant, attenuate, or reverse once the full variable set is included. Findings will be described as associations holding other included variables constant. No causal claims will be made.

<br>

# Prepping the data

The modeling dataset begins from the merged file containing both the disconnection-persistence categories (established in the bivariate analysis) and the full set of demographic, high school, and postsecondary variables used throughout this project. Three preparation steps are applied before modeling.

**Cohort restriction.** The sample is restricted to graduates from 2011 onward (`grad.year > 2010`). This mirrors the cohort restriction applied throughout this project's other analyses of non-college graduates: PSEO participation was not reliably tracked prior to 2011 (missingness runs 92-100% in the 2008-2010 cohorts, dropping to near-zero by 2014), and including those cohorts would risk misclassifying untracked PSEO participants as confirmed non-participants.

**Reference level assignment.** Categorical variables are releveled to match the reference categories used consistently throughout this project (male, White, non-economically-disadvantaged, non-Special-Education, non-English-learner, entirely rural district, no CTE involvement, non-participant/non-test-taker for the binary accomplishment variables). This keeps odds ratios in this analysis directly comparable in direction and interpretation to every other model built in this project.

**Missing-indicator recodes for two variables with structural or correlated missingness:**

- **ACT composite score.** Students who did not take the ACT have no composite score by definition — confirmed here to be a fully deterministic relationship with `took.ACT` (zero mismatches between the two). `ACTCompositeScore` is recoded to 0 for non-test-takers and entered into the model alongside `took.ACT`, so that `took.ACT` absorbs the effect of test participation itself while the filled score captures the achievement gradient among test-takers only.

**Dependent variable construction.** The two outcomes tested in this analysis are built directly from the bivariate persistence categories, collapsed to binary indicators:

- `persistent.non_sustained` — 1 if the graduate was non-sustained in 4 or 5 of their first 5 years post-education-exit, 0 otherwise.
- `persistent.no_record` — 1 if the graduate had no Minnesota employment record in 4 or 5 of those years, 0 otherwise.

Before fitting any models, missingness is checked across the full variable set to confirm the cohort restriction and missing-indicator recodes fully resolved known data-quality issues, and the English learner / MCA Math-missingness correlation is re-verified on this specific cohort-restricted sample to confirm the relationship under examination is actually present before testing whether the multivariate model absorbs it.

```{r}
master.disc <- master.original %>%
  filter(grad.year > 2010) %>%
  mutate(
    Gender              = fct_relevel(Gender, "M"),
    RaceEthnicity       = fct_relevel(RaceEthnicity, "White"),
    economic.status = as_factor(economic.status),
    economic.status     = fct_relevel(economic.status, "0"),
    SpecialEdStatus = as_factor(SpecialEdStatus),
    SpecialEdStatus     = fct_relevel(SpecialEdStatus, "0"),
    english.learner = as_factor(english.learner),
    english.learner     = fct_relevel(english.learner, "0"),
    Dem_Desc            = fct_relevel(Dem_Desc, "Entirely rural"),
    pseo.participant = as_factor(pseo.participant),
    pseo.participant    = fct_relevel(pseo.participant, "0"),
    took.ACT            = fct_relevel(took.ACT, "No"),
    ap.exam = as_factor(ap.exam),
    ap.exam             = fct_relevel(ap.exam, "0"),
    cte.achievement     = fct_relevel(cte.achievement, "No CTE"),
    attended.ps         = fct_relevel(attended.ps, "No"),
    
    # ACT missing-indicator recode - confirmed safe: missingness fully
    # explained by took.ACT (9,228 No/missing, 5,941 Yes/present, zero mismatches)
    ACTCompositeScore.filled = ifelse(is.na(ACTCompositeScore), 0, ACTCompositeScore),

    # Dependent variables: binary persistence indicators
    persistent.non_sustained = ifelse(non_sustained.cat == "Persistently non-sustained", 1, 0),
    persistent.no_record     = ifelse(no_record.cat == "Persistently no-record", 1, 0)
  )

nrow(master.disc)

model_vars_disc <- c("Gender", "RaceEthnicity", "economic.status", "SpecialEdStatus",
                     "english.learner", "Dem_Desc", "avg.wages.pct.state", "avg.unemp.rate",
                     "pseo.participant", "took.ACT", "ACTCompositeScore.filled", "ap.exam",
                     "cte.achievement", "attended.ps",
                     "persistent.non_sustained", "persistent.no_record")

colSums(is.na(master.disc[, model_vars_disc]))

table(master.disc$persistent.non_sustained)
table(master.disc$persistent.no_record)

master.disc %>%
  group_by(english.learner) %>%
  summarise(pct_missing_pseo = mean(is.na(pseo.participant)), n = n())


```

<br>

# Modeling: persistent non-sustained WF

## Model 1: Demographics and High School Characteristics

This model estimates the association between demographic characteristics and district context with persistent non-sustained employment (4-5 of the first 5 years post-education-exit), before any measure of academic engagement or preparation is introduced (n = 10,310).

**This model does not fully replicate the bivariate pattern, and that discrepancy is worth flagging directly rather than smoothing over.**

**Gender behaves as expected — strong and significant.** Female graduates show approximately 34% higher odds of persistent non-sustained employment relative to male graduates (OR ≈ 1.34, CI: 1.204-1.490, p \< .001), consistent with Gender being one of the two strongest bivariate predictors of instability.

**Economic status, by contrast, is not significant here** (OR ≈ 1.07, CI: 0.956-1.195, p = .243) — despite being the *other* strongest bivariate predictor of non-sustained persistence. This is a real departure from the bivariate finding, not a rounding difference. One plausible explanation: economic disadvantage's bivariate relationship with instability may be substantially confounded with race and disability status, both of which are now controlled for in this same model — but this is a hypothesis to note, not a conclusion to assert without further testing.

**Special Education status is significant here** (OR ≈ 1.36, CI: 1.189-1.562, p \< .001) — the opposite of what the bivariate analysis suggested, where Special Ed was one of the *weakest* predictors of instability specifically (its effect was concentrated almost entirely in no-record persistence, not non-sustained persistence). This reversal is worth watching closely as later blocks are added, since it may indicate a suppression effect — Special Ed's true relationship with instability may have been masked in the bivariate comparison by a confounding variable not yet identified.

**Race/Ethnicity shows significant effects for three of four groups, generally consistent with or stronger than the bivariate pattern.** American Indian graduates show higher odds (OR ≈ 1.37, consistent with the bivariate finding that AI graduates' disadvantage is instability-driven). Black graduates show the largest race-based effect in this model (OR ≈ 1.68), also consistent with the bivariate direction. Asian/Pacific Islander graduates show significantly *lower* odds (OR ≈ 0.56) — the same direction found bivariately (Asian/PI showed the lowest persistent-instability rate of any group). Hispanic graduates show no significant effect here (OR ≈ 1.04, CI crosses 1), consistent with the bivariate finding that this group's disadvantage was concentrated in no-record status rather than instability.

**District characteristics remain uniformly non-significant**, exactly as the bivariate analysis found: Dem_Desc (both categories), average local wages, and average local unemployment rate all show confidence intervals crossing 1 with no meaningful effect size.

**Overall Interpretation**

Two of the five demographic predictors — Economic status and Special Education status — do not behave as the bivariate analysis would predict once controlled for alongside the others, and in opposite directions from each other. Gender and the racial disparities largely replicate the bivariate pattern. District/geographic context remains entirely non-predictive, consistent with every other model in this project's disconnection work. The Economic status and Special Ed reversals are the two findings to track most closely once high school accomplishments are added in Model 2 — it's possible one or both of these patterns shift again once academic engagement variables, which correlate with both economic status and disability status, are accounted for.

**Model fit:** AIC = 9,092.4; residual deviance = 9,066.4 on 10,297 df (null deviance = 9,162.9 on 10,309 df); n = 10,310.

```{r}
model1_non_sustained <- glm(
  persistent.non_sustained ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    avg.wages.pct.state +
    avg.unemp.rate,
  family = binomial,
  data = master.disc
)

summary(model1_non_sustained)
nobs(model1_non_sustained)
AIC(model1_non_sustained)
exp(cbind(OR = coef(model1_non_sustained), confint(model1_non_sustained)))
```

<br>

## Model 2: Add High School Accomplishments

This model adds PSEO participation, ACT participation and score, AP exam, and CTE achievement to the demographic and district baseline (n = 10,128; 182 observations dropped for PSEO missingness, matching the expected small residual gap identified earlier).

Special Education status strengthens rather than attenuating — the reversal from Model 1 is not explained by academic engagement. OR increases slightly, from 1.36 in Model 1 to 1.45 here (CI: 1.255-1.676, p \< .001). This rules out the hypothesis raised after Model 1 that Special Ed's unexpected positive association with instability might be a proxy for weaker academic preparation — controlling for accomplishments directly does not reduce the effect at all.

Economic status remains non-significant (OR ≈ 1.12, CI: 0.995-1.254, p = .062), consistent with Model 1 — its bivariate relationship with instability continues not to hold up once other variables are controlled.

PSEO participation is significant, and in the adverse direction found bivariately. OR ≈ 1.21 (CI: 1.068-1.372, p = .003) — PSEO participants show higher, not lower, odds of persistent non-sustained employment. This directly confirms the bivariate finding that PSEO's protective effect on overall disconnection runs almost entirely through avoiding no-record status, while showing a mild adverse relationship with instability specifically.

ACT participation and score both fall short of conventional significance. Taking the ACT shows a protective direction (OR ≈ 0.694, CI: 0.468-1.027, p = .069), and ACT composite score shows the expected adverse direction among test-takers (OR ≈ 1.018 per point, CI: 0.999-1.038, p = .066) — but neither clears the p \< .05 threshold. Both are worth watching in Model 3 rather than dismissing outright, given how close each sits to significance.

CTE achievement and AP exam both remain non-significant. Neither CTE category holds up (Concentrator/Completor: OR ≈ 0.92, p = .266; Participant: OR ≈ 0.94, p = .352). AP exam similarly shows no significant relationship (OR ≈ 1.21, p = .128).

English learner status remains non-significant, in the same direction as Model 1 (OR ≈ 0.97, p = .797) — consistent with the bivariate finding that English learner status was never a strong predictor of instability specifically.

Gender, Race/Ethnicity (AI and Black), and district/local-context variables remain essentially unchanged in direction and significance from Model 1.

Overall Interpretation

Adding high school accomplishments significantly improves model fit (likelihood ratio test: Δdeviance = 26.06, df = 6, p \< .001; AIC decreases from 9,092.4 to 8,896.2). The Special Education reversal identified in Model 1 is confirmed, not explained away, by this block — it is not a proxy for academic underpreparation. PSEO shows a genuine reversal from its bivariate relationship once controlled for other accomplishments. ACT participation and score both trend in the expected protective/adverse directions but do not reach significance here. CTE and AP exam, both real bivariate predictors of instability, do not hold up once entered alongside the other accomplishment variables.

Model fit: AIC = 8,896.2 (down from 9,092.4 in Model 1); Δdeviance = 26.06 on 6 df, p \< .001; n = 10,128 in both models.

```{r}
model2_non_sustained <- glm(
  persistent.non_sustained ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    avg.wages.pct.state +
    avg.unemp.rate +
    pseo.participant +
    took.ACT +
    ACTCompositeScore.filled +
    ap.exam +
    cte.achievement, 
  family = binomial,
  data = master.disc
)

nobs(model1_non_sustained)
nobs(model2_non_sustained)
AIC(model1_non_sustained)
AIC(model2_non_sustained)
summary(model2_non_sustained)
exp(cbind(OR = coef(model2_non_sustained), confint(model2_non_sustained)))

model1_non_sustained_matched <- glm(
  persistent.non_sustained ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    avg.wages.pct.state +
    avg.unemp.rate,
  family = binomial,
  data = master.disc[!is.na(master.disc$pseo.participant), ]
)

nobs(model1_non_sustained_matched)  # should now match model2_non_sustained's 10,128

anova(model1_non_sustained_matched, model2_non_sustained, test = "Chisq")
```

<br>

## Model 3: Add Postsecondary Attempt

This model adds attended.ps to the full specification (n = 10,128, same complete-case sample as Model 2).

Attempting postsecondary education shows a strong, independent protective effect on top of demographics, district context, and high school accomplishments — confirming this variable adds real explanatory power beyond what CTE, PSEO, and ACT already capture. OR ≈ 0.715 (CI: 0.632-0.808, p \< .001): students who attempted postsecondary education, even without completing it, show roughly 29% lower odds of persistent non-sustained employment than those who never attempted it. This confirms the bivariate finding that attended.ps was the one variable affecting both disconnection mechanisms — its instability-side effect survives full multivariate control, not just partial control.

PSEO's adverse effect from Model 2 persists essentially unchanged (OR ≈ 1.22, CI: 1.076-1.384, p = .002) — even with postsecondary attempt now in the model, PSEO participation continues to show a mild, real association with higher odds of persistent instability, not lower.

Took-ACT and ACT score both weaken further and no longer approach significance. Took-ACT moves from marginal in Model 2 (p = .069) to clearly non-significant here (OR ≈ 0.734, CI: 0.495-1.086, p = .123). ACT composite score remains only marginal (OR ≈ 1.018, p = .074, CI: 0.998-1.037). This suggests some of ACT's apparent relationship with instability in Model 2 was capturing information that attended.ps explains more directly — sensible, since taking the ACT and attempting postsecondary education are closely related behaviors.

Special Education status remains essentially unchanged and still significant (OR ≈ 1.43, CI: 1.232-1.647, p \< .001) — confirming again that this effect is not explained by postsecondary attempt any more than it was explained by high school accomplishments in Model 2. This is now the third model in a row where Special Ed's adverse association with instability holds up.

Economic status and English learner status both remain non-significant (economic status: OR ≈ 1.11, p = .080; English learner: OR ≈ 0.95, p = .614), consistent with every prior model — neither shows any relationship with persistent instability once other variables are controlled for.

Gender and the Race/Ethnicity effects (AI, Black, Asian/PI) remain stable in direction and magnitude relative to Model 2. Hispanic graduates remain non-significant (OR ≈ 1.08, p = .719).

**Overall Interpretation**

Adding postsecondary attempt significantly improves model fit (likelihood ratio test: Δdeviance = 29.59, df = 1, p \< .001; AIC decreases from 8,896.2 to 8,868.7). This confirms attended.ps as a genuine, independent predictor of instability specifically, not merely a proxy for the academic-preparation variables already in the model — it earns its place in the specification on its own merits. Across all three models, Special Education status is the clearest and most stable finding of this analysis: a real, persistent, positive predictor of instability that survives every level of control tested, directly contradicting its weak bivariate relationship with this specific mechanism. Economic status and English learner status, by contrast, never emerge as significant predictors of instability at any stage, confirming the bivariate finding that whatever drives these two groups' overall disconnection, it is not this mechanism. ACT participation and score trend in the expected direction throughout but do not hold up as independent, reliable predictors once postsecondary attempt is accounted for.

Model fit: AIC = 8,868.7 (down from 8,896.2 in Model 2); Δdeviance = 29.59 on 1 df, p \< .001; n = 10,128 across all three models.

```{r}
model3_non_sustained <- glm(
  persistent.non_sustained ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    avg.wages.pct.state +
    avg.unemp.rate +
    pseo.participant +
    took.ACT +
    ACTCompositeScore.filled +
    ap.exam +
    cte.achievement +
    attended.ps,
  family = binomial,
  data = master.disc
)

nobs(model2_non_sustained)
nobs(model3_non_sustained)
AIC(model2_non_sustained)
AIC(model3_non_sustained)
summary(model3_non_sustained)
exp(cbind(OR = coef(model3_non_sustained), confint(model3_non_sustained)))

# likelihood ratio test comparing Model 2 to Model 3
anova(model2_non_sustained, model3_non_sustained, test = "Chisq")
```

<br>

## Summary: Multivariate Confirmation — Persistent Non-Sustained Employment

Summary: Multivariate Confirmation — Persistent Non-Sustained Employment

Across all three models, several bivariate findings held up, but one demographic effect reversed direction in a way worth treating as this analysis's central discovery.

Special Education status is a genuine, positive predictor of instability — the opposite of what the bivariate analysis suggested. Bivariately, Special Ed appeared to be almost entirely a no-record phenomenon with little relationship to instability specifically. Once demographics, district context, high school accomplishments, and postsecondary attempt are all controlled for, Special Ed shows a strong, stable, significant association with higher odds of persistent non-sustained employment (OR ≈ 1.36–1.45 across all three models) that never attenuates, no matter what else is added to the model. This is a real reversal, not a bivariate artifact resolving itself — Special Ed's relationship with instability appears to have been suppressed in the bivariate comparison, not masked by something the multivariate model then explains away.

Economic status and English learner status never emerge as predictors of instability at any stage, confirming the bivariate finding cleanly: whatever drives these two groups toward overall disconnection, it is not the instability mechanism.

PSEO participation behaves in a way that runs counter to its usual "broadly protective" role elsewhere in this project. It shows a small but real adverse association with instability (OR ≈ 1.21–1.22, consistent with the bivariate finding), holding steady across every model once other variables are controlled.

ACT participation and score both trend in the expected protective/adverse directions, but neither holds up as a reliable, independent predictor. Both sit at the edge of significance once high school accomplishments are added, and both fall further from significance once postsecondary attempt enters the model — suggesting whatever relationship they have with instability is substantially overlapping with attempting postsecondary education itself, rather than representing a distinct, independent effect.

Attempting postsecondary education is confirmed as an independent, genuine predictor of reduced instability, surviving full control alongside CTE, PSEO, and ACT (OR ≈ 0.72 in the final model) — this is the clearest bivariate finding to survive multivariate testing fully intact.

Race/Ethnicity effects for American Indian, Black, and Asian/Pacific Islander graduates all replicate their bivariate direction and remain significant across every model — American Indian and Black graduates show elevated odds of instability, Asian/Pacific Islander graduates show reduced odds, consistent with the bivariate finding that these groups' disadvantage (or advantage) is genuinely instability-driven. Hispanic graduates show no significant effect at any stage, consistent with the bivariate finding that this group's disadvantage is concentrated in no-record status rather than instability.

CTE achievement and AP exam, both real bivariate predictors of instability, do not survive multivariate control — neither reaches significance once entered alongside the other accomplishment variables, suggesting their bivariate relationships were substantially explained by other correlated variables now in the model.

Local geographic and economic context (Dem_Desc, wages, unemployment) remains entirely non-predictive across all three models, confirming this project's consistent finding that local labor market conditions do not meaningfully shape workforce instability for this population.

<br>

# Modeling: persistent No MN emp record

## Model 1: Demographics and High School Characteristics

This model estimates the association between demographic characteristics and district context with persistent no-record status (4-5 of the first 5 years with no Minnesota employment record), before any measure of academic engagement is introduced (n = 10,310).

**English learner status shows the strongest effect in the model, exactly as the bivariate analysis predicted.** OR ≈ 1.94 (CI: 1.631-2.318, p \< .001) — English learners show nearly double the odds of persistent no-record status relative to non-English-learners. This is the clearest confirmation so far that English learner status's disconnection story is genuinely a no-record phenomenon, and it survives full demographic control at nearly its full bivariate strength.

**Race/Ethnicity confirms the bivariate mechanism split precisely.** Hispanic (OR ≈ 2.42, CI: 1.745-3.336, p \< .001) and Asian/Pacific Islander graduates (OR ≈ 2.41, CI: 1.714-3.371, p \< .001) both show more than double the odds of persistent no-record status — matching the bivariate finding that these two groups' disadvantage concentrates in this mechanism specifically. Black graduates show a real but more moderate effect (OR ≈ 1.47, p = .003). American Indian graduates are only marginally significant (OR ≈ 1.18, p = .068) — consistent with the bivariate finding that AI graduates' disadvantage lives almost entirely in the instability mechanism, not this one.

**Special Education status is *not* significant here** (OR ≈ 1.10, CI: 0.959-1.252, p = .172) — a real surprise given that Special Ed was one of the two strongest bivariate predictors of no-record persistence. Combined with Model 1's finding for non-sustained employment (where Special Ed was strongly significant), this suggests Special Ed's disconnection story may be almost entirely captured by the instability mechanism once demographics are controlled for together, not the no-record mechanism the bivariate comparison initially suggested. This is worth tracking closely through Models 2 and 3.

**Economic status is significant, but in the protective direction** (OR ≈ 0.777, CI: 0.701-0.861, p \< .001) — economically disadvantaged graduates show *lower* odds of persistent no-record status. This matches a pattern already visible in the bivariate work (disadvantage showed a modestly *lower* no-record rate there too), now confirmed as a real, significant effect once other demographics are controlled for.

**Gender is significant and protective** (OR ≈ 0.887, CI: 0.802-0.981, p = .020) — female graduates show modestly lower odds of persistent no-record status, a real effect not clearly visible in the bivariate comparison (where Gender was one of the weakest predictors overall).

**District, wage, and unemployment variables remain largely non-significant**, though average local wages is worth flagging as a near-miss (OR ≈ 2.43, CI: 0.896-6.577, p = .081) — a large point estimate with a very wide confidence interval. Given that wages showed a clean null result throughout the bivariate analysis, this shouldn't be over-interpreted yet, but it's worth watching whether it firms up or fades as later blocks are added.

**Overall Interpretation**

This model confirms the central mechanism-split predictions for English learner status and for the Hispanic/Asian-PI/American-Indian racial pattern — all replicate their bivariate direction and magnitude closely. The one genuine surprise is Special Education status, which does not hold up as a no-record predictor here despite being one of the strongest bivariate predictors of this exact outcome — raising the question of whether Special Ed's disconnection story is really an instability story once demographics are considered together, not a no-record story as the bivariate comparison suggested. Economic status and Gender both emerge as real, if modest, protective predictors of no-record status — a nuance not fully visible bivariately.

**Model fit:** AIC = 10,201.6; residual deviance = 10,176 on 10,297 df (null deviance = 10,357 on 10,309 df); n = 10,310.

```{r}
model1_no_record <- glm(
  persistent.no_record ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    avg.wages.pct.state +
    avg.unemp.rate,
  family = binomial,
  data = master.disc
)

summary(model1_no_record)
nobs(model1_no_record)
AIC(model1_no_record)
exp(cbind(OR = coef(model1_no_record), confint(model1_no_record)))
```

<br>

## Model 2: Add High School Accomplishments

Model 2: Add High School Accomplishments

This model adds PSEO participation, ACT participation and score, AP exam, and CTE achievement to the demographic and district baseline (n = 10,128).

English learner status remains the strongest effect in the model, only modestly reduced from Model 1. OR ≈ 1.79 (CI: 1.489-2.148, p \< .001) — English learners continue to show substantially higher odds of persistent no-record status once high school accomplishments are accounted for.

Race/Ethnicity effects remain strong and largely intact. Hispanic graduates show OR ≈ 2.32 (CI: 1.663-3.223, p \< .001), and Asian/Pacific Islander graduates show OR ≈ 2.24 (CI: 1.579-3.156, p \< .001) — both remain more than double the odds of persistent no-record status relative to White graduates, consistent with Model 1. Black graduates strengthen to significance here (OR ≈ 1.41, CI: 1.081-1.826, p = .010). American Indian graduates remain non-significant (OR ≈ 1.15, p = .123), consistent with the bivariate finding that AI graduates' disadvantage concentrates in the instability mechanism, not this one.

Took-ACT shows a large, highly significant protective effect. OR ≈ 0.265 (CI: 0.179-0.391, p \< .001) — students who took the ACT show roughly 74% lower odds of persistent no-record status. ACT composite score shows the expected adverse direction among test-takers (OR ≈ 1.048 per point, p \< .001, CI: 1.028-1.068) — higher scores predicting somewhat higher odds of no-record status, consistent with the threshold-effect pattern found bivariately.

PSEO participation is significantly protective (OR ≈ 0.751, CI: 0.664-0.848, p \< .001), confirming the bivariate finding that PSEO's protective role for this population runs heavily through the no-record mechanism.

CTE achievement shows a strong, significant protective effect for both categories — Concentrator/Completor (OR ≈ 0.686, p \< .001) and Participant (OR ≈ 0.747, p \< .001) — a more substantial relationship than CTE showed against instability.

Average local unemployment rate becomes significant here (OR ≈ 0.923 per unit, CI: 0.884-0.964, p \< .001) — higher local unemployment predicting lower odds of persistent no-record status, the same counterintuitive direction found in the bivariate analysis.

Special Education status remains non-significant (OR ≈ 0.95, p = .501), extending the Model 1 finding — Special Ed shows no relationship with no-record status at this stage either, reinforcing that its disconnection story runs through the instability mechanism instead.

Economic status and Gender both remain significant and protective (economic status: OR ≈ 0.734, p \< .001; Gender: OR ≈ 0.872, p = .010), consistent with Model 1. AP exam remains non-significant (OR ≈ 1.01, p = .913).

Overall Interpretation

Adding high school accomplishments substantially improves model fit (likelihood ratio test: Δdeviance = 146.75, df = 6, p \< .001; AIC decreases from 10,201.6 to 9,843.3). English learner status and the Hispanic, Black, and Asian/Pacific Islander racial disparities identified in Model 1 all remain strong, significant predictors of persistent no-record status — none attenuate to non-significance once high school accomplishments are added. ACT participation and score, PSEO participation, and CTE achievement all confirm substantial, independent relationships with no-record status, consistent with the bivariate finding that these variables' broad protective effects for this population run heavily through avoiding no-record status specifically. Special Education status remains non-significant, extending the pattern from Model 1.

Model fit: AIC = 9,843.3 (down from 10,201.6 in Model 1); Δdeviance = 146.75 on 6 df, p \< .001; n = 10,128 in both models.

```{r}
model2_no_record <- glm(
  persistent.no_record ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    avg.wages.pct.state +
    avg.unemp.rate +
    pseo.participant +
    took.ACT +
    ACTCompositeScore.filled +
    ap.exam +
    cte.achievement,
  family = binomial,
  data = master.disc
)

nobs(model1_no_record)
nobs(model2_no_record)
AIC(model1_no_record)
AIC(model2_no_record)
summary(model2_no_record)
exp(cbind(OR = coef(model2_no_record), confint(model2_no_record)))

# Model 1 refit on Model 2's complete-case sample, for a valid LRT
# (same fix used for the non-sustained sequence, since PSEO's 182 missing
# rows will again shrink Model 2's sample relative to Model 1)
model1_no_record_matched <- glm(
  persistent.no_record ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    avg.wages.pct.state +
    avg.unemp.rate,
  family = binomial,
  data = master.disc[!is.na(master.disc$pseo.participant), ]
)

nobs(model1_no_record_matched)  # should match model2_no_record's n

anova(model1_no_record_matched, model2_no_record, test = "Chisq")
```

<br>

## Model 3: Add Postsecondary Attempt

This model adds attended.ps to the full specification (n = 10,128, same complete-case sample as Model 2).

Attempting postsecondary education shows a strong, independent protective effect — the second-largest effect in this entire model, after took.ACT. OR ≈ 0.530 (CI: 0.470-0.597, p \< 2e-16): students who attempted postsecondary education, even without completing it, show roughly 47% lower odds of persistent no-record status than those who never attempted it. This confirms attended.ps's bivariate strength on this outcome is real and independent, not an artifact of overlap with PSEO, ACT, or CTE.

Took-ACT remains the single largest effect in the model, essentially unchanged from Model 2 (OR ≈ 0.295, CI: 0.199-0.435, p \< .001). ACT composite score also remains stable (OR ≈ 1.047 per point, p \< .001, CI: 1.027-1.067). Neither is meaningfully explained by postsecondary attempt — both hold their own alongside it.

English learner status remains a strong, significant predictor, though it attenuates modestly from Model 2. OR ≈ 1.71 (CI: 1.419-2.051, p \< .001), down from 1.79 — a real but small reduction, not a collapse toward non-significance.

Race/Ethnicity effects hold essentially steady from Model 2. Hispanic (OR ≈ 2.29, p \< .001) and Asian/Pacific Islander (OR ≈ 2.12, p \< .001) graduates remain strongly elevated. Black graduates remain significant (OR ≈ 1.44, p = .007). American Indian graduates remain non-significant (OR ≈ 1.16, p = .103), consistent with every prior model.

PSEO participation and CTE achievement both remain essentially unchanged from Model 2 in direction, magnitude, and significance — none of these accomplishment variables' relationships with no-record status are meaningfully altered by adding postsecondary attempt to the model.

Special Education status remains non-significant (OR ≈ 0.93, p = .294), consistent with both prior models — Special Ed shows no relationship whatsoever with no-record status, at any level of control.

Economic status and Gender remain significant and protective (economic status: OR ≈ 0.722, p \< .001; Gender: OR ≈ 0.882, p = .018). Average local unemployment rate, significant in Model 2, weakens to marginal here (OR ≈ 0.963, CI: 0.922-1.006, p = .094) — worth noting as a real shift rather than treating the Model 2 result as the final word on this variable. AP exam remains non-significant (OR ≈ 0.95, p = .694).

Overall Interpretation

Adding postsecondary attempt significantly improves model fit (likelihood ratio test: Δdeviance = 113.57, df = 1, p \< .001; AIC decreases from 9,843.3 to 9,731.7), confirming attended.ps as a large, genuine, independent predictor of no-record status. Across the full three-model sequence, English learner status and the Hispanic, Black, and Asian/Pacific Islander racial disparities identified in Model 1 all remain significant predictors of persistent no-record status through every stage of control — none attenuate to non-significance. Took-ACT, ACT composite score, PSEO participation, and CTE achievement all confirm substantial, independent relationships with no-record status. Special Education status remains non-significant throughout the full sequence.

Model fit: AIC = 9,731.7 (down from 9,843.3 in Model 2); Δdeviance = 113.57 on 1 df, p \< .001; n = 10,128 across all three models.

```{r}
model3_no_record <- glm(
  persistent.no_record ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    avg.wages.pct.state +
    avg.unemp.rate +
    pseo.participant +
    took.ACT +
    ACTCompositeScore.filled +
    ap.exam +
    cte.achievement +
    attended.ps,
  family = binomial,
  data = master.disc
)

nobs(model2_no_record)
nobs(model3_no_record)
AIC(model2_no_record)
AIC(model3_no_record)
summary(model3_no_record)
exp(cbind(OR = coef(model3_no_record), confint(model3_no_record)))

# likelihood ratio test comparing Model 2 to Model 3
anova(model2_no_record, model3_no_record, test = "Chisq")
```

<br>

## Summary: Multivariate Confirmation — Persistent No Minnesota Employment Record

Across all three models, this analysis delivered a clear, stable central finding, along with several supporting results.

English learner status is a strong, persistent predictor of no-record status across every stage of control. OR moves from 1.94 (Model 1) to 1.79 (Model 2) to 1.71 (Model 3) — a modest, gradual decline, but English learners remain at nearly double the odds of persistent no-record status relative to non-English-learners even in the fully specified model (p \< .001 throughout). This is the clearest and most consequential finding in this analysis: English learners' disadvantage in disappearing from Minnesota's employment records is not explained away by high school accomplishments or postsecondary attempt — it holds up as an independent, substantial effect.

The same pattern holds for Hispanic and Asian/Pacific Islander graduates. Both groups show large, significant effects in the demographics-only model (Hispanic OR ≈ 2.42; Asian/PI OR ≈ 2.41) that remain strong and significant through the full model sequence (Hispanic OR ≈ 2.29-2.32; Asian/PI OR ≈ 2.12-2.24, both p \< .001 throughout). Black graduates also remain significant at every stage (OR ≈ 1.41-1.47, p \< .01 throughout). American Indian graduates remain non-significant or marginal across all three models (OR ≈ 1.15-1.18), consistent with the bivariate finding that AI graduates' disadvantage concentrates in the instability mechanism rather than this one.

Took-ACT is the single most powerful predictor found anywhere in this analysis. OR remains in the 0.26-0.30 range across both models where it appears — students who took the ACT show roughly 70-74% lower odds of persistent no-record status. ACT composite score confirms the expected threshold-effect pattern, with higher scores among test-takers predicting modestly higher odds of no-record status (OR ≈ 1.05 per point, p \< .001 in both models).

Attempting postsecondary education is confirmed as a large, genuine, independent predictor of reduced no-record risk, holding up fully alongside every other variable (OR ≈ 0.53 in the final model) — a larger effect than it showed in the non-sustained-employment sequence, consistent with the bivariate finding that attended.ps's protective effect concentrates more heavily in this mechanism.

PSEO participation and CTE achievement both confirm substantial, stable protective effects across every model in which they appear, consistent with the bivariate finding that these variables' broad protective role for this population runs heavily through avoiding no-record status specifically.

Special Education status shows no relationship with no-record status at any stage — a clean, consistent null result across all three models, confirming that Special Ed's disconnection story (established in the companion non-sustained-employment analysis) is an instability story, not a no-record story, despite the bivariate comparison initially suggesting otherwise.

Economic status and Gender both emerge as modest, genuine, stable protective predictors, holding up consistently across all three models. Average local unemployment rate is significant in Model 2 but weakens to marginal in Model 3 once postsecondary attempt is added — a real shift worth tracking rather than treating the Model 2 result as final. AP exam and district geographic classification remain non-significant throughout.

<br>

# Multivariate Confirmation: Side-by-Side Comparison of the Two Disconnection Mechanisms

**What the Bivariate Analysis Predicted vs. What the Multivariate Models Confirmed**

| Variable | Bivariate expectation | Confirmed for Non-Sustained? | Confirmed for No-Record? |
|----|----|----|----|
| Special Education status | No-record mechanism | **No — reversed.** Strong, stable predictor of instability instead (OR ≈ 1.36–1.45) | **No.** Non-significant at every stage |
| English learner status | No-record mechanism | Confirmed null (never significant) | **Confirmed, and robust.** Strong at every stage (OR ≈ 1.71–1.94), declining only modestly as controls are added |
| Economic status | Instability mechanism | **No.** Non-significant at every stage | **Reversed role — protective, no-record.** Significant, protective (OR ≈ 0.72–0.78) |
| Gender | Instability mechanism | Confirmed (OR ≈ 1.31–1.34, adverse) | **New finding.** Also significant here, protective (OR ≈ 0.87–0.89) |
| American Indian | Instability mechanism | Confirmed (OR ≈ 1.37–1.39) | Confirmed null (marginal at best) |
| Hispanic | No-record mechanism | Confirmed null (never significant) | **Confirmed, and robust.** Strong at every stage (OR ≈ 2.29–2.42) |
| Asian/Pacific Islander | No-record mechanism | **Reversed.** Significantly *reduced* odds of instability (OR ≈ 0.55–0.60) | **Confirmed, and robust.** Strong at every stage (OR ≈ 2.12–2.41) |
| Black | Both, moderately | Confirmed (OR ≈ 1.68–1.74) | **Confirmed, and robust.** Significant at every stage (OR ≈ 1.41–1.47) |
| Attended postsecondary | Both mechanisms | Confirmed, independent (OR ≈ 0.72) | Confirmed, independent and larger (OR ≈ 0.53) |
| ACT participation/score | No-record mechanism | **Weakens to non-significant** once postsecondary attempt is added | **Confirmed precisely.** Crossover pattern intact and strong throughout (OR ≈ 0.26–0.30 for participation) |
| PSEO participation | Ambiguous/adverse | Confirmed adverse (OR ≈ 1.21–1.22) | Confirmed protective (OR ≈ 0.75–0.76) |
| CTE achievement | Not tested against combined measure this way | Attenuates to non-significance | **Confirmed, strongly protective** (OR ≈ 0.69–0.75 across categories) |
| Local wages/unemployment/Dem_Desc | Null everywhere | Confirmed null throughout | Mostly null; unemployment rate is significant in Model 2 but **fades to marginal** once postsecondary attempt is added |

## **The Story This Comparison Tells**

**The core mechanism-split finding survives largely intact.** The bivariate analysis suggested Special Education and English learner status were both primarily no-record phenomena. The multivariate models confirm this only partially: **Special Education status turns out to be an instability phenomenon, not a no-record phenomenon at all** — the opposite of what the bivariate comparison suggested. **English learner status, by contrast, is confirmed as a genuine, robust no-record phenomenon** — its effect declines only modestly across the three models and remains highly significant throughout.

**Race/Ethnicity delivers the most important finding of this analysis: the Hispanic, Black, and Asian/Pacific Islander disparities in no-record status are real, independent, and do not attenuate.** All three groups remain significantly elevated at every stage of control, including after high school accomplishments and postsecondary attempt are both accounted for.

**Asian/Pacific Islander graduates show a striking divergence between the two mechanisms** — significantly *reduced* odds of instability, but among the *highest* odds of no-record status of any group. This confirms that different racial/ethnic groups experience genuinely different, sometimes opposite, forms of disadvantage depending on which outcome is examined.

**`attended.ps`, PSEO, and CTE achievement all confirm strong, stable roles in the no-record mechanism specifically** — three of the clearest, most consistent findings in this analysis, with no attenuation or reversal at any stage.

## What This Means for the Report

Three things are ready to go into the report as confirmed, multivariate-tested findings: (1) postsecondary attempt, PSEO participation, and CTE achievement's mechanism-specific protective effects on no-record status; (2) **English learner status and the Hispanic, Black, and Asian/Pacific Islander disparities in no-record status are robust, independent findings** — they hold up fully once high school accomplishments and postsecondary attempt are controlled for, and should be presented as confirmed disparities, not as effects requiring further explanation; and (3) Special Education status's correction from a no-record story to an instability story. Given that the central race/ethnicity and English learner findings replicate cleanly across bivariate and multivariate stages, the report can present these as a single, consistent story.
