---
title: "Non-post-secondary graduates"
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
library(htmlwidgets)
library(RColorBrewer)
library(readxl)
library(janitor)
library(lubridate)
library(systemfonts)
reset_font_cache()
library(ggtext)
library(gmodels)
library(fastDummies)
library(car)
library(glmnet)
library(glmnetUtils)
library(pscl)
library(sjPlot)
library(ggforce)
library(rpart)
library(rpart.plot)
library(tidymodels)
library(caret)
library(themis)
library(ranger)
library(openxlsx)
```

```{r themes and shapefiles, include=FALSE}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", linewidth  = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.key.size = unit(1, "lines"),
        legend.margin = margin(0,0,0,0),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.text = element_text(margin = margin(l = 2)),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", linewidth = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))


theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
    select(5,6) %>%
    unique() %>%
    mutate(edr = str_replace(edr, "  ", " "),
           planning.region = str_replace(planning.region, " Minnesota", ""),
           planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
           edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

counties.regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
  rename(mif = `MIF Region`) %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Name = str_replace(Name, "Mcleod", "McLeod"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc) ,
         edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"))

counties.regions.1 <- counties.regions %>%
  mutate(statefp = "27",
         project.pr = ifelse(edr %in% c("EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 4 - West Central"), "Northwest",
                             ifelse(edr == "EDR 3 - Arrowhead", "Northeast",
                                    ifelse(edr %in% c("EDR 5 - North Central", "EDR 7E- East Central", "EDR 7W- Central"), "Central",
                                           ifelse(edr %in% c("EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 8 - Southwest"), "Southwest",
                                                  ifelse(edr %in% c("EDR 9 - South Central", "EDR 10 - Southeast"), "Southern", as.character(planning.region)))))),
         project.pr = fct_relevel(project.pr, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southern"))


color.ruca <- c("Entirely rural" = "#009933", "Town/rural mix" = "#99CC33", "Urban/town/rural mix" = "#CC9966", "Entirely urban" = "#754C29", "Minnesota" = "black")

color.pr <- c("Northwest" = 	"#4575b4", "Northeast" = "grey", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.edr.simple <- c("EDR 1" = "#b3cde3", "EDR 2" = "#8c96c6", "EDR 3" = "#fe9929", "EDR 4" = "#8856a7", "EDR 5" = "#810f7c", "EDR 6E" = "#e5f5f9", "EDR 6W" = "#bdc9e1", "EDR 7E" = "#99d8c9", "EDR 7W" = "#2ca25f", "EDR 8" = "#74a9cf", "EDR 9" = "#0570b0", "EDR 10" = "#d7301f", "EDR 11" = "#d8b365", "Minnesota" = "black")

color.pr.edr <- c ("Northwest" = "#4575b4","Northeast" = "#e0f3f8", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black", "EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365")

mn_counties <- st_read("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Shapefiles/County shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE) %>%
  rename(countyfp = FIPS_CODE)
```

```{r master}
master.original <- read_csv("Data/SLEDS/Masters/Master-report-dataset.csv")

kable(names(master.original))

kable(head(master.original))
```

# Define the subset

One of the aspects noticed during analysis is the percentage of individuals that never graduated college that also had meaningful employment in the region. The table below shows that 58% of the high school graduates in Northeast Minnesota never graduated college.

```{r}
data <- master.original %>%
  filter(ps.grad != "Attending ps") %>%
  group_by(ps.grad) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  mutate(pct = n / sum(n))

datatable(data, class = "cell-border stripe", filter = "top", rownames = FALSE,
          options = list(columnDefs = list(list(className = "dt-center", targets = 1:2)))) %>%
  formatCurrency(2, "", digits = 0) %>%
  formatPercentage(3, digits = 1)


```

We are going to examine, more closely, this 58% of the population and see if there are patterns in which they end up in different employment outcomes 1-, 5- and 10-years after graduating high school.

<br>

# Employment outcomes distribution

The following tables provide the percentage of non-college graduates that were categorized in the various employment outcomes at 1-, 5-, and 10-years after graduating high school.

**1 Year After Graduation (n = 23,733)**

One year out, non-college graduates are concentrated most heavily in non-sustained workforce participation, which accounts for over a third of the population — a clear plateau above every other category. The remaining three-quarters split fairly evenly across the other three outcomes: roughly one in five have no Minnesota employment record, about one in five are still attending postsecondary, and just under one in five have already achieved sustained regional employment. Sustained employment elsewhere in Minnesota is rare at this stage (5.6%).

**5 Years After Graduation (n = 17,452)**

By five years, the picture consolidates substantially. **Sustained regional employment grows to 26.3%**, up from 19.2% at Year 1 — a meaningful gain in stability. The "attending ps" category nearly disappears (4.2%), as expected, since most who were going to enroll in postsecondary have done so by now (and, being non-completers by definition, this group either finished stopping out or moved into a different outcome). Non-sustained employment shrinks somewhat (27.3%, down from 35.0%), suggesting some early instability resolves into steadier work. **Notably, the "No MN employment record" category grows rather than shrinks (29.6%, up from 20.8%)** — this is worth flagging as a real pattern rather than noise: a meaningful share of this population appears to be leaving the state's tracked labor force (or the state itself) as time passes, not settling into it.

**10 Years After Graduation (n = 10,082)**

At ten years, **"No MN employment record" becomes the single largest outcome (36.7%)** — continuing to grow from 20.8% (Year 1) to 29.6% (Year 5) to 36.7% (Year 10). This is the dominant trend across the entire time horizon for this population: an increasing share of non-college graduates fall outside Minnesota's tracked employment system entirely as more time passes. Sustained regional employment holds essentially steady from Year 5 to Year 10 (26.3% → 25.6%), suggesting those who achieve regional attachment by mid-career tend to stay attached. Sustained employment elsewhere in Minnesota continues to grow modestly (12.6% → 15.6%), and non-sustained employment continues to shrink (27.3% → 22.1%).

**Overall Interpretation**

Three patterns emerge clearly from comparing all three time points:

1.  **Regional employment attachment stabilizes early and holds.** Sustained WF-NE roughly plateaus after Year 5 (26.3% → 25.6%), suggesting the Year 1–5 window is where most regional attachment is established, not the Year 5–10 window.
2.  **Non-sustained employment is largely a Year 1 phenomenon that resolves over time.** It falls steadily across all three points (35.0% → 27.3% → 22.1%), consistent with early-career instability giving way to steadier employment as workers age into the labor market.
3.  **The most important — and possibly under-examined — trend is the steady, uninterrupted growth of "No MN employment record"** (20.8% → 29.6% → 36.7%), which by Year 10 is the single largest outcome of any category. This is not a Year 1 artifact that resolves; it compounds over the full ten-year window. This raises a real question worth flagging directly, since your organization is "not a recommending organization" but does surface implications: is this growth driven by out-of-state migration, informal/gig work outside the tracked system, or something else? The data as shown here can describe the pattern but can't distinguish between these explanations — worth noting as an open question for the multivariate stage or for readers interpreting the finding.

::: panel-tabset
## 1 year after graduation

```{r}
data <- master.original %>%
  filter(ps.grad == "No") %>%
  filter(grad.year.1 != "After 2023") %>%
  group_by(grad.year.1) %>%
  summarize(n = n()) %>%
  ungroup %>%
  mutate(pct = n / sum(n)) %>%
  write_csv(., "Data/ChatGPT/Non-college grads ct/Distribution/one-year-grad.csv")

datatable(data, class = "cell-border stripe", filter = "top", rownames = FALSE,
          options = list(columnDefs = list(list(className = "dt-center", targets = 1:2)))) %>%
  formatCurrency(2, "", digits = 0) %>%
  formatPercentage(3, digits = 1)
```

## 5 years after graduation

```{r}
data <- master.original %>%
  filter(ps.grad == "No") %>%
  filter(grad.year.5 != "After 2023") %>%
  group_by(grad.year.5) %>%
  summarize(n = n()) %>%
  ungroup %>%
  mutate(pct = n / sum(n)) %>%
  write_csv(., "Data/ChatGPT/Non-college grads ct/Distribution/five-year-grad.csv")

datatable(data, class = "cell-border stripe", filter = "top", rownames = FALSE,
          options = list(columnDefs = list(list(className = "dt-center", targets = 1:2)))) %>%
  formatCurrency(2, "", digits = 0) %>%
  formatPercentage(3, digits = 1)
```

## 10 years after graduation

```{r}
data <- master.original %>%
  filter(ps.grad == "No") %>%
  filter(grad.year.10 != "After 2023") %>%
  group_by(grad.year.10) %>%
  summarize(n = n()) %>%
  ungroup %>%
  mutate(pct = n / sum(n)) %>%
  write_csv(., "Data/ChatGPT/Non-college grads ct/Distribution/ten-year-grad.csv")

datatable(data, class = "cell-border stripe", filter = "top", rownames = FALSE,
          options = list(columnDefs = list(list(className = "dt-center", targets = 1:2)))) %>%
  formatCurrency(2, "", digits = 0) %>%
  formatPercentage(3, digits = 1)

```
:::

<br>

# Methodology

The above tables show the importance of understanding how non-college grads are navigating their post-high school pathways with a significant portion of them being disconnected from the regional workforce. To better understand these pathways, we will explore the relationships between independent variables (Demographic, High school characteristics, and High school accomplishments) along with workforce outcomes at 1-, 5- and 10-years after graduating high school.

Since we provided the descriptive statistics above, we now need to analyze which factors relate most strongly to outcomes, and then see if we can use those factors to build a predictive model.

For exploring the relationships we will do the following;

- For **categorical predictors** (e.g., Race/Ethnicity, Economic status, CTE achievement, PSEO participation), use cross-tabs with Cramer’s V and chi-square.
- For **continuous predictors** (e.g., ACT, MCA scores, total CTE courses, county-level variables), use ANOVA to compare means across the employment states.
- Run these analyses separately for grad.year.1, grad.year.5, grad.year.10 to see how relationships shift over time.

<br>

## Dependent variable

The dependent variable we will be exploring here is `grad.year.x`, meaning we are going to see what is related to each of the workforce participation categories at 1-, 5-, and 10-years after graduating. The categories for the dependent variable are;

- Sustained WF - NE
- Sustained WF - MN
- Non-sustained WF
- No MN emp record

## Independent variables

We are going to continue using the variables that we have used throughout this project.

- **Demographic:**
  - `Gender`
  - `RaceEthnicity`,
  - `economic.status`,
  - `SpecialEdStatus`,
  - `english.learner`
- **High school characteristics:**
  - `Dem_Desc`,
  - `avg.unemp.rate`,
  - `avg.wages.pct.state`
- **High school accomplishments:**
  - `pseo.participant`,
  - `ap.exam`
  - `took.ACT`,
  - `ACTCompositeScore`,
  - `MCA.M`,
  - `MCA.R`,
  - `MCA.S`,
  - `cte.achievement`,
  - `total.cte.courses.taken`
- **Post-secondary**
  - **`attended.ps`**

<br>

# Prep Data

```{r}
model.vars <- c("Gender", "RaceEthnicity", "economic.status", "SpecialEdStatus", "english.learner", "Dem_Desc", "avg.wages.pct.state", "avg.unemp.rate", "pseo.participant", "ap.exam", "took.ACT", "ACTCompositeScore", "MCA.M", "MCA.R", "MCA.S", "cte.achievement", "total.cte.courses.taken", "attended.ps", "grad.year.1", "grad.year.5", "grad.year.10")

master <- master.original %>%
  filter(ps.grad == "No") %>%
  select(all_of(model.vars))

master.1 <- master %>%
  filter(!grad.year.1 %in% c("Attending ps", "After 2023")) %>%
  droplevels() %>%
  mutate(grad.year.1 = fct_relevel(grad.year.1, "Sustained WF - NE", "Sustained WF - MN", "Non-sustained WF", "No MN emp record"))
  
master.5 <- master %>%
  filter(!grad.year.5 %in% c("Attending ps", "After 2023")) %>%
  droplevels() %>%
  mutate(grad.year.5 = fct_relevel(grad.year.5, "Sustained WF - NE", "Sustained WF - MN", "Non-sustained WF", "No MN emp record"))

master.10 <- master %>%
  filter(!grad.year.10 %in% c("Attending ps", "After 2023")) %>%
  droplevels() %>%
  mutate(grad.year.10 = fct_relevel(grad.year.10, "Sustained WF - NE", "Sustained WF - MN", "Non-sustained WF", "No MN emp record"))
```

# Demographics

<br>

## `Gender`

Gender shows a small but statistically significant relationship with workforce participation outcomes at all three time points (Year 1: χ² = 100.41, df = 3, p \< .001, Cramér's V = 0.072; Year 5: χ² = 52.23, p \< .001, V = 0.056; Year 10: χ² = 47.74, p \< .001, V = 0.069). By conventional Cramér's V benchmarks, this is a **weak** relationship at every time point — statistically real given the large sample sizes, but small in practical magnitude. This is consistent with Gender's role throughout the rest of this project: a real but modest predictor, not a primary driver.

**A consistent, stable pattern holds across all three time horizons, rather than strengthening or weakening with time:**

- **Men show higher sustained regional employment (Sustained WF - NE) than women at every time point** — a gap of roughly 2–5 percentage points that persists from Year 1 (24.6% vs. 22.7%) through Year 5 (28.9% vs. 25.4%) to Year 10 (27.7% vs. 22.8%).
- **Women show higher non-sustained employment than men at every time point**, and this gap is the largest and most consistent pattern in the data — 47.5% vs. 40.4% at Year 1, 31.1% vs. 26.5% at Year 5, and 24.7% vs. 20.1% at Year 10. Notably, this gap doesn't shrink proportionally as non-sustained employment overall declines with time; it remains roughly 5–7 percentage points at every stage.
- **"No MN employment record" starts as a male-leaning category but converges by Year 10.** Men show meaningfully higher rates at Year 1 (27.7% vs. 23.3%) and modestly higher at Year 5 (31.5% vs. 30.1%), but by Year 10 the two genders are essentially identical (36.7% vs. 36.8%) — this gap is the one pattern here that changes shape over time rather than holding steady.
- **Sustained employment elsewhere in Minnesota shows almost no gender gap at any point** (6.5% vs. 7.2% at Year 1, narrowing to near-parity by Year 5 and Year 10) — this outcome doesn't appear to be gendered at all.

**Overall Interpretation**

Gender is a weak but persistent predictor of workforce pathway among non-college graduates, with a stable, unchanging signature over the ten-year window: men are somewhat more likely to land in sustained regional employment, and women are somewhat more likely to remain in non-sustained employment, at every single time point measured. Unlike some patterns elsewhere in this project that intensify or attenuate with time, this gender gap looks essentially fixed — established by Year 1 and still present, at a similar magnitude, at Year 10. The one gender difference that does shift over time is "no MN employment record," which starts higher for men and converges to parity by Year 10.

::: panel-tabset
#### 1-year after graduation

```{r}
data <- master.1

tab_xtab(var.row = data$grad.year.1,
         var.col = data$Gender,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/Non-college bivariate/Demographics/gender-1.html")

```

#### 5-years after graduation

```{r}
data <- master.5

tab_xtab(var.row = data$grad.year.5,
         var.col = data$Gender,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/Non-college bivariate/Demographics/gender-5.html")
```

#### 10-years after graduation

```{r}
data <- master.10

tab_xtab(var.row = data$grad.year.10,
         var.col = data$Gender,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/Non-college bivariate/Demographics/gender-10.html")


```
:::

<br>

## `RaceEthnicity`

Race/Ethnicity shows a statistically significant relationship with workforce outcomes at all three time points (Year 1: χ² = 133.08, df = 12, p \< .001, Cramér's V = 0.052; Year 5: χ² = 185.00, V = 0.061; Year 10: χ² = 120.28, V = 0.063). As with Gender, this is a **weak** relationship by conventional benchmarks at every time point, though marginally stronger than Gender's effect and, unlike Gender, it strengthens slightly rather than staying flat (V = 0.052 → 0.061 → 0.063 from Year 1 to Year 10).

**A note on cell sizes:** AI, Asian/PI, Black, and Hispanic groups are all much smaller than White at every time point, and shrink further by Year 10 (Asian/PI n=148, Hispanic n=137, Black n=258). Percentages for these groups should be read with real caution, especially at Year 10 — a handful of individuals can shift a percentage substantially in groups this size.

**No MN employment record shows the widest and most persistent racial disparity of any outcome, at every time point.**

- **Asian/PI graduates** show the highest "no MN employment record" rate of any group at all three horizons: 42.6% (Year 1) → 47.3% (Year 5) → 58.1% (Year 10) — a rate that not only stays highest but *grows* the most sharply of any group over time.
- **Hispanic graduates** show a similar, though less extreme, pattern: 39.2% → 50.2% → 51.1%.
- **White graduates**, by contrast, show the lowest "no MN employment record" rate at Year 1 and Year 5 (24.6%, 30.2%), rising only to 36.0% by Year 10 — a much flatter trajectory than Asian/PI or Hispanic graduates.

This divergence is the standout finding in this variable: by Year 10, Asian/PI graduates are leaving Minnesota's tracked employment system (or the state) at nearly 1.6 times the rate of White graduates (58.1% vs. 36.0%), a gap that was far smaller at Year 1 (42.6% vs. 24.6%, roughly 1.7x — so actually the *ratio* is fairly stable, but the *absolute gap* widens considerably: 18 points at Year 1 to 22 points at Year 10).

**Sustained regional employment (WF-NE) shows a moderate, fairly stable gap favoring White graduates.**

- White graduates consistently show the highest or near-highest Sustained WF-NE rate (24.7% → 28.5% → 26.5%).
- Asian/PI graduates show the lowest Sustained WF-NE rate at Year 1 (12.2%) and Year 10 (10.8%), though this improves somewhat at Year 5 (18.4%) — the smallest sample size of any group at Year 10 makes this fluctuation worth treating cautiously rather than as a clear trend reversal.
- Black graduates show a similar pattern to Asian/PI — consistently below White graduates at all three points (14.9% → 15.9% → 12.0%).

**Non-sustained employment shows the opposite pattern from "no MN employment record" — it's highest for AI graduates, not Asian/PI.**

- AI graduates show the highest non-sustained rate at every time point (44.2% → 36.3% → 29.8%), notably higher than any other group at Year 1 and Year 5.
- This is a distinct pattern from Asian/PI and Hispanic graduates, whose disadvantage shows up primarily in "no MN employment record" rather than "non-sustained" — suggesting AI graduates' early struggles look more like unstable-but-present employment, while Asian/PI and Hispanic graduates' struggles look more like disconnection from the tracked labor market entirely.

**Overall Interpretation**

Race/Ethnicity is a weak-to-modest but real and slightly strengthening predictor of workforce pathway among non-college graduates. The most important and consistent finding is that "no MN employment record" is substantially more common for Asian/PI and Hispanic graduates than for White graduates at every time point, and this gap widens rather than narrows over the ten-year window — the opposite trajectory from the overall population pattern (where "no MN emp record" grows for everyone, but grows fastest for these two groups specifically). AI graduates show a meaningfully different pattern: their disadvantage shows up primarily as *non-sustained* employment rather than disconnection from the tracked system, suggesting a distinct mechanism from what's driving the Asian/PI and Hispanic gaps. Given the small subgroup sizes — particularly by Year 10 — these patterns should be flagged as suggestive rather than definitive until confirmed in the multivariate stage, and any messaging to workforce/economic developers should note the sample-size caveat explicitly for the smaller racial groups.

::: panel-tabset
#### 1-year after graduation

```{r}
data <- master.1 %>%
  filter(RaceEthnicity != "Unknown")

tab_xtab(var.row = data$grad.year.1,
         var.col = data$RaceEthnicity,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/Non-college bivariate/Demographics/RaceEthnicity-1.html")

```

#### 5-years after graduation

```{r}
data <- master.5 %>%
  filter(RaceEthnicity != "Unknown")

tab_xtab(var.row = data$grad.year.5,
         var.col = data$RaceEthnicity,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/Non-college bivariate/Demographics/RaceEthnicity-5.html")

```

#### 10-years after graduation

```{r}
data <- master.10 %>%
  filter(RaceEthnicity != "Unknown")

tab_xtab(var.row = data$grad.year.10,
         var.col = data$RaceEthnicity,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/Non-college bivariate/Demographics/RaceEthnicity-10.html")

master.original %>%
  filter(grad.year.10 != "After 2023") %>%
  filter(ps.grad == "No") %>%
  filter(RaceEthnicity != "Unknown") %>%
  select(grad.year.10, RaceEthnicity) %>%
  write_csv(., "Data/ChatGPT/Non-college grads ct/CT/Demographics/RaceEthnicity-10.csv")
```
:::

<br>

## `economic.status`

Economic status shows a statistically significant relationship with workforce outcomes at all three time points, and unlike Gender (flat) or Race/Ethnicity (mild strengthening), this relationship **strengthens the most clearly and consistently of any demographic variable examined so far** (Year 1: χ² = 19.45, df = 3, p \< .001, Cramér's V = 0.032; Year 5: χ² = 82.21, V = 0.070; Year 10: χ² = 69.64, V = 0.083). At Year 1 this is a very weak relationship — among the weakest seen in this analysis — but by Year 10 it has grown into a modest, clearly real effect, more than doubling in strength over the ten-year window.

**Sustained regional employment and non-sustained employment drive this pattern — not "no MN employment record," which is a notable contrast to Race/Ethnicity.**

- **Sustained WF-NE:** economically disadvantaged graduates show consistently lower rates than non-disadvantaged graduates, and the gap widens over time — a modest 1.9-point gap at Year 1 (22.9% vs. 24.8%) grows to 4.2 points at Year 5 (25.3% vs. 29.5%) and 4.7 points at Year 10 (23.1% vs. 27.8%).
- **Non-sustained WF:** the mirror image — disadvantaged graduates show higher rates at every point, and this gap also widens over time: 3.0 points at Year 1 (44.8% vs. 41.8%), 5.7 points at Year 5 (31.3% vs. 25.6%), 6.0 points at Year 10 (25.3% vs. 19.3%).
- **Sustained WF-MN:** a smaller but consistent gap favoring non-disadvantaged graduates, also widening with time (7.0% vs. 6.9% at Year 1, essentially negligible, growing to 16.6% vs. 14.5% by Year 10).
- **No MN employment record: essentially flat between groups at every time point** — 26.4% vs. 25.4% at Year 1, 31.0% vs. 30.9% at Year 5, 36.3% vs. 37.2% at Year 10. This is the key structural difference from Race/Ethnicity, where "no MN emp record" was the dominant driver of the racial disparity. Economic status doesn't meaningfully predict disconnection from the tracked labor market — it predicts the *quality/stability* of employment among those who remain in it.

<u>**Overall Interpretation**</u>

Economic status is a real but modest predictor whose influence grows steadily over the ten-year window, unlike Gender (stable) or Race/Ethnicity (mildly strengthening). The mechanism here looks distinct from Race/Ethnicity's: economically disadvantaged non-college graduates are not meaningfully more likely to disappear from Minnesota's tracked employment system, but they are consistently and increasingly more likely to end up in non-sustained rather than sustained employment, whether in the region or elsewhere in the state. This suggests economic disadvantage operates primarily on employment *stability and quality* rather than on labor-market disconnection or geographic mobility — a meaningfully different story than what emerged for race/ethnicity, and one worth keeping distinct in the eventual synthesis rather than treating all demographic disparities as pointing to the same underlying mechanism.

::: panel-tabset
#### 1-year after graduation

```{r}
data <- master.1

tab_xtab(var.row = data$grad.year.1,
         var.col = data$economic.status,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/Non-college bivariate/Demographics/economic-status-1.html")
```

#### 5-years after graduation

```{r}
data <- master.5

tab_xtab(var.row = data$grad.year.5,
         var.col = data$economic.status,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/Non-college bivariate/Demographics/economic-status-5.html")
```

#### 10-years after graduation

```{r}
data <- master.10

tab_xtab(var.row = data$grad.year.10,
         var.col = data$economic.status,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/Non-college bivariate/Demographics/economic-status-10.html")
```
:::

<br>

## `SpecialEdStatus`

Special Education status shows the **strongest relationship of any demographic variable examined so far** at every time point (Year 1: χ² = 291.63, df = 3, p \< .001, Cramér's V = 0.124; Year 5: χ² = 143.63, V = 0.093; Year 10: χ² = 110.92, V = 0.105) — noticeably larger than Gender (max V = 0.072), Race/Ethnicity (max V = 0.063), or Economic Status (max V = 0.083) at their respective peaks. Unlike Economic Status, which strengthened steadily over time, this relationship is strongest at Year 1, dips at Year 5, and partially recovers by Year 10 — a U-shaped pattern rather than a steady trend in either direction.

**Sustained regional employment and "no MN employment record" both show large, persistent gaps — this variable affects both mechanisms at once, unlike Economic Status (which affected employment stability but not disconnection) or Race/Ethnicity (which primarily affected disconnection).**

- **Sustained WF-NE:** Special Ed students show substantially lower rates at every time point, and the gap is large and remarkably stable — 8.2 points at Year 1 (17.8% vs. 26.0%), 5.6 points at Year 5 (23.2% vs. 28.8%), 4.6 points at Year 10 (22.0% vs. 26.6%). This is the largest Sustained WF-NE gap found for any demographic variable at Year 1.
- **"No MN employment record":** Special Ed students show substantially *higher* rates at every point — 33.6% vs. 23.1% at Year 1 (a 10.5-point gap, the single largest gap of any outcome/variable combination seen in this analysis so far), narrowing somewhat to 35.5% vs. 29.5% at Year 5 (6.0 points) and 39.7% vs. 35.9% at Year 10 (3.8 points).
- **Sustained WF-MN:** consistently lower for Special Ed students, and this gap widens rather than narrows over time — 5.3% vs. 7.5% at Year 1 (2.2 points) growing to 9.5% vs. 14.3% at Year 5 (4.8 points) and 10.4% vs. 17.0% at Year 10 (6.6 points), the opposite trajectory from the other three outcomes.
- **Non-sustained WF:** the one outcome with almost no gap at Year 1 (43.4% vs. 43.4%, essentially identical), but a real gap emerges by Year 5 (31.8% vs. 27.4%) and persists at Year 10 (27.9% vs. 20.5%) — Special Ed students increasingly more likely to be non-sustained as time goes on, the opposite pattern from every other outcome examined so far, where gaps tend to be largest at Year 1 and shrink with time.

**Overall Interpretation**

Special Education status is the strongest demographic predictor of workforce pathway examined in this analysis, and its effect is distinctive in an important way: it depresses *both* sustained regional employment and drives *higher* rates of disconnection from the tracked labor market, whereas Economic Status affected only employment stability and Race/Ethnicity's disparity was concentrated almost entirely in disconnection. This suggests Special Ed status operates through a broader or more compounding set of barriers than either of the other two variables examined so far. The narrowing of the Sustained WF-NE and "no MN emp record" gaps over time is a modestly encouraging pattern — early disparities partially resolve — but this is offset by the Sustained WF-MN and non-sustained gaps, which both *widen* with time, meaning the overall disparity doesn't simply fade with time so much as shift in where it shows up. Given how large this effect is relative to the other demographics, Special Ed status is a strong candidate for one of the central findings in the eventual synthesis for this population.

::: panel-tabset
#### 1-year after graduation

```{r}
data <- master.1

tab_xtab(var.row = data$grad.year.1,
         var.col = data$SpecialEdStatus,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/Non-college bivariate/Demographics/SpecialEdStatus-1.html")
```

#### 5-years after graduation

```{r}
data <- master.5

tab_xtab(var.row = data$grad.year.5,
         var.col = data$SpecialEdStatus,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/Non-college bivariate/Demographics/SpecialEdStatus-5.html")
```

#### 10-years after graduation

```{r}
data <- master.10

tab_xtab(var.row = data$grad.year.10,
         var.col = data$SpecialEdStatus,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/Non-college bivariate/Demographics/SpecialEdStatus-10.html")

```
:::

<br>

## `english.learner`

English learner status shows one of the strongest relationships of any demographic variable examined in this analysis (Year 1: χ² = 342.33, df = 3, p \< .001, Cramér's V = 0.134 — the single largest V of any demographic variable at Year 1, surpassing even Special Ed's 0.124; Year 5: χ² = 143.44, V = 0.093; Year 10: χ² = 87.72, V = 0.093). Like Special Ed, this relationship is strongest at Year 1 and settles to a smaller, stable level by Year 5 and Year 10 — but unlike Special Ed's partial U-shape, English learner's effect holds essentially flat from Year 5 to Year 10 rather than recovering.

**A caution before the pattern itself: sample size shrinks sharply for English learners over time**, from n=1,638 at Year 1 to n=1,204 at Year 5 to just n=467 at Year 10. The Year 10 percentages in particular should be read as suggestive rather than precise — a relatively small number of individuals is driving those figures.

**"No MN employment record" shows the largest and most consistent gap of any outcome, and it widens sharply over time — a similar pattern to what emerged for Race/Ethnicity's Asian/PI and Hispanic groups.**

- English learners show substantially higher "no MN emp record" rates at every point: 44.0% vs. 24.2% at Year 1 (19.8-point gap — one of the largest single gaps found anywhere in this analysis), 43.5% vs. 30.0% at Year 5 (13.5 points), and 52.0% vs. 36.0% at Year 10 (16.0 points, and by Year 10 more than half of English learners fall into this category).

**Sustained regional employment shows a large, persistent gap that narrows somewhat but never closes.**

- English learners show consistently lower Sustained WF-NE rates: 13.1% vs. 24.8% at Year 1 (11.7-point gap), 19.1% vs. 28.1% at Year 5 (9.0 points), 15.4% vs. 26.1% at Year 10 (10.7 points) — this gap is remarkably stable in absolute size across all three time points, unlike most other variables examined so far where gaps tend to narrow meaningfully with time.

**Sustained WF-MN also shows a large, fairly stable relative gap** — English learners consistently show roughly half the rate of non-English-learners (4.2% vs. 7.2% at Year 1; 6.6% vs. 13.7% at Year 5; 5.6% vs. 16.1% at Year 10), with the gap widening in absolute terms even as it stays roughly proportional.

**Non-sustained WF is the one outcome where the direction is inconsistent across time** — English learners show *lower* non-sustained rates at Year 1 (38.7% vs. 43.8%) but *higher* rates at Year 5 (30.7% vs. 28.3%) and Year 10 (27.0% vs. 21.9%). This flip is worth noting but, given the shrinking sample size, should be treated cautiously rather than as a clear substantive reversal.

<u>**Overall Interpretation**</u>

English learner status shows the strongest Year 1 relationship of any demographic variable in this analysis, and — like Race/Ethnicity's pattern for Asian/PI and Hispanic graduates — its disparity is concentrated heavily in disconnection from the tracked labor market ("no MN emp record"), which not only persists but grows to affect over half of English learners by Year 10. This makes English learner status directionally similar to Race/Ethnicity's mechanism (disconnection-driven) rather than Economic Status's mechanism (stability-driven) or Special Ed's mechanism (both). Given the substantial overlap in likely underlying population between English learner status and certain Race/Ethnicity categories, and given this project's established finding elsewhere that English learner effects can be partly explained by other data patterns (e.g., testing-record completeness) once academic variables enter a multivariate model, this bivariate relationship should be treated as a strong candidate for attenuation once high school accomplishments are added at the multivariate stage — worth flagging explicitly rather than assuming the bivariate strength will carry through unchanged.

::: panel-tabset
#### 1-year after graduation

```{r}
data <- master.1

tab_xtab(var.row = data$grad.year.1,
         var.col = data$english.learner,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/Non-college bivariate/Demographics/english-learner-1.html")
```

#### 5-years after graduation

```{r}
data <- master.5

tab_xtab(var.row = data$grad.year.5,
         var.col = data$english.learner,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/Non-college bivariate/Demographics/english-learner-5.html")
```

#### 10-years after graduation

```{r}
data <- master.10

tab_xtab(var.row = data$grad.year.10,
         var.col = data$english.learner,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/Non-college bivariate/Demographics/english-learner-10.html")
```
:::

<br>

# High school characteristics

## `Dem_Desc`

Dem_Desc shows a statistically significant but consistently weak relationship with workforce outcomes at all three time points (Year 1: χ² = 45.79, df = 6, p \< .001, Cramér's V = 0.035; Year 5: χ² = 103.40, V = 0.056; Year 10: χ² = 61.07, V = 0.055). This is the **weakest of all high-school-level variables examined so far in this bivariate analysis** — even weaker than Gender at every time point — though it strengthens somewhat from Year 1 to Year 5 and then holds roughly flat into Year 10.

**Sustained regional employment shows a clean, monotonic gradient by district urbanicity — the clearest pattern in this variable, and one that grows more pronounced with time.**

- At Year 1, the gradient is modest: 20.0% (Entirely rural) → 23.2% (Town/rural mix) → 24.4% (Urban/town/rural mix), a 4.4-point spread.
- By Year 5, the gradient widens: 22.0% → 24.9% → 28.9%, a 6.9-point spread.
- By Year 10, it narrows slightly but remains clear: 20.0% → 23.2% → 26.9%, a 6.9-point spread.

This is a directionally consistent, if modest, finding: graduates from more urban districts are somewhat more likely to achieve sustained regional employment than graduates from entirely rural districts — the opposite of what one might naively expect if "regional" opportunity were concentrated in rural areas specifically. This is worth flagging since it mirrors a similar urban-favoring pattern found for this variable in the college-completer models.

**Sustained WF-MN shows the opposite gradient, and it's the strongest pattern in this variable.**

- Entirely rural graduates consistently show the *highest* rate of sustained employment elsewhere in Minnesota at every time point: 10.2% vs. 6.2% (urban) at Year 1, 20.9% vs. 11.9% at Year 5, and 24.0% vs. 14.3% at Year 10 — a gap that widens substantially over time (4.0 points → 9.0 points → 9.7 points).

Read together with the Sustained WF-NE pattern, this suggests entirely rural graduates aren't less likely to achieve *some form* of sustained Minnesota employment — they're more likely to find it *outside* their home region specifically, while urban-district graduates are more likely to stay within the region. This is a meaningfully different and more nuanced story than "rural graduates do worse."

**"No MN employment record" and non-sustained employment show much smaller, less consistent differences across district type** — neither shows a clear gradient the way Sustained WF-NE and Sustained WF-MN do (e.g., Year 1's "no MN emp record" ranges only 25.4%–26.1% across all three district types, essentially flat).

**Overall Interpretation**

District rural/urban classification is a real but weak predictor overall, and its effect is concentrated specifically in *where* sustained employment happens rather than *whether* it happens. Urban-district graduates lean toward staying in the region; entirely rural graduates lean toward sustained employment elsewhere in Minnesota rather than disconnection from the labor market. This is a more specific and, for a "not a recommending organization" audience, a more useful finding than a blanket urban-advantage/rural-disadvantage story — it points to a geographic *redirection* effect (rural graduates finding stable MN employment, just not regionally) rather than a geographic *disadvantage* effect. Given the overall weak Cramér's V at every time point, this variable should be framed as a secondary, refining factor rather than a headline finding in the eventual synthesis.

::: panel-tabset
#### 1-year after graduation

```{r}
data <- master.1

tab_xtab(var.row = data$grad.year.1,
         var.col = data$Dem_Desc,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/Non-college bivariate/High school characteristics/Dem_Desc-1.html")
```

#### 5-years after graduation

```{r}
data <- master.5

tab_xtab(var.row = data$grad.year.5,
         var.col = data$Dem_Desc,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/Non-college bivariate/High school characteristics/Dem_Desc-5.html")
```

#### 10-years after graduation

```{r}
data <- master.10

tab_xtab(var.row = data$grad.year.10,
         var.col = data$Dem_Desc,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/Non-college bivariate/High school characteristics/Dem_Desc-10.html")
```
:::

<br>

## `avg.unemp.rate`

This variable shows a statistically significant relationship with workforce outcomes at all three time points (Year 1: F(3,19114) = 5.92, p \< .001; Year 5: F(3,16716) = 19.20, p \< .001; Year 10: F(3,10078) = 7.07, p \< .001), but the effect size is very small throughout. Converting each model's sum of squares to eta-squared (variance explained): Year 1 ≈ 0.09%, Year 5 ≈ 0.34%, Year 10 ≈ 0.21%. These are trivial in practical terms even though the F-tests are significant — a predictable result given the very large sample sizes here. This variable should be treated as a weak, secondary factor at best, not a driver.

**The relationship changes shape substantially between Year 1 and Years 5/10 — this is the most interesting feature of the results, not the overall significance.**

**At Year 1**, the clearest pattern is that graduates who end up in **non-sustained employment** or with **no MN employment record** come from home areas with meaningfully *higher* local unemployment than graduates who achieve sustained regional employment (Tukey: Non-sustained − Sustained WF-NE = +0.096, p = .001; No MN emp record − Sustained WF-NE = +0.107, p = .001). Sustained WF-MN is not significantly different from any other group at this stage.

**By Year 5, this pattern essentially reverses and a different, stronger signal emerges: Sustained WF-MN becomes the clear outlier, coming from meaningfully higher-unemployment home areas than every other group** (Sustained WF-MN − Sustained WF-NE = +0.226, p \< .001; Sustained WF-MN − Non-sustained = +0.157, p \< .001; Sustained WF-MN − No MN emp record = +0.245, p \< .001). The Year 1 finding — that non-sustained and no-record graduates come from higher-unemployment areas — fades to non-significance relative to Sustained WF-NE by Year 5 (p = .06 and p = .91, respectively).

**Year 10 confirms the Year 5 pattern holds, though somewhat attenuated:** Sustained WF-MN remains significantly higher than both Sustained WF-NE (+0.146, p \< .001) and No MN emp record (−0.136, p \< .001), while other pairwise contrasts are no longer significant.

<u>**Overall Interpretation**</u>

Local unemployment rate is a weak predictor overall (all effect sizes under 0.5% of variance explained), but it tells a coherent and shifting story: at Year 1, higher home-area unemployment is modestly associated with landing in non-sustained work or falling off the employment record entirely — a plausible signal of a rougher initial transition for graduates from weaker local labor markets. By Year 5 and Year 10, however, the dominant and most robust pattern is different: graduates who end up in **sustained employment elsewhere in Minnesota** are the ones who came from the highest-unemployment home areas, consistently and significantly more so than every other outcome group. This is a sensible, substantively interpretable pattern — it suggests that for non-college graduates from weaker local labor markets, sustained employment is achievable, but increasingly by leaving the region for it, rather than by staying and finding sustained work regionally or falling out of the workforce. This nuance (a push-to-elsewhere effect rather than a push-to-instability effect) is worth carrying into the synthesis, even though the overall effect size is modest.

::: panel-tabset
#### 1-year after graduation

```{r}
data <- master.1

summary <- aov(avg.unemp.rate ~ grad.year.1, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college bivariate/High school characteristics/avg-unemp-rate-1.xlsx", overwrite = TRUE)

```

#### 5-years after graduation

```{r}
data <- master.5

summary <- aov(avg.unemp.rate ~ grad.year.5, data = data)


summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college bivariate/High school characteristics/avg-unemp-rate-5.xlsx", overwrite = TRUE)
```

#### 10-years after graduation

```{r}
data <- master.10

summary <- aov(avg.unemp.rate ~ grad.year.10, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college bivariate/High school characteristics/avg-unemp-rate-10.xlsx", overwrite = TRUE)

```
:::

<br>

## `avg.wages.pct.state`

This variable shows a statistically significant relationship with workforce outcomes at all three time points (Year 1: F(3,19114) = 15.79, p \< .001; Year 5: F(3,16716) = 19.47, p \< .001; Year 10: F(3,10078) = 6.55, p \< .001). As with local unemployment rate, the effect sizes are small (eta-squared: Year 1 ≈ 0.25%, Year 5 ≈ 0.35%, Year 10 ≈ 0.19%) despite the strong statistical significance — again a function of large sample size rather than a large practical effect. This should be treated as a weak, secondary predictor.

**Unlike local unemployment rate, this variable shows a single, highly consistent pattern across all three time points rather than a pattern that shifts with time — and it directly complements the unemployment finding.**

**Sustained WF-MN is significantly and consistently the lowest-wage-origin group at every single time point**, coming from home areas with meaningfully lower relative wages than all three other outcome groups:

- vs. Sustained WF-NE: −0.010 (Year 1, p \< .001), −0.011 (Year 5, p \< .001), −0.008 (Year 10, p \< .001)
- vs. Non-sustained WF: −0.010 (Year 1, p \< .001), −0.008 (Year 5, p \< .001), −0.006 (Year 10, p = .024)
- vs. No MN emp record: −0.009 (Year 1, p \< .001), −0.009 (Year 5, p \< .001), −0.007 (Year 10, p \< .001)

This holds at every time point with no exceptions and no weakening in direction (only a modest weakening in magnitude by Year 10).

**Among the other three groups, differences are small and mostly non-significant** — Sustained WF-NE, Non-sustained WF, and No MN emp record are not clearly distinguishable from one another on this variable at any time point (the one exception: Non-sustained vs. Sustained WF-NE reaches marginal significance at Year 5, p = .044, but the effect is tiny at −0.003 and doesn't hold at Year 1 or Year 10).

**Overall Interpretation**

This result reinforces and sharpens the local unemployment finding rather than adding a separate story: graduates who end up in sustained employment elsewhere in Minnesota (rather than in the region) consistently come from home areas with both higher unemployment *and* lower relative wages — a coherent "weak local labor market" signature that holds steady across all three time horizons, unlike the unemployment-rate pattern, which shifted between Year 1 and Year 5. Together, these two variables support a single clear narrative: graduates from areas with objectively weaker local labor market conditions (low relative wages, high unemployment) are more likely to find sustained employment specifically by leaving the region for elsewhere in Minnesota, not by settling for instability or disconnection. Given both variables point the same direction and reinforce each other so consistently, this "push toward MN-outside-region" pattern is one of the more robust structural findings in the high-school-characteristics group, even though each variable's individual effect size is small.

::: panel-tabset
#### 1-year after graduation

```{r}
data <- master.1

summary <- aov(avg.wages.pct.state ~ grad.year.1, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college bivariate/High school characteristics/avg-wages-pct-state-1.xlsx", overwrite = TRUE)

```

#### 5-years after graduation

```{r}
data <- master.5

summary <- aov(avg.wages.pct.state ~ grad.year.5, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college bivariate/High school characteristics/avg-wages-pct-state-5.xlsx", overwrite = TRUE)

```

#### 10-years after graduation

```{r}
data <- master.10

summary <- aov(avg.wages.pct.state ~ grad.year.10, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college bivariate/High school characteristics/avg-wages-pct-state-10.xlsx", overwrite = TRUE)

```
:::

<br>

# High school accomplishments

## `pseo.participant`

PSEO participation shows a statistically significant but consistently weak relationship with workforce outcomes at all three time points (Year 1: χ² = 61.99, df = 3, p \< .001, Cramér's V = 0.064; Year 5: χ² = 28.65, V = 0.050; Year 10: χ² = 28.45, V = 0.078). These figures are nearly identical to the uncorrected version from before the cohort restriction (V = 0.068, 0.052, 0.081) — confirming that the correction mattered for the *integrity* of the sample, but didn't meaningfully change the substantive result, since so few rows were actually affected once the true scope of the PSEO tracking gap was understood.

**The most notable finding here is what PSEO does *not* predict: sustained regional employment specifically.** Unlike CTE (in the college-completer models) or several other variables in this project, PSEO participation shows almost no difference in Sustained WF-NE rates at any time point — 24.9% vs. 24.1% at Year 1, 27.6% vs. 28.1% at Year 5, 25.6% vs. 26.0% at Year 10. Across all three years, participants and non-participants land in sustained regional employment at essentially the same rate.

**Where PSEO does show a real, if modest, relationship is in overall labor-market engagement — specifically, participants are more likely to be sustained-employed *somewhere* in Minnesota, and less likely to disconnect from the tracked system, especially as time passes.**

- **Sustained WF-MN:** PSEO participants show meaningfully higher rates at Year 5 (15.8% vs. 12.3%) and Year 10 (20.4% vs. 14.4%) — a real and widening gap, though not present at Year 1 (7.2% vs. 7.2%, identical).
- **No MN employment record:** PSEO participants show lower rates at Year 5 (29.0% vs. 30.9%) and Year 10 (34.5% vs. 37.1%) — modest but directionally consistent with reduced disconnection.
- **Year 1 shows a different, almost opposite pattern for these same two outcomes:** PSEO participants show *higher* non-sustained employment (47.4% vs. 42.4%) and *lower* "no MN emp record" (20.6% vs. 26.3%) — suggesting that immediately after graduation, PSEO participants are more likely to be actively working (even if not yet in stable jobs) rather than absent from the employment record, a pattern that later resolves into the Sustained WF-MN advantage seen at Year 5 and Year 10.

**Overall Interpretation**

PSEO participation does not appear to function as a regional anchor for non-college graduates — participants are no more likely than non-participants to end up in sustained employment specifically within the region, at any point in the ten-year window. Instead, its effect looks like a general labor-market engagement or "range-expanding" signal: participants are more likely to be actively working somewhere in Minnesota's tracked system (whether in the region or elsewhere), and less likely to disconnect from it entirely, with this effect strengthening over time in the Sustained WF-MN category specifically. This mirrors a pattern already identified elsewhere in this project — PSEO tends to expand a graduate's range of engagement and opportunity rather than specifically pulling them toward or keeping them in the region. Given the modest effect sizes throughout, this should be framed as a secondary, engagement-related finding rather than a strong regional-retention lever in the eventual synthesis for this population.

::: panel-tabset
#### 1-year after graduation

```{r}
model.vars <- c("Gender", "RaceEthnicity", "economic.status", "SpecialEdStatus", "english.learner", "Dem_Desc", "avg.wages.pct.state", "avg.unemp.rate", "pseo.participant", "ap.exam", "took.ACT", "ACTCompositeScore", "MCA.M", "MCA.R", "MCA.S", "cte.achievement", "total.cte.courses.taken", "attended.ps", "grad.year.1", "grad.year.5", "grad.year.10")

master.pseo <- master.original %>%
  filter(ps.grad == "No",
         grad.year > 2010 ) %>%
  select(all_of(model.vars))

master.pseo.1 <- master.pseo %>%
  filter(!grad.year.1 %in% c("Attending ps", "After 2023")) %>%
  droplevels() %>%
  mutate(grad.year.1 = fct_relevel(grad.year.1, "Sustained WF - NE", "Sustained WF - MN", "Non-sustained WF", "No MN emp record"))

master.pseo.5 <- master.pseo %>%
  filter(!grad.year.5 %in% c("Attending ps", "After 2023")) %>%
  droplevels() %>%
  mutate(grad.year.5 = fct_relevel(grad.year.5, "Sustained WF - NE", "Sustained WF - MN", "Non-sustained WF", "No MN emp record"))

master.pseo.10 <- master.pseo %>%
  filter(!grad.year.10 %in% c("Attending ps", "After 2023")) %>%
  droplevels() %>%
  mutate(grad.year.10 = fct_relevel(grad.year.10, "Sustained WF - NE", "Sustained WF - MN", "Non-sustained WF", "No MN emp record"))

data <- master.pseo.1 


tab_xtab(var.row = data$grad.year.1,
         var.col = data$pseo.participant,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/Non-college bivariate/High school accomplishments/pseo-participant-1.html")
```

#### 5-years after graduation

```{r}
data <- master.pseo.5

tab_xtab(var.row = data$grad.year.5,
         var.col = data$pseo.participant,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/Non-college bivariate/High school accomplishments/pseo-participant-5.html")
```

#### 10-years after graduation

```{r}
data <- master.pseo.10 

tab_xtab(var.row = data$grad.year.10,
         var.col = data$pseo.participant,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/Non-college bivariate/High school accomplishments/pseo-participant-10.html")

```
:::

<br>

## `ap.exam`

AP exam participation shows a statistically significant, consistently weak-to-modest relationship with workforce outcomes at all three time points (Year 1: χ² = 70.02, df = 3, p \< .001, Cramér's V = 0.061; Year 5: χ² = 51.43, V = 0.055; Year 10: χ² = 32.76, V = 0.057) — a stable effect size across the whole window, unlike several other variables examined so far.

**A quick note on scale:** AP exam-takers are a small minority of this non-completer population (850 of 19,118 at Year 1, roughly 4.4%, shrinking slightly to 516 of 10,082 by Year 10, about 5.1%). Percentages for this group should be read with some caution, though the samples are large enough (500+) that the significant findings below are still meaningfully powered.

**Year 1 shows a distinct, somewhat surprising pattern that does not persist into later years.** AP exam-takers who did *not* complete college show a dramatically higher non-sustained employment rate than non-takers (55.3% vs. 42.9%, a 12.4-point gap — the largest single gap found for this variable at any time point), alongside lower rates in both sustained categories (Sustained WF-NE: 14.4% vs. 24.2%; Sustained WF-MN: 4.5% vs. 7.0%). This is a striking early-career pattern: for the subset of AP-exam takers who end up as non-completers, the first year out looks distinctly rockier than for their non-AP peers — plausibly reflecting a population that expected or attempted a different (likely four-year) pathway and is still adjusting after that path didn't materialize.

**By Year 5, this pattern resolves into something that looks much more like the ACT crossover found elsewhere in this project — AP-taking is associated with leaving the region, not necessarily with instability.**

- **Sustained WF-NE:** 18.9% (AP) vs. 27.9% (non-AP) — a persistent, substantial gap.
- **Sustained WF-MN:** 18.3% (AP) vs. 12.9% (non-AP) — reversed from Year 1, now clearly *higher* for AP-takers.
- **Non-sustained WF:** 26.4% (AP) vs. 28.6% (non-AP) — the large Year 1 gap has now nearly disappeared and even reversed slightly.
- **No MN emp record:** 36.5% (AP) vs. 30.7% (non-AP) — AP-takers now show meaningfully higher disconnection rates.

**Year 10 confirms this later-stage pattern holds and strengthens somewhat:** Sustained WF-NE remains lower for AP-takers (20.0% vs. 25.9%), Sustained WF-MN remains notably higher (23.1% vs. 15.2% — the widest gap of the three time points), and "no MN emp record" remains higher (39.5% vs. 36.6%).

<u>**Overall Interpretation**</u>

AP exam participation among non-college graduates shows a two-phase story. At Year 1, AP-taking non-completers face a distinctly rougher immediate transition into the workforce — substantially higher non-sustained employment — than their non-AP peers, a pattern that doesn't appear for any other variable examined so far in this analysis and may reflect this being an atypical group (academically college-track students who didn't complete a credential). By Year 5 and Year 10, however, that early instability resolves, and the pattern shifts to look like the same "college-prep pulls away from the region" mechanism found with ACT scores throughout this project: AP-takers are less likely to be in sustained regional employment and more likely to be sustained-employed elsewhere in Minnesota or disconnected from the tracked system entirely. Given how small this subgroup is, the Year 1 finding in particular deserves a cautious framing in any synthesis — it's a real and statistically supported pattern, but based on a modest sample and worth flagging as an area where the mechanism (transitional instability vs. genuine underlying disadvantage) isn't fully clear from the bivariate data alone.

::: panel-tabset
#### 1-year after graduation

```{r}
data <- master.1

tab_xtab(var.row = data$grad.year.1,
         var.col = data$ap.exam,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/Non-college bivariate/High school accomplishments/ap-exam-1.html")
```

#### 5-years after graduation

```{r}
data <- master.5

tab_xtab(var.row = data$grad.year.5,
         var.col = data$ap.exam,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/Non-college bivariate/High school accomplishments/ap-exam-5.html")
```

#### 10-years after graduation

```{r}
data <- master.10

tab_xtab(var.row = data$grad.year.10,
         var.col = data$ap.exam,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/Non-college bivariate/High school accomplishments/ap-exam-10.html")

```
:::

<br>

## `took.ACT`

`took.ACT` shows one of the strongest and most consistent relationships of any variable examined in this bivariate analysis so far (Year 1: χ² = 287.35, df = 3, p \< .001, Cramér's V = 0.123; Year 5: χ² = 170.18, V = 0.101; Year 10: χ² = 133.44, V = 0.115). This is a genuinely modest-to-moderate effect by conventional benchmarks — noticeably stronger than most demographic variables and every high-school-characteristic variable examined earlier — and it holds essentially steady across the full ten-year window rather than fading like several other predictors have.

**The direction here is the opposite of what CTE showed in the college-completer models, and it's worth naming clearly: within this non-completer population, ACT test-takers consistently show *better* outcomes than non-test-takers, not worse.**

- **Sustained WF-NE:** ACT test-takers show modestly higher rates at every time point — 24.4% vs. 23.5% at Year 1, 28.8% vs. 26.5% at Year 5, 25.6% vs. 25.5% at Year 10 (this last gap essentially closes by Year 10).
- **Sustained WF-MN:** test-takers show a larger and *widening* advantage here — 7.7% vs. 6.5% at Year 1, 16.6% vs. 10.8% at Year 5, and 21.0% vs. 12.8% at Year 10 — the single clearest, most consistent gradient of any outcome for this variable.
- **No MN employment record:** test-takers show consistently *lower* rates — 18.9% vs. 29.7% at Year 1 (a 10.8-point gap, one of the largest gaps found for any variable at Year 1), narrowing somewhat but still substantial at Year 5 (27.1% vs. 33.6%) and Year 10 (34.8% vs. 37.7%).
- **Non-sustained WF:** test-takers show *higher* rates at Year 1 (49.1% vs. 40.3%) but this reverses by Year 5 (27.4% vs. 29.2%) and Year 10 (18.6% vs. 24.0%) — test-takers start out more likely to be in unstable work but end up less likely to be, a crossover pattern similar in shape to what we saw for AP exam.

**Taken together, this describes a population that starts out slightly rockier (Year 1: more non-sustained work) but converges toward, and ultimately exceeds, non-test-takers on every measure of stable engagement with the labor market.** By Year 5 and Year 10, ACT test-takers among non-completers are more likely to be in *any* form of sustained employment (regional or elsewhere in MN) and substantially less likely to have no employment record at all.

**Overall Interpretation**

Among non-college graduates specifically, simply having taken the ACT is one of the strongest positive predictors of long-term labor market engagement found in this bivariate analysis — the opposite of what might be assumed if ACT-taking were treated purely as a "college-track" signal that doesn't pay off without a completed degree. The pattern instead suggests ACT test-takers who don't complete college still carry forward some advantage — plausibly reflecting stronger underlying academic preparation, family/school support for engaging with standardized processes, or other correlated factors — that shows up as better long-term workforce attachment, particularly via employment *outside* the region (Sustained WF-MN), which shows the strongest and most steadily widening gap of any outcome. This is a genuinely different mechanism than what's been seen for AP exam participation or, in the college-completer models, for ACT *score* specifically — here it's mere participation, not performance, that's doing the work, and it's associated with broader labor-market success rather than regional retention specifically. Given that `ACTCompositeScore` hasn't been examined yet for this population, it will be worth checking whether score (among test-takers) tells a different, possibly opposing story — consistent with the crossover pattern found for this exact pairing of variables everywhere else in this project.

::: panel-tabset
#### 1-year after graduation

```{r}
data <- master.1

tab_xtab(var.row = data$grad.year.1,
         var.col = data$took.ACT,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/Non-college bivariate/High school accomplishments/took-ACT-1.html")
```

#### 5-years after graduation

```{r}
data <- master.5

tab_xtab(var.row = data$grad.year.5,
         var.col = data$took.ACT,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/Non-college bivariate/High school accomplishments/took-ACT-5.html")
```

#### 10-years after graduation

```{r}
data <- master.10

tab_xtab(var.row = data$grad.year.10,
         var.col = data$took.ACT,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/Non-college bivariate/High school accomplishments/took-ACT-10.html")

```
:::

<br>

## `ACTCompositeScore`

This analysis is restricted to the subset of non-college graduates who took the ACT (n = 6,786 at Year 1, 6,870 at Year 5, 3,431 at Year 10 — matching the "Yes" counts from the `took.ACT` cross-tabs). ACT score shows a significant relationship with workforce outcomes at every time point (Year 1: F(3,6782) = 40.47, p \< .001; Year 5: F(3,6866) = 31.71, p \< .001; Year 10: F(3,3427) = 17.62, p \< .001), and unlike the local labor-market variables examined earlier, the effect sizes here are genuinely modest rather than trivial (eta-squared: Year 1 ≈ 1.8%, Year 5 ≈ 1.4%, Year 10 ≈ 1.5%) — roughly 5-8 times larger than what local wages or unemployment rate showed.

**This produces the clearest, most complete crossover pattern found anywhere in this bivariate analysis, and it directly mirrors — but sharpens — the "took ACT vs. ACT score" split found throughout this project.** Recall that mere ACT *participation* was associated with *better* outcomes (higher sustained employment, lower disconnection). Among test-takers specifically, ACT *score* runs in the opposite direction:

**At Year 1, there is a complete, statistically clean ordering across all four groups, from lowest to highest mean ACT score:** Sustained WF-NE (lowest) \< Sustained WF-MN \< Non-sustained WF \< No MN employment record (highest) — every single pairwise comparison is statistically significant. In other words, among test-takers who didn't complete college, the graduates with the *lowest* ACT scores are the ones most likely to be sustained-employed regionally, while the *highest*-scoring test-takers are the most likely to show no Minnesota employment record at all.

**By Year 5, the ordering partially collapses into three tiers:** Sustained WF-NE remains clearly lowest; Non-sustained WF sits in the middle, now significantly higher than WF-NE but significantly lower than Sustained WF-MN; and Sustained WF-MN and No MN emp record are no longer distinguishable from each other, both sitting at the top.

**By Year 10, the pattern simplifies further into two clear tiers:** {Sustained WF-NE, Non-sustained WF} — no longer significantly different from each other, both low — versus {Sustained WF-MN, No MN emp record} — also not significantly different from each other, both high. The Sustained WF-NE vs. Sustained WF-MN gap remains the largest and most robust contrast across all three time points (+0.68 at Year 1, growing to +0.99 at Year 5 and +1.19 at Year 10 — the only contrast that not only stays significant but grows steadily larger over the full ten-year window).

**Overall Interpretation**

Among ACT test-takers who don't complete a postsecondary credential, higher scores consistently predict *not* remaining in sustained regional employment — the same direction found for ACT score in every other part of this project, but here the effect is cleaner and, at Year 1, complete across all four outcome categories at once. The single most robust, persistent finding is the gap between Sustained WF-NE and Sustained WF-MN, which not only holds at every time point but widens over the full ten years — higher-scoring test-takers who don't complete college are consistently and increasingly more likely to end up sustained-employed *outside* the region rather than within it. The Year 1 finding that the highest scorers are also the most likely to show no Minnesota employment record is worth flagging as a distinct and slightly concerning early pattern — this may reflect higher-scoring non-completers being more mobile or more likely to be job-searching outside the tracked system immediately after high school, though by Year 10 this specific group (No MN emp record) is statistically indistinguishable from Sustained WF-MN rather than uniquely extreme. Combined with the `took.ACT` finding, this variable pair tells a coherent, two-part story: simply engaging with the ACT is a broadly positive signal for this population, but among those who engage, higher scores specifically predict leaving the region rather than anchoring to it — the same "engagement helps, achievement pulls away" mechanism found throughout this entire project, now confirmed cleanly within the non-completer population as well.

::: panel-tabset
#### 1-year after graduation

```{r}
data <- master.1

summary <- aov(ACTCompositeScore ~ grad.year.1, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college bivariate/High school accomplishments/ACTCompositeScore-1.xlsx", overwrite = TRUE)

```

#### 5-years after graduation

```{r}
data <- master.5

summary <- aov(ACTCompositeScore ~ grad.year.5, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college bivariate/High school accomplishments/ACTCompositeScore-5.xlsx", overwrite = TRUE)

```

#### 10-years after graduation

```{r}
data <- master.10

summary <- aov(ACTCompositeScore ~ grad.year.10, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college bivariate/High school accomplishments/ACTCompositeScore-10.xlsx", overwrite = TRUE)

```
:::

<br>

## `MCA.M`

MCA Math shows a significant relationship with workforce outcomes at all three time points, and — unlike most variables examined so far — the effect size **grows meaningfully over time** rather than staying flat or shrinking (Year 1: F(3,17421) = 16.34, p \< .001, eta² ≈ 0.28%; Year 5: F(3,15373) = 22.16, p \< .001, eta² ≈ 0.43%; Year 10: F(3,9132) = 31.47, p \< .001, eta² ≈ 1.02% — a roughly 3.6x increase in variance explained from Year 1 to Year 10).

**A missingness note worth flagging before the results:** the samples here (n = 17,425 / 15,377 / 9,136) run 8-9% below the full cohort totals used elsewhere in this analysis (19,118 / 16,720 / 10,082) at every time point. This is a smaller, more consistent gap than PSEO's cohort-driven cliff, and doesn't show the same all-or-nothing pattern by year — but given this project has already found that MCA missingness can correlate meaningfully with English learner status in the college-completer population, it would be worth the same demographic crosstab check before this variable enters any multivariate model for this population.

**The pattern shifts meaningfully between Year 1 and Years 5/10 — similar in shape to what was found for local unemployment rate and AP exam.**

**At Year 1**, the only group that stands out is **No MN employment record**, which shows significantly lower MCA Math scores than all three other groups (vs. Sustained WF-NE: −0.116, p \< .001; vs. Sustained WF-MN: −0.187, p \< .001; vs. Non-sustained WF: −0.126, p \< .001). Sustained WF-NE, Sustained WF-MN, and Non-sustained WF are not distinguishable from one another at this stage.

**By Year 5 and Year 10, a different and strengthening pattern emerges: Sustained WF-MN becomes the clear high-achievement group, and Non-sustained WF becomes the clear low-achievement group — the widest and most robust gap in the entire table, and one that grows substantially with time.**

- Sustained WF-MN vs. Non-sustained WF: −0.226 at Year 5 (p \< .001), widening to −0.335 by Year 10 (p \< .001) — the single largest contrast found in this variable, and it grows by roughly 50% over five years.
- Sustained WF-MN also pulls significantly ahead of Sustained WF-NE at both points (+0.142 at Year 5, +0.198 at Year 10, also widening).
- Sustained WF-NE and No MN emp record are no longer distinguishable from each other at Year 5 or Year 10 (both sit in the middle), a shift from Year 1 where No MN emp record was the clear outlier on the low end.

**Overall Interpretation**

MCA Math proficiency's relationship with workforce outcomes strengthens over time and settles into a clear, persistent pattern: among non-college graduates, stronger academic preparation is most associated with finding sustained employment *outside* the region rather than within it, while weaker preparation is most associated with non-sustained (unstable) employment specifically — not with disconnection from the labor market, which sits in the middle rather than at either extreme. This is a meaningfully different shape than the ACT findings: ACT score's clearest signal was Sustained WF-NE vs. Sustained WF-MN specifically, with No MN emp record trending toward the high-score end at Year 1; MCA Math's clearest and most persistent signal is Sustained WF-MN vs. Non-sustained WF, a comparison more about employment *stability* than about disconnection. Combined with the earlier finding that CTE anchors regionally while academic/college-prep measures (ACT, AP) pull away from the region, this adds a further nuance: baseline academic proficiency itself — not just test-taking behavior — is part of the same broader pattern where stronger academic indicators correlate with success achieved elsewhere in Minnesota rather than within the region specifically.

::: panel-tabset
#### 1-year after graduating

```{r}
data <- master.1

summary <- aov(MCA.M ~ grad.year.1, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college bivariate/High school accomplishments/MCA-M-1.xlsx", overwrite = TRUE)

```

#### 5-years after graduation

```{r}
data <- master.5

summary <- aov(MCA.M ~ grad.year.5, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college bivariate/High school accomplishments/MCA-M-5.xlsx", overwrite = TRUE)
```

#### 10-years after graduation

```{r}
data <- master.10

summary <- aov(MCA.M ~ grad.year.10, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college bivariate/High school accomplishments/MCA-M-10.xlsx", overwrite = TRUE)
```
:::

<br>

## `MCA.R`

MCA Reading shows a significant relationship with workforce outcomes at all three time points, following an almost identical shape and trajectory to MCA Math (Year 1: F(3,17289) = 16.94, p \< .001, eta² ≈ 0.29%; Year 5: F(3,15214) = 16.43, p \< .001, eta² ≈ 0.32%; Year 10: F(3,9023) = 34.70, p \< .001, eta² ≈ 1.14%) — the same pattern of a modest effect that strengthens substantially by Year 10.

**Sample sizes here (n = 17,293 / 15,218 / 9,027) show almost exactly the same \~9-10% shortfall relative to the full cohort totals as MCA Math did** — consistent with these two variables sharing missingness for the same students, as expected given how closely related MCA subject scores tend to be. The same caution about checking this against demographic variables (particularly English learner status, given the pattern established elsewhere in this project) applies equally here.

**The pattern across time closely parallels MCA Math, reinforcing rather than adding a new signal.**

**At Year 1**, only **No MN employment record** stands out — significantly lower reading scores than all three other groups (vs. Sustained WF-NE: −0.116, p \< .001; vs. Sustained WF-MN: −0.161, p \< .001; vs. Non-sustained WF: −0.127, p \< .001) — an essentially identical shape to MCA Math's Year 1 result.

**By Year 5**, Sustained WF-MN emerges as the highest-scoring group, significantly ahead of Non-sustained WF (−0.178, p \< .001) and No MN emp record (−0.122, p \< .001), while Sustained WF-NE and No MN emp record are not clearly distinguished from each other.

**By Year 10, the pattern resolves into a complete, fully-separated four-way ordering — the clearest ordering found for either MCA subject:** Non-sustained WF (lowest) \< Sustained WF-NE \< No MN emp record \< Sustained WF-MN (highest), with every single pairwise contrast statistically significant. As with MCA Math, the widest and most robust gap is Sustained WF-MN vs. Non-sustained WF (−0.310, p \< .001) — nearly identical in size to MCA Math's equivalent Year 10 gap (−0.335).

**Overall Interpretation**

MCA Reading tells essentially the same story as MCA Math, both in direction and in how the pattern evolves over time — unsurprising given how closely correlated these two subject scores are known to be throughout this project (r ≈ 0.73 in the college-completer population). The consistent core finding across both subjects: Sustained WF-MN consistently represents the highest-achieving group and Non-sustained WF the lowest, with this gap widening substantially between Year 5 and Year 10. Given how similar MCA Math and MCA Reading's results are here, this is a strong early signal — consistent with what was found and acted on earlier in this project — that these two variables will likely need to be consolidated to just one representative subject once this moves to the multivariate confirmation stage, rather than treated as two independent sources of information.

::: panel-tabset
#### 1-year after graduating

```{r}
data <- master.1

summary <- aov(MCA.R ~ grad.year.1, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college bivariate/High school accomplishments/MCA-R-1.xlsx", overwrite = TRUE)
```

#### 5-years after graduation

```{r}
data <- master.5

summary <- aov(MCA.R ~ grad.year.5, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college bivariate/High school accomplishments/MCA-R-5.xlsx", overwrite = TRUE)
```

#### 10-years after graduation

```{r}
data <- master.10

summary <- aov(MCA.R ~ grad.year.10, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college bivariate/High school accomplishments/MCA-R-10.xlsx", overwrite = TRUE)
```
:::

<br>

## `MCA.S`

MCA Science shows a meaningfully different profile than MCA Math or MCA Reading, both in missingness and in results — worth treating this as a distinct case rather than assuming it will mirror the other two subjects.

**Missingness is substantially larger here, and grows sharply with the time horizon — a pattern worth flagging explicitly before interpreting the results.** Sample sizes are n = 14,796 (Year 1), 12,120 (Year 5), and 5,808 (Year 10) — shortfalls of 22.6%, 27.5%, and **42.4%** relative to the full cohort totals (19,118 / 16,720 / 10,082). This is a considerably bigger gap than MCA Math or Reading showed (\~9-10% at every time point), and unlike those two subjects, the gap here grows substantially as the time horizon lengthens — the same shape of problem (though smaller in scale) as the PSEO tracking-cliff issue found earlier in this project. This is consistent with what was found in this project's college-completer analysis, where MCA Science also showed dramatically higher missingness than Math or Reading. Given nearly half the Year 10 sample is missing this variable, **the Year 10 result here should be treated with real caution**, and this pattern is worth a dedicated missingness investigation (by cohort year and by demographic group) before this variable is considered for any multivariate model.

**At Year 1, MCA Science shows no significant relationship at all** (F(3,14792) = 2.33, p = .073) — the only academic-achievement variable examined in this entire bivariate analysis that fails to reach significance at the earliest time point. None of the six pairwise contrasts are significant either.

**By Year 5 and Year 10, a significant relationship emerges, but with a different shape than MCA Math or Reading showed.** Rather than Non-sustained WF standing alone at the bottom (as with Math/Reading), Science splits the four groups into two clear tiers:

- **Lower tier: Sustained WF-NE and Non-sustained WF** — not significantly different from each other at either Year 5 (p = .114) or Year 10 (p = .137).
- **Higher tier: Sustained WF-MN and No MN emp record** — also not significantly different from each other at either point (Year 5: p = .815; Year 10: p = .253).

The gap between tiers widens over time: Sustained WF-MN vs. Sustained WF-NE grows from +0.112 (Year 5) to +0.185 (Year 10); Sustained WF-MN vs. Non-sustained WF grows from −0.159 to −0.254.

<u>**Overall Interpretation**</u>

MCA Science partially confirms the pattern found in Math and Reading — Sustained WF-MN remains the highest-scoring group and the gap with lower-scoring groups widens over time — but the specific grouping differs: here, No MN employment record joins Sustained WF-MN at the high-scoring end rather than sitting in the middle or at the low end as it did for Math and Reading, while Sustained WF-NE and Non-sustained WF are statistically indistinguishable from each other throughout. Combined with Science's much larger and still-growing missingness problem, this variable looks like the weakest and least reliable of the three MCA subjects for this population — consistent with the earlier finding elsewhere in this project that MCA Math is the best candidate for a single representative academic-achievement variable, given its combination of lower missingness and a clean, consistent signal across all three time points (unlike Science, which fails to reach significance at Year 1 and carries a substantially bigger missing-data problem throughout).

::: panel-tabset
#### 1-year after graduation

```{r}
data <- master.1

summary <- aov(MCA.S ~ grad.year.1, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college bivariate/High school accomplishments/MCA-S-1.xlsx", overwrite = TRUE)
```

#### 5-years after graduation

```{r}
data <- master.5

summary <- aov(MCA.S ~ grad.year.5, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college bivariate/High school accomplishments/MCA-S-5.xlsx", overwrite = TRUE)
```

#### 10-years after graduation

```{r}
data <- master.10

summary <- aov(MCA.S ~ grad.year.10, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college bivariate/High school accomplishments/MCA-S-10.xlsx", overwrite = TRUE)
```
:::

<br>

## `cte.achievement`

CTE achievement shows a statistically significant, consistently weak-to-modest relationship with workforce outcomes at all three time points (Year 1: χ² = 168.91, df = 6, p \< .001, Cramér's V = 0.066; Year 5: χ² = 116.50, V = 0.059; Year 10: χ² = 79.24, V = 0.063) — a stable effect size across the full window, similar in magnitude to Special Ed status and stronger than most demographic variables.

**A clear, consistent, monotonic gradient shows up in every outcome at every time point: more CTE engagement predicts staying regional and less CTE engagement predicts disconnection — with no exceptions across all three years.**

**Sustained WF-NE increases steadily with CTE engagement level, at every time point:** \* Year 1: No CTE (20.3%) \< CTE Participant (23.8%) \< CTE Concentrator/Completer (27.8%) \* Year 5: No CTE (24.1%) \< CTE Participant (28.4%) \< CTE Concentrator/Completer (31.4%) \* Year 10: No CTE (21.7%) \< CTE Participant (26.1%) \< CTE Concentrator/Completer (29.7%)

**"No MN employment record" shows the mirror-image gradient, also holding at every time point:** \* Year 1: CTE Concentrator/Completer (23.0%) \< CTE Participant (24.4%) \< No CTE (29.4%) \* Year 5: CTE Concentrator/Completer (27.3%) \< CTE Participant (29.4%) \< No CTE (34.5%) \* Year 10: CTE Concentrator/Completer (32.9%) \< CTE Participant (35.8%) \< No CTE (40.5%)

**Sustained WF-MN and Non-sustained WF, by contrast, show little to no relationship with CTE level at any time point** — Sustained WF-MN is nearly flat across all three CTE categories at every year (e.g., Year 10: 15.2% / 15.6% / 15.9%), and Non-sustained WF shows only small, inconsistent differences. This is an important distinction: CTE's effect is concentrated specifically in the Sustained WF-NE / No MN emp record axis — staying regional vs. disconnecting entirely — not in overall employment stability broadly.

**Overall Interpretation**

CTE achievement shows the cleanest, most consistent dose-response relationship of any variable examined in this bivariate analysis: every step up in CTE engagement (No CTE → Participant → Concentrator/Completer) predicts higher odds of sustained regional employment and lower odds of falling off the Minnesota employment record entirely, and this holds without exception at Year 1, Year 5, and Year 10. This directly replicates the pattern found throughout the college-completer models in this project, where CTE consistently functioned as the strongest and most reliable regional-anchoring factor. Notably, CTE's effect here is specific to the regional-attachment/disconnection axis rather than to employment stability broadly (Sustained WF-MN and Non-sustained WF show little relationship with CTE level), which sharpens the interpretation: CTE appears to function specifically as a *regional* anchor for non-college graduates, not as a general predictor of workforce success. Given how consistent and interpretable this pattern is, CTE achievement is a strong candidate for one of the central, most confidently reportable findings in the eventual synthesis for this population.

::: panel-tabset
#### 1-year after graduation

```{r}
data <- master.1

tab_xtab(var.row = data$grad.year.1,
         var.col = data$cte.achievement,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/Non-college bivariate/High school accomplishments/cte-achievement-1.html")
```

#### 5-years after graduation

```{r}
data <- master.5

tab_xtab(var.row = data$grad.year.5,
         var.col = data$cte.achievement,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/Non-college bivariate/High school accomplishments/cte-achievement-5.html")

```

#### 10-years after graduation

```{r}
data <- master.10

tab_xtab(var.row = data$grad.year.10,
         var.col = data$cte.achievement,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/Non-college bivariate/High school accomplishments/cte-achievement-10.html")
```
:::

<br>

## `total.cte.courses.taken`

This continuous measure of CTE engagement shows a significant relationship with workforce outcomes at all three time points (Year 1: F(3,19114) = 60.00, p \< .001, eta² ≈ 0.93%; Year 5: F(3,16716) = 34.47, p \< .001, eta² ≈ 0.62%; Year 10: F(3,10078) = 16.50, p \< .001, eta² ≈ 0.49%). Unlike most variables examined in this analysis, the effect size here **decreases** over time rather than holding steady or growing — the opposite trajectory from `cte.achievement` (the categorical version), which held essentially flat across all three years. Even at Year 1, though, this is the single largest continuous-variable effect size found in this bivariate analysis, larger than ACT score, MCA Math, or MCA Reading at their peaks.

**The pattern here adds a genuine nuance beyond what the categorical `cte.achievement` variable showed — worth reading as a complement, not a duplicate.**

**At Year 1, the clearest finding is that "No MN employment record" is associated with dramatically fewer CTE courses than every other group**, and Sustained WF-NE and Sustained WF-MN are statistically tied at the high end (not significantly different from each other, p = .982): No MN emp record (lowest) \< Non-sustained WF \< {Sustained WF-NE ≈ Sustained WF-MN} (highest, tied).

**By Year 5 and Year 10, Sustained WF-NE pulls away as the clearly highest-CTE group on its own — a distinction that `cte.achievement`'s categorical version did not show as sharply.** Recall that the categorical CTE analysis found Sustained WF-MN essentially flat across all CTE levels, suggesting no relationship at all. Here, using course count, Sustained WF-MN is significantly *lower* than Sustained WF-NE at Year 5 (−0.239, p \< .001) and trending the same direction at Year 10 (−0.158, though not quite significant, p = .150). The revised ordering at Year 5/10: No MN emp record (lowest) \< {Sustained WF-MN ≈ Non-sustained WF} (middle tier, statistically tied) \< Sustained WF-NE (highest).

<u>**Overall Interpretation**</u>

This continuous measure confirms the same broad direction as `cte.achievement` — more CTE engagement associates with staying regional and less associates with disconnection — but reveals a sharper distinction that the categorical version smoothed over: by Year 5 and Year 10, it's specifically Sustained WF-NE that stands out as the uniquely CTE-intensive group, while Sustained WF-MN (elsewhere shown to be the highest-academic-achievement group on ACT score and MCA scores) actually shows *fewer* CTE courses than Sustained WF-NE. This reinforces a coherent overall story for this population: two distinct profiles lead to different forms of workforce success — a CTE-intensive profile that anchors specifically to the region, and an academically-stronger, less CTE-engaged profile that leads to sustained employment but more often *outside* the region. Given the redundancy with `cte.achievement` established earlier in this project, and the fact that this variable's effect size shrinks over time while the categorical version stays stable, `cte.achievement` remains the better candidate for the eventual multivariate model — but this continuous version has usefully surfaced the Sustained WF-NE/WF-MN distinction that the categorical analysis alone would have missed, and that nuance is worth preserving in the narrative even if only one version enters the formal model.

::: panel-tabset
#### 1-year after graduation

```{r}
data <- master.1

summary <- aov(total.cte.courses.taken ~ grad.year.1, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college bivariate/High school accomplishments/total-cte-courses-taken-1.xlsx", overwrite = TRUE)

```

#### 5-years after graduation

```{r}
data <- master.5

summary <- aov(total.cte.courses.taken ~ grad.year.5, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college bivariate/High school accomplishments/total-cte-courses-taken-5.xlsx", overwrite = TRUE)

```

#### 10-years after graduation

```{r}
data <- master.10

summary <- aov(total.cte.courses.taken ~ grad.year.10, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college bivariate/High school accomplishments/total-cte-courses-taken-10.xlsx", overwrite = TRUE)

```
:::

<br>

# Post-secondary

## `attended.ps`

This variable — whether a non-completer ever *attempted* postsecondary education at all — shows by far the strongest relationship of any variable examined in this entire bivariate analysis, at every time point (Year 1: χ² = 862.71, df = 3, p \< .001, Cramér's V = 0.212; Year 5: χ² = 412.44, V = 0.157; Year 10: χ² = 113.68, V = 0.106). Even at its weakest (Year 10), this exceeds the strongest effect found for any other variable in this analysis; at Year 1, it's nearly double the next-strongest relationship (English learner status, V = 0.134).

**This relationship weakens substantially over time — the sharpest decline of any variable examined**, from V = 0.212 at Year 1 down to V = 0.106 at Year 10 (roughly a 50% reduction). This is a distinct pattern from most other variables in this project, which either held steady or strengthened with time.

**Attempting postsecondary — even without completing — is associated with meaningfully better outcomes at every time point, especially early on.**

- **Sustained WF-NE:** those who attended show notably higher rates at every point — 35.6% vs. 19.9% at Year 1 (a 15.7-point gap, the largest single gap found anywhere in this bivariate analysis), narrowing to 32.1% vs. 23.8% at Year 5 and 28.2% vs. 22.9% at Year 10.
- **No MN employment record:** those who attended show dramatically lower rates at Year 1 (13.2% vs. 30.1%, a 16.9-point gap — the single largest gap of any outcome/variable pairing in this entire analysis), though this reverses direction by Year 5 (23.0% vs. 37.1% — still a large gap, same direction) and continues into Year 10 (31.7% vs. 41.8%).
- **Sustained WF-MN:** consistently higher for those who attended (10.1% vs. 5.8% at Year 1, narrowing to 16.8% vs. 14.4% by Year 10).
- **Non-sustained WF:** the only outcome where the direction is inconsistent — slightly *lower* for attendees at Year 1 (41.1% vs. 44.2%) but slightly *higher* by Year 5 (29.7% vs. 27.5%) and Year 10 (23.4% vs. 20.9%). This mirrors the same kind of early/late crossover seen for AP exam and ACT participation elsewhere in this analysis, though smaller in magnitude.

<u>**Overall Interpretation**</u>

Simply attempting postsecondary education — regardless of whether it was completed — is the single strongest predictor of workforce outcomes found anywhere in this bivariate analysis, particularly at Year 1, where it shows both the largest positive effect on sustained regional employment and the largest negative effect on disconnection from the tracked labor market of any variable/outcome pairing examined. This is a genuinely important finding for how this population should be understood: "non-completer" is not a single, uniform group — those who never attended postsecondary at all appear to fare meaningfully worse across nearly every outcome than those who attended but didn't finish, especially in the crucial first year after high school. The effect fades by roughly half by Year 10, suggesting this advantage is strongest as an early transition cushion rather than a permanent structural difference, though it remains the strongest single predictor even at that point. Given how large this effect is, and how directly it speaks to a distinction (attempted-but-not-completed vs. never-attempted) that the "non-completer" framing has treated as a single population throughout this analysis, this variable deserves prominent placement in the synthesis and is a strong candidate for inclusion in the eventual multivariate confirmation model.

::: panel-tabset
#### 1-year after graduation

```{r}
data <- master.1

tab_xtab(var.row = data$grad.year.1,
         var.col = data$attended.ps,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/Non-college bivariate/Post-secondary/attended-ps-1.html")
```

#### 5-years after graduation

```{r}
data <- master.5

tab_xtab(var.row = data$grad.year.5,
         var.col = data$attended.ps,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/Non-college bivariate/Post-secondary/attended-ps-5.html")

```

#### 10-years after graduation

```{r}
data <- master.10

tab_xtab(var.row = data$grad.year.10,
         var.col = data$attended.ps,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/Non-college bivariate/Post-secondary/attended-ps-10.html")
```
:::

<br>

# Bivariate Analysis Summary: Non-College Graduates

## Executive Summary

This bivariate analysis examined 20 independent variables across three demographic, high-school-characteristic, and high-school-accomplishment domains, tested against workforce participation outcomes at 1, 5, and 10 years after high school graduation, among the 58% of Northeast Minnesota graduates who never completed a postsecondary credential.

**Strength ranking (peak Cramér's V or eta² across the three time points):**

| Variable | Peak effect size | Pattern over time |
|------------------------|------------------------|------------------------|
| Attended postsecondary (Y/N) | V = 0.212 (strongest of any variable) | Weakens substantially (÷2) |
| English learner status | V = 0.134 | Strongest at Year 1, then flat |
| Special Ed status | V = 0.124 | U-shaped (strong → dip → partial recovery) |
| Took ACT (participation) | V = 0.123 | Stable, slightly strengthens |
| Race/Ethnicity | V = 0.063 | Mildly strengthens |
| Economic status | V = 0.083 | Strengthens steadily, most clearly of any demographic |
| CTE achievement (categorical) | V = 0.066 | Stable |
| Gender | V = 0.072 | Flat/stable |
| AP exam participation | V = 0.061 | Stable |
| Dem_Desc (district type) | V = 0.056 | Weakest of the high-school-characteristic variables |
| PSEO participation | V = 0.081 | Weak, modest strengthening |
| ACT composite score (test-takers) | eta² ≈ 1.4-1.8% | Stable, moderate |
| Total CTE courses (continuous) | eta² ≈ 0.5-0.9% | Weakens over time |
| MCA Math | eta² ≈ 0.3-1.0% | Strengthens substantially |
| MCA Reading | eta² ≈ 0.3-1.1% | Strengthens substantially (mirrors MCA Math) |
| MCA Science | eta² ≈ 0-1.2% | Not significant at Year 1; substantial missingness (up to 42%) |
| avg.wages.pct.state | eta² ≈ 0.2-0.35% | Stable, small |
| avg.unemp.rate | eta² ≈ 0.1-0.34% | Shifts in mechanism, not just magnitude |

**Two data-quality findings shape how every result above should be read:**

1.  **PSEO tracking gap (pre-2011):** required restricting all PSEO-dependent analyses to 2011+ cohorts, consistent with the fix applied throughout this project's college-completer models.
2.  **MCA Science missingness (up to 42% by Year 10):** far larger than MCA Math or Reading, and it grows with time horizon — this variable should be treated as the least reliable of the three MCA subjects for this population, mirroring the same finding in the college-completer analysis.

<br>

## The Story This Data Tells

Picture the 58% of Northeast Minnesota graduates who never finish a postsecondary credential. What separates the ones who build a stable life in the region from the ones who don't?

**The first and most important divide isn't about ability — it's about whether college was ever attempted at all.** Simply *attempting* postsecondary education, even without finishing, is the single strongest predictor found anywhere in this analysis — stronger than race, stronger than disability status, stronger than academic test scores. In the first year after high school alone, those who tried college are almost twice as likely to have found sustained regional employment, and less than half as likely to have vanished from the state's tracked labor force. "Non-completer" is not one population; it's at least two very different ones, and the gap between them is largest exactly when it matters most — the fragile first year of the transition.

**A second divide runs through academic engagement itself, and it splits the group in a genuinely two-directional way.** Career and technical education consistently and cleanly anchors graduates to the region — every step up in CTE involvement predicts staying, at every single time point measured, without exception. Simply taking the ACT does the same thing, somewhat surprisingly: test-takers who don't finish college still end up more engaged with the labor market, more often in some form of sustained employment, and less likely to disappear from the record. But *scoring well* on that same test tells the opposite story: among test-takers, the highest scorers are consistently the ones who end up employed outside the region rather than within it — and by the ten-year mark, this is one of the widest, most persistent gaps in the entire analysis. Academic proficiency (MCA Math and Reading) tells a related but distinct version of the same story: the strongest students who don't finish college are found in stable jobs, but increasingly outside the region as the years pass, while the weakest students cluster in unstable, non-sustained work. Put simply: this population splits into a *regionally-anchored, CTE-oriented pathway* and an *academically-stronger, more mobile pathway* — and the region keeps the first group but tends to lose the second.

**A third pattern, distinct from the first two, runs through economic and geographic circumstance.** Students from areas with weaker local labor markets — higher unemployment, lower relative wages — are increasingly likely, as time passes, to end up in sustained employment specifically *outside* the region rather than within it or disconnected from work entirely. This isn't a story of instability; it's a story of relocation. Struggling local economies appear to push their graduates toward opportunity elsewhere in the state, not toward failure.

**A fourth and more troubling pattern runs through demographics, and it deepens rather than resolves with time.** English learners and several racial/ethnic minority groups — particularly Asian/Pacific Islander and Hispanic graduates — show a disparity concentrated specifically in disconnection from Minnesota's tracked employment system, and this gap *widens*, not narrows, across the ten-year window; by Year 10, over half of English learners in this population have no Minnesota employment record at all. Special Education status shows perhaps the broadest disparity of any demographic factor — depressing regional attachment *and* elevating disconnection simultaneously — though unlike the racial/language disparities, its size is largest at Year 1 and partially resolves over time. Economic disadvantage tells a different story again: rather than predicting disconnection, it predicts employment *instability* specifically, a gap that grows steadily larger the longer these graduates are out of high school.

**Together, these four threads describe a population whose fate is shaped less by any single factor and more by which combination of doors they walked through.** Whether college was ever attempted, whether CTE or academic testing shaped their high school years, what the local economy looked like at home, and which demographic group they belong to — each of these operates somewhat independently, and several point toward genuinely different mechanisms (anchoring vs. mobility vs. instability vs. disconnection) rather than a single "doing well or not" axis. This is the central challenge — and opportunity — for the multivariate stage: confirming which of these bivariate patterns hold up as independent effects once they're all considered together, and which ones turn out to be standing in for each other.
