---
title: "Deeper-Dive Analysis Framework: Non-College Graduates Without Meaningful Workforce Participation"
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
library(htmlwidgets)
library(RColorBrewer)
library(readxl)
library(janitor)
library(lubridate)
library(systemfonts)
reset_font_cache()
library(ggtext)
library(gmodels)
library(fastDummies)
library(car)
library(glmnet)
library(glmnetUtils)
library(pscl)
library(sjPlot)
library(ggforce)
library(openxlsx)
```

```{r themes and shapefiles, include=FALSE}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", linewidth  = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.key.size = unit(1, "lines"),
        legend.margin = margin(0,0,0,0),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.text = element_text(margin = margin(l = 2)),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", linewidth = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))


theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
    select(5,6) %>%
    unique() %>%
    mutate(edr = str_replace(edr, "  ", " "),
           planning.region = str_replace(planning.region, " Minnesota", ""),
           planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
           edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

counties.regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
  rename(mif = `MIF Region`) %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Name = str_replace(Name, "Mcleod", "McLeod"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc) ,
         edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"))

counties.regions.1 <- counties.regions %>%
  mutate(statefp = "27",
         project.pr = ifelse(edr %in% c("EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 4 - West Central"), "Northwest",
                             ifelse(edr == "EDR 3 - Arrowhead", "Northeast",
                                    ifelse(edr %in% c("EDR 5 - North Central", "EDR 7E- East Central", "EDR 7W- Central"), "Central",
                                           ifelse(edr %in% c("EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 8 - Southwest"), "Southwest",
                                                  ifelse(edr %in% c("EDR 9 - South Central", "EDR 10 - Southeast"), "Southern", as.character(planning.region)))))),
         project.pr = fct_relevel(project.pr, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southern"))


color.ruca <- c("Entirely rural" = "#009933", "Town/rural mix" = "#99CC33", "Urban/town/rural mix" = "#CC9966", "Entirely urban" = "#754C29", "Minnesota" = "black")

color.pr <- c("Northwest" = 	"#4575b4", "Northeast" = "grey", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.edr.simple <- c("EDR 1" = "#b3cde3", "EDR 2" = "#8c96c6", "EDR 3" = "#fe9929", "EDR 4" = "#8856a7", "EDR 5" = "#810f7c", "EDR 6E" = "#e5f5f9", "EDR 6W" = "#bdc9e1", "EDR 7E" = "#99d8c9", "EDR 7W" = "#2ca25f", "EDR 8" = "#74a9cf", "EDR 9" = "#0570b0", "EDR 10" = "#d7301f", "EDR 11" = "#d8b365", "Minnesota" = "black")

color.pr.edr <- c ("Northwest" = "#4575b4","Northeast" = "#e0f3f8", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black", "EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365")

mn_counties <- st_read("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Shapefiles/County shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE) %>%
  rename(countyfp = FIPS_CODE)
```

```{r master}
master.original <- read_csv("Data/SLEDS/Masters/Master-report-dataset.csv") 

kable(names(master.original))

kable(head(master.original))
```

# Research Questions: Workforce Trajectories of Non-College Graduates in Northeast Minnesota

This phase of the analysis examines workforce participation trajectories among high school graduates in Northeast Minnesota who did not complete a postsecondary credential. Rather than relying on a single point-in-time workforce outcome, the analysis focuses on patterns of labor market attachment during the first five years after exiting formal education, with particular attention to persistence, instability, and regional attachment.

The research is guided by the following questions:

1.  Among non-college graduates, how persistent is being disconnected from sustained workforce participation either in the region or state?

2.  For non-college graduates who experience non-sustained workforce participation, where does employment occur when it is present?

3.  Among individuals categorized as having no Minnesota employment record during the first five years after high school, how persistent is this status?

4.  How are individual characteristics associated with workforce disconnection and subsequent employment trajectories?

<br>

# Study Population and Group Definitions

**Study Population**

The analytic population consists of individuals who:

- Graduated from a high school in Northeast Minnesota, and
- Did not complete a postsecondary credential (`ps.grad = No`), and,
- Have workforce participation records available for at least the first five years after high school graduation OR the first five years after exiting formal post-secondary education.

<br>

**Definition of Workforce Disconnection**

Workforce participation is measured annually during the first five years after high school or existing post-secondary attendance. Individuals are classified as **disconnected** in a given year if they are categorized as either:

- Non-sustained workforce participation, or
- No Minnesota employment record.

Using this annual classification, individuals are grouped based on the persistence of disconnection during years 1 through 5:

- **Persistently Disconnected:**\
  Disconnected in four or five of the first five years after graduation.

- **Disconnected (Broad Definition):**\
  Disconnected in three or more of the first five years after graduation.

- **Intermittently Disconnected:**\
  Disconnected in one or two of the first five years after graduation.

- \*\*Consistently attached:\*\* Have sustained workforce participation in all five years.

These definitions emphasize sustained labor market detachment rather than short-term transitions.

<br>

**Regional Attachment**

For years in which individuals are employed but not meaningfully attached to the workforce, employment location is classified as:

- Within Northeast Minnesota, or
- Outside Northeast Minnesota (elsewhere in Minnesota).

This distinction is used to assess whether workforce disconnection reflects regional underutilization or disengagement from the regional labor market.

<br>

**Framing Consideration**

Together, these definitions allow the analysis to move beyond point-in-time workforce outcomes and toward a trajectory-based understanding of early-career labor market experiences. This approach supports clearer interpretation of workforce instability, recovery, and regional retention among non-college graduates in Northeast Minnesota.

<br>

# Methodology and Analytic Steps

This analysis examines workforce participation trajectories among non-college graduates in Northeast Minnesota, with a focus on identifying patterns of workforce disconnection, regional attachment, and employment stability during the first five years after high school.

## Step 1: Data Preparation and Analytic Population

**1.1 Analytic Population**

The analytic population is restricted to individuals who:

- Graduated from a high school in Northeast Minnesota, and
- Did not complete a postsecondary credential (`ps.grad = No`), and
- Have complete workforce participation categorizations for years 1 through 5 after high school or,
- first 5 years after exiting attendance in a post-secondary institution.

**1.2 Annual Workforce Participation Classification**

For each individual, annual workforce participation status is identified for: - `grad.year.1` - `grad.year.2` - `grad.year.3` - `grad.year.4` - `grad.year.5`

Each year is classified into one of the following categories:

- Meaningful workforce participation
- Not meaningful workforce participation
- No Minnesota employment record.

It should be noted that "grad.year.x" here means years after exiting formal education, not years after graduating high school.

**1.3 Workforce Disconnection Indicators**

For each annual observation, a binary **disconnection indicator** is created:

- **Disconnected (year-level)** = Non-sustained WF - NE or MN, **or** No Minnesota emp record

Using these annual indicators, individuals are grouped based on persistence of disconnection across years 1–5:

- **Persistently Disconnected:** Disconnected in 4 or 5 of the first 5 years
- **Disconnected (Broad):** Disconnected in 3 of 5 years
- **Intermittently Disconnected:** Disconnected in 1–2 of the first 5 years
- **Consistently Attached:** Disconnected in 0 of the first 5 years

These categories emphasize sustained labor market attachment or detachment rather than point-in-time outcomes.

**1.4 Regional Location of Not Meaningful Employment**

For years in which individuals are categorized as **Non-sustained WF**, employment location is further classified as:

- Within Northeast Minnesota
- Outside Northeast Minnesota (elsewhere in Minnesota)
- Both (simultaneous employment inside and outside the region)

This classification allows identification of regional underemployment versus regional disengagement.

**1.5 Employment Intensity and Job Holding**

For each individual and each year (years 1–5), the following employment intensity measures are constructed:

- Number of distinct employers during the year
- Indicator for multiple job holding (two or more employers)

These measures are used to distinguish underemployment from precarious or fragmented employment.

<br>

## Step 2: Descriptive and Summary Statistics

**2.1 Distribution of Workforce Disconnection**

Summary statistics are produced to describe:

- The proportion of non-college graduates in each workforce disconnection category
- Differences in persistence and intermittency of disconnection

This establishes the overall scope and severity of workforce disconnection in early adulthood.

**2.2 Regional Attachment and Job Structure**

Among individuals experiencing workforce disconnection, summary statistics examine:

- The location of not meaningful employment (within region, outside region, both)
- Average and median number of employers per year
- Prevalence of multiple job holding

These summaries clarify whether disconnection reflects local underemployment, out-of-region work, or fragmented employment patterns.

<br>

## Step 3: Relationship Between Individual Characteristics and Disconnection Trajectories

Step 3 will focus on answering the question of which characteristics are associated with the disconnection persistence categories. This will be a bi-variate analysis.

**Independent Variables**

We will use the following variables that were selected based on our earlier multiple correspondence analysis.

- Demographics
  - `Gender`
  - `RaceEthnicity`
  - `economic.status`
  - `SpecialEdStatus`
  - `english.learner`
- High school characteristics
  - `Dem_Desc`
  - `avg.wages.pct.state`
  - `avg.unemp.rate`
- High school academics
  - `pseo.participant`
  - `ap.exam`
  - `took.ACT`
  - `ACTCompositeScore`
  - `MCA.M`
  - `MCA.R`
  - `MCA.S`
  - `cte.achievement`
  - `total.cte.courses.taken`
- Post-secondary
  - `attended.ps`

**3.2 Analytic Approach**

- Categorical variables are examined using cross-tabulations and measures of association (e.g., Cramer’s V).
- Continuous variables are examined using ANOVA.
- Interpretation emphasizes substantive differences and patterns rather than statistical significance alone.

**3.3 Stratified and Comparative Analyses**

Where sample sizes permit, additional analyses compare:

- Persistently disconnected versus intermittently disconnected individuals
- Individuals disconnected within the region versus outside the region
- Individuals who transition into meaningful workforce participation versus those who do not

These comparisons help identify which characteristics are associated with stability, recovery, and regional retention.

<br>

## Framing and Interpretation

This methodology prioritizes a trajectory-based understanding of workforce participation, recognizing that early-career labor market experiences are dynamic rather than static. By combining persistence measures, regional employment location, and job-holding intensity, the analysis aims to provide actionable insights for workforce development and labor force planning in Northeast Minnesota.

<br>

# Data prep

## Step 1: Analytic population

The first step is filtering for our population;

- `ps.grad` = "No"

```{r}
analytic.pop <- master.original %>%
  filter(ps.grad == "No",
         grad.year.5 != "After 2023") 

kable(head(analytic.pop))
kable(names(analytic.pop))
```

## Step 2: Identify Workforce State for 5 Years After Formal Education

Next, we need to pull out all the individuals in which we have five years of workforce categories after they either graduate from high school or the last year in which they exited post-secondary attendance.

After that, we will categorize their workforce participation state into the following;

- Sustained WF - NE
- Sustained WF - MN
- Non-sustained WF
- No MN emp record.

I already created the annual workforce participation states for every individuals one year after graduation through 15 years after graduation. I will bring that list in and filter it by our analytic population.

```{r}
annual.wf.states <- read_csv("Data/SLEDS/Masters/States/annual-seven-states.csv") %>%
  select(PersonID, grad.year, states) %>%
  mutate(grad.year = fct_relevel(grad.year, "grad.year.0", "grad.year.1", "grad.year.2", "grad.year.3", "grad.year.4", "grad.year.5", "grad.year.6", "grad.year.7", "grad.year.8", "grad.year.9", "grad.year.10", "grad.year.11", "grad.year.12", "grad.year.13", "grad.year.14", "grad.year.15")) 

no_ps_attendance <- analytic.pop %>%
  filter(attended.ps == "No") %>%
  select(PersonID) %>%
  left_join(annual.wf.states, by = "PersonID") %>%
  filter(states != "After 2023",
         states != "Attending ps",
         grad.year %in% c("grad.year.1", "grad.year.2", "grad.year.3", "grad.year.4", "grad.year.5")) %>%
  group_by(PersonID) %>%
  mutate(n = n()) %>%
  ungroup() %>%
  filter(n == 5) %>%
  mutate(states = as_factor(states))

last_ps_attendance <- analytic.pop %>%
  filter(attended.ps == "Yes") %>%
  select(PersonID) %>%
  left_join(annual.wf.states, by = "PersonID") %>%
  filter(states != "After 2023") %>%
  mutate(numeric.grad.year = str_sub(grad.year, -2, -1),
         numeric.grad.year = str_replace(numeric.grad.year, "\\.", ""),
         numeric.grad.year = as.integer(numeric.grad.year)) %>%
  filter(states == "Attending ps") %>%
  group_by(PersonID) %>%
  summarize(last_ps_year = max(numeric.grad.year, na.rm = TRUE)) %>%
  ungroup() 

ps_attendance <- annual.wf.states %>%
  right_join(last_ps_attendance, by = "PersonID") %>%
  filter(states != "After 2023") %>%
  mutate(numeric.grad.year = str_sub(grad.year, -2, -1),
         numeric.grad.year = str_replace(numeric.grad.year, "\\.", ""),
         numeric.grad.year = as.integer(numeric.grad.year),
         final.year = last_ps_year + 5) %>%
  group_by(PersonID) %>%
  filter(numeric.grad.year <= final.year,
         numeric.grad.year > last_ps_year) %>%
  mutate(n = n()) %>%
  ungroup() %>%
  filter(n == 5) %>%
  select(PersonID, grad.year, states, n)

master.annual.wf.five.year.states <- ps_attendance %>%
  rbind(no_ps_attendance) %>%
  mutate(states = str_replace(states, "Not meaningful, not attending ps", "Non-sustained WF"),
         states = str_replace(states, "No MN emp record, not attending ps", "No MN emp record"),
         states = ifelse(states %in% c("Meaningful emp County", "Meaningful emp EDR"), "Sustained WF - NE", states),
         states = str_replace(states, "Meaningful emp MN", "Sustained WF - MN"),
         states = fct_relevel(states, "Sustained WF - NE", "Sustained WF - MN", "Non-sustained WF", "No MN emp record")) %>%
  select(-n) %>%
  group_by(PersonID) %>%
  arrange(grad.year) %>%
  mutate(year.after.exit = seq(n())) %>%
  ungroup()


kable(head(master.annual.wf.five.year.states))

kable(names(master.annual.wf.five.year.states))

```

We now have a dataset consisting of `r comma(nrow(master.annual.wf.five.year.states))` individuals that meet the criteria that we want to examine along with their workforce categories from 1 to 5 years after exiting formal education.

<br>

## Step 3: Workforce Disconnection Indicators

Next we add another category to determine the level of disconnection, if any, for each individual. Disconnected is defined as any individual that was categorized as not having meaningful workforce participation or no Minnesota employment record in a given year one through five years after exiting formal education.

The level of disconnection is defined as the following;

- **Persistently Disconnected:** Disconnected in 4 or 5 of the first 5 years
- **Broadly Disconnected:** Disconnected in 3 of the first 5 years
- **Intermittently Disconnected:** Disconnected in 1–2 of the first 5 years
- **Consistently Attached:** Disconnected in 0 of the first 5 years

```{r}
disconnected.level <- master.annual.wf.five.year.states %>%
  mutate(disconnected = ifelse(states %in% c("Non-sustained WF", "No MN emp record"), "Disconnected", "Attached"),
         disconnected = as.factor(disconnected)) %>%
  group_by(PersonID, disconnected) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  complete(PersonID, disconnected, fill = list(n = 0)) %>%
  filter(disconnected == "Disconnected") %>%
  mutate(disconnected.cat = ifelse(n == 0, "Consistently attached",
                                   ifelse(n %in% c(1,2), "Intermittently disconnected",
                                          ifelse(n == 3, "Broadly disconnected",
                                                 ifelse(n %in% c(4,5), "Persistently disconnected", "A")))),
         disconnected.cat = fct_relevel(disconnected.cat, "Consistently attached", "Intermittently disconnected", "Broadly disconnected", "Persistently disconnected"))

kable(names(disconnected.level))
kable(head(disconnected.level))
```

The dataset now has `r comma(nrow(disconnected.level))` individuals that did not graduate college AND have 5 years of workforce categories immediately after exiting formal education categorized by workforce disconnection levels.

## **Step 4 & 5: Regional Location of Not Meaningful Employment**

For years in which individuals are categorized as **not meaningfully attached to the workforce**, employment location is further classified as: - Within Northeast Minnesota - Outside Northeast Minnesota (elsewhere in Minnesota) - Both (simultaneous employment inside and outside the region)

This classification allows identification of regional underemployment versus regional disengagement.

The population of individuals we need are the following;

- Did not graduate college
- For each individual categorize whether they had not meaningful workforce participation in the region, not in the region, or both.

We will use the `master.annual.wf.five.year.states` dataset as the starting point. We will have to have the year they graduated high school so when I match up employment records I'm only using employment records for the first 5 years after exiting formal education.

```{r}
hs.grad.year <- read_csv("Data/SLEDS/K12 Enrollment/K12Enrollment-master.csv")

analytic.pop.annual.wf.states.hs.grad.year <- master.annual.wf.five.year.states %>%
  left_join(hs.grad.year[,c(1,9)], by = "PersonID")  %>%
  rename(after.ed.year = grad.year.x,
         hs.grad.year = grad.year.y) %>%
  mutate(numeric.after.ed.year = str_sub(after.ed.year, -2, -1),
         numeric.after.ed.year = str_replace(numeric.after.ed.year, "\\.", ""),
         numeric.after.ed.year = as.integer(numeric.after.ed.year),
         emp.year = hs.grad.year + numeric.after.ed.year) 
  
kable(head(analytic.pop.annual.wf.states.hs.grad.year))

kable(names(analytic.pop.annual.wf.states.hs.grad.year))
```

Now we need to bring in the employment data for Northeast graduates and then filter for PersonID that are in the newly created dataset. The columns we are going to keep from the employment dataset are the following;

- PersonID: Identifies the graduate from Northeast Minnesota
- TimeID: Last day of fiscal quarter
- EmploymentOrganizationID: An internal identification for each employer.
- EmployeeQuarterlyHoursWorked: The total number of hours worked by an employee at an employer’s location during a specific calendar quarter
- EmployeeQuarterlyWages: The total wages earned by an employee at an employer’s location during a specific calendar quarter
- EmployeeTenure: The total number of quarters a worker has been employeed by an organization at a specific location, up to an including the current calendar quarter
- CountyName: The name of the county corresponding to the County and MNCounty Codes for this Location.

```{r}
employment.original <- read_csv("Data/SLEDS/Employment/Employment-Northeast.csv") %>%
  drop_na(PersonID)

emp.year <- employment.original %>%
  filter(!is.na(TimeID)) %>%
  mutate(emp.year = str_sub(TimeID, 1, 4),
         emp.year = as.integer(emp.year),
         quarter = str_sub(TimeID, 5,6),
         quarter = ifelse(quarter == "03", 1,
                          ifelse(quarter == "06", 2,
                                 ifelse(quarter == "09", 3,
                                        ifelse(quarter == "12", 4, 0))))) %>%
  select(PersonID, EmploymentOrganizationID, EmployeeQuarterlyHoursWorked, EmployeeTenure, CountyName, emp.year, quarter) %>%
  mutate(CountyName = str_replace(CountyName, " County", ""),
         CountyName = str_replace(CountyName, " County", ""),
         CountyName = str_replace(CountyName, "Saint Louis", "St. Louis"),
         CountyName = str_replace(CountyName, "City of Duluth", "St. Louis"),
         CountyName = str_replace(CountyName, "City of Minneapolis", "Hennepin"),
         CountyName = str_replace(CountyName, "City of St. Paul", "Ramsey"),
         CountyName = str_replace(CountyName, "Shakopee Mdewakanton Community", "Scott"),
         CountyName = str_replace(CountyName, "White Earth Tribal Counci", "Mahnomen")) %>%
  left_join(counties.regions.1[,c(1,2,10)], by = c("CountyName" = "Name"))

not.meaningful.pop.emp.year <- analytic.pop.annual.wf.states.hs.grad.year %>%
  filter(states == "Non-sustained WF") %>%
  select(PersonID, emp.year)

not.meaningful.emp.quarter <- not.meaningful.pop.emp.year %>%
  left_join(emp.year, by = c("PersonID", "emp.year")) %>%
  mutate(project.pr = ifelse(is.na(project.pr), "Unclassified", as.character(project.pr)),
         in_region = (project.pr == "Northeast"),
         unclassified_region = (project.pr == "Unclassified"),
         out_region = c(!in_region & !unclassified_region)) %>%
  group_by(PersonID, emp.year, quarter) %>%
  summarise(n.emp = n(),
         any_in_quarter = any(in_region, na.rm = TRUE),
         any_out_quarter = any(out_region, na.rm = TRUE),
         any_unclassified_quarter = any(unclassified_region, na.rm = TRUE),
         any_mixed_qtr =  (sum(c(any_in_quarter, any_out_quarter, any_unclassified_quarter), na.rm = TRUE) >= 2)) %>%
  ungroup()

not.meaningful.emp.annual <- not.meaningful.emp.quarter %>%
  group_by(PersonID, emp.year) %>%
  summarize(multiple_emp_same_quarter = any(n.emp > 1),
            max_emp_in_qtr = max(n.emp),
            any_in_region = any(any_in_quarter, na.rm = TRUE),
            any_out_region = any(any_out_quarter, na.rm = TRUE),
            any_unclassified_region = any(any_unclassified_quarter, na.rm = TRUE),
            any_mixed_region = any(any_mixed_qtr, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(region_location_year = case_when(any_in_region & !any_out_region & !any_unclassified_region ~ "in_only", !any_in_region & any_out_region & !any_unclassified_region ~ "out_only", !any_in_region & !any_out_region & any_unclassified_region  ~ "unclassified_only", TRUE ~ "mixed" )) %>%
  mutate(
    mixed_type = case_when(
      region_location_year != "mixed" ~ NA_character_,
      any_in_region & any_out_region & !any_unclassified_region ~ "in_out",
      any_in_region & any_unclassified_region & !any_out_region ~ "in_unclassified",
      any_out_region & any_unclassified_region & !any_in_region ~ "out_unclassified",
      any_in_region & any_out_region & any_unclassified_region  ~ "in_out_unclassified"
    ),
    both_structure = case_when(region_location_year != "mixed" ~ NA_character_,
                               any_mixed_region ~ "simultaneous",
                               !any_mixed_region ~ "sequential",
                               TRUE ~ NA_character_)) %>%
  select(-any_in_region, -any_out_region, -any_unclassified_region, - any_mixed_region)


kable(head(not.meaningful.emp.annual))

kable(names(not.meaningful.emp.annual))
```

We have created a dataset with `r comma(nrow(not.meaningful.emp.annual))` individuals that do not have meaningful workforce participation for any given year up to 5 years after graduating high school that also provides if they had multiple employers in one quarter, the max number of different employers in one quarter for that employment year, and whether the employer was located in the region, outside the region, unclassified, or a mixture of these categories. Another column further refines the "mixed" category into whether it was sequential - meaning the mix occurred in separate quarters - or simultaneous - meaning there was a mix within the same quarter.

Next, we will piece this back together with the entire analytic population and create a complete annual workforce state for all individuals that did not complete post-secondary and has 5 years of workforce categories after graduating high school.

```{r}
Master.not.meaningful <- not.meaningful.emp.annual %>%
  left_join(hs.grad.year[,c(1, 9)], by = "PersonID") %>%
  rename(hs.grad.year = grad.year) %>%
  mutate(grad.year = emp.year - hs.grad.year,
         grad.year = paste("grad.year.", grad.year, sep = "")) %>%
  complete(PersonID, grad.year) %>%
  mutate(grad.year = fct_relevel(grad.year, "grad.year.0", "grad.year.1", "grad.year.2", "grad.year.3", "grad.year.4", "grad.year.5", "grad.year.6", "grad.year.7", "grad.year.8", "grad.year.9", "grad.year.10", "grad.year.11", "grad.year.12", "grad.year.13", "grad.year.14", "grad.year.15"))


Master.annual.wf.states.disconnected <- master.annual.wf.five.year.states %>%
  left_join(Master.not.meaningful, by = c("PersonID", "grad.year")) %>%
  unite("wf.states.refined",
        states, region_location_year, mixed_type, both_structure,
        sep = "_",
        na.rm = TRUE) %>%
  mutate(year.after.exit = paste("after.ed.", year.after.exit, sep = "")) %>%
  select(PersonID, year.after.exit, wf.states.refined) %>%
  pivot_wider(names_from = year.after.exit, values_from = wf.states.refined) %>%
  left_join(disconnected.level[,c(1,4)], by = "PersonID")

kable(head(Master.annual.wf.states.disconnected)) 

kable(names(Master.annual.wf.states.disconnected))
```

Okay, now I need to bring this all together.

```{r}
master <- analytic.pop %>%
  right_join(Master.annual.wf.states.disconnected, by = "PersonID") %>%
  mutate(disconnected.cat = fct_relevel(disconnected.cat, "Consistently attached", "Intermittently disconnected", "Broadly disconnected", "Persistently disconnected"))

kable(head(master))

kable(names(master))

write_csv(master, "Data/Claude/Non-college deeper/Master-dataset.csv")
```

The data set has `r comma(nrow(master))` individuals in it. These individuals did not complete post-secondary and had workforce categories for 5 years after exiting formal education.

<br>

# Descriptive analysis

## **Distribution of Workforce Disconnection**

**More than half of non-college graduates in Northeast Minnesota are persistently disconnected from sustained workforce participation during their first five years out of school or postsecondary attendance.** This is the single largest category by a wide margin — more than double the next-largest group, and more than four times the size of the "Consistently attached" group.

**Sustained, uninterrupted workforce attachment is rare.** Only about 1 in 8 non-college graduates (11.9%) achieve sustained participation in every one of their first five years. This is essentially tied with the smallest disconnection category (Broadly disconnected, also 11.9%) — meaning a graduate in this population is about as likely to be attached in *every* year as they are to be disconnected in exactly three of five.

**Combining the two most severe categories, nearly two-thirds (65.1%) of non-college graduates spend at least three of their first five years disconnected from sustained workforce participation.** Only about a third (34.9%) spend three or more years attached (Intermittent + Consistently attached, loosely).

**Overall Interpretation**

This distribution reframes the scale of the workforce disconnection problem for non-college graduates in a way that point-in-time snapshots (like the earlier `grad.year.1/5/10` distributions) could not fully capture. Rather than disconnection being a transitional, early-career phenomenon that most people eventually move past, the data shows it is the *dominant, persistent experience* for a majority of this population — over half spend nearly their entire first five years disconnected. This is a markedly starker picture than the single-year distributions examined earlier in this project suggested, and it substantially raises the stakes for Step 3's question of which characteristics predict landing in the "Persistently disconnected" group specifically, since that group represents the majority experience for non-completers in this region, not an outlier one.

```{r}
data <- master %>%
  group_by(disconnected.cat) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  mutate(pct = n / sum(n),
         data_id = seq(n()),
         disconnected.cat = fct_relevel(disconnected.cat, "Consistently attached", "Intermittently disconnected", "Broadly disconnected",  "Persistently disconnected"))

write_csv(data, "Data/Claude/Non-college deeper/Descriptive stats/disconnected-cat.csv")

plot <- ggplot(data, aes(disconnected.cat, pct, fill = disconnected.cat)) +
  geom_col_interactive(position = "dodge", aes(data_id = data_id, tooltip = paste(disconnected.cat, "\nN: ", comma(n), "\n%: ", percent(pct, accuracy = .1), sep = ""))) +
  geom_label(aes(label = percent(pct, accuracy = .1)), show.legend = FALSE, color = "black", size = 5, position = position_dodge(width = .9)) +
  labs(x="", y = "", color="", title = "Percent share of non-college grads,\n2008 - 2018 high school grads")+
  scale_y_continuous(labels=scales::percent)+
  theme_bar+
  scale_fill_manual(values = brewer.pal(n = 5, "RdYlBu"),
                    guide = guide_legend(ncol = 3)) +
  theme(legend.position = "bottom",
        text = element_text(size = 16),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 8, height_svg = 8) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE))      


```

<br>

## **Regional Attachment and Job Structure**

Among individuals experiencing workforce disconnection, summary statistics examine: - The location of not meaningful employment (within region, outside region, both) - Average and median number of employers per year - Prevalence of multiple job holding

These summaries clarify whether disconnection reflects local underemployment, out-of-region work, or fragmented employment patterns.

Let's start with the distribution of the various workforce states with the newly defined categories within "Not meaningful" but not all of them because it will get too messy. Let's first do the following categories;

1.  Attending ps
2.  Meaningful emp region
3.  Meaningful emp MN
4.  Not meaningful
    1.  In only
    2.  Out only
    3.  Unclassified only
    4.  Mixed
5.  No MN emp record

**The majority of non-sustained employment in this population has at least some foothold in the regional labor market.** Separating "mixed" employment by whether it includes any in-region component shows that **81.1% of all non-sustained employment in Year 1 involves at least partial regional attachment**, declining gradually to 76.6% (Year 2), 74.6% (Year 3), 72.8% (Year 4), and 73.5% (Year 5). Only a small and stable share of non-sustained work — roughly 7-8% of the total population, or under 20% of non-sustained work specifically — happens entirely outside the region with zero regional component, and this stays essentially flat across all five years.

**This points to regional underutilization, not regional disengagement, as the dominant form of instability in this population.** The people captured in non-sustained employment are overwhelmingly not people who've left for opportunity elsewhere — they're present in the regional labor market, working, just not steadily or sufficiently to clear the "meaningful employment" threshold.

**The decline in this category over the five years reflects a shrinking pool, not people leaving the region for good.** As overall non-sustained employment shrinks (43.4% → 26.4% of the population from Year 1 to Year 5), the in-region-attached share shrinks proportionally, while remaining the dominant form of non-sustained work throughout. Meanwhile, "no MN employment record" keeps growing every year — the real destination for people moving out of non-sustained status increasingly appears to be dropping out of the tracked employment system altogether, not shifting into out-of-region-only work, which stays flat around 7-8% the entire time.

**Overall Interpretation**

Underutilization — partial, unstable, or insufficient regional employment — is the dominant experience among non-college graduates who aren't sustainably employed, at every point across the first five years, even as its absolute size shrinks. Combined with the fact that "no MN employment record" is the fastest-growing and eventually largest category overall, the fuller story is: a large share of this population remains meaningfully present in the regional labor force even when their employment doesn't count as "sustained," and the real risk that grows over time isn't people leaving the region for work elsewhere (which stays flat) — it's people falling out of Minnesota's tracked employment system entirely. This is an important distinction for workforce and economic developers: the underutilization piece points toward job quality, hours, and employer-engagement questions for people who are already here and already working, while the growing "no MN record" piece points toward a separate and less well-understood question about disengagement, migration, or informal work that this data alone can't resolve.

```{r}
wf.states.factors <- c("Attending ps", "Sustained WF - NE", "Sustained WF - MN", "Non-sustained WF_in_only", "Non-sustained WF_out_only", "Non-sustained WF_unclassified_only", "Non-sustained WF - mixed", "No MN emp record")

data <- Master.annual.wf.states.disconnected %>%
  select(PersonID:after.ed.5) %>%
  pivot_longer(names_to = "grad.year", values_to = "wf.states", 2:ncol(.)) %>%
  mutate(wf.states = ifelse(wf.states %in% c("Non-sustained WF_mixed_in_out_simultaneous", "Non-sustained WF_mixed_in_out_sequential", "Non-sustained WF_mixed_in_unclassified_simultaneous" , "Non-sustained WF_mixed_in_out_unclassified_simultaneous", "Non-sustained WF_mixed_out_unclassified_simultaneous", "Non-sustained WF_mixed_out_unclassified_sequential", "Non-sustained WF_mixed_in_unclassified_sequential", "Non-sustained WF_mixed_in_out_unclassified_sequential"), "Non-sustained WF - mixed", wf.states),
         wf.states = fct_relevel(wf.states, wf.states.factors)) %>%
  group_by(grad.year, wf.states) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  group_by(grad.year) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup() %>%
  mutate(data_id = seq(n()))

write_csv(data, "Data/Claude/Non-college deeper/Descriptive stats/disconnected-pattern.csv")

plot <- ggplot(data, aes(wf.states, pct, fill = wf.states)) +
  facet_wrap(~grad.year, ncol = 2) +
  geom_col_interactive(position = "dodge", aes(data_id = data_id, tooltip = paste(grad.year, "\nWorkforce category: ", wf.states, "\nN: ", comma(n), "\nPercent: ", percent(pct, accuracy = .1), sep = ""))) +
  geom_label(aes(label = percent(pct, accuracy = .1)), show.legend = FALSE, color = "white", size = 5, position = position_dodge(width = .9)) +
  labs(x="", y = "", color="", title = "Percent share of total building construction establishments")+
  scale_y_continuous(labels=scales::percent)+
  theme_bar+
  scale_fill_manual(values = c("#8c510a", "#bf812d", "#dfc27d", "#c7eae5", "#80cdc1", "#35978f", "#01665e", "black"),
                    guide = guide_legend(ncol = 3)) +
  theme(legend.position = "none",
        text = element_text(size = 16),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE))      

```

This breakdown splits the previously collapsed "mixed" category into its component parts (simultaneous vs. sequential in/out/unclassified combinations) and confirms the collapsing decision was reasonable — nothing here changes the Year 1-5 trend already established.

**Within "mixed" patterns, simultaneous in-and-out-of-region employment dominates over sequential.** At every time point, `mixed_in_out_simultaneous` (holding jobs in and out of the region during the same year) is 2-3x larger than `mixed_in_out_sequential` (e.g., Year 1: 6.2% vs. 2.8%; Year 5: 3.2% vs. 1.6%). Both decline steadily over time, consistent with the broader "mixed"/non-sustained shrinkage already noted.

**The "unclassified" sub-patterns are uniformly negligible** — every category involving "unclassified" is under 0.6% of the population at every time point, and the smallest (`mixed_in_out_unclassified_sequential`) drops to essentially zero by Year 5 (n = 1). These can be safely folded into their nearest real category without losing any meaningful information.

**Bottom line:** this level of granularity doesn't surface anything the collapsed 7-category table didn't already show — the simultaneous-vs-sequential distinction within "mixed" is the only piece with any real weight, and even that's a minor nuance on top of the main pattern (non-sustained employment of all types shrinking over time while sustained employment and "no MN record" both grow). I'd confirm this level of detail isn't needed going forward and stick with the collapsed categories from the previous table for the rest of this analysis, per your plan.

```{r}
refined.disconnected.factors <- c( "Non-sustained WF_in_only",  "Non-sustained WF_out_only", "Non-sustained WF_unclassified_only", "Non-sustained WF_mixed_in_out_simultaneous", "Non-sustained WF_mixed_in_unclassified_simultaneous", "Non-sustained WF_mixed_out_unclassified_simultaneous", "Non-sustained WF_mixed_in_out_unclassified_simultaneous","Non-sustained WF_mixed_in_out_sequential", "Non-sustained WF_mixed_in_unclassified_sequential", "Non-sustained WF_mixed_out_unclassified_sequential", "Non-sustained WF_mixed_in_out_unclassified_sequential", "No MN emp record")

data <- Master.annual.wf.states.disconnected %>%
  select(PersonID:after.ed.5) %>%
  pivot_longer(names_to = "grad.year", values_to = "wf.state", 2:ncol(.)) %>%
  filter(!wf.state %in% c("Meaningful emp region", "Meaningful emp MN", "Attending ps")) %>%
  mutate(wf.state = fct_relevel(wf.state, refined.disconnected.factors)) %>%
  group_by(grad.year, wf.state) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  group_by(grad.year) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup() %>%
  mutate(data_id = seq(n()))

write_csv(data, "Data/Claude/Non-college deeper/Descriptive stats/disconnected-refined-patterns.csv")

plot <- ggplot(data, aes(grad.year, pct, fill = wf.state)) +
  facet_wrap(~wf.state, ncol = 2) +
  geom_col_interactive(position = "dodge", aes(data_id = data_id, tooltip = paste(grad.year, "\nWorkforce category: ", wf.state, "\nN: ", comma(n), "\nPercent: ", percent(pct, accuracy = .1), sep = ""))) +
  geom_label(aes(label = percent(pct, accuracy = .1)), show.legend = FALSE, color = "white", size = 5, position = position_dodge(width = .9)) +
  labs(x="", y = "", color="", title = "Percent share of disconnected workforce states")+
  scale_y_continuous(labels=scales::percent)+
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 16),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 8, height_svg = 12) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE))      
```

<br>

## Summary: What the Disconnection Analysis Tells Us

1.  Disconnection is not a marginal or transitional experience for this population — it's the majority experience.\*\* 53.2% of non-college graduates are persistently disconnected (4-5 of their first 5 years), compared to just 11.9% who are consistently attached. Nearly two-thirds (65.1%) spend at least three of five years disconnected. Disconnection among non-college graduates is highly persistent, not a brief early-career phase that most people age out of.
2.  Underutilization — not disengagement — is the dominant form of instability while it lasts.\*\* Among the share of this population not sustainably employed in a given year, the large majority remain at least partially attached to the regional labor market. In Year 1, 81.1% of all non-sustained employment involves some in-region component; this declines gradually but stays above 73% through Year 5. Only a small, flat share (7-8% of the population) works entirely outside the region with no regional attachment at all. These are people who are present and working in the region — just not steadily or sufficiently to clear the "meaningful employment" bar. This is a job-quality and underemployment story, not an outmigration story.
3.  Two genuinely different trends are unfolding simultaneously, and they point in opposite directions.\*\* Sustained employment (regional and elsewhere in Minnesota) grows steadily every single year, from 30.2% of the population in Year 1 to 40.1% by Year 5 — real, consistent progress. At the same time, "no Minnesota employment record" also grows every year, from 26.4% to 33.4%, and grows faster than sustained employment does — overtaking every other single category by Year 5. As non-sustained/underutilized employment shrinks over time, it is not primarily converting into out-of-region work (which stays flat around 7-8%) — it's converting more into complete disappearance from the tracked system than into any other outcome.
4.  Together, these findings sharpen what "improvement over time" actually means for this population.\*\* It is not a simple story of instability resolving into stability. It's a population splitting into two increasingly distinct groups: one that's finding steadier footing (increasingly regional, but also increasingly happy to find that stability elsewhere in the state), and another that is increasingly vanishing from Minnesota's employment records altogether — a group whose size grows every year and whose underlying cause (true labor force exit, out-of-state migration, or informal/untracked work) can't be determined from this data alone.

**What this sets up for the rest of the analysis:** the data has now established that disconnection is persistent, that most non-sustained employment remains regionally rooted rather than reflecting flight to other labor markets, and that the fastest-growing risk is disappearance from the tracked system rather than instability or out-migration. The open pieces are:

1.  directly testing the persistence of "no MN employment record" as its own trajectory (the project's third research question), and
2.  the planned bivariate work on which characteristics predict landing in "Persistently disconnected" versus "Consistently attached."

Given how much of the disconnection story is really two distinct phenomena — regional underutilization and disappearance from the record — it may be worth testing predictors against each of these separately, rather than only against the combined four-category disconnection variable, since a single predictor could plausibly explain one of these patterns well and the other poorly.

<br>

# Separating Non-sustained and No MN Emp Record

It's clear from the descriptive statistics that it's necessary to separate the Non-sustained from the MN Employment records to get a sense of how these two overlap or not. We are going to categorize just like we did before, except we will have to separate categories of "disconnection";

1.  non-sustained.persistence.cat:
    - 0 years = none;
    - 1-2 = intermittent;
    - 3 = broad;
    - 4-5 = persistent
2.  no_record.persistence.cat
    - same four buckets, applied to the no-record count instead

## Data Prep

```{r}
non_sustained.level <- master.annual.wf.five.year.states %>%
  mutate(non_sustained = ifelse(states == "Non-sustained WF", "Non-sustained", "Other"),
         non_sustained = as.factor(non_sustained)) %>%
  group_by(PersonID, non_sustained) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  complete(PersonID, non_sustained, fill = list(n = 0)) %>%
  filter(non_sustained == "Non-sustained") %>%
  mutate(non_sustained.cat = ifelse(n == 0, "Never non-sustained",
                                    ifelse(n %in% c(1,2), "Intermittently non-sustained",
                                           ifelse(n == 3, "Broadly non-sustained",
                                                  ifelse(n %in% c(4,5), "Persistently non-sustained", "A")))),
         non_sustained.cat = fct_relevel(non_sustained.cat, "Never non-sustained", "Intermittently non-sustained", "Broadly non-sustained", "Persistently non-sustained"))

no_record.level <- master.annual.wf.five.year.states %>%
  mutate(no_record = ifelse(states == "No MN emp record", "No record", "Other"),
         no_record = as.factor(no_record)) %>%
  group_by(PersonID, no_record) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  complete(PersonID, no_record, fill = list(n = 0)) %>%
  filter(no_record == "No record") %>%
  mutate(no_record.cat = ifelse(n == 0, "Never no-record",
                                ifelse(n %in% c(1,2), "Intermittently no-record",
                                       ifelse(n == 3, "Broadly no-record",
                                              ifelse(n %in% c(4,5), "Persistently no-record", "A")))),
         no_record.cat = fct_relevel(no_record.cat, "Never no-record", "Intermittently no-record", "Broadly no-record", "Persistently no-record"))

master.sep <- Master.annual.wf.states.disconnected %>%
  left_join(non_sustained.level, by = "PersonID") %>%
  left_join(no_record.level, by = "PersonID") %>%
  rename(non.sustained.n = n.x,
         no.record.n = n.y)

write_csv(master.sep, "Data/Claude/Non-college deeper/master-separated.csv")

```

The table below provides the proportion of individuals that were in the Non-sustained WF category for the first five years after exiting formal education.

**Non-sustained employment touches the majority of this population at some point, but rarely as a permanent condition.** Roughly 7 in 10 non-college graduates (70.1%) experience at least one year of non-sustained employment during their first five years out, but the largest single group by far is "intermittently" affected (39.8%) — one or two years of instability, not an ongoing state.

**The distribution is heavily front-loaded toward the lighter end and thins out steadily as persistence increases.** Intermittent (39.8%) is more than 2.5 times the size of Persistently non-sustained (15.7%), and Broadly non-sustained (14.6%) and Persistently non-sustained (15.7%) are nearly the same size — meaning once someone crosses into "broad" territory (3 of 5 years), they're about as likely to escalate to "persistent" as to stay at "broad." This is a meaningfully different shape than a table where instability compounds — it looks more like a population where instability is common early but tends not to snowball into a fixed, years-long state for most people.

**Overall Interpretation**

This confirms the shape suggested by the earlier year-by-year state distribution: non-sustained employment is a widespread but largely resolving phenomenon for non-college graduates. Nearly a third of the population never experiences it at all, and among those who do, the single most common outcome is brief (1-2 years), not chronic. Only about 1 in 6 non-college graduates (15.7%) get stuck in non-sustained employment for the bulk of their first five years. This stands in contrast to what we'll want to compare directly against the no-record distribution — if that measure turns out to be more bimodal (concentrated at "never" and "persistent," with less in the middle), it would confirm that non-sustained employment behaves more like a transitional phase while no-record status behaves more like a stable end-state.

```{r}
non_sustained.dist <- master.sep %>%
  group_by(non_sustained.cat) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  mutate(pct = n / sum(n),
         pct = scales::percent(pct))

kable(non_sustained.dist, 
      format = "html",
      escape = FALSE,
      align = c("lcc")) %>%
  kable_styling(bootstrap_options = "striped",
                full_width = FALSE) 
```

The table below provides the proportion of individuals that were in the No Minnesota employment record category during the first five years after exiting formal education.

**This distribution has a genuinely different shape than non-sustained employment's — it's bimodal rather than front-loaded.** Just over half the population (51.0%) never falls off Minnesota's tracked employment system at all, while a substantial 21.3% are persistently absent from it (4-5 of 5 years) — nearly as large as the "never" group's complement, and larger than the two middle categories *combined* (20.5% + 7.2% = 27.7%). Compare this to non-sustained employment, where the largest group was the *middle* category (intermittent, 39.8%); here, the middle categories are comparatively small and the population clusters at the two extremes instead.

**Once someone starts accumulating no-record years, the pattern favors escalation rather than resolution.** Broadly no-record (7.2%, exactly 3 of 5 years) is the smallest of the four categories — smaller than Persistently no-record (21.3%), which is nearly three times its size. That's the opposite relationship from non-sustained employment, where "broad" and "persistent" were nearly equal in size. Here, once someone crosses the 3-year threshold, they are far more likely to continue on to 4-5 years than to stop at exactly 3 — a genuine escalation pattern, not a plateau.

**Overall Interpretation**

This confirms the hypothesis raised when comparing the two measures side by side: no-record status behaves like a stable end-state rather than a transitional phase, in direct contrast to non-sustained employment's more resolving, front-loaded pattern. The population splits into two fairly distinct groups — just over half who essentially never disappear from Minnesota's tracked employment system, and about a fifth who are persistently absent from it for most of their first five years — with comparatively few people passing through it briefly or getting stuck in the middle. Combined with the earlier finding that "no MN employment record" is the fastest-growing category over time in the year-by-year data, this reinforces that disappearance from the tracked system, once established, is the more durable and structural of the two disconnection mechanisms — a genuinely different (and arguably more concerning) phenomenon than the largely-resolving instability captured by non-sustained employment.

```{r}
no_record.dist <- master.sep %>%
  group_by(no_record.cat) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  mutate(pct = n / sum(n))

kable(no_record.dist, 
      format = "html",
      escape = FALSE,
      align = c("lcc")) %>%
  kable_styling(bootstrap_options = "striped",
                full_width = FALSE) 
```

The table below showsonly the people who fall into "Persistently disconnected" under our *original* combined measure — meaning across their 5 years, they were disconnected (non-sustained OR no-record, in any mix) in 4 or 5 of those years. That's 8,064 people. The question this table asks is: **for each of those 8,064 people, what actually made up their disconnected years — was it mostly unstable jobs, mostly no employment record at all, or a mix of both?**

The categories are defined as;

- **"Non-sustained driven"** = this person's disconnected years were mostly/entirely *unstable employment* (non-sustained), not disappearing from the record. Example: someone with 4 years of non-sustained work and 0 years of no-record. They were disconnected in the combined sense, but they were always at least working, just not steadily.
- **"No-record driven"** = the opposite — this person's disconnected years were mostly/entirely *no employment record at all*, not instability. Example: someone with 0 years non-sustained and 4 years no-record. They weren't cycling through shaky jobs; they mostly just weren't showing up in Minnesota's system at all.
- **"Mixed (neither individually persistent)"** = this person doesn't have 4+ years in *either* single category, but their two counts still *add up* to 4 or 5 disconnected years total. Example: 2 years non-sustained + 2 years no-record = 4 disconnected years total (enough to be "persistently disconnected" overall), but only 2 in each — not enough on its own to count as "persistent" in either individual measure. This person alternated between the two problems rather than being stuck in one.

**The single largest driver of the "persistently disconnected" majority is no-record status, not instability.** 40.1% of everyone classified as persistently disconnected got there primarily by disappearing from Minnesota's tracked employment system for most of their first five years — not by cycling through unstable jobs. Non-sustained employment alone accounts for a smaller share (29.5%).

**Nearly a third of the persistently disconnected group (30.4%) doesn't fit either single-mechanism story at all.** These 2,453 people never accumulated enough years in *either* category alone to qualify as persistent in that specific measure, yet their combined disconnected years (non-sustained + no-record) still reached 4 or 5 out of 5. This is the group entirely invisible in the original combined `disconnected.cat` measure and in either individual persistence measure on its own — people who bounce between unstable employment and complete absence from the record, rather than settling into one pattern.

**Overall Interpretation**

Breaking down the "persistently disconnected" majority (53.2% of the full population) shows it is not one uniform experience, but effectively three: two out of five persistently disconnected people (40.1%) are there primarily because they've disappeared from Minnesota's employment system, not because their employment was unstable; three out of ten (29.5%) are there because of genuinely unstable — but still present and trackable — employment; and the remaining three out of ten (30.4%) are a distinct, chronically-alternating group that neither of the two individual persistence measures alone would have identified. This last group deserves particular attention going into Step 3: since they don't show up as an extreme case on either separated measure individually, any bivariate analysis run only against `non_sustained.cat` or `no_record.cat` separately would miss them entirely, and they may warrant their own dedicated comparison (e.g., "mixed" vs. "single-mechanism" persistently disconnected) rather than being folded into either separated analysis.

```{r}
persistent_decomp <- master.sep %>%
  filter(disconnected.cat == "Persistently disconnected") %>%
  mutate(driver = case_when(
    non_sustained.cat == "Persistently non-sustained" & no_record.cat != "Persistently no-record" ~ "Non-sustained driven",
    no_record.cat == "Persistently no-record" & non_sustained.cat != "Persistently non-sustained" ~ "No-record driven",
    non_sustained.cat == "Persistently non-sustained" & no_record.cat == "Persistently no-record" ~ "Both persistent",
    TRUE ~ "Mixed (neither individually persistent)"
  )) %>%
  group_by(driver) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  mutate(pct = n / sum(n))

kable(persistent_decomp, 
      format = "html",
      escape = FALSE,
      align = c("lcc")) %>%
  kable_styling(bootstrap_options = "striped",
                full_width = FALSE) 
```

## What This Table Tells Us: The Story Behind "Persistent Disconnection"

| Driver                                  | n     | \%    |
|-----------------------------------------|-------|-------|
| No-record driven                        | 3,234 | 40.1% |
| Non-sustained driven                    | 2,377 | 29.5% |
| Mixed (neither individually persistent) | 2,453 | 30.4% |

**The "persistently disconnected" majority isn't one story — it's three roughly equal-sized ones.** Split almost evenly into thirds: about 4 in 10 got there mainly by disappearing from Minnesota's tracked employment system, about 3 in 10 got there mainly through cycling in and out of unstable jobs, and the remaining 3 in 10 got there through some combination of both, without either problem alone reaching the "persistent" threshold.

**The largest single group (40.1%) is defined by absence, not instability.** This is the most important finding here: the biggest driver of long-term disconnection for non-college graduates isn't unstable employment — it's simply not appearing in the state's employment records at all for most of their first five years. This reinforces something already suspected from the earlier year-by-year trends (where "no MN emp record" was the fastest-growing category) but makes it concrete: it's not just growing over time in the aggregate, it's the single most common reason any individual ends up persistently disconnected.

**The "mixed" group (30.4%) is a genuine third category, not a rounding artifact — and it's nearly as large as either pure group.** These are people who don't fit a clean story of "always unstable" or "always absent." They move between the two, which suggests a more chaotic, harder-to-pin-down relationship with the labor market than either single mechanism describes on its own.

**Overall Interpretation**

If you had to describe what "persistent disconnection" means for the majority of non-college graduates in Northeast Minnesota, the honest answer is: it means three different things, in roughly equal measure. Two out of five are people who've essentially fallen off the map — not unemployed in a way the data can see, just absent from it. Three out of ten are people who keep working but never find stability. And three out of ten are bouncing between the two. That's a meaningfully more complicated (and arguably more useful) picture than the single "53% are persistently disconnected" headline number suggested, and it argues for treating these as three distinct populations — potentially requiring three different kinds of intervention or outreach — rather than one undifferentiated group of "disconnected" graduates.

# Relationship Between Individual Characteristics and Disconnection Categories

The next sections focus on exploring what relationships exist between individual characteristics and the different ways in which disconnection from the workforce is measured;

- **Persistently Disconnected:** Disconnected in 4 or 5 of the first 5 years
- **Disconnected (Broad):** Disconnected in 3 of 5 years
- **Intermittently Disconnected:** Disconnected in 1–2 of the first 5 years
- **Consistently Attached:** Disconnected in 0 of the first 5 years

```{r}
updated.master <- master %>%
  left_join(master.sep[,c(1,8:13)], by = "PersonID")

write_csv(updated.master, "Data/Claude/Non-college deeper/updated-master.csv")
```

<br>

## Demographic variables

All five tables use the same n = 15,169 and the same four disconnection categories, so effect sizes are directly comparable. Ranked by Cramér's V: **Special Ed status (0.120) \> English learner (0.105) \> Economic status (0.061) ≈ Race/Ethnicity (0.060) \> Gender (0.047)**. This ordering is worth noting up front — it's the same relative ranking these variables showed against the earlier point-in-time `grad.year.x` outcomes for this population, suggesting these demographic patterns are stable across how "disconnection" gets measured.

**Special Education status shows the single largest, cleanest gap of any variable here.** Persistently disconnected: 63.2% of Special Ed students vs. 49.9% of others — a 13.3-point gap. Consistently attached shows the mirror pattern: 7.6% vs. 13.4%. This is consistent with Special Ed's role throughout this entire project as one of the most robust demographic predictors of workforce outcomes.

**English learner status shows the largest single percentage-point gap of any demographic table in this analysis.** Persistently disconnected: 71.4% of English learners vs. 51.7% of others — a 19.7-point gap, larger even than Special Ed's. Consistently attached: only 5.8% of English learners vs. 12.4% of others. Given this project's established finding that English learner effects can be partly explained by other factors (particularly missing academic records) once more variables enter a multivariate model, this large bivariate gap should be treated as a strong candidate for attenuation at the multivariate stage, not assumed to be an independent English-learner-specific effect.

**Race/Ethnicity shows large, consistent absolute gaps despite a modest overall Cramér's V.** Persistently disconnected ranges from 51.3% (White, the lowest) up to 73.2% (Hispanic, the highest), with Black (70.6%) and Asian/PI (69.8%) close behind — all three minority groups roughly 19-22 points above White. The modest V here (0.060) reflects White's overwhelming sample share (12,995 of 15,169, or 85.7%) diluting the statistic, not a weak underlying relationship — the substantive gaps between individual groups are actually larger in absolute terms than Economic status's gap, despite showing a similar V.

**Economic status and Gender both show real but comparatively modest effects.** Economic disadvantage: 55.4% persistently disconnected vs. 50.9% for non-disadvantaged (4.5-point gap). Gender: 51.7% (Male) vs. 55.1% (Female) persistently disconnected — a smaller, 3.4-point gap, and notably in a different direction than several other outcomes examined earlier in this project, where male graduates often showed worse regional/employment outcomes. Here, women show modestly *higher* persistent disconnection than men.

**Overall Interpretation**

Every demographic variable examined shows a statistically significant relationship with overall disconnection persistence, and the pattern of which variables matter most closely mirrors what's been found throughout this entire project: Special Ed status and English learner status are the strongest and most consistent demographic predictors of poor workforce outcomes, Race/Ethnicity shows substantively large gaps for minority groups even when its summary statistic looks modest, and Economic status and Gender show smaller, though still real, effects. Given the sheer size of the English learner and racial/ethnic gaps here — several exceeding 15-20 percentage points — and given this project's repeated finding that some of these gaps get explained (though not eliminated) once academic and high-school-accomplishment variables are added, these should be flagged as the variables most likely to show meaningful attenuation once we reach the multivariate stage, particularly English learner status given the established masking relationship with missing MCA scores.

#### `Gender`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = master$Gender, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/Demographics/Gender.html")
```

<br>

#### `RaceEthnicity`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = updated.master$RaceEthnicity, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/Demographics/RaceEthnicity.html")
```

<br>

#### `economic.status`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = updated.master$economic.status, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/Demographics/economic-status.html")
```

<br>

#### `SpecialEdStatus`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = updated.master$SpecialEdStatus, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/Demographics/SpecialEdStatus.html")
```

<br>

#### `english.learner`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = updated.master$english.learner, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/Demographics/english-learner.html")
```

<br>

## High school characteristics

**Both local labor market variables show no significant relationship with disconnection persistence at all** — average local unemployment rate (F(3,15165) = 1.73, p = .159) and average local wages as % of state (F(3,15165) = 1.91, p = .126) both fail to reach significance, and none of the six pairwise Tukey contrasts for either variable are significant. This is a clean null result, not a weak-but-present one.

**This is a genuinely different finding than what these same two variables showed throughout the rest of this project.** Local unemployment and wages consistently mattered for the college-completer models (DV1a-DV4) and showed real, sometimes large effects in the earlier non-completer bivariate work against the point-in-time `grad.year.x` outcomes (recall: local unemployment rate showed a real, if modest, relationship with the location of sustained employment, and local wages showed one of the largest effects found anywhere in that analysis for regional completion). Here, against overall disconnection persistence specifically, neither shows any relationship at all.

**District rural/urban classification (Dem_Desc) does show a significant relationship, but it is the weakest significant effect found in this entire project.** χ² = 18.75, df = 6, p = .005, Cramér's V = 0.025 — significant only because of the very large sample size (n = 15,169), not because the relationship carries any real substantive weight. Persistently disconnected rates rise modestly from Entirely rural (48.3%) to Town/rural mix (53.4%) to Urban/town/rural mix (53.5%) — a 5.2-point spread, smaller than nearly every demographic gap examined in the previous set of tables, and the reverse direction of what one might expect (rural areas showing *less* persistent disconnection, not more).

**Overall Interpretation**

High school geographic and economic context — one of the more consistently important variable groups throughout this entire project — shows essentially no meaningful relationship with overall disconnection persistence for non-college graduates. This is a real and interesting divergence from the earlier point-in-time analysis (where local wages and unemployment did matter, particularly for *where* sustained employment happened) and from the college-completer models (where these same variables were often decisive). One plausible explanation: `disconnected.cat` combines two different disconnection mechanisms (non-sustained employment and no-record status) that we've already established behave quite differently and may respond to different predictors — it's possible local economic context matters for one mechanism specifically (e.g., wages predicting *where* someone works, which matters less once you're just asking whether someone is disconnected at all, regardless of location) but gets washed out when both are combined into a single persistence measure. This is exactly the kind of divergence the planned comparison across `disconnected.cat`, `non_sustained.cat`, and `no_record.cat` was designed to catch — worth watching closely once we reach the separated measures to see if local economic context reappears as a meaningful predictor there.

#### `Dem_Desc`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = updated.master$Dem_Desc, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/High school characteristics/Dem_Desc.html")
```

<br>

#### `avg.unemp.rate`

```{r}
summary <- aov(avg.unemp.rate ~ disconnected.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Disconnected cat/High school characteristics/avg-unemp-rate.xlsx", overwrite = TRUE)
```

<br>

#### `avg.wages.pct.state`

```{r}
summary <- aov(avg.wages.pct.state ~ disconnected.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Disconnected cat/High school characteristics/avg-wages-pct-state.xlsx", overwrite = TRUE)
```

<br>

## High school academics

**Effect size ranking (Cramér's V / eta-squared):** Took ACT (V=0.075) ≈ CTE achievement (V=0.069) \> AP exam (V=0.051) \> PSEO (V=0.034, caveat above) for the categorical variables; ACT composite score (eta²≈1.9%) is by far the strongest continuous predictor, followed by total CTE courses (eta²≈0.9%), with MCA Math (0.15%), MCA Reading (0.08%), and MCA Science (0.08%, only barely significant at p=.033) all showing negligible practical effect despite technical statistical significance.

**PSEO participation shows a small but measurable protective association.** Persistently disconnected: 50.5% of PSEO participants vs. 53.9% of non-participants — a modest 3.4-point gap in the expected direction, consistent with PSEO's broadly protective (if usually weak) role throughout this project. This is the weakest effect of any high-school-accomplishment variable examined here (V = 0.034), and, as noted above, its sample reflects listwise deletion of missing values rather than an explicit cohort restriction — though given that gap is small (\~224 records, 2.2% of the sample), the result itself is unlikely to change materially if re-run on a fully restricted sample.

**CTE engagement shows the same clean, monotonic, protective gradient found throughout this entire project.** Persistently disconnected rates: CTE Concentrator/Completer (47.5%) \< CTE Participant (51.8%) \< No CTE (58.0%) — a 10.5-point spread, the cleanest dose-response relationship in this table. Total CTE courses (continuous) confirms this: Persistently disconnected students have significantly fewer CTE courses than every other group, and the "Consistently attached" and "Intermittently disconnected" groups are statistically tied at the high end.

**Took ACT (participation) shows the same protective effect found in the earlier point-in-time analysis of this same population.** Persistently disconnected: 48.6% of ACT test-takers vs. 56.1% of non-test-takers — a 7.5-point gap favoring test-takers, consistent with ACT participation functioning as a broad marker of engagement/stability for non-completers, not just a college-prep signal that fails to pay off.

**ACT score (among test-takers) and AP exam participation both point the opposite direction — the same crossover pattern established earlier in this project.** Among test-takers, ACT score is significantly *higher* for the Persistently disconnected group than every other group (+1.26 vs. Consistently attached, +1.45 vs. Intermittently disconnected, both p \< .001) — the Consistently attached and Intermittently disconnected groups are statistically tied at the *low* end of ACT scores. AP exam shows the same direction categorically: Persistently disconnected is 62.6% among AP-takers vs. 52.7% among non-takers — AP-takers are *more* likely to be persistently disconnected, not less.

**MCA scores show weak, largely one-sided effects, and MCA Science barely clears significance at all.** Only the Persistently disconnected vs. Consistently attached contrast reaches significance for MCA Math (−0.123, p \< .001) and MCA Reading (−0.080, p = .012) — Persistently disconnected students score modestly lower, but no other pairwise contrast is significant for either subject, and the overall effect sizes are negligible (under 0.2% variance explained for both). MCA Science's overall F-test is barely significant (p = .033) and only one contrast (Intermittently disconnected vs. Consistently attached) reaches significance — this is the weakest and least reliable of the three MCA subjects here, consistent with its established missingness problems elsewhere in this project.

**Overall Interpretation**

This set of results replicates, almost variable-for-variable, the same "dual pathways" pattern found throughout this entire project: CTE engagement and simple ACT participation both function as broad protective factors against workforce disconnection, while ACT *score* and AP exam participation — markers of stronger academic achievement and college-prep orientation — actually predict *worse* disconnection outcomes for this specifically non-completing population. PSEO participation fits the protective side of this story as well, though as the weakest effect of the group (V = 0.034) — a modest reminder that not every college-preparatory activity carries the same "pulls away from stability" signal as ACT score or AP exam; PSEO, like CTE and ACT participation, appears to function more as a marker of general engagement than of academic ambition specifically. This is a coherent and now well-established story: for students who don't complete college, technical/vocational engagement and basic test-taking or coursework behavior anchor them to stable workforce participation, while stronger academic credentials without a completed degree appear to correlate with *more* persistent disconnection — plausibly reflecting a mismatch between academic readiness and the outcome actually achieved. MCA scores add comparatively little independent signal here, with Math and Reading showing only a modest gap concentrated at the extremes (Persistently disconnected vs. Consistently attached) and Science barely registering at all.

#### `pseo.participant`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = updated.master$pseo.participant, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/High school accomplishments/pseo-participant.html")
```

<br>

#### `ap.exam`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = updated.master$ap.exam, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/High school accomplishments/ap-exam.html")
```

<br>

#### `took.ACT`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = updated.master$took.ACT, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/High school accomplishments/took-ACT.html")
```

<br>

#### `ACTCompositeScore`

```{r}
summary <- aov(ACTCompositeScore ~ disconnected.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Disconnected cat/High school accomplishments/ACTCompositeScore.xlsx", overwrite = TRUE)
```

<br>

#### `MCA.M`

```{r}
summary <- aov(MCA.M ~ disconnected.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Disconnected cat/High school accomplishments/MCA-M.xlsx", overwrite = TRUE)
```

<br>

#### `MCA.R`

```{r}
summary <- aov(MCA.R ~ disconnected.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Disconnected cat/High school accomplishments/MCA-R.xlsx", overwrite = TRUE)
```

<br>

#### `MCA.S`

```{r}
summary <- aov(MCA.S ~ disconnected.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Disconnected cat/High school accomplishments/MCA-S.xlsx", overwrite = TRUE)
```

<br>

#### `cte.achievement`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = updated.master$cte.achievement, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/High school accomplishments/cte-achievement.html")
```

<br>

#### `total.cte.courses.taken`

```{r}
summary <- aov(total.cte.courses.taken ~ disconnected.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Disconnected cat/High school accomplishments/total-cte-courses-taken.xlsx", overwrite = TRUE)
```

<br>

## Post-secondary

Attending postsecondary at all — regardless of never completing it — shows by far the strongest relationship of any variable examined against this outcome (χ² = 577.90, df = 3, p \< .001, Cramér's V = 0.195), more than double the next-strongest variable (Special Ed status, V = 0.120) and nearly triple ACT participation and CTE achievement.

**The gap is large and moves the same population from majority-persistent to roughly a coin flip.** Persistently disconnected: 60.2% among those who never attempted postsecondary vs. 41.7% among those who did — an 18.5-point gap. Put differently: someone who never tried college is about 1.4x as likely to be persistently disconnected as someone who attended but didn't finish.

**The pattern holds cleanly across every category, not just at the extremes.** Consistently attached is roughly twice as common among those who attended (17.4%) as those who didn't (8.6%), and Intermittently disconnected is also meaningfully more common among attendees (28.4% vs. 19.7%) — while Broadly disconnected is nearly identical between the two groups (12.5% vs. 11.5%). The entire gap is concentrated in the "Consistently attached" and "Persistently disconnected" tails, with the middle categories barely differing — suggesting attempting postsecondary mostly determines whether someone avoids the *most severe* end of disconnection, more than it shifts people gradually along the whole spectrum.

**Overall Interpretation**

This confirms, in the persistence framing, exactly what the earlier point-in-time bivariate analysis found for this same population: simply attempting postsecondary education — even without completing it — is the single strongest predictor of workforce outcomes examined anywhere in this project, stronger than every demographic characteristic, high school context variable, and academic achievement measure combined. This reinforces that "non-completer" is not one uniform population; whether someone ever tried college at all appears to function as the dividing line separating a group that's largely protected from persistent disconnection from one that overwhelmingly is not. Given this is by a wide margin the strongest bivariate relationship found in this entire disconnection analysis, `attended.ps` should be treated as a central variable in any eventual multivariate confirmation of these findings, not just one item among many.

#### `attended.ps`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = updated.master$attended.ps, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/Post-secondary/attended-ps.html")
```

<br>

## Summary: Bivariate Predictors of Overall Disconnection Persistence (`disconnected.cat`)

| Variable | Effect size | Direction |
|------------------------|------------------------|------------------------|
| **Attended postsecondary (attempt only)** | **V = 0.195** (strongest overall) | Attempting protects against persistence |
| Special Ed status | V = 0.120 | SpecialEd → more persistent |
| English learner status | V = 0.105 | EL → more persistent |
| ACT composite score (test-takers) | eta² ≈ 1.9% | Higher score → more persistent |
| Took ACT (participation) | V = 0.075 | Participation → less persistent |
| CTE achievement (categorical) | V = 0.069 | More CTE → less persistent |
| Economic status | V = 0.061 | Disadvantage → more persistent |
| Race/Ethnicity | V = 0.060 | Minority groups → more persistent |
| Total CTE courses (continuous) | eta² ≈ 0.9% | More courses → less persistent |
| AP exam participation | V = 0.051 | AP → more persistent |
| Gender | V = 0.047 | Female → modestly more persistent |
| PSEO participation | V = 0.034 | PSEO → modestly less persistent |
| Dem_Desc (district type) | V = 0.025 | Urban → very slightly more persistent |
| MCA Math | eta² ≈ 0.15% | Negligible |
| MCA Reading | eta² ≈ 0.08% | Negligible |
| MCA Science | eta² ≈ 0.08% | Negligible, barely significant (p=.033) |
| Average local wages | Not significant (p = .126) | — |
| Average local unemployment rate | Not significant (p = .159) | — |

**The Story This Data Tells**

**One variable dominates everything else: simply attempting postsecondary education, whether or not it was ever completed, is by far the strongest predictor of long-term disconnection persistence found anywhere in this analysis.** At nearly double the next-strongest effect, `attended.ps` isn't just the top variable — it reframes the whole population. "Non-completer" is not one group; someone who tried college and stopped out looks statistically much more like a "successful" workforce participant than someone who never tried at all.

**Two demographic characteristics — Special Education status and English learner status — stand out as the strongest individual-level barriers, each showing 13-20 point gaps in persistent disconnection.** Both replicate patterns established throughout this entire project. Race/Ethnicity shows a more modest summary statistic but conceals large, real gaps for individual minority groups (Hispanic, Black, and Asian/PI graduates all show persistent-disconnection rates roughly 19-22 points above White graduates) — the modest V reflects White students' large sample share, not a weak underlying relationship.

**A clean "dual pathways" pattern runs through every high-school-accomplishment variable, replicating what this project has found again and again:** CTE engagement and simple ACT participation both protect against persistent disconnection, while ACT *score* and AP exam participation — markers of stronger academic achievement rather than just engagement — actually predict *worse* outcomes for this specifically non-completing population. This is now a thoroughly confirmed pattern across every outcome examined in this entire project: for non-completers, technical/vocational engagement and basic test-taking behavior anchor people to stability, while stronger academic credentials without a finished degree correlate with more persistent disconnection, plausibly reflecting a mismatch between preparation and outcome. PSEO participation fits this same protective side of the story, though weakly.

**High school geographic and economic context — one of the most consistently important variable groups everywhere else in this project — shows essentially no relationship with overall disconnection persistence.** Average local wages and unemployment rate are both flatly non-significant, and district rural/urban classification, while technically significant, is the weakest effect found anywhere in this analysis (V = 0.025) and driven purely by the large sample size rather than any real substantive gap. This is a genuine departure from earlier findings in this project and raises a real question: does local economic context matter for *some* forms of disconnection but not others, in a way the combined measure is hiding?

**What This Sets Up**

The near-total absence of a local-context effect here is the most important open question heading into the separated analyses. Since local wages and unemployment showed real relationships with *where* people worked in the earlier point-in-time analysis, but show nothing here against overall persistence, one clear hypothesis to test: local economic conditions may specifically predict *no-record* persistence (people leaving a weak local labor market entirely) but have little bearing on *non-sustained* persistence (people who stay and struggle regardless of local conditions) — or vice versa. This is exactly the kind of divergence the `non_sustained.cat` / `no_record.cat` comparison was built to catch, and it's the first thing worth checking once those two tables are in hand.

<br>

# Relationship Between Individual Characteristics and Non-sustaining Categories

The next sections focus on exploring what relationships exist between individual characteristics and the frequency of having non-sustained workforce participation in the first 5 years after formal education.

- **Never non-sustained** — 0 of the first 5 years (post-high-school or post-postsecondary-exit) were classified as "Non-sustained WF"
- **Intermittently non-sustained** — 1-2 of the first 5 years were "Non-sustained WF"
- **Broadly non-sustained** — exactly 3 of the first 5 years were "Non-sustained WF"
- **Persistently non-sustained** — 4-5 of the first 5 years were "Non-sustained WF"

<br>

## Demographic variables

**The ranking of demographic variables shifts dramatically compared to the combined `disconnected.cat` measure — this is exactly the divergence the separation was designed to catch.** Against `non_sustained.cat`: Economic status (V=0.083) \> Gender (V=0.079) \> Special Ed status (V=0.045) \> Race/Ethnicity (V=0.041) \> English learner (V=0.033). Compare this to the `disconnected.cat` ranking, where Special Ed (0.120) and English learner (0.105) were by far the two strongest predictors, and Economic status (0.061) and Gender (0.047) were comparatively modest. **The two strongest demographic predictors of overall disconnection are now the two weakest predictors of instability specifically** — a complete reordering, not just a rescaling.

**This strongly suggests Special Ed and English learner status drive overall disconnection primarily through no-record status, not through unstable employment.** Special Ed's gap here is real but modest (Persistently non-sustained: 18.2% vs. 14.9%, a 3.3-point gap) — a fraction of the 13.3-point gap seen in the combined measure. English learner's gap is smaller still (17.9% vs. 15.5%), and its "Never non-sustained" rate is actually *higher* for English learners than non-English-learners (33.5% vs. 29.6%) — meaning English learners are somewhat *less* likely to experience unstable-but-present employment, the opposite direction from what the combined measure's large gap implied. Both findings point the same way: whatever is driving these two groups' strong overall disconnection numbers, it isn't instability.

**Economic status and Gender, by contrast, show their strongest relationships specifically here — concentrated in the instability mechanism.** Economic disadvantage: Persistently non-sustained 16.9% vs. 14.4%, and more strikingly, Never non-sustained 26.3% vs. 33.5% — a real, substantial gap in who avoids instability altogether. Gender shows a similar pattern: Persistently non-sustained 17.9% (female) vs. 14.1% (male), and Never non-sustained 27.2% vs. 31.8% — women are both more likely to experience persistent instability and less likely to avoid it altogether. Notably, gender's effect size here (V=0.079) is nearly *double* its effect size against the combined measure (V=0.047) — instability is where gender's relationship to workforce outcomes actually lives, and it was being diluted in the combined measure.

**Race/Ethnicity reveals the clearest individual-group divergence of the whole set: Asian/PI graduates.** In the combined measure, Asian/PI showed the second-highest persistent-disconnection rate of any group (69.8%). Here, against non-sustained specifically, Asian/PI shows the *lowest* persistently-non-sustained rate of any racial group (12.1%, below even White at 15.2%) and by far the highest "Never non-sustained" rate (40.5%, roughly 10 points above every other group). Black graduates show the opposite pattern — the highest persistently-non-sustained rate (22.3%) of any group, a genuine instability-specific finding not fully visible in the combined measure.

**Overall Interpretation**

This is the clearest confirmation yet that `disconnected.cat` was masking meaningfully different underlying stories for different groups. Special Ed status and English learner status — the two strongest predictors of overall disconnection — turn out to have little to do with employment instability specifically; their disconnection is concentrated in no-record status (to be confirmed once we run these same variables against `no_record.cat`). Economic status and Gender, conversely, show their real signal here, in instability specifically, and were understated by the combined measure. Race/Ethnicity shows the most striking individual reversal: Asian/PI graduates are the *least* likely of any group to experience unstable employment, despite being among the *most* likely to be persistently disconnected overall — meaning their disconnection story is almost certainly a no-record story, not an instability story, while Black graduates show the opposite: a genuine instability-driven pattern that the combined measure partially obscured.

#### `Gender`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = master$Gender, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/Demographics/Gender.html")
```

<br>

#### `RaceEthnicity`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = updated.master$RaceEthnicity, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/Demographics/RaceEthnicity.html")
```

<br>

#### `economic.status`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = updated.master$economic.status, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/Demographics/economic-status.html")
```

<br>

#### `SpecialEdStatus`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = updated.master$SpecialEdStatus, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/Demographics/SpecialEdStatus.html")
```

<br>

#### `english.learner`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = updated.master$english.learner, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/Demographics/english-learner.html")
```

<br>

## High school characteristics

**All three high-school-characteristic variables show essentially no relationship with non-sustained employment persistence — an even more complete null than what these variables showed against the combined `disconnected.cat` measure.**

- **Dem_Desc:** not significant (χ² = 6.91, df = 6, p = .329, Cramér's V = 0.015) — a real step down from the already-weak V = 0.025 found against the combined measure, and the highest p-value of any variable examined in this analysis so far.
- **Average local unemployment rate:** not significant overall (F(3,15165) = 2.45, p = .062), though one pairwise contrast reaches significance (Intermittently non-sustained vs. Never non-sustained, +0.070, p = .038) — a borderline, inconsistent result that shouldn't be read as a real effect given the overall test fails to clear significance.
- **Average local wages:** not significant at all (F(3,15165) = 0.74, p = .529), and no individual contrast comes close to significance either. This is the cleanest null result of any variable examined in either the combined or separated analyses so far.

<u>**Overall Interpretation**</u>

Local geographic and economic context has no meaningful relationship with employment instability specifically. Combined with the earlier finding that these same variables were also non-significant (or barely significant, in Dem_Desc's case) against the combined `disconnected.cat` measure, this rules out one branch of the hypothesis raised earlier: local economic conditions do not appear to drive the instability side of disconnection at all. That leaves the no-record side as the more likely place local context might matter, if it matters anywhere in this population — recall that in the earlier point-in-time analysis, local wages and unemployment showed real relationships with *where* sustained employment occurred (regional vs. elsewhere in Minnesota), which is conceptually closer to a no-record/migration question than an instability question. This makes the upcoming `no_record.cat` results for these same three variables the more informative test of whether local context matters anywhere in this disconnection framework.

#### `Dem_Desc`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = updated.master$Dem_Desc, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/High school characteristics/Dem_Desc.html")
```

<br>

#### `avg.unemp.rate`

```{r}
summary <- aov(avg.unemp.rate ~ non_sustained.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Non-sustained cat/High school characteristics/avg-unemp-rate.xlsx", overwrite = TRUE)
```

<br>

#### `avg.wages.pct.state`

```{r}
summary <- aov(avg.wages.pct.state ~ non_sustained.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Non-sustained cat/High school characteristics/avg-wages-pct-state.xlsx", overwrite = TRUE)
```

<br>

## High school academics

**A note on scale carried over from before:** `pseo.participant`'s n = 10,352 reflects the same near-complete, mostly-benign PSEO tracking gap discussed earlier — not a new concern.

**Several variables show dramatically weaker effects here than against `disconnected.cat` — and in a few cases, the direction actually reverses. This is the richest set of divergences found yet in this comparison.**

**ACT score's crossover effect nearly disappears.** Against the combined measure, ACT score showed a strong, clean effect (eta² ≈ 1.9%, Persistently disconnected significantly higher than every other group). Here, the overall effect shrinks by nearly 90% (eta² ≈ 0.21%), and only one contrast survives (Persistently vs. Intermittently non-sustained, +0.621, p = .003) — the clean, complete ordering found against the combined measure is gone. **This strongly confirms the hypothesis raised earlier: ACT score's relationship with overall disconnection is mostly a no-record story, not an instability story** — something to verify directly once we reach `no_record.cat`.

**Took-ACT and PSEO participation both flip from protective to flat-or-reversed.** Against `disconnected.cat`, both showed real protective effects (took.ACT: 7.5-point gap favoring test-takers; PSEO: 3.4-point gap favoring participants). Here: took.ACT shows almost no gap in Persistently non-sustained (16.0% vs. 15.5%) and, more tellingly, test-takers show a *lower* "Never non-sustained" rate (27.4% vs. 31.5%) — the opposite of protective. PSEO shows a similar reversal: Persistently non-sustained is actually *higher* for participants (17.2% vs. 15.6%). **Both of these variables' protective effects against overall disconnection appear to run almost entirely through avoiding no-record status, not through avoiding instability** — a second and third confirmation of the same pattern found for ACT score.

**CTE shows a genuinely more complex, non-monotonic story here than the clean gradient found against the combined measure.** `cte.achievement`'s effect shrinks (V = 0.039 vs. 0.069) and its shape changes: at the "Persistently non-sustained" end, the expected protective gradient holds (CTE Concentrator/Completer 13.9% \< Participant 15.8% \< No CTE 16.8%), but at "Intermittently non-sustained," the direction *reverses* — CTE Concentrators/Completers show the *highest* intermittent rate (43.2%), not the lowest. `total.cte.courses.taken` confirms this isn't noise: Intermittently non-sustained shows significantly *more* CTE courses than every other category, including Never non-sustained. **This suggests CTE engagement protects specifically against severe/persistent instability, while plausibly contributing to brief, intermittent instability** — possibly reflecting the short-term employment transitions inherent to CTE work-based-learning placements (internships, apprenticeship-style courses ending and restarting).

**AP exam is the one variable that holds up nearly unchanged.** Persistently non-sustained: 20.7% (AP) vs. 15.4% (non-AP), V = 0.042 — comparable in both direction and magnitude to its effect against the combined measure (V = 0.051). AP exam appears to be a genuine instability-specific predictor, not one whose effect runs through no-record status.

**MCA scores show a genuine subject-specific divergence: Science is now significant, Math and Reading are not.** This is the reverse of the pattern found against `disconnected.cat`, where Math and Reading (weakly) reached significance and Science barely did. Here, MCA Science shows a clean, real effect (eta² ≈ 0.19%, p \< .001): "Never non-sustained" scores significantly higher than all three other categories, which aren't distinguished from each other — any instability at all associates with lower Science scores, not a dose-response pattern. Math (p = .069) and Reading (p = .230) are both non-significant here.

**Overall Interpretation**

This is the most revealing comparison yet between the combined and separated measures. Three variables that looked like broad protective factors against overall disconnection — ACT score's crossover, ACT participation, and PSEO participation — turn out to have little or nothing to do with employment instability specifically; their apparent effect on `disconnected.cat` is almost certainly concentrated in no-record status instead, a hypothesis the upcoming `no_record.cat` analysis can confirm directly. AP exam is the one accomplishment variable that clearly *does* predict instability on its own terms. CTE reveals a genuinely more nuanced mechanism than the clean story from the combined measure suggested: it protects against the most severe instability while being associated with more frequent brief/intermittent instability, a distinction the combined measure's monotonic gradient had completely flattened. MCA Science, not Math or Reading, is the academic-achievement measure that actually matters for instability specifically — the opposite of what mattered for overall disconnection.

#### `pseo.participant`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = updated.master$pseo.participant, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/High school accomplishments/pseo-participant.html")
```

<br>

#### `ap.exam`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = updated.master$ap.exam, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/High school accomplishments/ap-exam.html")
```

<br>

#### `took.ACT`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = updated.master$took.ACT, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/High school accomplishments/took-ACT.html")
```

<br>

#### `ACTCompositeScore`

```{r}
summary <- aov(ACTCompositeScore ~ non_sustained.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Non-sustained cat/High school accomplishments/ACTCompositeScore.xlsx", overwrite = TRUE)
```

<br>

#### `MCA.M`

```{r}
summary <- aov(MCA.M ~ non_sustained.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Non-sustained cat/High school accomplishments/MCA-M.xlsx", overwrite = TRUE)
```

<br>

#### `MCA.R`

```{r}
summary <- aov(MCA.R ~ non_sustained.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Non-sustained cat/High school accomplishments/MCA-R.xlsx", overwrite = TRUE)
```

<br>

#### `MCA.S`

```{r}
summary <- aov(MCA.S ~ non_sustained.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Non-sustained cat/High school accomplishments/MCA-S.xlsx", overwrite = TRUE)
```

<br>

#### `cte.achievement`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = updated.master$cte.achievement, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/High school accomplishments/cte-achievement.html")
```

<br>

#### `total.cte.courses.taken`

```{r}
summary <- aov(total.cte.courses.taken ~ non_sustained.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Non-sustained cat/High school accomplishments/total-cte-courses-taken.xlsx", overwrite = TRUE)
```

<br>

## Post-secondary

**This is the single largest divergence found in this entire comparison, and it confirms the emerging pattern across nearly every other variable in this table.** Against `disconnected.cat`, `attended.ps` was by far the dominant predictor (V = 0.195, nearly triple the next-strongest variable). Here, its effect drops to V = 0.077 — less than 40% of its original size, and no longer anywhere near the strongest predictor in the table (now comparable to Economic status and Gender, both around 0.08).

**The most telling detail: "Never non-sustained" shows almost no gap at all between attendees and non-attendees (30.2% vs. 29.7%).** Compare this to the combined measure, where "Consistently attached" showed attendees at roughly *double* the rate of non-attendees (17.4% vs. 8.6%). **This means attempting postsecondary does essentially nothing to prevent instability from occurring at all** — the massive protective effect seen against overall disconnection must be running almost entirely through avoiding no-record status, not through avoiding unstable employment. This is now the fourth variable (after ACT score, ACT participation, and PSEO) to show this same pattern, making it look less like a series of individual coincidences and more like a structural feature of this population: no-record status, not instability, is where most of these "protective" variables' real effect lives.

**Where attendance does matter here is in preventing instability from becoming chronic, not in preventing it altogether — the same reallocation pattern found for CTE.** Attendees show a notably *higher* Intermittently non-sustained rate (43.6% vs. 37.6%) but a notably *lower* Persistently non-sustained rate (12.8% vs. 17.4%) and a modestly lower Broadly non-sustained rate (13.4% vs. 15.3%). In other words: attendees are just as likely to experience *some* instability as non-attendees, but when they do, it tends to resolve rather than persist — echoing exactly the pattern found for CTE achievement and total CTE courses.

<u>**Overall Interpretation**</u>

`attended.ps`'s enormous effect on overall disconnection turns out to be concentrated almost entirely in the no-record mechanism, not instability — attendees are no more likely to avoid unstable employment altogether than non-attendees, but they are considerably more likely to have that instability resolve rather than become chronic, and (presumably, pending the `no_record.cat` results) considerably less likely to disappear from the tracked system entirely. This is a genuinely important refinement of what was, until now, the single strongest finding in this whole disconnection analysis: attempting college doesn't buy someone stability from day one, it appears to buy them resilience against instability turning permanent, and — more importantly — protection against falling out of the system altogether.

#### `attended.ps`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = updated.master$attended.ps, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/Post-secondary/attended-ps.html")
```

<br>

## Summary: Bivariate Predictors of Non-Sustained Employment Persistence (`non_sustained.cat`)

| Variable | Effect size | Direction |
|------------------------|------------------------|------------------------|
| Economic status | V = 0.083 | Disadvantage → more instability |
| Gender | V = 0.079 | Female → more instability |
| Attended postsecondary (attempt only) | V = 0.077 | Attendance → less *persistent* instability, but not less instability overall |
| ACT composite score (test-takers) | eta² ≈ 0.21% | Weak; higher score → modestly more persistent instability |
| Took ACT (participation) | V = 0.045 | Weak/reversed — no longer protective |
| SpecialEd status | V = 0.045 | Weak here (was strongest against combined measure) |
| PSEO participation | V = 0.049 | Weak/reversed — no longer protective |
| Race/Ethnicity | V = 0.041 | Weak summary stat, hides a real Asian/PI reversal |
| AP exam participation | V = 0.042 | AP → more instability (holds up unchanged) |
| CTE achievement (categorical) | V = 0.039 | Non-monotonic — protects against severe, not mild, instability |
| Total CTE courses (continuous) | eta² ≈ 0.24% | Inverse-U — more courses → more *intermittent*, less *persistent* instability |
| English learner status | V = 0.033 | Weak here (was 2nd strongest against combined measure) |
| MCA Science | eta² ≈ 0.19% | Any instability → lower score; no dose-response |
| MCA Math | Not significant (p=.069) | — |
| MCA Reading | Not significant (p=.230) | — |
| Dem_Desc | Not significant (p=.329) | — |
| Average local wages | Not significant (p=.529) | — |
| Average local unemployment rate | Not significant (p=.062) | — |

**The Story This Data Tells**

**This comparison against `disconnected.cat` produced the clearest confirmation yet that overall disconnection persistence was hiding two genuinely different mechanisms.** Four separate variables — ACT composite score, ACT participation, PSEO participation, and (most dramatically) attempting postsecondary at all — showed dramatically weaker, flat, or even reversed relationships with instability specifically, despite being among the strongest predictors of overall disconnection. This isn't a coincidence repeated four times; it's a structural pattern. **These variables' apparent protective power against disconnection overall is running almost entirely through avoiding no-record status, not through avoiding unstable employment** — a hypothesis the `no_record.cat` analysis will now be able to test directly and, most likely, confirm.

**The two demographic variables that mattered most for overall disconnection barely matter here, and vice versa.** Special Ed status and English learner status — the two strongest predictors of `disconnected.cat` — collapse to two of the weakest predictors of instability specifically. Economic status and Gender, comparatively modest against the combined measure, become the *strongest* predictors here, with Gender's effect size nearly doubling. This is a clean reallocation: whatever drives Special Ed and English learner students toward overall disconnection isn't instability; whatever drives economically disadvantaged and female graduates toward disconnection largely *is*.

**Race/Ethnicity's most striking finding is a reversal for Asian/PI graduates specifically.** Despite showing one of the highest persistent-disconnection rates overall, Asian/PI graduates show the *lowest* rate of persistent instability and the highest rate of avoiding instability altogether — meaning their overall disconnection story is almost certainly a no-record story, not an instability one. Black graduates show the opposite: a genuine, instability-specific disadvantage not fully visible in the combined measure.

**A second, more subtle pattern emerged specifically for CTE-related variables and postsecondary attendance: they don't prevent instability from happening, they prevent it from becoming chronic.** Both CTE achievement/course count and `attended.ps` show a genuine reallocation from "persistent" toward "intermittent" instability, rather than a clean reduction in instability overall. This is a real, substantively interesting nuance the combined measure's clean monotonic story didn't reveal.

**High school geographic and economic context shows no relationship with instability whatsoever** — all three variables (Dem_Desc, local wages, local unemployment) are non-significant, continuing the pattern already found against the combined measure and strengthening the case that local context, if it matters at all in this framework, matters for the no-record mechanism specifically.

**MCA subjects flip which one matters.** Against the combined measure, Math and Reading were (weakly) significant and Science was borderline; here, only Science is significant, and Math/Reading are not. This is a genuine, if modest, divergence worth noting but not over-interpreting given how small all three effects are.

**What This Sets Up**

Nearly every major finding in this table points toward the same next step: **`no_record.cat` should show the mirror image of what's missing here.** Specifically watch for: ACT score, ACT participation, PSEO, and `attended.ps` all showing much *larger* effects against no-record status than they did here; Special Ed and English learner status re-emerging as strong predictors; and local economic/geographic context finally showing a real relationship, if it's going to show one anywhere. If that pattern holds, it will confirm that `disconnected.cat`'s "persistent disconnection" majority was really two distinct populations all along — one defined by chronic instability (driven by economic status, gender, AP exam, and to some extent CTE/attendance patterns), and one defined by disappearing from the tracked system entirely (driven by demographic and academic-achievement/college-engagement variables).

<br>

# Relationship Between Individual Characteristics and having No Minnesota Employment Categories

The next sections focus on exploring what relationships exist between individual characteristics and the frequency of having no Minnesota employment record in the first 5 years after formal education.

- **Never no-record** — 0 of the first 5 years (post-high-school or post-postsecondary-exit) had no Minnesota employment record
- **Intermittently no-record** — 1-2 of the first 5 years had no Minnesota employment record
- **Broadly no-record** — exactly 3 of the first 5 years had no Minnesota employment record
- **Persistently no-record** — 4-5 of the first 5 years had no Minnesota employment record

<br>

## Demographic variables

**This table confirms almost every hypothesis raised by the `non_sustained.cat` comparison, variable for variable.**

**Special Ed status and English learner status re-emerge as strong predictors, nearly matching their strength against the combined measure.** Special Ed: V = 0.087 here vs. 0.045 against `non_sustained.cat` — essentially a full reversal of where its effect lives. Persistently no-record: 25.6% (Special Ed) vs. 19.9% (others), and Never no-record: 43.7% vs. 53.4% — a genuine, substantial gap. English learner status shows the same pattern even more sharply: V = 0.105 here (identical to its effect against the combined measure) vs. 0.033 against instability. Persistently no-record: 35.2% (English learners) vs. 20.2% (others) — nearly double, and Never no-record: just 36.4% vs. 52.2%. **This confirms directly what the instability analysis only implied: Special Ed and English learner students' overall disconnection is almost entirely a no-record phenomenon, not an instability one.**

**Gender flips to essentially the weakest demographic predictor here, exactly as predicted.** V = 0.025 (barely significant, p = .021), compared to V = 0.079 against instability. The gap in Persistently no-record is small (22.2% male vs. 20.1% female) and modest in every category. Gender's relationship with workforce outcomes lives almost entirely in the instability mechanism, not disconnection from the record.

**Economic status shows a similar, if less extreme, reversal.** V = 0.050 here vs. 0.083 against instability — roughly the mirror image of Gender's pattern. Economic disadvantage still shows a real gap (Persistently no-record: 20.2% vs. 22.4%, though in this table disadvantage actually shows a *slightly lower* no-record rate, worth noting as a genuine, if small, reversal in direction, not just magnitude) — economic status's disconnection story is predominantly about instability, only weakly about disappearing from the record.

**Race/Ethnicity reveals the sharpest and most important individual-group finding in this entire analysis: Hispanic and Asian/PI graduates show the two highest persistent no-record rates of any group, by a wide margin.** Hispanic: 41.2% persistently no-record — the highest of any group. Asian/PI: 40.5% — nearly identical, and both roughly double White's rate (20.6%). This is a dramatic confirmation of the hypothesis raised in the instability analysis: Asian/PI graduates showed the *lowest* persistent instability rate of any group, yet were among the *highest* overall-disconnection groups — this table shows exactly where that disconnection lives. Black graduates show a more moderate elevation (27.2%), while American Indian graduates are nearly identical to White (20.3% vs. 20.6%) — a genuine divergence from the instability table, where AI graduates showed the highest persistent instability rate of any group. **Put together: AI graduates' disadvantage is an instability story; Hispanic and Asian/PI graduates' disadvantage is overwhelmingly a no-record story; Black graduates show a real, if more moderate, elevation in both.**

**Overall Interpretation**

This table delivers exactly the confirmation the separated analysis was built to find. Every demographic variable that looked weak against instability (Special Ed, English learner) is strong here, and every variable that looked strong against instability (Gender, Economic status) is comparatively weak here — a clean, near-total reallocation of effect across the two mechanisms. The Race/Ethnicity findings are the most substantively important: this data shows that different racial/ethnic groups experience genuinely different *forms* of workforce disconnection, not just different overall rates — American Indian graduates' disadvantage is concentrated in unstable-but-present employment, while Hispanic and Asian/PI graduates' disadvantage is concentrated in disappearing from Minnesota's tracked employment system entirely. This is a meaningfully more precise and more actionable finding than the combined measure could offer, and it argues for tailoring outreach or intervention strategies differently across these groups rather than treating "workforce disconnection" as a single, uniform experience.

#### `Gender`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = master$Gender, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/Demographics/Gender.html")
```

<br>

#### `RaceEthnicity`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = updated.master$RaceEthnicity, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/Demographics/RaceEthnicity.html")
```

<br>

#### `economic.status`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = updated.master$economic.status, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/Demographics/economic-status.html")
```

<br>

#### `SpecialEdStatus`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = updated.master$SpecialEdStatus, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/Demographics/SpecialEdStatus.html")
```

<br>

#### `english.learner`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = updated.master$english.learner, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/Demographics/english-learner.html")
```

<br>

## High school characteristics

**Contrary to the hypothesis raised after the `non_sustained.cat` results, local economic context still does not show a strong or consistent relationship here — this needs to be stated plainly rather than reframed to fit expectations.**

**Average local wages remains completely non-significant** (F(3,15165) = 0.67, p = .573, eta² ≈ 0.01%) — the third and final null result for this variable across all three dependent variables in this analysis. Local wage levels appear to have no bearing on any form of disconnection examined in this entire project's non-completer analysis.

**Average local unemployment rate does reach significance here, but the pattern is small and doesn't fit a simple story.** F(3,15165) = 6.21, p \< .001, though eta² is still tiny (≈0.12%). The Tukey contrasts show something genuinely counterintuitive: **Persistently no-record shows a significantly *lower* average local unemployment rate than both Intermittently no-record (−0.137, p \< .001) and Broadly no-record (−0.123, p = .044)** — the opposite of what you'd expect if a weak local labor market were pushing people out of the tracked system entirely. Intermittently no-record shows the *highest* rate, modestly above Never no-record (+0.075, p = .040). This is not a clean dose-response relationship, and given how small the effect size is, I'd be cautious about building a narrative around it rather than treating it as a genuine, well-supported pattern.

**Dem_Desc reaches significance but remains one of the weakest effects found anywhere in this analysis** (V = 0.021, p = .038, barely below the .05 threshold). The direction is consistent with the combined measure: Persistently no-record rises modestly with urbanicity (18.8% Entirely rural → 21.4% Town/rural mix → 21.5% Urban/town/rural mix), a small gradient in the same direction found against `disconnected.cat`.

**Overall Interpretation**

The hypothesis that local economic and geographic context would finally show a real relationship here — given that wages and unemployment mattered substantially for *where* sustained employment happened in the earlier point-in-time analysis — does not hold up. Local wages remain flatly irrelevant across all three disconnection measures, and while local unemployment rate does reach statistical significance here, the effect is tiny and the direction (persistent no-record status associated with *lower*, not higher, local unemployment) runs counter to a simple "weak local economy pushes people out of the system" story. This is worth reporting honestly as a negative or inconclusive finding rather than forcing an explanation: high school geographic and economic context, across all three framings of disconnection examined in this analysis, is simply not where the meaningful predictive signal lives for this population. Whatever drives long-term workforce disconnection for non-college graduates in Northeast Minnesota, it is not primarily rooted in the local labor market conditions of the high school they attended.

#### `Dem_Desc`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = updated.master$Dem_Desc, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/High school characteristics/Dem_Desc.html")
```

<br>

#### `avg.unemp.rate`

```{r}
summary <- aov(avg.unemp.rate ~ no_record.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/No record cat/High school characteristics/avg-unemp-rate.xlsx", overwrite = TRUE)
```

<br>

#### `avg.wages.pct.state`

```{r}
summary <- aov(avg.wages.pct.state ~ no_record.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/No record cat/High school characteristics/avg-wages-pct-state.xlsx", overwrite = TRUE)
```

<br>

## High school academics

**This table delivers the clearest, most complete confirmation yet of the hypotheses built up across this entire comparison — nearly every variable's "true home" is now visible.**

**Took ACT (participation) shows its strongest effect of any table in this entire analysis, confirming it is almost entirely a no-record phenomenon.** V = 0.098 here — actually *larger* than its effect against the combined measure (V = 0.075) and dramatically larger than against instability (V = 0.045, where the direction had even reversed). Persistently no-record: 16.8% (test-takers) vs. 24.2% (non-test-takers) — a 7.4-point gap. Never no-record: 56.3% vs. 47.6%. **The earlier finding that ACT participation showed a flat-or-reversed relationship with instability, combined with this result, confirms conclusively: ACT participation protects against disappearing from the tracked system, and has essentially nothing to do with avoiding unstable employment.**

**PSEO participation shows the same pattern, and even more starkly than took.ACT relative to its own combined-measure effect.** V = 0.070 here vs. 0.034 against the combined measure and a reversed, adverse direction against instability. Persistently no-record: 16.9% (PSEO) vs. 22.8% (non-PSEO) — a 5.9-point gap, nearly double what appeared in the combined measure. **PSEO's entire protective effect against overall disconnection lives here.**

**ACT composite score confirms the same story with a distinctive "threshold" shape.** eta² ≈ 1.50% here (vs. 1.87% against the combined measure, vs. a mere 0.21% against instability) — nearly all of ACT score's crossover effect is concentrated in no-record status. The shape is notable: Never no-record shows a significantly *lower* score than all three other categories (all three contrasts vs. Never are significant, roughly +0.93 to +1.33), but Intermittently, Broadly, and Persistently no-record are not distinguished from each other. This is a threshold effect, not a dose-response — any amount of no-record experience associates with a meaningfully higher ACT score among test-takers, but more years of it doesn't compound the gap further.

**CTE (both achievement and course count) shows a real, substantial effect here, comparable in strength to the combined measure and clearly stronger than its effect on instability.** `cte.achievement`: V = 0.058 (vs. 0.069 combined, 0.039 instability), with Persistently no-record showing a clean gradient — CTE Concentrator/Completer (18.1%) \< Participant (19.0%) \< No CTE (25.0%), a 6.9-point spread, the widest specific gap in this table. `total.cte.courses.taken` confirms this: eta² ≈ 0.66%, with a real, largely monotonic protective pattern (Persistently no-record shows significantly fewer courses than every other category). This differs from the instability findings, where CTE showed a non-monotonic, reallocation pattern rather than a clean reduction — **CTE's protective effect is concentrated more in preventing disappearance from the record than in preventing chronic instability.**

**AP exam shows its weakest effect of any table so far, and a genuinely different shape.** V = 0.030 (vs. 0.051 combined, 0.042 instability) — AP exam's effect is actually more balanced between the two mechanisms than most other variables, tilting slightly toward instability rather than no-record. Notably, Persistently no-record is nearly identical between AP-takers and non-takers (21.3% vs. 21.3%) — AP's effect here is concentrated entirely in the "Never" and "Intermittently" categories, not at the severe end.

**MCA subjects show small, largely consistent effects across all three, similar in size to what was found against the combined measure** — all three reach significance (eta² ≈ 0.09-0.11%), with Math and Reading showing the expected direction (lower scores among no-record groups) and Science showing, curiously, the opposite direction for the Persistently no-record contrast specifically (+0.081, p = .005) — a small, likely not very meaningful reversal given Science's well-established missingness problems in this project, worth noting but not over-interpreting.

**Overall Interpretation**

This table completes the picture the entire `non_sustained.cat`/`no_record.cat` comparison was designed to reveal. ACT participation, PSEO participation, and (from the earlier table) attempting postsecondary at all are now confirmed as variables whose apparent "broad protective" effect against overall disconnection is almost entirely a no-record story — none of them meaningfully reduce instability, but all three substantially reduce the risk of disappearing from Minnesota's tracked employment system. CTE shows a more genuinely dual-mechanism effect: real in both, but somewhat stronger here. AP exam is the interesting exception — its effect is the most evenly split of any accomplishment variable, and notably weaker here than elsewhere, suggesting it isn't primarily a no-record-driving factor either, but a somewhat diffuse predictor of both mild instability and mild no-record risk without much force behind either.

#### `pseo.participant`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = updated.master$pseo.participant, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/High school accomplishments/pseo-participant.html")
```

<br>

#### `ap.exam`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = updated.master$ap.exam, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/High school accomplishments/ap-exam.html")
```

<br>

#### `took.ACT`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = updated.master$took.ACT, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/High school accomplishments/took-ACT.html")
```

<br>

#### `ACTCompositeScore`

```{r}
summary <- aov(ACTCompositeScore ~ no_record.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/No record cat/High school accomplishments/ACTCompositeScore.xlsx", overwrite = TRUE)
```

<br>

#### `MCA.M`

```{r}
summary <- aov(MCA.M ~ no_record.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/No record cat/High school accomplishments/MCA-M.xlsx", overwrite = TRUE)
```

<br>

#### `MCA.R`

```{r}
summary <- aov(MCA.R ~ no_record.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/No record cat/High school accomplishments/MCA-R.xlsx", overwrite = TRUE)
```

<br>

#### `MCA.S`

```{r}
summary <- aov(MCA.S ~ no_record.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/No record cat/High school accomplishments/MCA-S.xlsx", overwrite = TRUE)
```

<br>

#### `cte.achievement`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = updated.master$cte.achievement, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/High school accomplishments/cte-achievement.html")
```

<br>

#### `total.cte.courses.taken`

```{r}
summary <- aov(total.cte.courses.taken ~ no_record.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/No record cat/High school accomplishments/total-cte-courses-taken.xlsx", overwrite = TRUE)
```

<br>

## Post-secondary

**As predicted, this is the largest effect found against `no_record.cat` — though the story is "primarily no-record," not "exclusively no-record" as with ACT and PSEO.** V = 0.148 here, compared to V = 0.077 against instability and V = 0.195 against the combined measure. This confirms `attended.ps`'s effect is concentrated in the no-record mechanism (roughly double the strength of its instability effect), but unlike ACT participation or PSEO — where the instability effect nearly vanished or reversed entirely — `attended.ps` retained a real, if smaller, protective effect on instability too (recall: Persistently non-sustained was 12.8% for attendees vs. 17.4% for non-attendees). This makes `attended.ps` the one variable in this entire analysis that shows a genuinely dual-mechanism effect rather than one that collapses almost entirely into a single pathway.

**The gap here is the largest specific percentage-point gap found for this variable against either separated measure.** Persistently no-record: 15.1% (attended) vs. 25.1% (never attended) — a 10-point gap. Never no-record: 59.8% vs. 45.6% — attendees are roughly 1.3x as likely to never disappear from the tracked system at all.

**Overall Interpretation**

This closes out the picture for `attended.ps`: attempting postsecondary education, even without completing it, delivers its largest protective benefit against disappearing from Minnesota's tracked employment system entirely, but it also provides a real, smaller protective benefit against chronic instability — the only high-school/postsecondary-accomplishment variable examined in this entire analysis to show meaningful, non-trivial effects on both mechanisms simultaneously rather than concentrating almost entirely in one. This reinforces `attended.ps`'s standing as the single most important variable in this whole disconnection analysis: not just because its effect size is the largest, but because it's the one variable whose benefit doesn't have a major blind spot on either side of the non-sustained/no-record divide.

#### `attended.ps`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = updated.master$attended.ps, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/Post-secondary/attended-ps.html")
```

<br>

## Summary: Bivariate Predictors of No-Record Persistence (`no_record.cat`)

| Variable | Effect size | Direction |
|------------------------|------------------------|------------------------|
| Attended postsecondary (attempt only) | V = 0.148 | Strongest single predictor; attendance protects |
| English learner status | V = 0.105 | EL → much more likely persistently no-record |
| Took ACT (participation) | V = 0.098 | Participation → strongly protective |
| Special Ed status | V = 0.087 | SpecialEd → more persistently no-record |
| ACT composite score (test-takers) | eta² ≈ 1.50% | Threshold effect — any no-record experience → higher score |
| PSEO participation | V = 0.070 | PSEO → strongly protective |
| Race/Ethnicity | V = 0.060 | Hispanic & Asian/PI → dramatically elevated |
| CTE achievement (categorical) | V = 0.058 | More CTE → less no-record |
| Economic status | V = 0.050 | Disadvantage → modestly more no-record |
| Total CTE courses (continuous) | eta² ≈ 0.66% | More courses → less no-record |
| AP exam participation | V = 0.030 | Weak; balanced across both mechanisms |
| Gender | V = 0.025 | Weak, barely significant |
| Dem_Desc | V = 0.021 | Weakest significant effect; barely significant |
| MCA Math / Science | eta² ≈ 0.11% each | Negligible |
| MCA Reading | eta² ≈ 0.09% | Negligible |
| Average local unemployment rate | eta² ≈ 0.12% | Significant but tiny, non-monotonic, counterintuitive direction |
| Average local wages | Not significant (p = .573) | — |

**The Story This Data Tells**

**This table resolves nearly every open question raised by the `non_sustained.cat` comparison — it is the mirror image, almost variable for variable, of what was missing there.** Four variables that showed dramatically weakened, flat, or reversed effects against instability — ACT composite score, ACT participation, PSEO participation, and `attended.ps` — all re-emerge here as among the strongest predictors in the entire analysis. This confirms, rather than just suggests, that overall disconnection persistence was masking two genuinely separate populations: one whose disadvantage is rooted in unstable-but-present employment, and one whose disadvantage is rooted in disappearing from Minnesota's tracked employment system altogether. Most of the variables long assumed to be broadly "protective" for non-completers turn out to protect against only one of these two things, not both.

**Attempting postsecondary education remains the single most important variable in this entire analysis, and it's now clear why it dominated the combined measure.** Its effect here (V = 0.148) is nearly double its effect against instability (V = 0.077) — but unlike ACT and PSEO, it didn't collapse to near-zero on the instability side. `attended.ps` is the one variable in this whole analysis that delivers a genuine, non-trivial benefit on both mechanisms simultaneously, which is exactly why it towered over every other predictor in the combined `disconnected.cat` measure.

**Special Education and English learner status — the two strongest predictors of overall disconnection — are confirmed here as being almost entirely no-record phenomena.** Both variables, which nearly vanished against instability, come roaring back here at effect sizes matching or exceeding their combined-measure strength. Whatever barriers these two groups face in the workforce, they manifest overwhelmingly as absence from the tracked system, not as cycling through unstable jobs.

**Race/Ethnicity produces the most substantively important and actionable finding across this entire three-part analysis.** Hispanic and Asian/PI graduates show persistent no-record rates around double White graduates' rate — the widest individual-group gap found anywhere in this project — while American Indian graduates, who showed the *highest* persistent instability rate of any group, look no different from White graduates here. This means three different racial/ethnic groups are experiencing three genuinely different forms of disadvantage: AI graduates face an instability problem, Hispanic and Asian/PI graduates face a disconnection-from-the-system problem, and Black graduates show a real, more moderate elevation on both.

**High school geographic and economic context is confirmed, across all three dependent variables, to be essentially irrelevant to this population's workforce disconnection.** Average local wages is non-significant everywhere; average local unemployment rate reaches statistical significance only here, but the effect is tiny and runs in a counterintuitive direction (persistent no-record status associated with *lower*, not higher, local unemployment) that shouldn't be over-interpreted. Whatever drives long-term disconnection for this population, it is not rooted in the economic conditions of the high school district they came from.

**What This Sets Up**

Taken together, the three-part comparison (`disconnected.cat`, `non_sustained.cat`, `no_record.cat`) has produced a genuinely richer and more actionable picture than the original combined measure could have: (1) economic status and gender predict instability specifically; (2) Special Ed and English learner status, along with ACT/PSEO/postsecondary-attempt variables, predict no-record status specifically; (3) `attended.ps` and CTE are the two variables that meaningfully address both mechanisms; and (4) Race/Ethnicity reveals that different groups face genuinely different forms of disadvantage that a single combined outcome measure would have blurred together. This three-way comparison — not any single table — is the real deliverable of this bivariate stage, and it should directly shape how the multivariate confirmation models are structured: likely as two separate model sequences (mirroring the DV1a/DV1b-style split used earlier in this project), one for instability persistence and one for no-record persistence, rather than a single combined disconnection model.

<br>

# Full Synthesis: Workforce Disconnection Among Non-College Graduates in Northeast Minnesota

## Executive Summary

This analysis examined the 58% of Northeast Minnesota high school graduates who never completed a postsecondary credential, tracing their workforce participation across the first five years after exiting formal education. Three major findings anchor this work:

1.  **Disconnection is not a transitional phase — it is the majority experience.** 53.2% of non-college graduates are persistently disconnected from sustained employment (4-5 of their first 5 years), and only 11.9% are consistently attached.

2.  **"Disconnection" is not one phenomenon — it is at least two, and they behave completely differently.** Separating unstable-but-present employment (non-sustained work) from complete absence from Minnesota's tracked system (no-record status) revealed that these two mechanisms have almost entirely different predictors, different trajectories over time, and different underlying stories.

3.  **Underutilization, not disengagement, is the dominant experience for those who aren't stably employed** — 73-81% of non-sustained employment involves at least partial regional attachment, meaning most instability reflects people working in the region, just not enough to be "sustained," not people leaving for opportunity elsewhere.

<br>

## The Story This Data Tells

### Part 1: The scale and shape of the problem

Picture the roughly 15,000 non-college graduates who make up this population. In their first year out, they're widely dispersed — a little over a third in unstable work, roughly a fifth each in no-record status, still-attending-postsecondary, and sustained regional employment. But this snapshot obscures the real story: **tracked over five years, this population sorts into a stark and durable pattern.** Just over half end up persistently disconnected; barely one in eight ever achieve full, sustained attachment. Two very different trajectories are unfolding simultaneously — sustained employment does grow steadily every year (30.2% → 40.1% of the population from Year 1 to Year 5) — but "no Minnesota employment record" grows even faster, overtaking every other category by Year 5. Non-sustained employment, meanwhile, consistently shrinks — but critically, when we looked closely at *where* that non-sustained employment happens, we found the large majority (73-81%, holding steady across all five years) involves at least partial regional attachment. This is underutilization, not outmigration: most people who aren't stably employed are still here, working, just not enough.

### Part 2: Two distinct disconnection mechanisms hiding inside one number

The single most important methodological move in this analysis was splitting "disconnected" into its two components — **non-sustained employment** and **no employment record at all** — and testing each independently. This wasn't a formality; it fundamentally changed the picture:

- **The two measures have genuinely different shapes.** Non-sustained persistence is front-loaded toward the intermittent middle (39.8% of the population) and shrinks steadily — a phenomenon that resolves over time. No-record persistence is bimodal (51.0% never, 21.3% persistent) and, once someone crosses into the "broad" category, they're far more likely to escalate into "persistent" than to plateau — a phenomenon that compounds.
- **Decomposing the "persistently disconnected" majority revealed three roughly equal populations, not one.** 40.1% of the persistently disconnected got there mainly through no-record status, 29.5% mainly through instability, and 30.4% through a genuine, chronic alternation between both that neither individual measure alone would have identified.
- **Nearly every bivariate predictor split cleanly onto one side of this divide or the other**, and in several cases the split was total.

### Part 3: Which characteristics drive which mechanism

**Demographics split almost perfectly across the two mechanisms.** Special Education status and English learner status — the two strongest predictors of overall disconnection — turned out to be driven almost entirely by no-record status; both showed weak, sometimes reversed relationships with instability. Economic status and Gender showed the opposite pattern: comparatively modest against the combined measure, but the strongest demographic predictors of instability specifically, with Gender's effect size nearly doubling once isolated.

**Race/Ethnicity revealed the most consequential and actionable divergence in the whole analysis.** Three different racial/ethnic groups face three different forms of disadvantage: American Indian graduates show the highest persistent-instability rate of any group but look no different from White graduates on no-record status; Hispanic and Asian/PI graduates show the opposite — roughly double the persistent no-record rate of White graduates, but Asian/PI graduates actually show the *lowest* persistent-instability rate of any group; Black graduates show a real, more moderate elevation on both. Treating "disconnection" as one undifferentiated outcome would have completely obscured these three distinct patterns.

**High school accomplishments split into three groups of variables, not two.** ACT participation, PSEO participation, and ACT composite score all turned out to be almost purely no-record predictors — their apparent broad protective effect against overall disconnection nearly vanished or reversed once instability was isolated. AP exam participation was the reverse case — a real instability predictor with almost no no-record effect. CTE engagement and, most importantly, **attempting postsecondary education at all** were the two variables that showed genuine, meaningful effects on *both* mechanisms — `attended.ps` in particular remained the single strongest predictor in this entire analysis precisely because it's the only variable that doesn't have a blind spot on either side of the divide.

**High school geographic and economic context was, across all three framings of disconnection, essentially irrelevant.** Local wages showed no relationship with any outcome examined. Local unemployment rate reached significance only against no-record status, but the effect was tiny and ran in a counterintuitive direction. Whatever drives long-term disconnection for this population, it is not rooted in the local economic conditions of the high school they attended — a genuine and somewhat surprising departure from what these same variables showed throughout the rest of this project's college-completer analyses.

### Part 4: A unifying picture

Put all of this together, and a clear structural story emerges. **This population's long-term workforce outcomes are shaped by two largely independent forces**: whether someone ever engaged with the postsecondary system at all (attendance, ACT, PSEO) — which mainly determines whether they stay visible in Minnesota's tracked employment system — and a separate set of demographic and programmatic factors (economic status, gender, CTE, AP) that mainly determine whether the employment they do have is stable. Geography and local economic conditions, so influential elsewhere in this project, play almost no role here. And underneath both mechanisms sits a genuine equity story: different demographic groups aren't just disconnected at different *rates*, they're disconnected in different *ways* — a distinction that matters enormously for what kind of intervention would actually help.

<br>

## What This Sets Up for the Multivariate Stage

This bivariate work points toward a clear multivariate design, directly analogous to the DV1a/DV1b structure used earlier in this project's college-completer analysis:

1.  **Two separate model sequences**, not one combined model — one for `non_sustained.cat` persistence, one for `no_record.cat` persistence — since a single combined model would recreate the same masking problem this entire analysis was built to solve.
2.  **`attended.ps` belongs in both models** as a foundational predictor, given it's the only variable shown to matter for both mechanisms.
3.  **Expect substantial attenuation for Special Ed and English learner status once academic/accomplishment variables enter the no-record model** — mirroring the exact pattern found throughout the rest of this project (e.g., the MCA-missingness masking effect discovered in the college-completer work).
4.  **Local geographic/economic variables can likely be deprioritized or dropped** from both models given their consistent null results across all three outcome framings — though worth retaining at least once to formally confirm rather than assume.
5.  **Race/Ethnicity should be watched closely for interaction effects** — given that AI, Hispanic/Asian-PI, and Black graduates show genuinely different mechanisms of disadvantage, a single additive coefficient per race category may not fully capture what's happening, and this is worth flagging as a candidate for deeper stratified analysis (per the original methodology's Step 3.3) even if it's not built into the initial multivariate pass.
