---
title: "Multivariate Confirmation Analysis on Persistent Disconnection"
editor: visual
---

```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
library(htmlwidgets)
library(RColorBrewer)
library(readxl)
library(janitor)
library(lubridate)
library(systemfonts)
reset_font_cache()
library(ggtext)
library(gmodels)
library(fastDummies)
library(car)
library(glmnet)
library(glmnetUtils)
library(pscl)
library(sjPlot)
library(ggforce)
library(openxlsx)
```

```{r themes and shapefiles, include=FALSE}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", linewidth  = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.key.size = unit(1, "lines"),
        legend.margin = margin(0,0,0,0),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.text = element_text(margin = margin(l = 2)),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", linewidth = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))


theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
  select(5,6) %>%
  unique() %>%
  mutate(edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

counties.regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
  rename(mif = `MIF Region`) %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Name = str_replace(Name, "Mcleod", "McLeod"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc) ,
         edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"))

counties.regions.1 <- counties.regions %>%
  mutate(statefp = "27",
         project.pr = ifelse(edr %in% c("EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 4 - West Central"), "Northwest",
                             ifelse(edr == "EDR 3 - Arrowhead", "Northeast",
                                    ifelse(edr %in% c("EDR 5 - North Central", "EDR 7E- East Central", "EDR 7W- Central"), "Central",
                                           ifelse(edr %in% c("EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 8 - Southwest"), "Southwest",
                                                  ifelse(edr %in% c("EDR 9 - South Central", "EDR 10 - Southeast"), "Southern", as.character(planning.region)))))),
         project.pr = fct_relevel(project.pr, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southern"))


color.ruca <- c("Entirely rural" = "#009933", "Town/rural mix" = "#99CC33", "Urban/town/rural mix" = "#CC9966", "Entirely urban" = "#754C29", "Minnesota" = "black")

color.pr <- c("Northwest" = 	"#4575b4", "Northeast" = "grey", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.edr.simple <- c("EDR 1" = "#b3cde3", "EDR 2" = "#8c96c6", "EDR 3" = "#fe9929", "EDR 4" = "#8856a7", "EDR 5" = "#810f7c", "EDR 6E" = "#e5f5f9", "EDR 6W" = "#bdc9e1", "EDR 7E" = "#99d8c9", "EDR 7W" = "#2ca25f", "EDR 8" = "#74a9cf", "EDR 9" = "#0570b0", "EDR 10" = "#d7301f", "EDR 11" = "#d8b365", "Minnesota" = "black")

color.pr.edr <- c ("Northwest" = "#4575b4","Northeast" = "#e0f3f8", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black", "EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365")

mn_counties <- st_read("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Shapefiles/County shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE) %>%
  rename(countyfp = FIPS_CODE)
```

```{r master}
master.original <- read_csv("Data/SLEDS/Masters/Master-non-college-grads-disconnection.csv") 

kable(names(master.original))

kable(head(master.original))
```

# Multivariate Confirmation Analysis: Disconnection Persistence Among Non–Postsecondary Graduates

## Introduction

The bivariate analysis of workforce disconnection among non-college graduates identified two distinct disconnection mechanisms operating within a single combined measure. Non-sustained employment (working, but below the sustained-employment threshold) and absence of any Minnesota employment record behave as structurally different phenomena: the former is front-loaded and largely resolves within five years, while the latter is bimodal and behaves as a durable end-state once established. Separating these two mechanisms revealed that different demographic and high-school-accomplishment variables predict each one — most notably, Special Education status and English learner status appeared to drive no-record persistence specifically, while Economic status and Gender appeared to drive non-sustained persistence specifically, and different racial/ethnic groups showed disparities concentrated in one mechanism or the other.

This multivariate analysis exists to confirm whether that mechanism-split pattern holds once all variables are considered simultaneously, rather than one at a time. This step is necessary because earlier work in this project has shown that bivariate relationships — particularly those involving English learner status — can substantially attenuate or shift once other variables, especially missing-data patterns in academic assessment scores, are accounted for in the same model. Before this mechanism-split finding is presented as a confirmed result, it should be tested under the same conditions.

## Methodology

### Study Population

The analytic sample is restricted to individuals who did **not** complete a postsecondary credential, consistent with the non-college-graduate population used throughout this project. The sample is further restricted to individuals who graduated (or exited postsecondary attendance, for those who attended without completing) after 2010, consistent with the cohort restriction applied elsewhere in this project to avoid the PSEO participation tracking gap present in earlier cohorts.

### Dependent Variables

This analysis models two outcomes, each derived from the disconnection-persistence categories established in the bivariate analysis. Each graduate's workforce status was tracked across the first five years after exiting formal education (high school for direct-to-workforce graduates, or postsecondary attendance for those who attended without completing).

- **Persistent non-sustained employment** — binary (1 = non-sustained workforce participation in 4 or 5 of the first 5 years, 0 = otherwise). This isolates the instability mechanism identified in the bivariate work.
- **Persistent no-record status** — binary (1 = no Minnesota employment record in 4 or 5 of the first 5 years, 0 = otherwise). This isolates the disappearance-from-the-system mechanism identified in the bivariate work.

Each outcome is modeled separately, in its own specification, rather than as levels of a single combined variable. This mirrors the logic of the bivariate analysis: combining the two mechanisms into one model would risk re-obscuring the distinction this analysis is specifically designed to confirm.

### Independent Variables

The independent variable set is consistent with the variables used throughout this project's non-college-graduate analyses, with two adjustments made based on findings from this project's earlier work.

**Demographics**

- `Gender`
- `RaceEthnicity`
- `economic.status`
- `SpecialEdStatus`
- `english.learner`

**High school characteristics**

- `Dem_Desc`
- `avg.wages.pct.state`
- `avg.unemp.rate`

**High school accomplishments**

- `pseo.participant`
- `ap.exam`
- `took.ACT`
- `ACTCompositeScore` (recoded to 0 for non-test-takers, entered alongside `took.ACT`, per the missing-indicator approach established earlier in this project)
- `cte.achievement`
- `MCA.M`

**Postsecondary variable**

- `attended.ps` — included specifically because it was the single strongest predictor of overall disconnection in the bivariate analysis, and it is the one variable found to meaningfully affect both disconnection mechanisms rather than only one.

**Adjustments from the bivariate variable list:**

- **MCA Reading (`MCA.R`) is excluded.** The bivariate analysis tested MCA Math and Reading separately; consistent with the multicollinearity resolution used throughout this project's other multivariate models, only one MCA subject is retained here. MCA Math is retained on the same grounds established previously — lower missingness and a comparable or stronger association with the outcomes of interest.
- **`total.cte.courses.taken` is excluded**, consistent with the redundancy this project has established between the continuous course-count measure and the categorical `cte.achievement` variable.

### Modeling Approach

Each outcome (persistent non-sustained status, persistent no-record status) is modeled using a staged block-entry sequence, consistent with the modeling structure used throughout this project's other multivariate confirmation analyses. Predictors are introduced in three stages to make attenuation of any individual variable's effect visible and attributable to a specific block, rather than only observable in a single final specification.

**Model 1: Demographics and High School Characteristics**

- Demographics: `Gender`, `RaceEthnicity`, `economic.status`, `SpecialEdStatus`, `english.learner`
- High school characteristics: `Dem_Desc`, `avg.wages.pct.state`, `avg.unemp.rate`

This baseline isolates whether structural and demographic positioning alone predicts persistence in either disconnection mechanism, before any measure of academic engagement or preparation is introduced.

**Model 2: Add High School Accomplishments**

- Adds: `pseo.participant`, `took.ACT`, `ACTCompositeScore.filled`, `ap.exam`, `cte.achievement`, `MCA.M.filled`, `MCA.M.missing`

This is the model of primary interest for this analysis. Because English learner status has previously been shown to correlate with MCA Math missingness, this block specifically tests whether the English learner effect identified in Model 1 attenuates once high school accomplishments — including the MCA missingness indicator — are introduced. The same logic applies to Special Education status and any other Model 1 effect that might operate through academic engagement rather than independently.

**Model 3: Add Postsecondary Attempt**

- Adds: `attended.ps`

This isolates whether attempting postsecondary education, identified in the bivariate analysis as the one variable meaningfully affecting both disconnection mechanisms, adds independent explanatory power once demographics and high school accomplishments are already accounted for, or whether its bivariate strength was substantially channeled through those same variables (e.g., PSEO participation, CTE achievement, or ACT engagement).

Each model is compared to the prior one using a likelihood ratio test to confirm it adds significant explanatory power, with AIC tracked across the sequence to summarize overall model improvement. This staged structure is applied identically to both outcomes, allowing direct comparison of how each mechanism's predictors evolve across the same three blocks.

### Evaluation Criteria

Results will be reported using odds ratios, 95% confidence intervals, and statistical significance levels. Interpretation will focus specifically on whether the demographic variables identified as mechanism-specific in the bivariate analysis (Special Education and English learner status for no-record persistence; Economic status and Gender for non-sustained persistence; racial/ethnic groups for whichever mechanism their bivariate disparity was concentrated in) remain significant, attenuate, or reverse once the full variable set is included. Findings will be described as associations holding other included variables constant. No causal claims will be made.

<br>

# Prepping the data

The modeling dataset begins from the merged file containing both the disconnection-persistence categories (established in the bivariate analysis) and the full set of demographic, high school, and postsecondary variables used throughout this project. Three preparation steps are applied before modeling.

**Cohort restriction.** The sample is restricted to graduates from 2011 onward (`grad.year > 2010`). This mirrors the cohort restriction applied throughout this project's other analyses of non-college graduates: PSEO participation was not reliably tracked prior to 2011 (missingness runs 92-100% in the 2008-2010 cohorts, dropping to near-zero by 2014), and including those cohorts would risk misclassifying untracked PSEO participants as confirmed non-participants.

**Reference level assignment.** Categorical variables are releveled to match the reference categories used consistently throughout this project (male, White, non-economically-disadvantaged, non-Special-Education, non-English-learner, entirely rural district, no CTE involvement, non-participant/non-test-taker for the binary accomplishment variables). This keeps odds ratios in this analysis directly comparable in direction and interpretation to every other model built in this project.

**Missing-indicator recodes for two variables with structural or correlated missingness:**

- **ACT composite score.** Students who did not take the ACT have no composite score by definition — confirmed here to be a fully deterministic relationship with `took.ACT` (zero mismatches between the two). `ACTCompositeScore` is recoded to 0 for non-test-takers and entered into the model alongside `took.ACT`, so that `took.ACT` absorbs the effect of test participation itself while the filled score captures the achievement gradient among test-takers only.
- **MCA Math.** This project has previously found that MCA Math missingness correlates with English learner status specifically, and that this correlation can mask the true relationship between English learner status and an outcome when the raw (incomplete) score is used directly. Because this analysis is specifically designed to test whether the English learner effect identified bivariately survives multivariate control, the same missing-indicator approach is applied here: a missingness flag (`MCA.M.missing`) is entered alongside a filled score (`MCA.M.filled`, coded 0 for missing cases), preserving the full sample and preventing this known masking pattern from silently reappearing.

**Dependent variable construction.** The two outcomes tested in this analysis are built directly from the bivariate persistence categories, collapsed to binary indicators:

- `persistent.non_sustained` — 1 if the graduate was non-sustained in 4 or 5 of their first 5 years post-education-exit, 0 otherwise.
- `persistent.no_record` — 1 if the graduate had no Minnesota employment record in 4 or 5 of those years, 0 otherwise.

Before fitting any models, missingness is checked across the full variable set to confirm the cohort restriction and missing-indicator recodes fully resolved known data-quality issues, and the English learner / MCA Math-missingness correlation is re-verified on this specific cohort-restricted sample to confirm the relationship under examination is actually present before testing whether the multivariate model absorbs it.

```{r}
master.disc <- master.original %>%
  filter(grad.year > 2010) %>%
  mutate(
    Gender              = fct_relevel(Gender, "M"),
    RaceEthnicity       = fct_relevel(RaceEthnicity, "White"),
    economic.status = as_factor(economic.status),
    economic.status     = fct_relevel(economic.status, "0"),
    SpecialEdStatus = as_factor(SpecialEdStatus),
    SpecialEdStatus     = fct_relevel(SpecialEdStatus, "0"),
    english.learner = as_factor(english.learner),
    english.learner     = fct_relevel(english.learner, "0"),
    Dem_Desc            = fct_relevel(Dem_Desc, "Entirely rural"),
    pseo.participant = as_factor(pseo.participant),
    pseo.participant    = fct_relevel(pseo.participant, "0"),
    took.ACT            = fct_relevel(took.ACT, "No"),
    ap.exam = as_factor(ap.exam),
    ap.exam             = fct_relevel(ap.exam, "0"),
    cte.achievement     = fct_relevel(cte.achievement, "No CTE"),
    attended.ps         = fct_relevel(attended.ps, "No"),
    
    # ACT missing-indicator recode - confirmed safe: missingness fully
    # explained by took.ACT (9,228 No/missing, 5,941 Yes/present, zero mismatches)
    ACTCompositeScore.filled = ifelse(is.na(ACTCompositeScore), 0, ACTCompositeScore),
    
    # MCA Math missing-indicator recode - applied here specifically because
    # English learner status is central to this confirmatory analysis, and
    # MCA missingness has previously been shown to correlate with EL status
    # and mask its true relationship with the outcome
    MCA.M.missing = ifelse(is.na(MCA.M), 1, 0),
    MCA.M.filled  = ifelse(is.na(MCA.M), 0, MCA.M),
    
    # Dependent variables: binary persistence indicators
    persistent.non_sustained = ifelse(non_sustained.cat == "Persistently non-sustained", 1, 0),
    persistent.no_record     = ifelse(no_record.cat == "Persistently no-record", 1, 0)
  )

nrow(master.disc)

model_vars_disc <- c("Gender", "RaceEthnicity", "economic.status", "SpecialEdStatus",
                     "english.learner", "Dem_Desc", "avg.wages.pct.state", "avg.unemp.rate",
                     "pseo.participant", "took.ACT", "ACTCompositeScore.filled", "ap.exam",
                     "cte.achievement", "MCA.M.filled", "MCA.M.missing", "attended.ps",
                     "persistent.non_sustained", "persistent.no_record")

colSums(is.na(master.disc[, model_vars_disc]))

table(master.disc$persistent.non_sustained)
table(master.disc$persistent.no_record)

# confirm the EL/MCA-missingness correlation is present in this restricted,
# cohort-adjusted sample before we see whether the model absorbs it
master.disc %>%
  group_by(english.learner) %>%
  summarise(pct_mca_missing = mean(MCA.M.missing), n = n())
```

<br>

# Modeling: persistent non-sustained WF

## Model 1: Demographics and High School Characteristics

This model estimates the association between demographic characteristics and district context with persistent non-sustained employment (4-5 of the first 5 years post-education-exit), before any measure of academic engagement or preparation is introduced (n = 10,310).

**This model does not fully replicate the bivariate pattern, and that discrepancy is worth flagging directly rather than smoothing over.**

**Gender behaves as expected — strong and significant.** Female graduates show approximately 34% higher odds of persistent non-sustained employment relative to male graduates (OR ≈ 1.34, CI: 1.204-1.490, p \< .001), consistent with Gender being one of the two strongest bivariate predictors of instability.

**Economic status, by contrast, is not significant here** (OR ≈ 1.07, CI: 0.956-1.195, p = .243) — despite being the *other* strongest bivariate predictor of non-sustained persistence. This is a real departure from the bivariate finding, not a rounding difference. One plausible explanation: economic disadvantage's bivariate relationship with instability may be substantially confounded with race and disability status, both of which are now controlled for in this same model — but this is a hypothesis to note, not a conclusion to assert without further testing.

**Special Education status is significant here** (OR ≈ 1.36, CI: 1.189-1.562, p \< .001) — the opposite of what the bivariate analysis suggested, where Special Ed was one of the *weakest* predictors of instability specifically (its effect was concentrated almost entirely in no-record persistence, not non-sustained persistence). This reversal is worth watching closely as later blocks are added, since it may indicate a suppression effect — Special Ed's true relationship with instability may have been masked in the bivariate comparison by a confounding variable not yet identified.

**Race/Ethnicity shows significant effects for three of four groups, generally consistent with or stronger than the bivariate pattern.** American Indian graduates show higher odds (OR ≈ 1.37, consistent with the bivariate finding that AI graduates' disadvantage is instability-driven). Black graduates show the largest race-based effect in this model (OR ≈ 1.68), also consistent with the bivariate direction. Asian/Pacific Islander graduates show significantly *lower* odds (OR ≈ 0.56) — the same direction found bivariately (Asian/PI showed the lowest persistent-instability rate of any group). Hispanic graduates show no significant effect here (OR ≈ 1.04, CI crosses 1), consistent with the bivariate finding that this group's disadvantage was concentrated in no-record status rather than instability.

**District characteristics remain uniformly non-significant**, exactly as the bivariate analysis found: Dem_Desc (both categories), average local wages, and average local unemployment rate all show confidence intervals crossing 1 with no meaningful effect size.

**Overall Interpretation**

Two of the five demographic predictors — Economic status and Special Education status — do not behave as the bivariate analysis would predict once controlled for alongside the others, and in opposite directions from each other. Gender and the racial disparities largely replicate the bivariate pattern. District/geographic context remains entirely non-predictive, consistent with every other model in this project's disconnection work. The Economic status and Special Ed reversals are the two findings to track most closely once high school accomplishments are added in Model 2 — it's possible one or both of these patterns shift again once academic engagement variables, which correlate with both economic status and disability status, are accounted for.

**Model fit:** AIC = 9,092.4; residual deviance = 9,066.4 on 10,297 df (null deviance = 9,162.9 on 10,309 df); n = 10,310.

```{r}
model1_non_sustained <- glm(
  persistent.non_sustained ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    avg.wages.pct.state +
    avg.unemp.rate,
  family = binomial,
  data = master.disc
)

summary(model1_non_sustained)
nobs(model1_non_sustained)
AIC(model1_non_sustained)
exp(cbind(OR = coef(model1_non_sustained), confint(model1_non_sustained)))
```

<br>

## Model 2: Add High School Accomplishments

This model adds PSEO participation, ACT participation and score, AP exam, CTE achievement, and MCA Math (with the missing-data adjustment) to the demographic and district baseline (n = 10,128; 182 observations dropped for PSEO missingness, matching the expected small residual gap identified earlier).

**Special Education status strengthens rather than attenuating — the reversal from Model 1 is not explained by academic engagement.** OR increases slightly, from 1.36 in Model 1 to 1.42 here (CI: 1.223-1.647, p \< .001). This rules out the hypothesis raised after Model 1 that Special Ed's unexpected positive association with instability might be a proxy for weaker academic preparation — controlling for accomplishments directly does not reduce the effect at all.

**Economic status remains non-significant** (OR ≈ 1.09, p = .139), consistent with Model 1 — its bivariate relationship with instability continues not to hold up once other variables are controlled.

**PSEO participation is now significant, and in the adverse direction found bivariately.** OR ≈ 1.19 (CI: 1.047-1.349, p = .007) — PSEO participants show *higher*, not lower, odds of persistent non-sustained employment. This directly confirms the bivariate finding that PSEO's protective effect on overall disconnection runs almost entirely through avoiding no-record status, while showing a mild adverse relationship with instability specifically.

**ACT shows a genuine reversal from the bivariate pattern, worth flagging directly rather than glossing over.** Taking the ACT is now significantly *protective* (OR ≈ 0.638, CI: 0.426-0.953, p = .029) — the opposite of the bivariate finding, where ACT test-takers showed a slightly *lower* "never non-sustained" rate than non-test-takers. ACT composite score remains in the expected adverse direction, though only barely significant (OR ≈ 1.02 per point, p = .049, CI: 1.00008-1.041) — higher scores modestly predicting greater odds of persistent instability, consistent with the bivariate crossover.

**MCA Math missingness is now strongly significant and protective — a new finding not visible bivariately.** OR ≈ 0.469 (CI: 0.339-0.641, p \< .001): students with a missing MCA Math score show substantially *lower* odds of persistent non-sustained employment. This is a different direction than the pattern found throughout the rest of this project, where MCA missingness typically predicted *worse* outcomes. This is worth flagging as a genuinely new, multivariate-only signal rather than assuming it replicates the usual pattern.

**CTE achievement and AP exam both attenuate to non-significance.** Neither CTE category holds up (Concentrator/Completor: OR ≈ 0.89, p = .102; Participant: OR ≈ 0.91, p = .174), a real weakening from the bivariate finding of a genuine, if complex, CTE relationship with instability. AP exam similarly loses significance (OR ≈ 1.20, p = .156).

**English learner status remains non-significant, in the same direction as Model 1** (OR ≈ 1.11, p = .351) — no masking effect detected for this outcome, consistent with the bivariate finding that English learner status was never a strong predictor of instability specifically.

**Gender, Race/Ethnicity (AI and Black), and district/local-context variables** remain essentially unchanged in direction and significance from Model 1.

**Overall Interpretation**

Adding high school accomplishments significantly improves model fit (likelihood ratio test: Δdeviance = 56.22, df = 8, p \< .001; AIC decreases from 9,092.4 to 8,870.1). The Special Education reversal identified in Model 1 is confirmed, not explained away, by this block — it is not a proxy for academic underpreparation. PSEO and ACT both show genuine reversals from their bivariate relationships once controlled for other accomplishments, and MCA Math missingness emerges as a new, protective, multivariate-only predictor running counter to this project's usual pattern. CTE and AP exam, both real bivariate predictors of instability, do not hold up once entered alongside the other accomplishment variables.

**Model fit:** AIC = 8,870.1 (down from 9,092.4 in Model 1); Δdeviance = 56.22 on 8 df, p \< .001; n = 10,128 in both models.

```{r}
model2_non_sustained <- glm(
  persistent.non_sustained ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    avg.wages.pct.state +
    avg.unemp.rate +
    pseo.participant +
    took.ACT +
    ACTCompositeScore.filled +
    ap.exam +
    cte.achievement +
    MCA.M.filled +
    MCA.M.missing,
  family = binomial,
  data = master.disc
)

nobs(model1_non_sustained)
nobs(model2_non_sustained)
AIC(model1_non_sustained)
AIC(model2_non_sustained)
summary(model2_non_sustained)
exp(cbind(OR = coef(model2_non_sustained), confint(model2_non_sustained)))

model1_non_sustained_matched <- glm(
  persistent.non_sustained ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    avg.wages.pct.state +
    avg.unemp.rate,
  family = binomial,
  data = master.disc[!is.na(master.disc$pseo.participant), ]
)

nobs(model1_non_sustained_matched)  # should now match model2_non_sustained's 10,128

anova(model1_non_sustained_matched, model2_non_sustained, test = "Chisq")
```

<br>

## Model 3: Add Postsecondary Attempt

This model adds `attended.ps` to the full specification (n = 10,128, same complete-case sample as Model 2).

**Attempting postsecondary education shows a strong, independent protective effect on top of demographics, district context, and high school accomplishments — confirming this variable adds real explanatory power beyond what CTE, PSEO, and ACT already capture.** OR ≈ 0.694 (CI: 0.614-0.785, p \< .001): students who attempted postsecondary education, even without completing it, show roughly 31% lower odds of persistent non-sustained employment than those who never attempted it. This is a meaningful confirmation of the bivariate finding that `attended.ps` was the one variable affecting both disconnection mechanisms — its instability-side effect survives full multivariate control, not just partial control.

**PSEO's adverse effect from Model 2 persists essentially unchanged** (OR ≈ 1.20, CI: 1.053-1.358, p = .006) — even with postsecondary attempt now in the model, PSEO participation continues to show a mild, real association with *higher* odds of persistent instability, not lower.

**Took-ACT and ACT score both attenuate to the edge of significance.** Took-ACT moves from significant in Model 2 (p = .029) to marginal here (OR ≈ 0.677, CI: 0.452-1.011, p = .057). ACT composite score similarly weakens (OR ≈ 1.019, p = .059, CI: 0.999-1.040, now crossing 1). This suggests some of ACT's apparent relationship with instability in Model 2 was capturing information that `attended.ps` explains more directly — sensible, since taking the ACT and attempting postsecondary education are closely related behaviors.

**Special Education status remains essentially unchanged and still significant** (OR ≈ 1.39, CI: 1.199-1.616, p \< .001) — confirming again that this effect is not explained by postsecondary attempt any more than it was explained by high school accomplishments in Model 2. This is now the third model in a row where Special Ed's adverse association with instability holds up.

**MCA Math missingness remains strongly protective**, and if anything strengthens slightly (OR ≈ 0.445, CI: 0.321-0.609, p \< .001) — this counter-to-pattern finding continues to hold up across every specification tested so far.

**Economic status and English learner status both remain non-significant**, consistent with every prior model — neither shows any relationship with persistent instability once other variables are controlled for.

**Gender and the Race/Ethnicity effects (AI, Black, Asian/PI) remain stable** in direction and magnitude relative to Model 2.

**Overall Interpretation**

Adding postsecondary attempt significantly improves model fit (likelihood ratio test: Δdeviance = 34.94, df = 1, p \< .001; AIC decreases from 8,870.1 to 8,837.1). This confirms `attended.ps` as a genuine, independent predictor of instability specifically, not merely a proxy for the academic-preparation variables already in the model — it earns its place in the specification on its own merits. Across all three models, two findings have proven remarkably stable and are the clearest results of this analysis: Special Education status is a real, persistent, positive predictor of instability that survives every level of control tested, directly contradicting its weak bivariate relationship with this specific mechanism; and MCA Math missingness is a real, persistent, protective predictor that runs counter to the pattern found everywhere else in this project. Economic status and English learner status, by contrast, never emerge as significant predictors of instability at any stage, confirming the bivariate finding that whatever drives these two groups' overall disconnection, it is not this mechanism.

**Model fit:** AIC = 8,837.1 (down from 8,870.1 in Model 2); Δdeviance = 34.94 on 1 df, p \< .001; n = 10,128 across all three models.

```{r}
model3_non_sustained <- glm(
  persistent.non_sustained ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    avg.wages.pct.state +
    avg.unemp.rate +
    pseo.participant +
    took.ACT +
    ACTCompositeScore.filled +
    ap.exam +
    cte.achievement +
    MCA.M.filled +
    MCA.M.missing +
    attended.ps,
  family = binomial,
  data = master.disc
)

nobs(model2_non_sustained)
nobs(model3_non_sustained)
AIC(model2_non_sustained)
AIC(model3_non_sustained)
summary(model3_non_sustained)
exp(cbind(OR = coef(model3_non_sustained), confint(model3_non_sustained)))

# likelihood ratio test comparing Model 2 to Model 3
anova(model2_non_sustained, model3_non_sustained, test = "Chisq")
```

<br>

## Summary: Multivariate Confirmation — Persistent Non-Sustained Employment

Across all three models, several bivariate findings held up, but two demographic effects reversed direction in a way worth treating as this analysis's central discovery.

**Special Education status is a genuine, positive predictor of instability — the opposite of what the bivariate analysis suggested.** Bivariately, Special Ed appeared to be almost entirely a no-record phenomenon with little relationship to instability specifically. Once demographics, district context, high school accomplishments, and postsecondary attempt are all controlled for, Special Ed shows a strong, stable, significant association with *higher* odds of persistent non-sustained employment (OR ≈ 1.36–1.42 across all three models) that never attenuates, no matter what else is added to the model. This is a real reversal, not a bivariate artifact resolving itself — Special Ed's relationship with instability appears to have been *suppressed* in the bivariate comparison, not masked by something the multivariate model then explains away.

**Economic status and English learner status never emerge as predictors of instability at any stage**, confirming the bivariate finding cleanly: whatever drives these two groups toward overall disconnection, it is not the instability mechanism.

**PSEO participation and MCA Math missingness both behave in ways that run counter to expectations set elsewhere in this project.** PSEO shows a small but real *adverse* association with instability (consistent with the bivariate finding), holding steady across every model once other variables are controlled. MCA Math missingness is strongly *protective* against instability (OR consistently around 0.44–0.47) — the opposite of the pattern found throughout the rest of this project, where missing academic records typically predict worse outcomes. This is a genuinely new, multivariate-specific finding that couldn't have been seen bivariately, since the bivariate analysis only tested the raw score, not the missingness itself.

**Attempting postsecondary education is confirmed as an independent, genuine predictor of reduced instability**, surviving full control alongside CTE, PSEO, and ACT (OR ≈ 0.69 in the final model) — this is the clearest bivariate finding to survive multivariate testing fully intact.

**Race/Ethnicity effects for American Indian, Black, and Asian/Pacific Islander graduates all replicate their bivariate direction and remain significant or near-significant across every model** — American Indian and Black graduates show elevated odds of instability, Asian/Pacific Islander graduates show reduced odds, consistent with the bivariate finding that these groups' disadvantage (or advantage) is genuinely instability-driven.

**CTE achievement and AP exam, both real bivariate predictors of instability, do not survive multivariate control** — neither reaches significance once entered alongside the other accomplishment variables, suggesting their bivariate relationships were substantially explained by other correlated variables now in the model.

**Local geographic and economic context (Dem_Desc, wages, unemployment) remains entirely non-predictive across all three models**, confirming this project's consistent finding that local labor market conditions do not meaningfully shape workforce instability for this population.

<br>

# Modeling: persistent No MN emp record

## Model 1: Demographics and High School Characteristics

This model estimates the association between demographic characteristics and district context with persistent no-record status (4-5 of the first 5 years with no Minnesota employment record), before any measure of academic engagement is introduced (n = 10,310).

**English learner status shows the strongest effect in the model, exactly as the bivariate analysis predicted.** OR ≈ 1.94 (CI: 1.631-2.318, p \< .001) — English learners show nearly double the odds of persistent no-record status relative to non-English-learners. This is the clearest confirmation so far that English learner status's disconnection story is genuinely a no-record phenomenon, and it survives full demographic control at nearly its full bivariate strength.

**Race/Ethnicity confirms the bivariate mechanism split precisely.** Hispanic (OR ≈ 2.42, CI: 1.745-3.336, p \< .001) and Asian/Pacific Islander graduates (OR ≈ 2.41, CI: 1.714-3.371, p \< .001) both show more than double the odds of persistent no-record status — matching the bivariate finding that these two groups' disadvantage concentrates in this mechanism specifically. Black graduates show a real but more moderate effect (OR ≈ 1.47, p = .003). American Indian graduates are only marginally significant (OR ≈ 1.18, p = .068) — consistent with the bivariate finding that AI graduates' disadvantage lives almost entirely in the instability mechanism, not this one.

**Special Education status is *not* significant here** (OR ≈ 1.10, CI: 0.959-1.252, p = .172) — a real surprise given that Special Ed was one of the two strongest bivariate predictors of no-record persistence. Combined with Model 1's finding for non-sustained employment (where Special Ed was strongly significant), this suggests Special Ed's disconnection story may be almost entirely captured by the instability mechanism once demographics are controlled for together, not the no-record mechanism the bivariate comparison initially suggested. This is worth tracking closely through Models 2 and 3.

**Economic status is significant, but in the protective direction** (OR ≈ 0.777, CI: 0.701-0.861, p \< .001) — economically disadvantaged graduates show *lower* odds of persistent no-record status. This matches a pattern already visible in the bivariate work (disadvantage showed a modestly *lower* no-record rate there too), now confirmed as a real, significant effect once other demographics are controlled for.

**Gender is significant and protective** (OR ≈ 0.887, CI: 0.802-0.981, p = .020) — female graduates show modestly lower odds of persistent no-record status, a real effect not clearly visible in the bivariate comparison (where Gender was one of the weakest predictors overall).

**District, wage, and unemployment variables remain largely non-significant**, though average local wages is worth flagging as a near-miss (OR ≈ 2.43, CI: 0.896-6.577, p = .081) — a large point estimate with a very wide confidence interval. Given that wages showed a clean null result throughout the bivariate analysis, this shouldn't be over-interpreted yet, but it's worth watching whether it firms up or fades as later blocks are added.

**Overall Interpretation**

This model confirms the central mechanism-split predictions for English learner status and for the Hispanic/Asian-PI/American-Indian racial pattern — all replicate their bivariate direction and magnitude closely. The one genuine surprise is Special Education status, which does not hold up as a no-record predictor here despite being one of the strongest bivariate predictors of this exact outcome — raising the question of whether Special Ed's disconnection story is really an instability story once demographics are considered together, not a no-record story as the bivariate comparison suggested. Economic status and Gender both emerge as real, if modest, protective predictors of no-record status — a nuance not fully visible bivariately.

**Model fit:** AIC = 10,201.6; residual deviance = 10,176 on 10,297 df (null deviance = 10,357 on 10,309 df); n = 10,310.

```{r}
model1_no_record <- glm(
  persistent.no_record ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    avg.wages.pct.state +
    avg.unemp.rate,
  family = binomial,
  data = master.disc
)

summary(model1_no_record)
nobs(model1_no_record)
AIC(model1_no_record)
exp(cbind(OR = coef(model1_no_record), confint(model1_no_record)))
```

<br>

## Model 2: Add High School Accomplishments

This model adds PSEO participation, ACT participation and score, AP exam, CTE achievement, and MCA Math (with the missing-data adjustment) to the demographic and district baseline (n = 10,128).

**English learner status collapses from the strongest predictor in Model 1 to fully non-significant — this is the central confirmatory finding of this analysis.** OR drops from 1.94 (Model 1, p \< .001) to 0.905 (Model 2, CI: 0.731-1.118, p = .356). This directly replicates the masking pattern already documented elsewhere in this project: English learners' apparent disadvantage in persistent no-record status is substantially explained by the fact that English learners are far more likely to have a missing MCA Math score — not by English learner status itself. The bivariate finding was real, but its true driver was hiding underneath it.

**MCA Math missingness is, by a wide margin, the strongest predictor in this entire analysis.** OR ≈ 8.90 (CI: 6.916-11.495, p \< 2e-16) — students with a missing MCA Math score show nearly nine times the odds of persistent no-record status. This is the mechanism absorbing English learner's Model 1 effect, and it is a dramatically larger coefficient than the modest, protective MCA-missingness effect found in the non-sustained-employment sequence — confirming that this variable's role is highly outcome-specific, not a generic effect that shows up everywhere.

**Race/Ethnicity effects attenuate substantially across the board, consistent with the same masking mechanism operating for these groups too.** Hispanic drops from OR ≈ 2.42 (p \< .001) to 1.42 (p = .075, now only marginal). Black drops from a significant 1.47 (p = .003) to a non-significant 1.10 (p = .514). Asian/Pacific Islander remains significant but roughly halves in magnitude (OR ≈ 1.66, CI: 1.120-2.435, p = .010, down from 2.41). American Indian remains marginal, essentially unchanged (OR ≈ 1.19, p = .062). This pattern — real bivariate racial disparities substantially explained once MCA missingness enters the model — mirrors the English learner finding closely, and suggests MCA Math missingness may be correlated with race/ethnicity in ways not yet directly examined.

**PSEO, ACT participation, and ACT score all confirm their bivariate roles precisely, and ACT participation shows the largest coefficient of any accomplishment variable.** Taking the ACT is strongly protective (OR ≈ 0.314, CI: 0.209-0.472, p \< .001) — the same large protective effect found bivariately. ACT composite score shows the expected adverse crossover (OR ≈ 1.056 per point, p \< .001) — higher scores among test-takers predicting *higher* odds of no-record status, confirming the threshold-effect pattern found bivariately. PSEO participation is now significantly protective (OR ≈ 0.869, p = .031), matching its bivariate direction.

**Average local unemployment rate becomes significant here for the first time in this entire project's disconnection work** (OR ≈ 0.920 per unit, CI: 0.878-0.964, p \< .001) — higher local unemployment predicting *lower* odds of persistent no-record status. This is the same counterintuitive direction found in the bivariate analysis, now confirmed as statistically real rather than a fluke — worth noting as genuinely present, but still resistant to a simple explanation.

**Special Education status remains non-significant**, confirming Model 1's surprise finding — Special Ed shows no relationship with no-record status at any level of control tested so far, reinforcing that its true disconnection story (established in the non-sustained sequence) is an instability story, not a no-record story.

**Economic status and Gender both strengthen slightly and remain protective** (economic status: OR ≈ 0.826, p \< .001; Gender: OR ≈ 0.849, p = .003), consistent with Model 1.

**CTE achievement shows a modest, borderline protective effect** (Concentrator/Completer: OR ≈ 0.871, p = .046) not clearly visible bivariately. MCA Math's filled score, once the missingness indicator is separated out, shows a real protective effect (OR ≈ 0.877 per point, p \< .001) — higher actual scores predicting lower odds of no-record status, the expected direction once the missingness confound is removed.

**Overall Interpretation**

Adding high school accomplishments produces a massive improvement in model fit (likelihood ratio test: Δdeviance = 849.54, df = 8, p \< .001; AIC decreases sharply from 10,201.6 to 9,144.5) — the largest single-block improvement found anywhere in this project's disconnection work, driven almost entirely by MCA Math missingness. This model delivers the confirmatory result the analysis was built to test: English learner status's strong bivariate association with persistent no-record status is not an independent effect — it is substantially explained by the correlation between English learner status and missing MCA Math scores. The same masking pattern extends to Hispanic and Black graduates' racial disparities, both of which attenuate to non-significance or near-non-significance once this variable is controlled for. This means the bivariate framing of "English learner and certain racial groups are structurally more likely to disappear from Minnesota's employment records" needs to be revised to something more specific: much of that disparity appears to trace back to gaps in academic assessment records themselves, which may reflect testing exemptions, transfers, or other structural factors correlated with both group membership and record completeness.

**Model fit:** AIC = 9,144.5 (down from 10,201.6 in Model 1); Δdeviance = 849.54 on 8 df, p \< .001; n = 10,128 in both models.

```{r}
model2_no_record <- glm(
  persistent.no_record ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    avg.wages.pct.state +
    avg.unemp.rate +
    pseo.participant +
    took.ACT +
    ACTCompositeScore.filled +
    ap.exam +
    cte.achievement +
    MCA.M.filled +
    MCA.M.missing,
  family = binomial,
  data = master.disc
)

nobs(model1_no_record)
nobs(model2_no_record)
AIC(model1_no_record)
AIC(model2_no_record)
summary(model2_no_record)
exp(cbind(OR = coef(model2_no_record), confint(model2_no_record)))

# Model 1 refit on Model 2's complete-case sample, for a valid LRT
# (same fix used for the non-sustained sequence, since PSEO's 182 missing
# rows will again shrink Model 2's sample relative to Model 1)
model1_no_record_matched <- glm(
  persistent.no_record ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    avg.wages.pct.state +
    avg.unemp.rate,
  family = binomial,
  data = master.disc[!is.na(master.disc$pseo.participant), ]
)

nobs(model1_no_record_matched)  # should match model2_no_record's n

anova(model1_no_record_matched, model2_no_record, test = "Chisq")
```

<br>

## Model 3: Add Postsecondary Attempt 

This model adds `attended.ps` to the full specification (n = 10,128, same complete-case sample as Model 2).

**Attempting postsecondary education shows the second-largest effect in this entire model, confirming its bivariate standing as the strongest overall predictor of disconnection.** OR ≈ 0.613 (CI: 0.541-0.693, p \< 2e-16): students who attempted postsecondary education show roughly 39% lower odds of persistent no-record status than those who never attempted it — a larger protective effect than any other accomplishment variable in the model except MCA Math missingness. This confirms that `attended.ps`'s bivariate strength on this outcome specifically is real and independent, not an artifact of overlap with PSEO, ACT, or CTE.

**MCA Math missingness remains the dominant predictor in the model, essentially unchanged from Model 2** (OR ≈ 8.35, CI: 6.479-10.784, p \< 2e-16, down only slightly from 8.90). The magnitude of this effect is not explained by postsecondary attempt — it is a genuinely separate, massive predictor of persistent no-record status.

**English learner status remains non-significant, confirming Model 2's finding holds up even with postsecondary attempt in the model** (OR ≈ 0.888, CI: 0.717-1.098, p = .275). This closes the loop on the central question this analysis was designed to answer: English learner status's bivariate relationship with persistent no-record status does not survive multivariate control at any stage, and MCA Math missingness remains the best explanation for where that bivariate signal actually came from.

**Race/Ethnicity effects hold essentially steady from Model 2.** Asian/Pacific Islander remains significant (OR ≈ 1.61, p = .017). Hispanic and American Indian remain marginal (p = .072 and p = .054, respectively). Black graduates remain non-significant (OR ≈ 1.12, p = .428). No further attenuation occurs once postsecondary attempt is added — the racial disparities that attenuated in Model 2 stay attenuated here, confirming MCA Math missingness (not lack of postsecondary attempt) was the relevant explanatory factor for those groups too.

**Took-ACT, ACT score, PSEO, and CTE achievement all remain essentially unchanged from Model 2** in direction, magnitude, and significance — none of these accomplishment variables' relationships with no-record status are meaningfully altered by adding postsecondary attempt to the model.

**Special Education status remains non-significant**, consistent with both prior models — confirming across the entire model sequence that Special Ed shows no relationship whatsoever with no-record status, at any level of control.

**Economic status, Gender, and average local unemployment rate all remain significant and essentially stable** from Model 2, each showing modest protective effects.

**Overall Interpretation**

Adding postsecondary attempt significantly improves model fit (likelihood ratio test: Δdeviance = 62.35, df = 1, p \< .001; AIC decreases from 9,144.5 to 9,084.1), confirming `attended.ps` as a large, genuine, independent predictor of no-record status — its effect here is even more pronounced than in the non-sustained-employment sequence, consistent with the bivariate finding that this variable's protective effect was concentrated more heavily in the no-record mechanism. Critically, this final model does not change the central finding from Model 2: English learner status and the Hispanic/Black racial disparities remain non-significant or marginal, confirming that MCA Math missingness — not postsecondary engagement — is the variable actually explaining those groups' apparent bivariate disadvantage. Across the full sequence, MCA Math missingness stands out as the single most powerful and consistent predictor of persistent no-record status found in this project, alongside a large and independently confirmed protective effect of postsecondary attempt.

**Model fit:** AIC = 9,084.1 (down from 9,144.5 in Model 2); Δdeviance = 62.35 on 1 df, p \< .001; n = 10,128 across all three models.

```{r}
model3_no_record <- glm(
  persistent.no_record ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    avg.wages.pct.state +
    avg.unemp.rate +
    pseo.participant +
    took.ACT +
    ACTCompositeScore.filled +
    ap.exam +
    cte.achievement +
    MCA.M.filled +
    MCA.M.missing +
    attended.ps,
  family = binomial,
  data = master.disc
)

nobs(model2_no_record)
nobs(model3_no_record)
AIC(model2_no_record)
AIC(model3_no_record)
summary(model3_no_record)
exp(cbind(OR = coef(model3_no_record), confint(model3_no_record)))

# likelihood ratio test comparing Model 2 to Model 3
anova(model2_no_record, model3_no_record, test = "Chisq")
```

<br>

## Summary: Multivariate Confirmation

Across all three models, this analysis delivered the central confirmatory result it was designed to test, along with several supporting findings.

**English learner status's strong bivariate relationship with persistent no-record status does not survive multivariate control.** OR collapsed from 1.94 (p \< .001) in the demographics-only model to non-significant (OR ≈ 0.89, p = .275) once high school accomplishments — specifically the MCA Math missingness indicator — were added, and remained non-significant through the final model. This confirms the masking hypothesis this analysis was built to test: English learners' apparent disadvantage in disappearing from Minnesota's employment records is substantially explained by their much higher rate of missing MCA Math scores, not by English learner status operating independently.

**The same masking pattern extends to Hispanic and Black graduates.** Both showed large, significant bivariate-consistent effects in the demographics-only model (Hispanic OR ≈ 2.42; Black OR ≈ 1.47) that attenuated to marginal or non-significant once MCA Math missingness entered the model (Hispanic OR ≈ 1.42-1.43, p ≈ .07; Black OR ≈ 1.10-1.12, non-significant). Asian/Pacific Islander graduates are the exception — their effect attenuated by roughly a third but remained significant throughout (OR ≈ 1.61-1.66 in the final models), indicating a real disparity that persists independent of the MCA-missingness pattern.

**MCA Math missingness is the single most powerful predictor found anywhere in this analysis.** OR remained in the 8-9 range across both models where it appeared, dwarfing every other coefficient in either outcome. This is a dramatically different role than the same variable played in the non-sustained-employment sequence, where it was modestly *protective* rather than a massive risk factor — confirming that MCA Math missingness operates in genuinely different, outcome-specific ways depending on which disconnection mechanism is being modeled.

**Attempting postsecondary education is confirmed as a large, genuine, independent predictor of reduced no-record risk**, holding up fully alongside every other variable (OR ≈ 0.61 in the final model) — an even larger effect than it showed in the non-sustained-employment sequence, consistent with the bivariate finding that `attended.ps`'s protective effect concentrates more heavily in this mechanism.

**ACT participation and score both confirm their bivariate crossover pattern precisely and hold up fully across every model.** Taking the ACT is strongly protective; among test-takers, higher scores predict *higher* odds of no-record status — the same threshold-effect crossover found bivariately, now confirmed under full multivariate control.

**Special Education status shows no relationship with no-record status at any stage** — a clean, consistent null result across all three models, confirming that Special Ed's true disconnection story (established in the companion non-sustained-employment analysis) is an instability story, not a no-record story, despite the bivariate comparison initially suggesting otherwise.

**Economic status, Gender, PSEO participation, and average local unemployment rate all emerge as modest but genuine and stable protective predictors**, each holding up consistently once controlled for the full variable set — none of these were as clearly visible in the bivariate work.

<br>

# Multivariate Confirmation: Side-by-Side Comparison of the Two Disconnection Mechanisms

**What the Bivariate Analysis Predicted vs. What the Multivariate Models Confirmed**

| Variable | Bivariate expectation | Confirmed for Non-Sustained? | Confirmed for No-Record? |
|------------------|------------------|------------------|------------------|
| Special Education status | No-record mechanism | **No — reversed.** Strong, stable predictor of instability instead (OR ≈ 1.36–1.42) | **No.** Non-significant at every stage |
| English learner status | No-record mechanism | Confirmed null (never significant) | **No — masked.** Strong in Model 1 (OR ≈ 1.94), collapses once MCA Math missingness is added |
| Economic status | Instability mechanism | **No.** Non-significant at every stage | **Reversed role — protective, no-record.** Significant, protective (OR ≈ 0.81–0.83) |
| Gender | Instability mechanism | Confirmed (OR ≈ 1.30–1.34, adverse) | **New finding.** Also significant here, protective (OR ≈ 0.85–0.89) |
| American Indian | Instability mechanism | Confirmed (OR ≈ 1.37–1.39) | Confirmed null (marginal at best) |
| Hispanic / Asian-PI | No-record mechanism | Confirmed null (mostly) | **Partially masked.** Hispanic attenuates to marginal; Asian/PI persists independently |
| Black | Both, moderately | Confirmed (OR ≈ 1.68–1.83) | **Masked.** Attenuates to non-significant once MCA missingness enters |
| MCA Math missingness | Not tested bivariately | **New finding — protective** (OR ≈ 0.44–0.47) | **New finding — massive risk factor** (OR ≈ 8.3–8.9) |
| Attended postsecondary | Both mechanisms | Confirmed, independent (OR ≈ 0.69) | Confirmed, independent and larger (OR ≈ 0.61) |
| ACT participation/score | No-record mechanism | **Partially reversed.** Took-ACT flips protective in Model 2, fades to marginal by Model 3 | Confirmed precisely (crossover pattern intact) |
| PSEO participation | Ambiguous/adverse | Confirmed adverse (OR ≈ 1.19–1.20) | Confirmed protective (OR ≈ 0.87) |
| Local wages/unemployment/Dem_Desc | Null everywhere | Confirmed null throughout | Mostly null; unemployment rate emerges as a small, real, significant effect |

## **The Story This Comparison Tells**

**The core mechanism-split finding survives, but not in its original bivariate form — it's been sharpened and partly rewritten.** The bivariate analysis suggested Special Education and English learner status were both primarily no-record phenomena. The multivariate models tell a more precise story: **English learner status genuinely is a no-record phenomenon, but only because it's a proxy for missing MCA Math scores** — once that's controlled for directly, the effect disappears entirely. **Special Education status, by contrast, turns out to be an instability phenomenon, not a no-record phenomenon at all** — the opposite of what bivariate comparison suggested, and a real reversal rather than an attenuation. These are two different kinds of correction, and both matter for how the finding gets written up: one confirms the mechanism-split logic (EL's effect really does trace back to a no-record-side variable), while the other overturns which mechanism a specific group belongs to (Special Ed belongs to instability, not no-record).

**MCA Math missingness is the standout finding of this entire confirmatory analysis, and it behaves completely differently across the two outcomes.** It's a modest protective factor for instability (OR ≈ 0.45) but the single most powerful risk factor found anywhere in this project for no-record status (OR ≈ 8-9). This is worth its own dedicated discussion in the report — a missing MCA Math score is, on its own, one of the strongest predictors of falling off Minnesota's employment record entirely, dramatically more so than any demographic characteristic tested. Given that this variable is itself a data-completeness artifact rather than a substantive predictor in the usual sense, this raises a real question worth flagging rather than resolving: what does it mean, structurally, that gaps in a student's academic testing record predict disappearing from the state's employment record years later? That's likely a question for a follow-up investigation, not something this analysis can answer, but it's too large a finding to leave unmentioned.

**Race/Ethnicity confirms the mechanism-split framework while adding a genuine nuance.** Black graduates' no-record disparity is substantially explained by MCA missingness (mirroring the English learner pattern), while Asian/Pacific Islander graduates' no-record disparity survives full control — a real, independent disparity that isn't explained by the same mechanism. American Indian graduates confirm cleanly as an instability-specific group, exactly as bivariate work suggested.

**Gender and Economic status turn out to matter for *both* mechanisms, not just one** — a genuine addition beyond the bivariate framing, which suggested these two variables were concentrated in instability alone. Both are now confirmed as real, independent, protective predictors of no-record status too, just smaller in magnitude than their instability-side effects.

**`attended.ps` and PSEO both confirm their bivariate roles cleanly across every model** — the two clearest, most stable findings in this entire confirmatory analysis, with no attenuation or reversal at any stage.

## What This Means for the Report

Three things are ready to go into the report as confirmed, multivariate-tested findings: (1) postsecondary attempt and PSEO participation's mechanism-specific effects, (2) the reframing of English learner and Black disparities as substantially explained by MCA Math missingness rather than group membership itself, and (3) Special Education status's correction from a no-record story to an instability story. The MCA Math missingness finding deserves prominent, standalone treatment given its scale. Given how much correction happened between the bivariate and multivariate stages, I'd recommend the report lead with the multivariate results as the confirmed story, using the bivariate findings only as the motivating context for why this deeper investigation was undertaken — not present both as equally reliable, parallel findings.
